### 2015-07-02 PAIS.tex
init

\documentclass[final]{siamltex}
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins}
\setlength{\parindent}{0in}
\usepackage{amsmath,amsfonts,amsbsy}
\newcommand{\RARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longrightarrow}^{#3}\; #2}
\newcommand{\RARRlong}[3]{#1
  \;\displaystyle\mathop{-\!\!\!-\!\!\!-\!\!\!-\!\!\!-\!\!\!\!\displaystyle
  \longrightarrow}^{#3}\; #2}
\newcommand{\LARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longleftarrow}^{#3}\; #2}
\newcommand{\LRARR}[4]{{\mbox{ \raise 0.4 mm \hbox{$#1$}}} \;
  \mathop{\stackrel{\displaystyle\longrightarrow}\longleftarrow}^{#3}_{#4}
  \; {\mbox{\raise 0.4 mm\hbox{$#2$}}}}
\newcommand{\bX}{{\bf X}}
\newcommand{\vecx}{{\mathbf x}}
\newcommand{\vecy}{{\mathbf y}}
\newcommand{\vecz}{{\mathbf z}}
\newcommand{\vecq}{{\mathbf q}}
\newcommand{\bs}{{\mathbf s}}
\newcommand{\vecr}{{\mathbf r}}
\newcommand{\vecX}{{\mathbf X}}
\newcommand{\vecv}{{\mathbf v}}
\newcommand{\tick}{\ding{52}}
\newcommand{\cross}{\ding{54}}
\newcommand{\vecn}{{\mathbf n}}
\newcommand{\vecp}{{\mathbf p}}
\newcommand{\cT}{{\mathcal T}}
\newcommand{\dt}{{\mbox{d}t}}
\newcommand{\dx}{{\mbox{d} \vecx}}
\newcommand{\boldnu}{{\boldsymbol \nu}}
\newcommand{\er}{{\mathbb R}}
\renewcommand{\div}{{\rm div}}
\newcommand{\bnu}{{\bf \nu}}
\newcommand{\divergence}{\mathop{\mbox{div}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\FP}{P_{\rm{FP}}}
\newcommand{\ME}{P_{\rm{ME}}}
\newcommand{\MEs}{P_{\rm{ME}_{S}}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\X}{{\mathbf X}}
\newcommand{\Y}{{\mathbf Y}}
\newcommand{\W}{{\mathbf W}}
\newcommand{\data}{D}
\newcommand{\neff}{n_{\text{eff}}}
\newcommand{\E}{{\mathbb E}}
\newcommand{\picturesAB}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(a)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(b)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\newcommand{\picturesCD}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(c)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(d)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are commonly used and powerful
  family of numerical methods for sampling from complex probability
  distributions. As the applications that these methods are being
  applied to increase in size and complexity, the need for efficient
  methods which can exploit the parallel architectures which are
  prevalent in high performance computing increases. In this paper, we
  aim to develop a framework for scalable parallel MCMC algorithms. At
  each iteration, an importance sampling proposal distribution is
  formed using the current states of all of the chains within an
  ensemble. Once weighted samples have been produced from this, a
  state-of-the-art resampling method is then used to create an evenly
  weighted sample ready for the next iteration. We demonstrate that
  this parallel adaptive importance sampling (PAIS) method outperforms
  naive parallelisation of serial MCMC methods using the same number
  of processors. This demonstrates the added
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
%\subsection{Lit. review}
%\subsection{Optimal transport resampler}
%\subsection{Function Space MCMC}
Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
that allow us to sample from complex probability distributions. MCMC
methods were first developed in the 70s\cite{hastings1970monte}, and with the development of
faster more powerful computers, have become ever more important in a
whole range of fields in statistics, science and engineering. However,
despite the advances in hardware that have made it possible to fully
characteries the posterior distributions of many problems, it remains
unfeasible to use MCMC methods for many more. In particular, when
considering Bayesian inverse problems, each MCMC step may involve the
numerical solution of one or more PDE. As many samples are usually
required before Monte Carlo error is reduced to acceptable levels,
these types of problem remain frustratingly out of our grasp.

Many advances have been made in the field of MCMC, to design ever more
complex methods, which more intelligently propose moves which leads to
faster converging methods. Function space versions of standard methods
such as the random walk Metropolis-Hastings (RWMH) algorithm or the
Metropolis adjusted Langevin algorithm (MALA), whose convergence rates
are independent of dimension have been
developed\cite{cotter2013mcmc}. The hybrid (or Hamiltonian) Monte
Carlo (HMC) method uses Hamiltonian dynamics in order to propose and
accept moves to states which are a long way away from the current
position\cite{sexton1992hamiltonian}, and function space analogues of
this have also been proposed\cite{beskos2011hybrid}. Riemannian
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.  These
methods allow us to more fully explore the posterior distribution, at
the cost of fewer iterations of the method.

Simultaneously, great strides are continually being made in the
development of computing hardware. Moore's law, which predicted that
the number of transistors that it is possible to fit on a single
microchip will double every two years, has been largely followed
since the early 70s\cite{moore1998cramming}. However, the main revolution with regards to
scientific computing in recent decades has been the increasing focus on parallel
architectures. The efficient exploitation of these facilities is the key to
solving many of the computational challenges that we currently face.

As such, the development of efficient parallel MCMC algorithms is an
important area for research. Since MCMC methods can be trivially
parallelised by simply running many independent chains in parallel,
the focus needs to be on the development of methods which gain some
added benefit through parallelism. One class of parallel MCMC method
uses multiple proposals, with only one of these proposals being
accepted. Examples of this approach include multiple try
MCMC\cite{liu2000multiple} and ensemble MCMC\cite{neal2011mcmc}. In
\cite{calderhead2014general}, a general construction for the
parallelisation of MCMC methods was presented, which demonstrated
speed ups of up to two orders of magnitude when compared with serial methods.

In this paper, we present a framework for parallelisation of
importance sampling, which allows us to use any of the current
Metropolis-based methodologies in order to create an efficient target
proposal from the current state of all of the chains in the
ensemble. In Section \ref{sec:Prelim} we outline some preliminaries,
including the general set up for Bayesian inverse problems, the
preconditioned Crank-Nicolson Langevin (pCNL) algorithm and a
brief review of the optimal transport resampler,  both of which we will be
employing within the algorithm. In Section \ref{sec:MCMC}, we explore the common problems with MCMC
methods, which we intend to address with the development of our new
framework. We describe this framework in Section \ref{Sec:PAIS}. In
Section \ref{Sec:Num}, we present some numerical experiments which
demonstrate the savings that can be made by employing this approach as
opposed to a naive/trivial parallelisation of existing MCMC
methods. In Section \ref{Sec:Conc}, we will summarise our results and
suggest some areas for future investigation.

\section{Preliminaries}
\subsection{Bayesian inverse problems}
In this paper, we particularly focus on the use of MCMC methods in characterising
posterior probability distributions in Bayesian inverse problems. We
wish to learn about a particular unknown quantity $u$, of which we are
able to make direct or indirect noisy observations. For now
we say that $u$ is a member of a Hilbert
space $X$.

The parameter $u$ is observed
through the observation operator $\mathcal{G}:X \to
\mathbb{R}^d$. Since observations are rarely, if ever, perfect, we
assume that these measurements $\data$ are subject to Gaussian noise, so that
\begin{equation}\label{eqn:obs}
	\data = \mathcal{G}(u) + \varepsilon, \qquad \varepsilon \sim \mu_{\varepsilon} = \mathcal{N}(0,\Sigma).
\end{equation}
For example, if $u$ are the rates of reactions in a chemical system, $\G$ might return the times at which each reaction occurs, or some summary of this information.

These modelling assumptions allow us to construct the 
likelihood of observing the data $\data$ given the parameter $u =
u^*$. Rearranging \eqref{eqn:obs} and using the distribution of $\varepsilon$, we get:
\begin{equation}\label{eqn:like}
	\mathbb{P}(\data|u=u^*) \propto \exp \left ( -\frac{1}{2} \|\mathcal{G}(u^*)
	  - \data\|_\Sigma^2 \right ) = \exp\left(-\Phi(u^*)\right),
\end{equation}
where $\| x - y \|_\Sigma$ is the Mahalanobis distance between $x$ and $y$.

As discussed in \cite{stuart2010inverse,cotter2009bayesian},
in order for this inverse problem to be well-posed in the Bayesian
sense, we require the posterior distribution, $\mu_Y$, to be absolutely
continuous with respect to the prior, $\mu_0$. A
minimal regularity prior can be chosen informed by regularity results
of the observational operator $\mathcal{G}$. Given such a prior, then
the Radon-Nikodym derivative of the posterior measure, $\mu_Y$, with
respect to the prior measure, $\mu_0$, is proportional to the
likelihood:
\begin{equation}\label{eqn:RND}
	\frac{d\mu_Y}{d\mu_0} \propto \exp \left ( -\Phi(u^*) \right ).
\end{equation}

\subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}
In recent years, work has been carried out to frame MCMC proposal
distributions on function space\cite{cotter2013mcmc}. These new
discretisations perform comparably with the original versions in low
dimensions, so we choose one of these distributions to consider. If gradient information regarding the observation operator is
available, then a range of MCMC methods are available which exploit this information to improve mixing rates. One
example of such an algorithm is MALA. In \cite{cotter2013mcmc}, a function space version of this
method was presented, the pCNL algorithm, and is described in full in Table
\ref{tab:pCNL}. The proposal used in this method comes about through
a Crank-Nicolson approximation of the Langevin SDE, whose invariant
measure is the posterior measure $\mu_Y$.

\begin{table}
\begin{mdframed}
\begin{algorithmic}
\STATE $X = x_0$
\FOR{$i=1,2,3,\ldots$}
\STATE $Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
2\delta\mathcal{C}\nabla \Phi(u)+
\sqrt{8\delta} W\right] $, $W \sim \mu_0$
\STATE $a(X_{i-1},Y) = \min \left \{ 1,  \exp(\Phi(X_{i-1}) - \Phi(Y) ) \right \}$.
\STATE $u \sim U([0,1])$
\IF{$u<a(X_{i-1},Y)$}
\STATE $X_i = Y$
\ELSE 
\STATE $X_i = X_{i-1}$
\ENDIF
\ENDFOR
\end{algorithmic}
\end{mdframed}\caption{A pseudo-code representation of the preconditioned Crank-Nicolson Langevin
   (pCNL) algorithm. $\delta \in (0,2]$ is a step size parameter.}
\label{tab:pCNL}
\end{table}

\subsection{Particle filters and resamplers}\label{sec:filters}
In several applications, data must be assimilated in an ``online''
fashion, with up to date observations of the studied system being made
available on a regular basis. In these contexts, such as in weather forecasting or
oceanography, data is incorporated using a filtering
methodology. One popular filtering method is the particle filter, the
first of which was dubbed the Bootstrap
filter\cite{gordon1993novel}. In this method, a set of weighted particles is
used to represent the posterior distribution. The positions of the
particles are updated using the model dynamics. Then, when more
observations are made available, the relative weights of the particles
are updated to take account of this data, using Bayes' formula. Other filtering methods, such
as the Kalman filter\cite{kalman1960new} and ensemble Kalman filter\cite{evensen1994sequential}, have also been developed which are often used within the data
assimilation community.

One advantage of the particle filter is that there are convergence
results for this method as the number of particles is increased. The
downside is that, the effective sample size decreases at each iteration. One way to tackle
this is to employ a resampling scheme. The aim of a successful
resample is to take your unevenly weighted ensemble and return a new
ensemble of particles with even weights which is highly correlated to the original samples.

The Ensemble Transform (ET) method proposed by Reich \cite{reich2013nonparametric} makes use of optimal transportation as described in \cite{villani2003topics,villani2008optimal}. The transform takes a sample of weighted particles $\{y_i\}_{i=1}^M$ from $\mu_Y$ and converts it into a sample of evenly weighted particles $\{x_i\}_{i=1}^M$ from $\mu_X$, by means of defining a coupling $T^*$ between $Y$ and $X$. Given that a trivial coupling $T^t$ always exists in the space of transference plans, $\Pi(\mu_{X}, \mu_{Y})$, we can find a coupling $T^*$ which maximises the correlation between $X$ and $Y$ \cite{cotter2012ensemble}. This coupling is the solution to a linear programming problem in $M^2$ variables with $2M-1$ constraints as defined in \cite{reich2013nonparametric}. Maximising the correlation ensures that the new sample is as much like the original sample as possible with the additional property that the sample is evenly weighted.

As proposed in \cite{reich2013nonparametric} a Monte Carlo algorithm can be implemented to resample from a weighted ensemble. We create a weighted sample, then solve the optimal transport problem which produces the coupling described above, we can draw a new sample from the evenly weighted distribution. Reich suggests using the mean of the evenly weighted distribution to produce a consistent estimator.

Analysis of this method shows that as the ensemble size increases, the statistics of the evenly weighted sample approach those of the posterior distribution, at least well enough for a proposal distribution as described in Section~\ref{sec:PAIS_alg}. The histogram of the evenly weighted sample exhibits small oscillations in the tails of the posterior, and also struggles to deal with discontinuities.

Ideally the dimension of particles in a filter will not affect the convergence of the method. Unfortunately, the approximation of continuous density functions is computationally infeasible in high dimensions and the required ensemble size scales exponentially with the dimension of the problem\cite{silverman1986density,snyder2008obstacles}.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=\textwidth]{figures/ET_dim_scaling}
\caption{Demonstration of the ETMC algorithm rapidly failing as the dimension increases. The relative error shown is that between the sampled mean and the true mean.}
\label{fig:ET_dim_scaling}
\end{center}
\end{figure}

Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm loses its ability to accurately represent the mean of the posterior distribution as the dimension increases. Clearly for a simple Gaussian problem, we can not resample in much more than 10 dimensions and hope to recover a representative sample.

\subsection{Deficiencies of Metropolis-type MCMC schemes}
All MCMC methods are trivially parallelisable. One can take a method
and simply implement it simultaneously over a set of processes. All
of the states of all of the processes can be recorded, and in the time
that it takes one process to draw $N$ samples, $M$ processes can draw
$NM$ samples.

However, we argue that this is a far from optimal
scenario. First of all, unless we have a lot of information about the posterior, we will begin the algorithm a long way from statistical equilibrium. Some initial iterations then are not samples from the posterior, and must be thrown away. This process is known as the burn-in. In a trivially parallelised scenario, each process must perform this process independently, while also tuning the scaling parameter.

Moreover, many MCMC algorithms suffer from poor mixing, especially in
multimodal systems. The time for an MCMC trajectory to switch between modes can be
large and given that a large number of switches are required before we have a
good idea of the relative probability densities of these different
regions, it can be prohibitively expensive.

Another aspect of Metropolis-type samplers is that information
computed about a proposed state is simply lost if we choose to reject
that proposal in the Metropolis step. An advantage of importance
samplers is that no evaluations of $\mathcal{G}$ are ever wasted
since all samples are saved along with their relative weighting.

Moreover, a trivially parallelised MCMC scheme is exactly that -
trivial. Intuition suggests that we can gain some speed up by sharing information across the processes and that is
exactly what we wish to demonstrate in this paper.

These deficiencies of the trivial method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(PAIS). In the next section we will introduce the method in its most
general form. We will then introduce the version that we have
implemented, which utilises the resampler recently suggested by
Reich\cite{reich2013nonparametric} described in Section \ref{sec:filters}, and the pCNL proposal distribution.

\section{The Parallel Adaptive Importance Sampler \allowbreak (PAIS)}\label{sec:PAIS_alg}

In a trivially parallelised MCMC scheme, if we have enough processes
all sampling from the target density at the same time, then we can
make a very rough approximation of that density by taking the current
states of all of the processes as a representative sample. In the
PAIS, the proposal density for the importance sampler stage is a
mixture of all of the proposal densities for a chosen MCMC scheme,
using the current states. If the proposal in the MCMC scheme is
appropriate and our number of samples is big enough, we might hope
that this mixture of densities is a good enough approximation of the
target density, and that new samples drawn from
the mixture proposal are approximately draws from the target
measure. If this is the case, then the variance of importance weights
for these samples will be small, leading to a highly efficient
sampler.

The problem at this stage, is that our current states all have varying
importance, but in order for us to iterate this whole process again,
we require them to be the same. Therefore, we employ a resampler to
take our new set of samples with differing weights, and return a set
of states with equal weights, which are chosen so that the correlation
between the two sets is maximised. At this stage we can start the whole
process again. The algorithm is given in more detail in Table \ref{tab:PAIS}.

We wish to sample states $x \in X$ from a posterior
probability distribution $\mu_Y$. Since we have $M$ processes, we
represent the current state of all of the Markov chains as a vector
$\X = [x_1,x_2,\ldots,x_M]^T$. We are also given a transition kernel
$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example.

\begin{table}
\begin{mdframed}
\begin{algorithmic}
\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
\FOR{$i=0,1,2, ..., N$}
\STATE $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^T, \quad y_j^{(i)} \sim
\nu(\cdot;x_j^{(i)})$
\STATE $\chi(y;\X^{(i)}) = \frac{1}{M}
\sum_{j=1}^M \nu(y;x_j^{(i)})$.
\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
\STATE Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
\ENDFOR 
\end{algorithmic}
\end{mdframed}

\caption{A pseudo-code representation of the Parallel Adaptive
  Importance Sampler (PAIS).}
\label{tab:PAIS}
\end{table}

Since the resampling does not give us a statistically identical sample
to that which is inputted, we cannot assume that the samples $\X^{(i)}$
are samples from the posterior. Therefore, as with serial
importance samplers, the weighted samples $(\W^{(i)},\Y^{(i)})_{i=1}^N$ are stored.

The key here is to choose a suitable transition kernel $\nu$ such that
if $X^{(i)}$ is a decent respresentative sample of the posterior,
the mixture density $\chi(\cdot;\X^{(i)})$ is a reasonable
approximation of the posterior distribution. If this is the case,
the newly proposed states $\Y^{(i)}$ will also be a good (and relatively
independent) sample of the posterior with low variance in the
weights $\W^{(i)}$.

In Section \ref{sec:num}, we will demonstrate how the algorithm
performs, using pCNL algorithm. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
processes can improve on the original MCMC algorithm and lead to
convergence in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost.

\section{Automated tuning of Algorithm Parameters}\label{sec:adap_delta}

Efficient selection of scaling parameters in MCMC algorithms is critical to achieving optimal mixing rates and hence achieving fast convergence to the target density. One significant difficulty with this approach, as with many kernel
approximation approaches, is finding an appropriate $\nu$ such that
$\chi$ is a close approximation to the posterior density $\pi$ in the
right situation. If the proposal distribution is too overdispersed, then the algorithm will propose states a long way from the current state but there will be a relatively low acceptance rate, impacting the quality of inferences on the posterior. Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the space so the mixing rate, and hence convergence rate, will be slow. It is therefore necessary to find a proposal distribution which is slightly overdispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.

Proposal distributions which are slightly over-dispersed as described above, can be found by tuning the variance of the proposal distribution during the burn-in phase of the algorithm. Algorithms which use this method to find optimal proposal distributions are known as adaptive MCMC algorithms. Adaptive MCMC algorithms will converge to the stationary distribution $\pi(\cdot)$, irrespective of whether the adaptive parameter itself converges to an optimal value, under the following conditions\cite{roberts2007coupling,roberts2009examples}.

Condition 1: The adaptive parameter exhibits diminishing adaptation, which says that the amount of movement in the adaptive parameter decreases with the length of the chain.

Condition 2: The smallest number of steps for which the kernel has sufficiently converged, from an initial state $x$, is finite, by bounded covergence.

Adaptively choosing $\delta$ with a large initial guess allows us to first explore the state space, searching for multiple modes, then reducing $\delta$ to an optimal value for efficient sampling helps combat these problems. This process is effective in the PAIS algorithm since tuning $\delta$ using all of the chains is faster than in a single chain. The resampling step causes the processes to spread out across the regions of high density so that when $\delta$ is reduced for locally efficient sampling, it is not required for the chains to constantly search out new modes.

In many MCMC algorithms such as the Random Walk Metropolis-Hastings (RWMH) algorithm, the optimal scaling parameter can be found by searching for the parameter value which gives an optimal acceptance rate, e.g. for near Gaussian targets we have 23.4\% for RWMH and 57.4\% for MALA\cite{roberts2001optimal}. Unlike Metropolis-Hastings algorithms, the PAIS algorithm does not accept or reject proposed values, so we need another method of measuring the optimality of $\delta$. Section~\ref{sec:statistics} gives some possible methods of tuning $\delta$.

\subsection{Statistics for Determining the Optimal Scaling Parameter}\label{sec:statistics}

\subsubsection{Determining optimal scaling parameter $\delta$ using error analysis}

MCMC algorithms can be assessed by comparing their approximation of the posterior to the analytic distribution. To assess this, a distance metric must be decided on, such as the relative error between the sample moments and the true moments, or the relative L2 error between the true density, $\pi(x|D)$, and the constructed histogram. The relative error in the $m$-th moment is
\[
	\left|\frac{N^{-1}\sum_{i=1}^N \! x_i^m - \mathbb{E}[X^m]}{\mathbb{E}[X^m]}\right|,
\]
where $\{x_i\}_{i=1}^N$ is a sample of size $N$ produced by the algorithm. The relative L2 error between a continuous function to a piecewise constant function, $e$, is
\begin{equation}\label{eqn:L2_error}
	e^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a\right]^2,
\end{equation}
where the points $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional regions the histogram is defined over, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of a bin, and $B_i$ is the value of the $i$th bin.

These statistics are usually not practical for finding optimal values of $\delta$ since they require knowledge of the analytic solution, and require that the algorithm be run for a long time to build up a sufficiantly large sample.

Another related statistic, is to use the variance of the estimate of the first moment $\hat{\mu}$. Assuming that the algorithm is converging to the invariant distribution, convergence will be fastest when the variance of the estimate is minimised. This fact follows from the convergence rate of Monte Carlo algorithms, $\sigma/\sqrt{n}$. The proposal distribution which minimises the variance, $\sigma^2$, of $\hat{\mu}$, is the most desirable. This variance statistic often converges faster than the moment itself, and does not require any knowledge of the true value of the moment.

The following statistics can be used for importance sampling algorithms.

\subsubsection{Determining optimal $\delta$ using the variance of the weights}

Importance samplers assign a weight to each sample they produce, based on a ratio of the posterior to the proposal at that point. This type of sampler is most efficient when the posterior is proportional to the proposal distribution, in this case the weights are constant, and so the variance of the weights, $var(w(y))$, is zero. Hence, the optimal value of $\delta$, is
\[
	\delta^*_{\text{var}} = \arg\min_{\delta} \text{var}(w(y)).
\]
The mean of the estimator $\text{var}(w(y))$ is a smooth function, and so it is used to tune $\delta$ during the burn-in phase of MCMC algorithms. The variance of the estimator can be large, especially far away from the optimal value, so it can take a large number of samples to calculate descent directions.

\subsubsection{Determining optimal $\delta$ using the effective sample size}\label{sec:ess}

The effective sample size, $\neff$, can be used to assess the efficiency of importance samplers. Ideally, in each iteration, we would like all $M$ of our samples to provide us with new information about the posterior distribution. In practise, we are unlikely to achieve a ratio of exactly 1.

The effective sample size can be defined in the following ways:
\[
	\neff = \frac{\left(\sum_{i=1}^M \! w_i\right)^2}{\sum_{i=1}^M \! w_i^2} = \frac{M\E(w)^2}{\E(w^2)} = M\left(1 - \frac{\mbox{var}(w)}{\mathbb{E}(w^2)}\right).
\]
From the last definition we can see that when the variance of the weights is zero, which is our ideal scenario, $\neff = M$. Also, in a neighbourhood around $\delta_{\text{var}}^*$, $\neff$ decreases, when the variance increases. So if $v(w(y))$ can be equal to zero for a particular problem and proposal distribution, maximising the effective sample size is equivalent to minimising the variance of the weights. In the PAIS algorithm we have a dynamic proposal which is perturbed at every iteration. Hence, our proposal distribution cannot be exactly proportional to the posterior.

The statistic $\neff$ converges faster than the variance of the weights, and so is preferable as a means of tuning $\delta$. While $\delta_{\text{var}}^*$ and the optimal value found using the effective sample size, $\delta_{\text{eff}}^*$, coincide when $v(w(y)) = 0$, this is not usually possible, so the methods will find different optimal values of $\delta^*$.

The $\neff$ statistic also has another useful property; if we imagine the algorithm in the burn-in phase, for example, we have $M$ processes in the tail of a Gaussian curve searching for the area of high density. If the processes are evenly spaced, then the particle closest to the mean will have an exponentially higher weight assigned to it. The effective sample size in this scenario will be close to 1. As the algorithm burns in, the processes find the flatter area near the mean so the number of samples contributing information to the posterior will increase. By this argument we can see that rising $\neff$ signals the end of the burn-in period.

\subsubsection{Behaviour of the effective sample size for varying ensemble size}

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{figures/neff-M}
\caption{Left: A contour plot showing how $\delta_{\text{eff}}$ varies with the ensemble size. The black line highlights the optimal value of $\delta$. Right: The value of the effective sample size ratio for the optimal $\delta$ at each ensemble size. i.e. the value of the effective sample size ratio along the black line in the left figure. Set up for this problem is given in Section~\ref{sec:problem 2}.}
\label{fig:neff-M}
\end{center}
\end{figure}

If we are to use the effective sample size ratio, we need to look at how it behaves in different situations. In the next section we see how it behaves for different observation operators. Here we look at how the statistic behaves as we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results for the second problem discussed in \ref{sec:problem 2} when using the PAIS-pCNL algorithm to explore the posterior distribution. The left part of the figure shows that as the ensemble size increases, the value of $\delta$ which gives the optimal effective sample size ratio decreases. This is to be expected since we are trying to fit more kernels into the posterior distribution, so the variance of each kernel needs to be decreased.

\subsection{The Adaptive Algorithm}\label{sec:adapt}

A popular method for an adaptive MCMC algorithm is to view the scaling parameter as a random variable which we can sample during the course of the MCMC iterations. However, it can be slow and we may need an uninformative prior. Alternatively, the parameter may be randomly sampled from some interval at various points during the chain, a benefit being that it allows some exploration of the state space but never converges to an optimal value of the scaling parameter. We choose to use a divide and conquer scheme which optimises the effective sample size. Some more sophisticated examples are described in \cite{roberts2009examples} and \cite{Ji2013adaptive}. The proposed algorithm, is not presented as an optimal strategy but as an example of the benefits of tuning the algorithm using the effective sample size statistic.

In the adaptive algorithm described in Table~\ref{tab:adapt} we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm. We must choose some sequence of iterations at which to update $\delta^{(j)}$, $\{n_k\}_{k=1}$, which can be decided using a sequence in which the terms grow exponentially further apart. The algorithm is given for the PAIS method but a similar method is used to adaptively calculate $\delta^*$ for the pCNL algorithm.

When implemented, there are a number of parameters which must be chosen to allow efficient tuning of $\delta$. Firstly, the initial value of $\delta$ should be chosen so that the chains spread out quickly across the state space. If we always choose $\delta^{(0)}=2$ the chains may miss the high density region completely, so a smaller value of $\delta$ must be chosen. This can be seen for example in the optimisation graph for the problem described in Section~\ref{sec:problem 1}, Figure~\ref{fig:MH1 opt delta}, where there is a sudden jump at $\delta\approx 0.1$.

Another choice to be made is which statistic $T(\delta)$ to optimise. In the examples which follow, we have optimised the naively parallelised pCNL algorithm using the optimal acceptance rate $\hat{\alpha} = 0.75$. This value was found by using the L2 error minimums to calculate $\delta^*$. This means that we are minimising the statistic
\[
	T_{\text{MH}}(\delta) = \left| \frac{N_{\text{acc}}(\delta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\delta)$ is the number of accepted moves and $N_{\text{total}}$ is the total number of samples produced. For the PAIS algorithm, we are maximising the effective sample size as discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\delta) = \neff(\delta)$.

In the numerics that follow in Section \ref{sec:num}, we utilise the
adaptive scheme in Table \ref{tab:adapt} for finding the optimum value of $\delta$ in
the proposal step in the PAIS.

\begin{table}
\begin{mdframed}
\begin{algorithmic}
\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
\STATE Choose $\delta^{(0)} \in (0, 2]$, $\delta_L^{(0)} = 0.99\delta^{(0)}$, $\delta_U^{(0)} = 1.01\delta^{(0)} \wedge 2$.
\FOR{$i=0,1,2, \ldots, N$}
\STATE Sample $y_j^{(i)}~\sim~\nu(\cdot;x_j^{(i)},\delta_L^{(i)})$ for $j=1, \ldots,M/2$.
\STATE Sample $y_j^{(i)}~\sim~\nu(\cdot;x_j^{(i)},\delta_U^{(i)})$ for $j=M/2+1, \ldots, M$.
\STATE $\chi(y;\X^{(i)}) = M^{-1}
\left(\sum_{j=1}^{M/2} \nu(y;x_j^{(i)},\delta_L^{(i)})+\sum_{j=M/2+1}^{M} \nu(y;x_j^{(i)},\delta_U^{(i)})\right)$.
\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
\IF{$i$ is in $\{n_k\}_{k=1}$}
	\STATE For $w_{kj} = w_j^{(k)}$, $N$ iterations since last update,
	\STATE $T_L = NM(\sum_{k=i-N}^N\sum_{j=1}^{M/2} w_{kj})^2/(\sum_{k=i-N}^N\sum_{j=1}^{M/2} w_{kj}^2)$.
	\STATE $T_U = NM(\sum_{k=i-N}^N\sum_{j=M/2+1}^M w_{kj})^2/(\sum_{k=i-N}^N\sum_{j=M/2+1}^M w_{kj}^2)$.
	\STATE $\delta^{(i+1)} = \delta^{(i)} - \Delta\D T$.
\ELSE
	\STATE $\delta^{(i+1)} = \delta^{(i)}$.
\ENDIF
\IF{$i < K$}
	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=1}^{M/2} \rightarrow (\frac{1}{M/2}, x_j^{(i+1)})_{j=1}^{M/2}$
	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=M/2+1}^M \rightarrow (\frac{1}{M/2}, x_j^{(i+1)})_{j=M/2+1}^M$
\ELSE
	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=1}^M \rightarrow (\frac{1}{M}, x_j^{(i+1)})_{j=1}^M$
\ENDIF
\ENDFOR
\end{algorithmic}
\end{mdframed}\caption{A pseudo-code representation of the adaptive
  PAIS algorithm.}
\label{tab:adapt}
\end{table}

\section{Numerical Examples}\label{sec:num}

\subsection{Sampling from a one dimensional Gaussian distribution}\label{sec:problem 1}

In this example we look at how PAIS compares to naively parallelised MCMC. We compare the pCNL algorithm \cite{cotter2013mcmc} with its PAIS variant, the PAIS-pCNL algorithm. We consider several statistics for measuring the efficiency of the PAIS algorithm compared to existing Metropolis-Hastings algorithms when applied to a simple one dimensional Gaussian posterior.

\subsubsection{Target distribution}

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow a Gaussian distribution. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the resulting form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_\Sigma - \frac{1}{2}\big\|u\big\|^2_{\mathcal{T}}\right),
\end{equation}
where $\Sigma$ and $\mathcal{T}$ are the covariances of the observational noise and prior distributions respectively. In the numerics which follow, we choose $\Sigma = \sigma^2I_N$ and $\mathcal{T} = \tau^2I_d$ with $\tau^2 =2$ and $\sigma^2 = 0.1$, and we observe $u=-2.5$ with noise drawn from the prior, resulting in $D = -2.6738662$. These values result in a posterior density in which the vast majority of the density is contained inside the high density region of the prior. This means that it should be straightforward for the algorithm to find the stationary distribution. The Kullback-Leibler (KL) divergence for this problem is $D_{KL}(\mu_Y || \mu_0) = 2.67$.

\subsubsection{Numerical implementation}\label{sec:Implementation P1}

In each of the following simulations, we perform three tasks. First we calculate the optimal value of $\delta$ by optimising the statistics described in Section~\ref{sec:statistics}. Once we have these parameters, we run the algorithms and compare the convergence speeds of the algorithms. Finally, we implement the adaptive algorithms described in Section~\ref{sec:adapt} and compare the convergence of these algorithms with the nonadaptive algorithms.

{\bf (1) Finding the optimal parameters}: To find the optimal parameters we choose 32 values of $\delta$ evenly spaced in $[10^{-5}, 2]$. We run the PAIS-pCNL algorithm for 10,000 iterations and pCNL for 100,000 iterations, each with $M=50$ processes. We took 32 repeats of both algorithms and then used the medians of the statistics to find the optimal parameters.

{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} and Section~\ref{sec:problem 2} for 10 million iterations, again with 50 processes. The algorithms are run using the optimal parameters found in (1). The relative L2 error (Equation~\ref{eqn:L2_error}) is used as a measure of accuracy. The simulation for each algorithm was repeated 24 times.

{\bf (3) Measuring convergence of adaptive algorithms}: We run the adaptive algorithms under the same conditions as the nonadaptive algorithms, and again use the relative L2 error to compare efficiency. The initial value of $\delta$ is given in the discussion of each simulation.

\subsubsection{Numerical optimal values of $\delta$}\label{sec:Optimal values P1}

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{"figures/pCNL1 opt delta"}
\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}.}
\label{fig:P1 opt delta}
\end{center}
\end{figure}

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|c|r|}
	\hline
	Statistic											& pCNL \\ \hline
	$\delta_{\text{L2}}^*$								& 3.7e-3 \\
	$\delta_{\text{var}(\hat{\mu})}^*$					& 5.8e-2 \\
	Acceptance Rate ($\delta_{\text{L2}}^*$)				& 9.9e-1 \\
	Acceptance Rate ($\delta_{\text{var}(\hat{\mu})}^*$)	& 7.4e-1 \\
	\hline
	\end{tabular}
    \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Statistic							& PAIS-pCNL \\ \hline
	$\delta_{\text{eff}}^*$				& 1.5e-2 \\
	$\delta_{\text{var}(w(y))}^*$		& 6.4e-2 \\
	$\delta_{\text{L2}}^*$				& 1.7e-2 \\
	\hline
	\end{tabular}
    \end{minipage}
	\caption{Optimal values of $\delta$ summarised from Figure~\ref{fig:P1 opt delta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
	\label{table:prob1 opt delta}
\end{table}

Figure~\ref{fig:P1 opt delta} (left) shows the two values of $\delta$ which may be optimal for the pCNL algorithm. The first estimate comes from the relative L2 error, and the second comes from the variance of the estimate of the mean. Heuristically, the optimal acceptance rates of the function space algorithms are expected to be slightly higher than their finite dimensional versions. In light of this, it is clear that the variance of the estimate of the mean is a better indicator of the scaling parameter. The results in Figure~\ref{fig:P1 opt delta} are summarised in Table~\ref{table:prob1 opt delta} (left).

Figure~\ref{fig:P1 opt delta} (right) shows the effective sample size ratio compared to the error analysis and the variance of the weights. It is clear to see that the effective sample size does a much better job of finding the optimal value of $\delta$ as given by the error analysis than the variance of the weights does.

\subsubsection{Convergence of pCNL vs PAIS-pCNL}

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{"figures/pCNL1_error"}
\caption{Relative error in the moments (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iterations for problem~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1} (2, 3).}
\label{fig:MH1 L2}
\end{center}
\end{figure}

Figure~\ref{fig:MH1 L2} shows that the PAIS-pCNL algorithm converges to the posterior distribution faster than the standard function space pCNL algorithm, in both L2 error and relative error in the moments. To reach a desired tolerance, after convergence the PAIS-pCNL algorithm requires 40\% fewer iterations than the pCNL algorithm.

The adaptive algorithms are run with initial values of $\delta_0=0.1$ for the APAIS-pCNL algorithm, and $\delta_0=2$ for the ApCNL algorithm. From Figure~\ref{fig:MH1 L2} we can see that the APAIS-pCNL algorithm converges at least as quickly as the PAIS-pCNL algorithm. The ApCNL algorithm initially has trouble converging to the posterior; during the initial burn-in phase, some chains find themselves a long way out in the tails where due to the high gradient they will overshoot the high density region and reject almost all proposed values.

\subsubsection{Scaling of the PAIS algorithm with ensemble size}

Throughout this paper, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales if we were to increase the ensemble size, and if there is some limit below which the algorithm fails. We implement the problem in Section~\ref{sec:problem 1}, using the pCNL algorithm with ensemble sizes ranging from $M=2$ up to $M=494$.

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{"figures/PAIS_saving"}
\caption{Ratio of PAIS-pCNL samples required to reach the same tolerance as the pCNL algorithm.}
\label{fig:PAIS_saving}
\end{center}
\end{figure}

Figure~\ref{fig:PAIS_saving} was produced using the method of finding optimal $\delta$ described in Section~\ref{sec:Implementation P1}(1), then running 32 repeats at each ensemble size. The convergence rates are then found by regressing through the data. The graph is still very noisy but demonstrates that increasing the sample size continues to reduce the number of iterations required. When $M<8$, the algorithm takes a long time to reach stationarity.

\subsection{Sampling from a one dimensional Gaussian with a higher KL divergence}\label{sec:problem 2}

In this example we use the same setup used in Section~\ref{sec:problem 1}. We choose a posterior distribution which has most of its mass far out in the tails of the prior distribution. The KL divergence is $D_{KL}(\mu_Y||\mu_0) = 4.670$. This means that the algorithm will have to ``work harder'' to find the area of high probability in the posterior density.

\subsubsection{Target distribution}

As in the previous example, we use the identity observation operator $\G(u)=u$, which results in the Gaussian posterior s in Equation~\ref{eqn:Gaussian posterior}. However, this time we choose $\sigma^2=0.01$ and $\tau^2=0.01$ so the posterior distribution is $\N(D/(1+\sigma^2/\tau^2), \tau^2\sigma^2/(\sigma^2+\tau^2)) = \N(D/2, 0.005)$ which for large $D$ is a long way out in the tail of the prior with a very small variance. For the simulations which follow we observe a reading of $u=4$, with noise drawn from $\N(0, \sigma^2)$, resulting in $D = 3.9979631942$.

\subsubsection{Numerical optimal values of $\delta$}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{"figures/P2 opt delta"}
\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 2}. The setup is as in Section~\ref{sec:Implementation P1}.}
\label{fig:P2 opt delta}
\end{center}
\end{figure}

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|c|r|}
	\hline
	Statistic											& pCNL \\ \hline
	$\delta_{\text{L2}}^*$								& 8.6e-2 \\
	$\delta_{\text{var}(\hat{\mu})}^*$					& 9.1e-1 \\
	Acceptance Rate ($\delta_{\text{L2}}^*$	)			& 9.9e-1 \\
	Acceptance Rate ($\delta_{\text{var}(\hat{\mu})}^*$)	& 8.1e-1 \\
	\hline
	\end{tabular}
    \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Statistic							& PAIS-pCNL \\ \hline
	$\delta_{\text{eff}}^*$				& 2.6e-1 \\
	$\delta_{\text{var}(w(y))}^*$		& 2.6e-1 \\
	$\delta_{\text{L2}}^*$				& 2.8e-1 \\
	\hline
	\end{tabular}
    \end{minipage}
	\caption{Optimal values of $\delta$ summarised from Figure~\ref{fig:P2 opt delta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
	\label{table:P2 opt delta}
\end{table}

We find the optimal values of $\delta$ using the same methods described in Sections~\ref{sec:Implementation P1} and \ref{sec:Optimal values P1}. The results are displayed in Figure~\ref{fig:P2 opt delta} and Table~\ref{table:P2 opt delta}. There is a huge difference between the two error estimates of $\delta$ for pCNL, although the corresponding variance graph is very flat making it sensitive to Monte Carlo error. The variance of the mean estimate has an 81\% acceptance rate, which is larger than the pCNL acceptance rate found previously.

For the PAIS-pCNL algorithm, it is again clear that the effective sample size ratio is a useful statistic for judging the optimal value of $\delta$. The relative L2 error estimate of $\delta^*$, shown in Table~\ref{table:P2 opt delta}, is slightly higher than the other two estimates, but from the graph there again seems to be a fairly flat wide minimum which is in the same region as the optimal effective sample size ratio.

\subsubsection{Convergence of pCNL vs PAIS-pCNL}

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/pCNL2_error"}
\caption{Relative error in the moments (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iteratons for problem~\ref{sec:problem 2}. The setup is as described in Section~\ref{sec:Implementation P1} (2,3).}
\label{fig:MH2 L2}
\end{center}
\end{figure}

Figure~\ref{fig:MH2 L2} that the PAIS-pCNL algorithm, converges to the posterior distribution faster than the pCNL algorithm. The adaptive algorithms both struggle with the first moment for the first million iterations, but produce better estimates after 10 million iterations. We see that to obtain a given tolerance level, the PAIS-pCNL algorithm requires 35\% fewer iterations than the pCNL algorithm.

\subsection{Sampling from Bimodal Distributions}

In this section we investigate the behaviour of the PAIS algorithm when applied to symmetric bimodal problems. We see that the resampling step allows the algorithm to redistribute chains to new modes as they are found. This means that we expect the number of chains in a mode to be related to the probability mass in that mode. As a result reconstructed posteriors with disproportional modes, as is familiar with the Metropolis-Hastings algorithms, are not produced. We again look at an `easy' problem, BM(1), which has a KL divergence of 0.880, and a `harder' problem, BM(2), which has a KL divergence of 3.647. Problem BM(1) has two modes which are separated by a smaller energy barrier. In BM(2) we increase the distance between the two modes which has the effect of increasing the required energy to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

\subsubsection{Target Distribution}

The following setup is the same for both problems. We consider an observation operator $\G(u) = u^2$, and assign the prior $u \sim \mu_0 = \N(0, \tau^2=0.25)$. We assume that a noisy reading, $D$, is taken according to $D = \G(u) + \varepsilon$, where $\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$. This results in the non-Gaussian posterior
\[
	\pi(u|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|u^2 - D\|^2 - \frac{1}{2\tau^2}\|u\|^2\right).
\]
To create the `easy' problem we say that the true value of $\G(u) = 0.75$, and the `hard' problem is generated using $\G(u) = 2$. In the numerics which follow we draw noise from $\mu_\varepsilon$ to generate our data point, we obtain $D = 0.92131223$ for BM(1) and $D = 1.948664$ for BM(2).

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/posteriors3"}
\caption{Posterior distributions for problem BM(1) and BM(2). Problem BM(1) has a noisy data value of  0.921312 which results in an energy barrier which is relatively easy to cross, as well as their common prior distribution.}
\label{fig:problem 3 posteriors}
\end{center}
\end{figure}

\subsubsection{Numerical Implementation}\label{sec:Implementation P2}

The numerical implementation for most of the following simulations follow the same setup as described in Section~\ref{sec:Implementation P1}. The only exception is that the convergence plots for both adaptive and nonadaptive algorithms are run for $10^6$ iterations instead of $10^7$.

\subsubsection{Calculating values of Optimal $\delta^*$}\label{sec:BM1_opt_delta}

Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for pCNL and we use the effective sample size to find the optimal values for PAIS-pCNL. Table~\ref{table:BM_opt_delta} gives the optimal values of $\delta$ for both problems.

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|c|r|r|}
	\hline
	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$ \\ \hline
	pCNL								& 1.9e-1					& - \\
	PAIS-pCNL							& -						& 3.9e-2\\
	\hline
	\end{tabular}
    \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$	& $\delta^*_{\text{L2}}$ \\ \hline
	pCNL								& 5.8e-2					& - 						& 9.1e-1\\
	PAIS-pCNL							& -						& 2.6e-2 					& 2.6e-2\\
	\hline
	\end{tabular}
    \end{minipage}
	\caption{Optimal values of $\delta$ for BM(1) (left) and BM(2) (right).}
	\label{table:BM_opt_delta}
\end{table}

We expect the PAIS algorithm to generally have higher values of the scaling parameter $\delta$. In BM(1), the pCNL algorithm has the higher value of $\delta$, this is because the prior is more efficient to sample from than any single proposal kernel. The PAIS-pCNL algorithm samples more efficiently from its mixture $\chi$.

Problem BM(2) is much harder than BM(1); transitions between the modes are extremely unlikely. This means that we need to consider the density on two levels; we should consider the overall convergence to two correctly proportioned modes as well as local convergence to smooth modes.

To get correctly proportioned modes with the pCNL algorithm it is important that the chains can transition between the modes. Otherwise mode height will be decided by chance. This means that $\delta$ must be large. The prior distribution $\mu_0$ is not a good approximation of the posterior distribution in this case leading to an inefficient method of sampling.

We can achieve these two regimes in pCNL by tuning $\delta$ using the acceptance rate for local convergence, and by L2 error for global convergence. Similarly in PAIS-pCNL we can use the effective sample size for local convergence, and the L2 error for global convergence.

From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime, meaning that both will result in inefficient sampling. The PAIS-pCNL algorithm manages to sample the detail and the large scale behaviour with the same value of $\delta^*$: a clear advantage to using this algorithm for this problem.

\subsubsection{Convergence of pCNL vs PAIS-pCNL}

We see a significant speed with the PAIS-pCNL algorithm for BM(1). Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates.

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM1_L2"}
\caption{Error analysis for the PAIS-pCNL and pCNL algorithms. The solid blue line, and dashed red line compare the algorithms with fixed optimal scaling parameters, and the blue crossed and magenta crossed lines compare the adaptive algorithms. The setup is as described in Section~\ref{sec:Implementation P2} (2,3).}
\label{fig:BM1_L2}
\end{center}
\end{figure}

We can see that the adaptive algorithms compare closely with the respective nonadaptive algorithms and the improvement PAIS offers remains significant. The PAIS-pCNL algorithm requires 45\% fewer iterations than the standard pCNL algorithm.

For BM(2) the algorithms are run with the global optimal value of $\delta^*$, and with the local optimal value of $\delta^*$. Figure~\ref{fig:BM2_L2} shows that the algorithms using the globally optimal $\delta^*$ convergence diagnostics converge slowly towards the true posterior after a long burn-in period, whereas the algorithms using the local optimal $\delta^*$ converge quickly, but get stuck in one mode meaning that the convergence rate flattens out.

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM2_L2"}
\caption{pCNL and PAIS-pCNL convergence statistics using locally and globally optimal $\delta$. The setup is as described in Section~\ref{sec:Implementation P2} (2).}
\label{fig:BM2_L2}
\end{center}
\end{figure}

The adaptive algorithm as stated in Section~\ref{sec:adapt} is one way of combining the two regimes. Another method uses a small number of `scout' chains with a large $\delta$ to continually search out new modes. Other methods of mode searches are described in \cite{lan2013wormhole}, and the regeneration method is applicable\cite{nummelin1984general}.

Figure~\ref{fig:BM2_AL2} shows the success of the (A)PAIS-pCNL algorithms in converging to the posterior, compared to the (A)pCNL algorithms. We can see that using the pCNL algorithm for this problem would be infeasible.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM2_AL2"}
\caption{Convergence graphs for Problem BM(2), the (A)pCNL and (A)PAIS-pCNL algorithms have been run with optimal $\delta^*$ for the L2 error and the adaptive algorithm described in Section~\ref{sec:adapt}. The setup is as described in Section~\ref{sec:Implementation P2} (3).}
\label{fig:BM2_AL2}
\end{center}
\end{figure}

\subsubsection{A useful property of the PAIS algorithm for multimodal distributions}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM2_suction"}
\caption{This figure demonstrates the redistribution property of the PAIS algorithm. Initially there is one chain in the positive mode, and 49 chains in the negative mode.}
\label{fig:BM2_suction}
\end{center}
\end{figure}

The biggest issue for the Metropolis-Hastings algorithms when sampling from a posterior such as the one in BM(2) is that it is unlikely that the same number of chains will occur in each of the modes, and since there is no interaction between the chains, there is no way to remedy this problem. The PAIS algorithm tackles this problem with its resampling step. The algorithm uses its dynamic kernel to build up an approximation of the posterior at each iteration, and then compares this to the posterior distribution via the weights function. Any large discrepancy in the approximation will result in a large or small weight being assigned to the relevant chain, meaning the chain will either pull other chains towards it or be sucked towards a chain with a larger weight. In this way, the algorithm allows chains to `teleport' to regions of the posterior which are in need of more weight. Figure~\ref{fig:BM2_suction} shows Problem BM(2) with initially 1 chain in the positive mode, and 49 chains in the negative mode. It takes only a handful of iterations for the algorithm to balance out the chains into 25 chains in each mode. The chains switch modes without having to climb the energy gradient in the middle.

\section{Concluding Remarks}

We have explored the application of parallelised MCMC algorithms in low dimensional Inverse problems. We have demonstrated numerically that these algorithms converge faster than the analogous Metropolis-Hastings algorithms, further experimentation with the Random Walk Metropolis-Hastings (RWMH), Metropolis Adjusted Langevin Algorithm (MALA) and preconditioned Crank-Nicolson (pCN) proposals has yielded similar results.

One exception where the proposal distribution makes a difference to
the convergence of the PAIS algorithm is where the KL divergence is
large e.g. the PAIS-pCN algorithm struggles to sample from the
distribution in Section~\ref{sec:problem 2}, while the pCN manages to
produce a sample, albeit with a slow burn-in period. Additionally the
PAIS-RWMH algorithm has no problems sampling from this distribution.

\section{More conclusions to add}

Incomplete PAIS (i.e. each processor only needs last reported value of
the other chains)

Using less than optimal resampler (lower cost, Schefzik, Roman,
Thordis L. Thorarinsdottir, and Tilmann Gneiting. "Uncertainty
quantification in complex simulation models using ensemble copula
coupling." Statistical Science 28.4 (2013): 616-640.) COLIN TO DO THIS

Use of other MCMC proposals (Riemann, HMC, etc)

Importantly, we have compared the efficiency of our parallel scheme
with a naive parallelisation of serial methods. Thus our increase in
efficiency is over and above an $N$-fold increase, where $N$ is the
number of cores or processors at our disposal. Our approach
demonstrates a better-than-linear speed-up with the number
processors/cores used. Thus, our approach is not only embarrasingly
parallel (as coined by Cleve Moler in \cite{moler1986matrix}), but
humiliatingly so.

%It is commonly agreed that is preferable to run one MCMC long chain
%than to run many shorter ones - this is largely due to the burn-in
%period that is required.

\bibliographystyle{siam}
\bibliography{refs}

\end{document}

### 2015-07-02 PAIS.tex
meeting corrections

@@
 
-The effective sample size can be defined in the following ways:
+The effective sample size can be defined in the following way:
 \[
-	\neff = \frac{\left(\sum_{i=1}^M \! w_i\right)^2}{\sum_{i=1}^M \! w_i^2} = \frac{M\E(w)^2}{\E(w^2)} = M\left(1 - \frac{\mbox{var}(w)}{\mathbb{E}(w^2)}\right).
+	\neff = \frac{\left(\sum_{i=1}^M \! w_i\right)^2}{\sum_{i=1}^M \! w_i^2} \approx \frac{M\E(w)^2}{\E(w^2)} = M\left(1 - \frac{\mbox{var}(w)}{\mathbb{E}(w^2)}\right).
 \]
-From the last definition we can see that when the variance of the weights is zero, which is our ideal scenario, $\neff = M$.
+The second two expressions are true when $M\rightarrow\infty$.
+From the last expression we see that when the variance of the weights is zero, which is our ideal scenario, $\neff = M$.
 Also, in a neighbourhood around $\delta_{\text{var}}^*$, $\neff$ decreases, when the variance increases.
@@
 
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/pCNL1 opt delta"}
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.45\textwidth]{"figures/pCNL1o"}
+\includegraphics[width=0.45\textwidth]{"figures/PAISpCNL1o"}
 \caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 1}.
@@
 \label{fig:P1 opt delta}
-\end{center}
 \end{figure}
@@
 
-Figure~\ref{fig:P1 opt delta} (left) shows the two values of $\delta$ which may be optimal for the pCNL algorithm.
+Figure~\ref{fig:P1 opt delta} (left) shows the two values of $\delta$ which may be optimal for the pCNL algorithm, found at the turning points.
 The first estimate comes from the relative L2 error, and the second comes from the variance of the estimate of the mean.
-Heuristically, the optimal acceptance rates of the function space algorithms are expected to be slightly higher than their finite dimensional versions.
-In light of this, it is clear that the variance of the estimate of the mean is a better indicator of the scaling parameter.
+The optimal acceptance rates of the function space algorithms are expected to be slightly higher than their finite dimensional versions.
+Since the L2 error estimate of the histogram gives an acceptance rate which is near to 100\%, we say that the optimal acceptance rate is that as given by the variance of the mean, roughly 75\%.
 The results in Figure~\ref{fig:P1 opt delta} are summarised in Table~\ref{table:prob1 opt delta} (left).
@@
 Figure~\ref{fig:P1 opt delta} (right) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
-It is clear to see that the effective sample size does a much better job of finding the optimal value of $\delta$ as given by the error analysis than the variance of the weights does.
+Although the L2 error graph is noisy, the maximum in the effective sample size is close to the minimum in the L2 error.
+The minimum in the variance of the weights however is a long way away.
+We choose the effective sample size as the best estimator of the optimal scaling parameter because of this, and because the effective sample size statistic converges to a smooth graph faster than either the variance of the weights or the L2 error.
 
@@
 
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/pCNL1_error"}
-\caption{Relative error in the moments (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iterations for problem~\ref{sec:problem 1}.
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.45\textwidth]{"figures/pCNL1l"}
+\includegraphics[width=0.45\textwidth]{"figures/pCNL1m"}
+\caption{Relative error in the first moment (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iterations for problem~\ref{sec:problem 1}.
 The setup is as in Section~\ref{sec:Implementation P1} (2, 3).}
 \label{fig:MH1 L2}
-\end{center}
 \end{figure}
@@
 
-The adaptive algorithms are run with initial values of $\delta_0=0.1$ for the APAIS-pCNL algorithm, and $\delta_0=2$ for the ApCNL algorithm.
+Both adaptive algorithms are run with initial values of $\delta_0=0.1$.
 From Figure~\ref{fig:MH1 L2} we can see that the APAIS-pCNL algorithm converges at least as quickly as the PAIS-pCNL algorithm.
@@
 
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/P2 opt delta"}
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.45\textwidth]{"figures/pCNL2 opt delta"}
+\includegraphics[width=0.45\textwidth]{"figures/PAIS pCNL2 opt delta"}
 \caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 2}.
@@
 \label{fig:P2 opt delta}
-\end{center}
 \end{figure}
@@
 The relative L2 error estimate of $\delta^*$, shown in Table~\ref{table:P2 opt delta}, is slightly higher than the other two estimates, but from the graph there again seems to be a fairly flat wide minimum which is in the same region as the optimal effective sample size ratio.
+The variance of the weights becomes so small in the critical region that we get numerical zeroes, this means that we could not use this method to tune the scaling parameter adaptively.
 
@@
 
-\begin{figure}[htpb]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/pCNL2_error"}
-\caption{Relative error in the moments (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iteratons for problem~\ref{sec:problem 2}.
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.45\textwidth]{"figures/pCNL2l"}
+\includegraphics[width=0.45\textwidth]{"figures/pCNL2m"}
+\caption{Relative error in the first moment (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iteratons for problem~\ref{sec:problem 2}.
 The setup is as described in Section~\ref{sec:Implementation P1} (2,3).}
 \label{fig:MH2 L2}
-\end{center}
 \end{figure}
@@
       \centering
-        \begin{tabular}{|c|r|r|}
+        \begin{tabular}{|l|r|r|}
 	\hline

### 2015-08-05 PAIS.tex
Conclusions done
Conclusions done, now sending to Colin.

@@
 \documentclass[final]{siamltex}
+%test change
 \usepackage{cite}
@@
   naive parallelisation of serial MCMC methods using the same number
-  of processors. This demonstrates the added
+  of processors, for low dimensional problems, and in fact shows
+  better than linear improvements in convergence rates with respect to
+  the number of processors used.
 \end{abstract}
@@
 position\cite{sexton1992hamiltonian}, and function space analogues of
-this have also been proposed\cite{beskos2011hybrid}. Riemannian
+this have also been proposed\cite{beskos2011hybrid}. Riemann
 manifold Monte Carlo methods exploit the Riemann geometry of the
@@
 proposal from the current state of all of the chains in the
-ensemble. In Section \ref{sec:Prelim} we outline some preliminaries,
+ensemble. In Section \ref{Sec:Prelim} we outline some preliminaries,
 including the general set up for Bayesian inverse problems, the
@@
 brief review of the optimal transport resampler,  both of which we will be
-employing within the algorithm. In Section \ref{sec:MCMC}, we explore the common problems with MCMC
-methods, which we intend to address with the development of our new
-framework. We describe this framework in Section \ref{Sec:PAIS}. In
+employing within the algorithm. We describe the Parallel Adaptive
+Importance Sampler (PAIS) in Section \ref{Sec:PAIS}, and in Section
+\ref{Sec:adapt} we describe how algorithmic parameters can be
+automatically tuned to provide optimal convergence. In
 Section \ref{Sec:Num}, we present some numerical experiments which
@@
 
-\section{Preliminaries}
+\section{Preliminaries}\label{Sec:Prelim}
+In this Section we will introduce preliminary topics and algorithms
+that will be referred to throughout the paper.
 \subsection{Bayesian inverse problems}
@@
 
-\subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}
+\subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}\label{Sec:pCNL}
 In recent years, work has been carried out to frame MCMC proposal
@@
 discretisations perform comparably with the original versions in low
-dimensions, so we choose one of these distributions to consider. If gradient information regarding the observation operator is
+dimensions. If gradient information regarding the observation operator is
 available, then a range of MCMC methods are available which exploit this information to improve mixing rates. One
@@
 
-Analysis of this method shows that as the ensemble size increases, the statistics of the evenly weighted sample approach those of the posterior distribution, at least well enough for a proposal distribution as described in Section~\ref{sec:PAIS_alg}.
+Analysis of this method shows that as the ensemble size increases, the statistics of the evenly weighted sample approach those of the posterior distribution, at least well enough for a proposal distribution as described in Section~\ref{Sec:PAIS}.
 The histogram of the evenly weighted sample exhibits small oscillations in the tails of the posterior, and also struggles to deal with discontinuities.
@@
 
-Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm loses its ability to accurately represent the mean of the posterior distribution as the dimension increases.
-Clearly for a simple Gaussian problem, we can not resample in much more than 10 dimensions and hope to recover a representative sample.
+Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm
+loses its ability to accurately preserve the mean of the posterior
+distribution as the dimension increases. Clearly for a simple Gaussian
+problem, we can not resample in much more than 10 dimensions and hope
+to recover a representative sample, without hugely increasing the
+number of particles in the ensemble. As the cost of the ETMC is
+$\mathcal{O}(n^2)$, this does limit the dimension for which this
+approach can be used.
 
@@
 This process is known as the burn-in.
-In a trivially parallelised scenario, each process must perform this process independently, while also tuning the scaling parameter.
+In a trivially parallelised scenario, each process must perform this process independently.
 
@@
 implemented, which utilises the resampler recently suggested by
-Reich\cite{reich2013nonparametric} described in Section \ref{sec:filters}, and the pCNL proposal distribution.
-
-\section{The Parallel Adaptive Importance Sampler \allowbreak (PAIS)}\label{sec:PAIS_alg}
-
-In a trivially parallelised MCMC scheme, if we have enough processes
-all sampling from the target density at the same time, then we can
-make a very rough approximation of that density by taking the current
-states of all of the processes as a representative sample. In the
-PAIS, the proposal density for the importance sampler stage is a
-mixture of all of the proposal densities for a chosen MCMC scheme,
-using the current states. If the proposal in the MCMC scheme is
-appropriate and our number of samples is big enough, we might hope
-that this mixture of densities is a good enough approximation of the
-target density, and that new samples drawn from
-the mixture proposal are approximately draws from the target
-measure. If this is the case, then the variance of importance weights
-for these samples will be small, leading to a highly efficient
-sampler.
-
-The problem at this stage, is that our current states all have varying
-importance, but in order for us to iterate this whole process again,
-we require them to be the same. Therefore, we employ a resampler to
-take our new set of samples with differing weights, and return a set
-of states with equal weights, which are chosen so that the correlation
-between the two sets is maximised. At this stage we can start the whole
-process again. The algorithm is given in more detail in Table \ref{tab:PAIS}.
+Reich\cite{reich2013nonparametric} described in Section
+\ref{sec:filters}, and the pCNL proposal distribution described in
+Section \ref{Sec:pCNL}
+
+\section{The Parallel Adaptive Importance Sampler \allowbreak (PAIS)}\label{Sec:PAIS}
+
+% In a trivially parallelised MCMC scheme, if we have enough processes
+% all sampling from the target density at the same time, then we can
+% make a very rough approximation of that density by taking the current
+% states of all of the processes as a representative sample. In the
+% PAIS, the proposal density for the importance sampler stage is a
+% mixture of all of the proposal densities for a chosen MCMC scheme,
+% using the current states. If the proposal in the MCMC scheme is
+% appropriate and our number of samples is big enough, we might hope
+% that this mixture of densities is a good enough approximation of the
+% target density, and that new samples drawn from
+% the mixture proposal are approximately draws from the target
+% measure. If this is the case, then the variance of importance weights
+% for these samples will be small, leading to a highly efficient
+% sampler.
+
+% The problem at this stage, is that our current states all have varying
+% importance, but in order for us to iterate this whole process again,
+% we require them to be the same. Therefore, we employ a resampler to
+% take our new set of samples with differing weights, and return a set
+% of states with equal weights, which are chosen so that the correlation
+% between the two sets is maximised. At this stage we can start the whole
+% process again.
+
+Important samplers can be a very efficient method for sampling from a
+probability distribution. A proposal density is chosen from which we
+can draw samples, and each sample is assigned a weight given by the
+ratio of the target density and the proposal density at that
+point. They are efficient when the proposal density is concentrated in
+similar areas to the target density, and incredibly inefficient when
+this is not the case. The aim of the PAIS is to use the states of a
+set of parallel MCMC chains to construct a proposal distribution which
+will be as close as possible to the target density. Given enough
+processors, the states of all of these chains at one point
+in time may be reasonably representative sample of the target density.
+
+The proposal distribution could be constructed in many different ways,
+but we choose to use a mixture distribution, made up of MCMC proposals
+(in this paper, specifically the pCNL proposal from Section \ref{Sec:pCNL}). Once the proposal is
+constructed, we can sample a new set of states from the proposal
+distribution, and each is assigned a weight given by the ratio of the target
+density and the proposal mixture distribution density. Assuming that
+our proposal distribution is a good one, then the variance of the
+weights will be small, and we will have many useful samples. Finally, we
+need to create a set of evenly weighted samples which best represent
+this set of weighted samples. This is achieved by implementing the
+ETMC algorithm. Once we once again have a set of evenly weighted
+samples that we believe represents the target distribution well, we
+can iterate the process once again. The algorithm is given in more detail in Table \ref{tab:PAIS}.
 
@@
 $\X = [x_1,x_2,\ldots,x_M]^T$. We are also given a transition kernel
-$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example.
+$\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
+the pCNL proposal presented in Section \ref{Sec:pCNL}.
 
@@
 are samples from the posterior. Therefore, as with serial
-importance samplers, the weighted samples $(\W^{(i)},\Y^{(i)})_{i=1}^N$ are stored.
+importance samplers, the weighted samples
+$(\W^{(i)},\Y^{(i)})_{i=1}^N$, or their statistics, are stored.
 
 The key here is to choose a suitable transition kernel $\nu$ such that
-if $X^{(i)}$ is a decent respresentative sample of the posterior,
+if $X^{(i)}$ is a decent representative sample of the posterior,
 the mixture density $\chi(\cdot;\X^{(i)})$ is a reasonable
@@
 
-In Section \ref{sec:num}, we will demonstrate how the algorithm
+In Section \ref{Sec:Num}, we will demonstrate how the algorithm
 performs, using pCNL algorithm. We do not claim that this choice is optimal, but
@@
 
-\section{Automated tuning of Algorithm Parameters}\label{sec:adap_delta}
+\section{Automated tuning of Algorithm Parameters}\label{Sec:adapt}
 
 Efficient selection of scaling parameters in MCMC algorithms is critical to achieving optimal mixing rates and hence achieving fast convergence to the target density.
-One significant difficulty with this approach, as with many kernel
-approximation approaches, is finding an appropriate $\nu$ such that
-$\chi$ is a close approximation to the posterior density $\pi$ in the
-right situation.
-If the proposal distribution is too overdispersed, then the algorithm will propose states a long way from the current state but there will be a relatively low acceptance rate, impacting the quality of inferences on the posterior.
+One significant difficulty is finding an appropriate $\nu$ such that
+$\chi$ is a close approximation to the posterior density $\pi$.
+If the proposal distribution is too over-dispersed, then the algorithm will propose states a long way from the current state but there will be a relatively low acceptance rate, impacting the quality of inferences on the posterior.
 Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the space so the mixing rate, and hence convergence rate, will be slow.
-It is therefore necessary to find a proposal distribution which is slightly overdispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.
-
-Proposal distributions which are slightly over-dispersed as described above, can be found by tuning the variance of the proposal distribution during the burn-in phase of the algorithm.
-Algorithms which use this method to find optimal proposal distributions are known as adaptive MCMC algorithms.
-Adaptive MCMC algorithms will converge to the stationary distribution $\pi(\cdot)$, irrespective of whether the adaptive parameter itself converges to an optimal value, under the following conditions\cite{roberts2007coupling,roberts2009examples}.
-
-Condition 1: The adaptive parameter exhibits diminishing adaptation, which says that the amount of movement in the adaptive parameter decreases with the length of the chain.
-
-Condition 2: The smallest number of steps for which the kernel has sufficiently converged, from an initial state $x$, is finite, by bounded covergence.
-
-Adaptively choosing $\delta$ with a large initial guess allows us to first explore the state space, searching for multiple modes, then reducing $\delta$ to an optimal value for efficient sampling helps combat these problems.
-This process is effective in the PAIS algorithm since tuning $\delta$ using all of the chains is faster than in a single chain.
-The resampling step causes the processes to spread out across the regions of high density so that when $\delta$ is reduced for locally efficient sampling, it is not required for the chains to constantly search out new modes.
+It is therefore necessary to find a proposal distribution which is slightly over-dispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.
+
+Proposal distributions which are slightly over-dispersed as described
+above, can be found by tuning the variance of the proposal
+distribution $\nu$ during the burn-in phase of the
+algorithm. Algorithms which use this method to find optimal proposal
+distributions are known as adaptive MCMC algorithms. Adaptive MCMC
+algorithms will converge to the stationary distribution $\pi(\cdot)$,
+irrespective of whether the adaptive parameter itself converges to an
+optimal value, under the following
+conditions\cite{roberts2007coupling,roberts2009examples}: 
+\begin{enumerate}
+\item The adaptive parameter exhibits diminishing adaptation, which says that the amount of movement in the adaptive parameter decreases with the length of the chain. 
+\item The smallest number of steps for which the kernel has sufficiently converged, from an initial state $x$, is finite, by bounded convergence.
+\end{enumerate}
+
+The variance of the proposal distributions that are used within the
+proposal mixture distribution plays a key role in how well the
+proposal distribution represents the target distribution, and
+therefore how quickly the algorithm converges. A large over-dispersed
+proposal distribution will sample a lot from the tails of the
+distribution, but may also find other regions of high density with
+respect to the target distribution. A very small variance in the
+proposal distributions will lead to a very rough proposal mixture
+distribution. In practise, we wish to find the happy medium in which
+the proposal distribution is slightly more dispersed than the target
+density, whilst also being a good representation of the regions which
+have high density with respect to the target distribution.
+
+Adaptively choosing the variance of the proposal distributions with a
+large initial guess allows us to first explore the state space
+globally, searching for multiple modes. Reducing the proposal variances to an
+optimal value then allows us to explore each region efficiently. The
+fact that we have multiple chains allows us to assess quickly and
+effectively what the optimal variance of the proposal distributions
+should be.
 
@@
 
-These statistics are usually not practical for finding optimal values of $\delta$ since they require knowledge of the analytic solution, and require that the algorithm be run for a long time to build up a sufficiantly large sample.
-
-Another related statistic, is to use the variance of the estimate of the first moment $\hat{\mu}$.
+These statistics are not practical for finding optimal values of
+$\delta$ since they require knowledge of the analytic solution, and
+require that the algorithm be run for a long time to build up a
+sufficiently large sample. However they can be used to assess the
+ability of other statistical measures to find the optimal proposal
+variances in a controlled setting.
+
+Another related statistic, is the variance of the sample estimate of the first moment $\hat{\mu}$.
 Assuming that the algorithm is converging to the invariant distribution, convergence will be fastest when the variance of the estimate is minimised.
@@
 \]
-The second two expressions are true when $M\rightarrow\infty$.
-From the last expression we see that when the variance of the weights is zero, which is our ideal scenario, $\neff = M$.
-Also, in a neighbourhood around $\delta_{\text{var}}^*$, $\neff$ decreases, when the variance increases.
+The second two expressions are true when $M\rightarrow\infty$. From
+the last expression we see that when the variance of the weights is
+zero, $\neff = M$, which is our ideal scenario.
+In a neighbourhood around $\delta_{\text{var}}^*$, $\neff$ decreases, when the variance increases.
 So if $v(w(y))$ can be equal to zero for a particular problem and proposal distribution, maximising the effective sample size is equivalent to minimising the variance of the weights.
 In the PAIS algorithm we have a dynamic proposal which is perturbed at every iteration.
-Hence, our proposal distribution cannot be exactly proportional to the posterior.
 
@@
 
-If we are to use the effective sample size ratio, we need to look at how it behaves in different situations.
-In the next section we see how it behaves for different observation operators.
+If we are to use the effective sample size ratio as an indicator of
+how to tune the proposal variance, we need to look at how it behaves
+in different situations.
+In Section \ref{Sec:Num} we see how it behaves for different observation operators.
 Here we look at how the statistic behaves as we vary the ensemble size, $M$.
@@
 
-A popular method for an adaptive MCMC algorithm is to view the scaling parameter as a random variable which we can sample during the course of the MCMC iterations.
-However, it can be slow and we may need an uninformative prior.
+A popular approach for adaptive MCMC algorithms is to view the scaling
+parameter as a random variable which we can sample during the course
+of the MCMC iterations. However, it can be slow to converge to the
+optimal value, and we may need an
+uninformative prior for the scaling parameter.
 Alternatively, the parameter may be randomly sampled from some interval at various points during the chain, a benefit being that it allows some exploration of the state space but never converges to an optimal value of the scaling parameter.
@@
 
-In the adaptive algorithm described in Table~\ref{tab:adapt} we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm.
+From hereon in, we will assume that we are using the proposal from the
+pCNL algorithm given in Section \ref{Sec:pCNL}, which is given by:
+\begin{equation}Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
+2\delta\mathcal{C}\nabla \Phi(u)+
+\sqrt{8\delta} W\right], \qquad W \sim \mu_0,
+\end{equation}
+where $X$ is our current state and $Y$ is our proposed state. The
+scaling parameter $\delta$ dictates the variance in the proposal distribution.
+
+ In the adaptive algorithm described in Table~\ref{tab:adapt} in
+ Appendix \ref{Sec:App}, we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm.
 We must choose some sequence of iterations at which to update $\delta^{(j)}$, $\{n_k\}_{k=1}$, which can be decided using a sequence in which the terms grow exponentially further apart.
@@
 
-When implemented, there are a number of parameters which must be chosen to allow efficient tuning of $\delta$.
-Firstly, the initial value of $\delta$ should be chosen so that the chains spread out quickly across the state space.
-If we always choose $\delta^{(0)}=2$ the chains may miss the high density region completely, so a smaller value of $\delta$ must be chosen.
-This can be seen for example in the optimisation graph for the problem described in Section~\ref{sec:problem 1}, Figure~\ref{fig:MH1 opt delta}, where there is a sudden jump at $\delta\approx 0.1$.
-
-Another choice to be made is which statistic $T(\delta)$ to optimise.
+When implemented, there are a number of parameters which must be
+chosen to allow efficient tuning of $\delta$. Firstly, the initial
+value of $\delta$ should be chosen so that the chains spread out
+quickly across the state space. However, if it is chosen to be too
+large, then the method may become unstable and inefficient.
+
+\section{Numerical Examples}\label{Sec:Num}
+
+\subsection{Sampling from a one dimensional Gaussian distribution}\label{sec:problem 1}
+
+In this example we look at how PAIS compares to naively parallelised
+MCMC. We compare the pCNL algorithm \cite{cotter2013mcmc} with its
+PAIS variant, the PAIS-pCNL algorithm. We consider several statistics
+for measuring the efficiency of the PAIS algorithm compared to
+existing Metropolis-Hastings algorithms when applied to a simple one
+dimensional Gaussian posterior.
+
+Since we are comparing against trivially parallelised MCMC algorithms,
+we also need to decide which statistic $T(\delta)$ to optimise for
+these approaches.
 In the examples which follow, we have optimised the naively parallelised pCNL algorithm using the optimal acceptance rate $\hat{\alpha} = 0.75$.
@@
 \]
-where $N_{\text{acc}}(\delta)$ is the number of accepted moves and $N_{\text{total}}$ is the total number of samples produced.
-For the PAIS algorithm, we are maximising the effective sample size as discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\delta) = \neff(\delta)$.
-
-In the numerics that follow in Section \ref{sec:num}, we utilise the
-adaptive scheme in Table \ref{tab:adapt} for finding the optimum value of $\delta$ in
-the proposal step in the PAIS.
+where $N_{\text{acc}}(\delta)$ is the number of accepted moves and
+$N_{\text{total}}$ is the total number of samples produced. For the
+PAIS algorithm, we are maximising the effective sample size as
+discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\delta) =
+\neff(\delta)$.
+
+\subsubsection{Target distribution}
+
+Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions.
+Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the resulting form
+\begin{equation}\label{eqn:Gaussian posterior}
+	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_\Sigma - \frac{1}{2}\big\|u\big\|^2_{\mathcal{T}}\right),
+\end{equation}
+where $\Sigma$ and $\mathcal{T}$ are the covariances of the
+observational noise and prior distributions respectively. In the
+numerics which follow, we choose $\Sigma = \sigma^2I_N$ and
+$\mathcal{T} = \tau^2I_d$ with $\tau^2 =2$ and $\sigma^2 = 0.1$, and
+we observe $u=-2.5$ noisily such that
+\[
+D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\Sigma).
+\]
+These values result in a posterior density in which the vast majority
+of the density is contained inside the high density region of the
+prior. This means that it should be straightforward for the algorithm
+to find the stationary distribution. The Kullback-Leibler (KL)
+divergence, which gives us a measure of how different the prior and
+posterior are, is $D_{KL}(\mu_Y || \mu_0) = 2.67$ for this problem.
+
+\subsubsection{Numerical implementation}\label{sec:Implementation P1}
+
+In each of the following simulations, we perform three tasks.
+First we calculate the optimal value of $\delta$ by optimising the statistics described in Section~\ref{sec:statistics}.
+Once we have these parameters, we run the algorithms and compare the convergence speeds of the algorithms.
+Finally, we implement the adaptive algorithms described in Section~\ref{sec:adapt} and compare the convergence of these algorithms with the nonadaptive algorithms.
+
+{\bf (1) Finding the optimal parameters}: To find the optimal
+parameters we choose 32 values of $\delta$ evenly spaced on a log scale
+between $[10^{-5}, 2]$.
+We run the PAIS-pCNL algorithm for 10,000 iterations and pCNL for 100,000 iterations, each with $M=50$ processes.
+We took 32 repeats of both algorithms and then used the medians of the statistics to find the optimal parameters.
+
+{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} and Section~\ref{sec:problem 2} for 10 million iterations, again with 50 processes.
+The algorithms are run using the optimal parameters found in (1).
+The relative L2 error (Equation~\ref{eqn:L2_error}) is used as a measure of accuracy.
+The simulation for each algorithm was repeated 24 times.
+
+{\bf (3) Measuring convergence of adaptive algorithms}: We run the
+adaptive algorithms under the same conditions as the nonadaptive
+algorithms, and again use the relative L2 error to compare
+efficiency. The initial value of $\delta$ is given in the discussion
+of each simulation.
+
+\subsubsection{Optimal values of $\delta$}\label{sec:Optimal values P1}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.45\textwidth]{"figures/pCNL1o"}
+\includegraphics[width=0.45\textwidth]{"figures/PAISpCNL1o"}
+\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 1}.
+The setup is as in Section~\ref{sec:Implementation P1}.}
+\label{fig:P1 opt delta}
+\end{figure}
+
+\begin{table}[!htb]
+    \begin{minipage}{.5\linewidth}
+      \centering
+        \begin{tabular}{|c|r|}
+	\hline
+	Statistic											& pCNL \\ \hline
+	$\delta_{\text{L2}}^*$								& 3.7e-3 \\
+	$\delta_{\text{var}(\hat{\mu})}^*$					& 5.8e-2 \\
+	Acceptance Rate ($\delta_{\text{L2}}^*$)				& 9.9e-1 \\
+	Acceptance Rate ($\delta_{\text{var}(\hat{\mu})}^*$)	& 7.4e-1 \\
+	\hline
+	\end{tabular}
+    \end{minipage}%
+    \begin{minipage}{.5\linewidth}
+      \centering
+        \begin{tabular}{|l|r|r|}
+	\hline
+	Statistic							& PAIS-pCNL \\ \hline
+	$\delta_{\text{eff}}^*$				& 1.5e-2 \\
+	$\delta_{\text{var}(w(y))}^*$		& 6.4e-2 \\
+	$\delta_{\text{L2}}^*$				& 1.7e-2 \\
+	\hline
+	\end{tabular}
+    \end{minipage}
+	\caption{Optimal values of $\delta$ summarised from Figure~\ref{fig:P1 opt delta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
+	\label{table:prob1 opt delta}
+\end{table}
+
+Figure~\ref{fig:P1 opt delta} (left) shows the two values of $\delta$ which may be optimal for the pCNL algorithm, found at the turning points.
+The first estimate comes from the relative L2 error, and the second comes from the variance of the estimate of the mean.
+The optimal acceptance rates of the function space algorithms are expected to be slightly higher than their finite dimensional versions.
+Since the L2 error estimate of the histogram gives an acceptance rate which is near to 100\%, we say that the optimal acceptance rate is that as given by the variance of the mean, roughly 75\%.
+The results in Figure~\ref{fig:P1 opt delta} are summarised in Table~\ref{table:prob1 opt delta} (left).
+
+Figure~\ref{fig:P1 opt delta} (right) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
+Although the L2 error graph is noisy, the maximum in the effective sample size is close to the minimum in the L2 error.
+The minimum in the variance of the weights however is a long way away.
+We choose the effective sample size as the best estimator of the optimal scaling parameter because of this, and because the effective sample size statistic converges to a smooth graph faster than either the variance of the weights or the L2 error.
+
+\subsubsection{Convergence of pCNL vs PAIS-pCNL}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.45\textwidth]{"figures/pCNL1l"}
+\includegraphics[width=0.45\textwidth]{"figures/pCNL1m"}
+\caption{Relative error in the first moment (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iterations for problem~\ref{sec:problem 1}.
+The setup is as in Section~\ref{sec:Implementation P1} (2, 3).}
+\label{fig:MH1 L2}
+\end{figure}
+
+Figure~\ref{fig:MH1 L2} shows that the PAIS-pCNL algorithm converges to the posterior distribution faster than the standard function space pCNL algorithm, in both L2 error and relative error in the moments.
+To reach a desired tolerance, the PAIS-pCNL algorithm requires 40\% fewer iterations than the pCNL algorithm.
+
+Both adaptive algorithms are run with initial values of $\delta_0=0.1$.
+From Figure~\ref{fig:MH1 L2} we can see that the APAIS-pCNL algorithm converges at least as quickly as the PAIS-pCNL algorithm.
+The ApCNL algorithm initially has trouble converging to the posterior; during the initial burn-in phase, some chains find themselves a long way out in the tails where due to the high gradient they will overshoot the high density region and reject almost all proposed values.
+
+\subsubsection{Scaling of the PAIS algorithm with ensemble size}
+
+Throughout this paper, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales if we were to increase the ensemble size, and if there is some limit below which the algorithm fails.
+We implement the problem in Section~\ref{sec:problem 1}, using the pCNL algorithm with ensemble sizes ranging from $M=2$ up to $M=494$.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=\textwidth]{"figures/PAIS_saving"}
+\caption{Ratio of PAIS-pCNL samples required to reach the same tolerance as the pCNL algorithm.}
+\label{fig:PAIS_saving}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:PAIS_saving} was produced using the method of finding
+optimal $\delta$ described in Section~\ref{sec:Implementation P1}(1),
+then running 32 repeats at each ensemble size. The convergence rates
+are then found by regressing through the data. The graph is still very
+noisy but demonstrates that increasing the ensemble size continues to
+reduce the number of iterations required in comparison with naive
+MCMC. When $M<8$, the algorithm takes a long time to reach
+stationarity. This indicates superlinear improvement of PAIS with
+respect to ensemble size, in terms of the number of iterations
+required, which is a demonstration of our belief that parallelism of
+MCMC should give us added value over and above that provided by naive parallelism.
+
+\subsection{Sampling from a one dimensional Gaussian with a higher KL divergence}\label{sec:problem 2}
+
+In this example we use the same setup used in Section~\ref{sec:problem 1}.
+We choose a posterior distribution which has most of its mass far out in the tails of the prior distribution.
+The KL divergence is $D_{KL}(\mu_Y||\mu_0) = 4.670$.
+This means that the algorithm will have to ``work harder'' to find the area of high probability in the posterior density.
+
+\subsubsection{Target distribution}
+
+As in the previous example, we use the identity observation operator
+$\G(u)=u$, which results in the Gaussian posteriors in
+Equation~\ref{eqn:Gaussian posterior}. However, this time we choose
+$\sigma^2=0.01$ and $\tau^2=0.01$ so the posterior distribution is
+$\N(D/(1+\sigma^2/\tau^2), \tau^2\sigma^2/(\sigma^2+\tau^2)) = \N(D/2,
+0.005)$ which for large $D$ is a long way out in the tail of the prior
+with a very small variance. For the simulations which follow we
+observe a reading of $u=4$, with observational noise drawn from $\N(0, \sigma^2)$.
+
+\subsubsection{Optimal values of $\delta$}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.45\textwidth]{"figures/pCNL2 opt delta"}
+\includegraphics[width=0.45\textwidth]{"figures/PAIS pCNL2 opt delta"}
+\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 2}.
+The setup is as in Section~\ref{sec:Implementation P1}.}
+\label{fig:P2 opt delta}
+\end{figure}
+
+\begin{table}[!htb]
+    \begin{minipage}{.5\linewidth}
+      \centering
+        \begin{tabular}{|c|r|}
+	\hline
+	Statistic											& pCNL \\ \hline
+	$\delta_{\text{L2}}^*$								& 8.6e-2 \\
+	$\delta_{\text{var}(\hat{\mu})}^*$					& 9.1e-1 \\
+	Acceptance Rate ($\delta_{\text{L2}}^*$	)			& 9.9e-1 \\
+	Acceptance Rate ($\delta_{\text{var}(\hat{\mu})}^*$)	& 8.1e-1 \\
+	\hline
+	\end{tabular}
+    \end{minipage}%
+    \begin{minipage}{.5\linewidth}
+      \centering
+        \begin{tabular}{|l|r|r|}
+	\hline
+	Statistic							& PAIS-pCNL \\ \hline
+	$\delta_{\text{eff}}^*$				& 2.6e-1 \\
+	$\delta_{\text{var}(w(y))}^*$		& 2.6e-1 \\
+	$\delta_{\text{L2}}^*$				& 2.8e-1 \\
+	\hline
+	\end{tabular}
+    \end{minipage}
+	\caption{Optimal values of $\delta$ summarised from Figure~\ref{fig:P2 opt delta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
+	\label{table:P2 opt delta}
+\end{table}
+
+We find the optimal values of $\delta$ using the same methods described in Sections~\ref{sec:Implementation P1} and \ref{sec:Optimal values P1}.
+The results are displayed in Figure~\ref{fig:P2 opt delta} and Table~\ref{table:P2 opt delta}.
+There is a huge difference between the two error estimates of $\delta$ for pCNL, although the corresponding variance graph is very flat making it sensitive to Monte Carlo error.
+The variance of the mean estimate has an 81\% acceptance rate, which is larger than the pCNL acceptance rate found previously.
+
+For the PAIS-pCNL algorithm, it is again clear that the effective sample size ratio is a useful statistic for judging the optimal value of $\delta$.
+The relative L2 error estimate of $\delta^*$, shown in Table~\ref{table:P2 opt delta}, is slightly higher than the other two estimates, but from the graph there again seems to be a fairly flat wide minimum which is in the same region as the optimal effective sample size ratio.
+The variance of the weights becomes so small in the critical region that we get numerical zeroes, which means that we could not use this method to tune the scaling parameter adaptively.
+
+\subsubsection{Convergence of pCNL vs PAIS-pCNL}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.45\textwidth]{"figures/pCNL2l"}
+\includegraphics[width=0.45\textwidth]{"figures/pCNL2m"}
+\caption{Relative error in the first moment (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iterations for problem~\ref{sec:problem 2}.
+The setup is as described in Section~\ref{sec:Implementation P1} (2,3).}
+\label{fig:MH2 L2}
+\end{figure}
+
+Figure~\ref{fig:MH2 L2} shows that the PAIS-pCNL algorithm, converges to the posterior distribution faster than the pCNL algorithm.
+The adaptive algorithms both struggle with the first moment for the first million iterations, but produce better estimates after 10 million iterations.
+We see that to obtain a given tolerance level, the PAIS-pCNL algorithm requires 35\% fewer iterations than the pCNL algorithm.
+
+\subsection{Sampling from Bimodal Distributions}
+
+In this section we investigate the behaviour of the PAIS algorithm
+when applied to bimodal problems. Metropolis-Hastings methods can
+struggle with multimodal problems, particularly where switches between
+the modes are rare, resulting incorrectly proportioned modes in the
+histograms produced. With the PAIS algorithm, we see that the
+resampling step allows redistribute chains to new modes as they are
+found. This means that we expect the number of chains in a mode to be
+approximately proportional to the probability mass in that mode.
+As a result, reconstructed posteriors with disproportional modes, as is familiar with the Metropolis-Hastings algorithms, are not produced.
+We again look at an `easy' problem, BM(1), which has a KL divergence of 0.880, and a `harder' problem, BM(2), which has a KL divergence of 3.647.
+Problem BM(1) has two modes which are separated by a smaller energy barrier.
+In BM(2) we increase the distance between the two modes which has the effect of increasing the required energy to jump between modes.
+These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.
+
+\subsubsection{Target Distribution}
+
+The following setup is the same for both problems.
+We consider an observation operator $\G(u) = u^2$, and assign the prior $u \sim \mu_0 = \N(0, \tau^2=0.25)$.
+We assume that a noisy reading, $D$, is taken according to $D = \G(u) + \varepsilon$, where $\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$.
+This results in the non-Gaussian posterior
+\[
+	\pi(u|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|u^2 - D\|^2 - \frac{1}{2\tau^2}\|u\|^2\right).
+\]
+To create the `easy' problem we say that the true value of $\G(u) = 0.75$, and the `hard' problem is generated using $\G(u) = 2$.
+In the numerics which follow we draw noise from $\mu_\varepsilon$ to generate our data point.
+
+\begin{figure}[htpb]
+\begin{center}
+\includegraphics[width=\textwidth]{"figures/posteriors3"}
+\caption{Posterior distributions for problem BM(1) and BM(2).
+Problem BM(1) has a noisy data value of  0.921312 which results in an energy barrier which is relatively easy to cross, as well as their common prior distribution.}
+\label{fig:problem 3 posteriors}
+\end{center}
+\end{figure}
+
+\subsubsection{Numerical Implementation}\label{sec:Implementation P2}
+
+The numerical implementation for most of the following simulations follow the same setup as described in Section~\ref{sec:Implementation P1}.
+The only exception is that the convergence plots for both adaptive and nonadaptive algorithms are run for $10^6$ iterations instead of $10^7$.
+
+\subsubsection{Calculating values of Optimal $\delta^*$}\label{sec:BM1_opt_delta}
+
+Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for pCNL and we use the effective sample size to find the optimal values for PAIS-pCNL.
+Table~\ref{table:BM_opt_delta} gives the optimal values of $\delta$ for both problems.
+
+\begin{table}[!htb]
+    \begin{minipage}{.5\linewidth}
+      \centering
+        \begin{tabular}{|l|r|r|}
+	\hline
+	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$ \\ \hline
+	pCNL								& 1.9e-1					& - \\
+	PAIS-pCNL							& -						& 3.9e-2\\
+	\hline
+	\end{tabular}
+    \end{minipage}%
+    \begin{minipage}{.5\linewidth}
+      \centering
+        \begin{tabular}{|l|r|r|r|}
+	\hline
+	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$	& $\delta^*_{\text{L2}}$ \\ \hline
+	pCNL								& 5.8e-2					& - 						& 9.1e-1\\
+	PAIS-pCNL							& -						& 2.6e-2 					& 2.6e-2\\
+	\hline
+	\end{tabular}
+    \end{minipage}
+	\caption{Optimal values of $\delta$ for BM(1) (left) and BM(2) (right).}
+	\label{table:BM_opt_delta}
+\end{table}
+
+In problem BM(1), the pCNL algorithm has the
+higher value of $\delta$, which corresponds to a close to independence
+sampling, effectively sampling from the prior. This is because the
+regions with the majority of the target density are well covered by
+the prior. The PAIS-pCNL algorithm samples more efficiently from its
+mixture $\chi$, and so a lower value of $\delta$ is more efficient.
+
+Problem BM(2) is much harder than BM(1); transitions between the modes
+are extremely unlikely for the standard pCNL algorithm. This means that we need to consider the
+convergence on two levels; we should consider the algorithm's ability
+to find all the modes, and to sample them thoroughly and in the correct proportions.
+
+To get correctly proportioned modes with the pCNL algorithm it is
+important that the chains can transition between the modes, which
+means that $\delta$ must be large. However, this leads to a lower
+acceptance rate, and so we sacrifice convergence locally. The prior distribution $\mu_0$ is
+not a good approximation of the posterior distribution, and so too
+large a value of $\delta$ (which leads to an independence sampler
+using the prior as a proposal distribution) leads to an inefficient
+method sampling. For these reasons, the pCNL algorithm is very slow to
+converge for problems of this type.
+
+We can achieve these two regimes in pCNL by tuning $\delta$ using the acceptance rate for local convergence, and by L2 error for global convergence.
+Similarly in PAIS-pCNL we can use the effective sample size for local convergence, and the L2 error for global convergence.
+
+From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime, meaning that both will result in inefficient sampling.
+The PAIS-pCNL algorithm manages to sample the detail and the large scale behaviour with the same value of $\delta^*$: a clear advantage to using this algorithm for this problem.
+
+\subsubsection{Convergence of pCNL vs PAIS-pCNL}
+
+We see a significant speed with the PAIS-pCNL algorithm for BM(1). Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates.
+
+\begin{figure}[htpb]
+\begin{center}
+\includegraphics[width=\textwidth]{"figures/BM1_L2"}
+\caption{Error analysis for the PAIS-pCNL and pCNL algorithms for
+  problem BM(1).
+The solid blue line, and dashed red line compare the algorithms with fixed optimal scaling parameters, and the blue crossed and magenta crossed lines compare the adaptive algorithms.
+The setup is as described in Section~\ref{sec:Implementation P2} (2,3).}
+\label{fig:BM1_L2}
+\end{center}
+\end{figure}
+
+We can see that the adaptive algorithms compare closely with the
+respective nonadaptive algorithms and the improvement PAIS offers
+remains significant. The PAIS-pCNL algorithm requires 45\% fewer
+iterations than the naively parallelised pCNL algorithm.
+
+For BM(2) the algorithms are run with the global optimal value of $\delta^*$, and with the local optimal value of $\delta^*$.
+Figure~\ref{fig:BM2_L2} shows that the algorithms using the globally optimal $\delta^*$ convergence diagnostics converge slowly towards the true posterior after a long burn-in period, whereas the algorithms using the local optimal $\delta^*$ converge quickly, but get stuck in one mode meaning that the convergence rate flattens out.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=\textwidth]{"figures/BM2_L2"}
+\caption{pCNL and PAIS-pCNL convergence statistics using locally and
+  globally optimal $\delta$, for problem BM(2). The setup is as described in Section~\ref{sec:Implementation P2} (2).}
+\label{fig:BM2_L2}
+\end{center}
+\end{figure}
+
+The adaptive algorithm as stated in Section~\ref{sec:adapt} is one way of combining the two regimes.
+Another method uses a small number of `scout' chains with a large $\delta$ to continually search out new modes.
+Other methods of mode searches are described in \cite{lan2013wormhole}, and the regeneration method is applicable\cite{nummelin1984general}.
+
+Figure~\ref{fig:BM2_AL2} shows the success of the (A)PAIS-pCNL algorithms in converging to the posterior, compared to the (A)pCNL algorithms.
+We can see that using the pCNL algorithm for this problem would be infeasible.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{"figures/BM2_AL2"}
+\caption{Convergence graphs for problem BM(2), the (A)pCNL and (A)PAIS-pCNL algorithms have been run with optimal $\delta^*$ for the L2 error and the adaptive algorithm described in Section~\ref{sec:adapt}.
+The setup is as described in Section~\ref{sec:Implementation P2} (3).}
+\label{fig:BM2_AL2}
+\end{center}
+\end{figure}
+
+\subsubsection{A useful property of the PAIS algorithm for multimodal distributions}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{"figures/BM2_suction"}
+\caption{This figure demonstrates the redistribution property of the PAIS algorithm. Initially there is one chain in the positive mode, and 49 chains in the negative mode.}
+\label{fig:BM2_suction}
+\end{center}
+\end{figure}
+
+The biggest issue for the Metropolis-Hastings algorithms when sampling
+from a posterior such as the one in BM(2) is that it is unlikely that
+the correct ratio of chains will occur in each of the modes, and since there is no interaction between the chains, there is no way to remedy this problem.
+The PAIS algorithm tackles this problem with its resampling step.
+The algorithm uses its dynamic kernel to build up an approximation of the posterior at each iteration, and then compares this to the posterior distribution via the weights function.
+Any large discrepancy in the approximation will result in a large or small weight being assigned to the relevant chain, meaning the chain will either pull other chains towards it or be sucked towards a chain with a larger weight.
+In this way, the algorithm allows chains to `teleport' to regions of the posterior which are in need of more exploration.
+Figure~\ref{fig:BM2_suction} shows Problem BM(2) with initially 1 chain in the positive mode, and 49 chains in the negative mode.
+It takes only a handful of iterations for the algorithm to balance out the chains into 25 chains in each mode.
+The chains switch modes without having to climb the energy gradient in the middle.
+
+\section{Discussion and Conclusions}\label{Sec:Conc}
+
+We have explored the application of parallelised MCMC algorithms in
+low dimensional Inverse problems. We have demonstrated numerically
+that these algorithms converge faster than the analogous naively parallelised
+Metropolis-Hastings algorithms. Further experimentation with the
+Random Walk Metropolis-Hastings (RWMH), Metropolis Adjusted Langevin
+Algorithm (MALA) and preconditioned Crank-Nicolson (pCN) proposals has
+yielded similar results\cite{Paul}.%  Importantly, we have demonstrated
+% that the convergence rate improves superlinearly with respect to the
+% number of processors used.
+
+Importantly, we have compared the efficiency of our parallel scheme
+with a naive parallelisation of serial methods. Thus our increase in
+efficiency is over and above an $N$-fold increase, where $N$ is the
+number of cores or processors at our disposal. Our approach
+demonstrates a better-than-linear speed-up with the number
+processors/cores used. Thus, our approach is not only embarrassingly
+parallel (as coined by Cleve Moler in \cite{moler1986matrix}), but
+humiliatingly so.
+
+The PAIS has a number of favourable features, for example the
+algorithm's ability to redistribute, through the resampling regime,
+the chains to regions which require more exploration. This allows the
+method to be used to sample from complex multimodal distribution.
+
+The PAIS approach is limited by the curse of dimensionality, since the
+cost of the resampling algorithm becomes prohibitively large. The cost
+of the resampler is $\mathcal{O}(n^2)$, where $n$ is the number of
+particles in the ensemble. Other resamplers exist which are not as
+costly, but for which there is some sacrifice in terms of optimality
+of the output. For example, the resampler proposed in
+\cite{schefzik2013uncertainty}  has computational complexity
+$\mathcal{O}(n)$. This might allow the PAIS to be used in higher
+dimensional problems, and is an avenue for future investigation.
+
+Another strength of the PAIS is that it can also be used with any MCMC
+proposal. There are a growing number of increasing sophisticated MCMC
+algorithms (HMC, Riemann manifold MCMC etc) which could be incorporated into this framework, leading to
+even more efficient algorithms, and this is another opportunity for
+future work.
+
+One disadvantage of parallelised algorithms is that often different
+processors will complete their tasks in different amount of times, for
+a number of reasons, often due to communication between the
+processors. One approach which could be applied to the PAIS to avoid
+this is for each processor to immediately start a new iteration, only
+using the last values of the other processors that were communicated
+to it. This incomplete approach would still be valid, and could lead
+to more efficient use of the computer architecture.
+
+\bibliographystyle{siam}
+\bibliography{refs}
+
+\begin{appendix}
+\section{The adaptive PAIS algorithm}\label{Sec:App}
 
@@
 \end{table}
-
-\section{Numerical Examples}\label{sec:num}
-
-\subsection{Sampling from a one dimensional Gaussian distribution}\label{sec:problem 1}
-
-In this example we look at how PAIS compares to naively parallelised MCMC.
-We compare the pCNL algorithm \cite{cotter2013mcmc} with its PAIS variant, the PAIS-pCNL algorithm.
-We consider several statistics for measuring the efficiency of the PAIS algorithm compared to existing Metropolis-Hastings algorithms when applied to a simple one dimensional Gaussian posterior.
-
-\subsubsection{Target distribution}
-
-Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow a Gaussian distribution.
-Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the resulting form
-\begin{equation}\label{eqn:Gaussian posterior}
-	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_\Sigma - \frac{1}{2}\big\|u\big\|^2_{\mathcal{T}}\right),
-\end{equation}
-where $\Sigma$ and $\mathcal{T}$ are the covariances of the observational noise and prior distributions respectively.
-In the numerics which follow, we choose $\Sigma = \sigma^2I_N$ and $\mathcal{T} = \tau^2I_d$ with $\tau^2 =2$ and $\sigma^2 = 0.1$, and we observe $u=-2.5$ with noise drawn from the prior, resulting in $D = -2.6738662$.
-These values result in a posterior density in which the vast majority of the density is contained inside the high density region of the prior.
-This means that it should be straightforward for the algorithm to find the stationary distribution.
-The Kullback-Leibler (KL) divergence for this problem is $D_{KL}(\mu_Y || \mu_0) = 2.67$.
-
-\subsubsection{Numerical implementation}\label{sec:Implementation P1}
-
-In each of the following simulations, we perform three tasks.
-First we calculate the optimal value of $\delta$ by optimising the statistics described in Section~\ref{sec:statistics}.
-Once we have these parameters, we run the algorithms and compare the convergence speeds of the algorithms.
-Finally, we implement the adaptive algorithms described in Section~\ref{sec:adapt} and compare the convergence of these algorithms with the nonadaptive algorithms.
-
-{\bf (1) Finding the optimal parameters}: To find the optimal parameters we choose 32 values of $\delta$ evenly spaced in $[10^{-5}, 2]$.
-We run the PAIS-pCNL algorithm for 10,000 iterations and pCNL for 100,000 iterations, each with $M=50$ processes.
-We took 32 repeats of both algorithms and then used the medians of the statistics to find the optimal parameters.
-
-{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} and Section~\ref{sec:problem 2} for 10 million iterations, again with 50 processes.
-The algorithms are run using the optimal parameters found in (1).
-The relative L2 error (Equation~\ref{eqn:L2_error}) is used as a measure of accuracy.
-The simulation for each algorithm was repeated 24 times.
-
-{\bf (3) Measuring convergence of adaptive algorithms}: We run the adaptive algorithms under the same conditions as the nonadaptive algorithms, and again use the relative L2 error to compare efficiency.
-The initial value of $\delta$ is given in the discussion of each simulation.
-
-\subsubsection{Numerical optimal values of $\delta$}\label{sec:Optimal values P1}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.45\textwidth]{"figures/pCNL1o"}
-\includegraphics[width=0.45\textwidth]{"figures/PAISpCNL1o"}
-\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 1}.
-The setup is as in Section~\ref{sec:Implementation P1}.}
-\label{fig:P1 opt delta}
-\end{figure}
-
-\begin{table}[!htb]
-    \begin{minipage}{.5\linewidth}
-      \centering
-        \begin{tabular}{|c|r|}
-	\hline
-	Statistic											& pCNL \\ \hline
-	$\delta_{\text{L2}}^*$								& 3.7e-3 \\
-	$\delta_{\text{var}(\hat{\mu})}^*$					& 5.8e-2 \\
-	Acceptance Rate ($\delta_{\text{L2}}^*$)				& 9.9e-1 \\
-	Acceptance Rate ($\delta_{\text{var}(\hat{\mu})}^*$)	& 7.4e-1 \\
-	\hline
-	\end{tabular}
-    \end{minipage}%
-    \begin{minipage}{.5\linewidth}
-      \centering
-        \begin{tabular}{|l|r|r|}
-	\hline
-	Statistic							& PAIS-pCNL \\ \hline
-	$\delta_{\text{eff}}^*$				& 1.5e-2 \\
-	$\delta_{\text{var}(w(y))}^*$		& 6.4e-2 \\
-	$\delta_{\text{L2}}^*$				& 1.7e-2 \\
-	\hline
-	\end{tabular}
-    \end{minipage}
-	\caption{Optimal values of $\delta$ summarised from Figure~\ref{fig:P1 opt delta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
-	\label{table:prob1 opt delta}
-\end{table}
-
-Figure~\ref{fig:P1 opt delta} (left) shows the two values of $\delta$ which may be optimal for the pCNL algorithm, found at the turning points.
-The first estimate comes from the relative L2 error, and the second comes from the variance of the estimate of the mean.
-The optimal acceptance rates of the function space algorithms are expected to be slightly higher than their finite dimensional versions.
-Since the L2 error estimate of the histogram gives an acceptance rate which is near to 100\%, we say that the optimal acceptance rate is that as given by the variance of the mean, roughly 75\%.
-The results in Figure~\ref{fig:P1 opt delta} are summarised in Table~\ref{table:prob1 opt delta} (left).
-
-Figure~\ref{fig:P1 opt delta} (right) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
-Although the L2 error graph is noisy, the maximum in the effective sample size is close to the minimum in the L2 error.
-The minimum in the variance of the weights however is a long way away.
-We choose the effective sample size as the best estimator of the optimal scaling parameter because of this, and because the effective sample size statistic converges to a smooth graph faster than either the variance of the weights or the L2 error.
-
-\subsubsection{Convergence of pCNL vs PAIS-pCNL}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.45\textwidth]{"figures/pCNL1l"}
-\includegraphics[width=0.45\textwidth]{"figures/pCNL1m"}
-\caption{Relative error in the first moment (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iterations for problem~\ref{sec:problem 1}.
-The setup is as in Section~\ref{sec:Implementation P1} (2, 3).}
-\label{fig:MH1 L2}
-\end{figure}
-
-Figure~\ref{fig:MH1 L2} shows that the PAIS-pCNL algorithm converges to the posterior distribution faster than the standard function space pCNL algorithm, in both L2 error and relative error in the moments.
-To reach a desired tolerance, after convergence the PAIS-pCNL algorithm requires 40\% fewer iterations than the pCNL algorithm.
-
-Both adaptive algorithms are run with initial values of $\delta_0=0.1$.
-From Figure~\ref{fig:MH1 L2} we can see that the APAIS-pCNL algorithm converges at least as quickly as the PAIS-pCNL algorithm.
-The ApCNL algorithm initially has trouble converging to the posterior; during the initial burn-in phase, some chains find themselves a long way out in the tails where due to the high gradient they will overshoot the high density region and reject almost all proposed values.
-
-\subsubsection{Scaling of the PAIS algorithm with ensemble size}
-
-Throughout this paper, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales if we were to increase the ensemble size, and if there is some limit below which the algorithm fails.
-We implement the problem in Section~\ref{sec:problem 1}, using the pCNL algorithm with ensemble sizes ranging from $M=2$ up to $M=494$.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/PAIS_saving"}
-\caption{Ratio of PAIS-pCNL samples required to reach the same tolerance as the pCNL algorithm.}
-\label{fig:PAIS_saving}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:PAIS_saving} was produced using the method of finding optimal $\delta$ described in Section~\ref{sec:Implementation P1}(1), then running 32 repeats at each ensemble size.
-The convergence rates are then found by regressing through the data.
-The graph is still very noisy but demonstrates that increasing the sample size continues to reduce the number of iterations required.
-When $M<8$, the algorithm takes a long time to reach stationarity.
-
-\subsection{Sampling from a one dimensional Gaussian with a higher KL divergence}\label{sec:problem 2}
-
-In this example we use the same setup used in Section~\ref{sec:problem 1}.
-We choose a posterior distribution which has most of its mass far out in the tails of the prior distribution.
-The KL divergence is $D_{KL}(\mu_Y||\mu_0) = 4.670$.
-This means that the algorithm will have to ``work harder'' to find the area of high probability in the posterior density.
-
-\subsubsection{Target distribution}
-
-As in the previous example, we use the identity observation operator $\G(u)=u$, which results in the Gaussian posterior s in Equation~\ref{eqn:Gaussian posterior}.
-However, this time we choose $\sigma^2=0.01$ and $\tau^2=0.01$ so the posterior distribution is $\N(D/(1+\sigma^2/\tau^2), \tau^2\sigma^2/(\sigma^2+\tau^2)) = \N(D/2, 0.005)$ which for large $D$ is a long way out in the tail of the prior with a very small variance.
-For the simulations which follow we observe a reading of $u=4$, with noise drawn from $\N(0, \sigma^2)$, resulting in $D = 3.9979631942$.
-
-\subsubsection{Numerical optimal values of $\delta$}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.45\textwidth]{"figures/pCNL2 opt delta"}
-\includegraphics[width=0.45\textwidth]{"figures/PAIS pCNL2 opt delta"}
-\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 2}.
-The setup is as in Section~\ref{sec:Implementation P1}.}
-\label{fig:P2 opt delta}
-\end{figure}
-
-\begin{table}[!htb]
-    \begin{minipage}{.5\linewidth}
-      \centering
-        \begin{tabular}{|c|r|}
-	\hline
-	Statistic											& pCNL \\ \hline
-	$\delta_{\text{L2}}^*$								& 8.6e-2 \\
-	$\delta_{\text{var}(\hat{\mu})}^*$					& 9.1e-1 \\
-	Acceptance Rate ($\delta_{\text{L2}}^*$	)			& 9.9e-1 \\
-	Acceptance Rate ($\delta_{\text{var}(\hat{\mu})}^*$)	& 8.1e-1 \\
-	\hline
-	\end{tabular}
-    \end{minipage}%
-    \begin{minipage}{.5\linewidth}
-      \centering
-        \begin{tabular}{|l|r|r|}
-	\hline
-	Statistic							& PAIS-pCNL \\ \hline
-	$\delta_{\text{eff}}^*$				& 2.6e-1 \\
-	$\delta_{\text{var}(w(y))}^*$		& 2.6e-1 \\
-	$\delta_{\text{L2}}^*$				& 2.8e-1 \\
-	\hline
-	\end{tabular}
-    \end{minipage}
-	\caption{Optimal values of $\delta$ summarised from Figure~\ref{fig:P2 opt delta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
-	\label{table:P2 opt delta}
-\end{table}
-
-We find the optimal values of $\delta$ using the same methods described in Sections~\ref{sec:Implementation P1} and \ref{sec:Optimal values P1}.
-The results are displayed in Figure~\ref{fig:P2 opt delta} and Table~\ref{table:P2 opt delta}.
-There is a huge difference between the two error estimates of $\delta$ for pCNL, although the corresponding variance graph is very flat making it sensitive to Monte Carlo error.
-The variance of the mean estimate has an 81\% acceptance rate, which is larger than the pCNL acceptance rate found previously.
-
-For the PAIS-pCNL algorithm, it is again clear that the effective sample size ratio is a useful statistic for judging the optimal value of $\delta$.
-The relative L2 error estimate of $\delta^*$, shown in Table~\ref{table:P2 opt delta}, is slightly higher than the other two estimates, but from the graph there again seems to be a fairly flat wide minimum which is in the same region as the optimal effective sample size ratio.
-The variance of the weights becomes so small in the critical region that we get numerical zeroes, this means that we could not use this method to tune the scaling parameter adaptively.
-
-\subsubsection{Convergence of pCNL vs PAIS-pCNL}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.45\textwidth]{"figures/pCNL2l"}
-\includegraphics[width=0.45\textwidth]{"figures/pCNL2m"}
-\caption{Relative error in the first moment (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iteratons for problem~\ref{sec:problem 2}.
-The setup is as described in Section~\ref{sec:Implementation P1} (2,3).}
-\label{fig:MH2 L2}
-\end{figure}
-
-Figure~\ref{fig:MH2 L2} that the PAIS-pCNL algorithm, converges to the posterior distribution faster than the pCNL algorithm.
-The adaptive algorithms both struggle with the first moment for the first million iterations, but produce better estimates after 10 million iterations.
-We see that to obtain a given tolerance level, the PAIS-pCNL algorithm requires 35\% fewer iterations than the pCNL algorithm.
-
-\subsection{Sampling from Bimodal Distributions}
-
-In this section we investigate the behaviour of the PAIS algorithm when applied to symmetric bimodal problems.
-We see that the resampling step allows the algorithm to redistribute chains to new modes as they are found.
-This means that we expect the number of chains in a mode to be related to the probability mass in that mode.
-As a result reconstructed posteriors with disproportional modes, as is familiar with the Metropolis-Hastings algorithms, are not produced.
-We again look at an `easy' problem, BM(1), which has a KL divergence of 0.880, and a `harder' problem, BM(2), which has a KL divergence of 3.647.
-Problem BM(1) has two modes which are separated by a smaller energy barrier.
-In BM(2) we increase the distance between the two modes which has the effect of increasing the required energy to jump between modes.
-These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.
-
-\subsubsection{Target Distribution}
-
-The following setup is the same for both problems.
-We consider an observation operator $\G(u) = u^2$, and assign the prior $u \sim \mu_0 = \N(0, \tau^2=0.25)$.
-We assume that a noisy reading, $D$, is taken according to $D = \G(u) + \varepsilon$, where $\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$.
-This results in the non-Gaussian posterior
-\[
-	\pi(u|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|u^2 - D\|^2 - \frac{1}{2\tau^2}\|u\|^2\right).
-\]
-To create the `easy' problem we say that the true value of $\G(u) = 0.75$, and the `hard' problem is generated using $\G(u) = 2$.
-In the numerics which follow we draw noise from $\mu_\varepsilon$ to generate our data point, we obtain $D = 0.92131223$ for BM(1) and $D = 1.948664$ for BM(2).
-
-\begin{figure}[htpb]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/posteriors3"}
-\caption{Posterior distributions for problem BM(1) and BM(2).
-Problem BM(1) has a noisy data value of  0.921312 which results in an energy barrier which is relatively easy to cross, as well as their common prior distribution.}
-\label{fig:problem 3 posteriors}
-\end{center}
-\end{figure}
-
-\subsubsection{Numerical Implementation}\label{sec:Implementation P2}
-
-The numerical implementation for most of the following simulations follow the same setup as described in Section~\ref{sec:Implementation P1}.
-The only exception is that the convergence plots for both adaptive and nonadaptive algorithms are run for $10^6$ iterations instead of $10^7$.
-
-\subsubsection{Calculating values of Optimal $\delta^*$}\label{sec:BM1_opt_delta}
-
-Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for pCNL and we use the effective sample size to find the optimal values for PAIS-pCNL.
-Table~\ref{table:BM_opt_delta} gives the optimal values of $\delta$ for both problems.
-
-\begin{table}[!htb]
-    \begin{minipage}{.5\linewidth}
-      \centering
-        \begin{tabular}{|l|r|r|}
-	\hline
-	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$ \\ \hline
-	pCNL								& 1.9e-1					& - \\
-	PAIS-pCNL							& -						& 3.9e-2\\
-	\hline
-	\end{tabular}
-    \end{minipage}%
-    \begin{minipage}{.5\linewidth}
-      \centering
-        \begin{tabular}{|l|r|r|r|}
-	\hline
-	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$	& $\delta^*_{\text{L2}}$ \\ \hline
-	pCNL								& 5.8e-2					& - 						& 9.1e-1\\
-	PAIS-pCNL							& -						& 2.6e-2 					& 2.6e-2\\
-	\hline
-	\end{tabular}
-    \end{minipage}
-	\caption{Optimal values of $\delta$ for BM(1) (left) and BM(2) (right).}
-	\label{table:BM_opt_delta}
-\end{table}
-
-We expect the PAIS algorithm to generally have higher values of the scaling parameter $\delta$.
-In BM(1), the pCNL algorithm has the higher value of $\delta$, this is because the prior is more efficient to sample from than any single proposal kernel.
-The PAIS-pCNL algorithm samples more efficiently from its mixture $\chi$.
-
-Problem BM(2) is much harder than BM(1); transitions between the modes are extremely unlikely.
-This means that we need to consider the density on two levels; we should consider the overall convergence to two correctly proportioned modes as well as local convergence to smooth modes.
-
-To get correctly proportioned modes with the pCNL algorithm it is important that the chains can transition between the modes.
-Otherwise mode height will be decided by chance.
-This means that $\delta$ must be large.
-The prior distribution $\mu_0$ is not a good approximation of the posterior distribution in this case leading to an inefficient method of sampling.
-
-We can achieve these two regimes in pCNL by tuning $\delta$ using the acceptance rate for local convergence, and by L2 error for global convergence.
-Similarly in PAIS-pCNL we can use the effective sample size for local convergence, and the L2 error for global convergence.
-
-From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime, meaning that both will result in inefficient sampling.
-The PAIS-pCNL algorithm manages to sample the detail and the large scale behaviour with the same value of $\delta^*$: a clear advantage to using this algorithm for this problem.
-
-\subsubsection{Convergence of pCNL vs PAIS-pCNL}
-
-We see a significant speed with the PAIS-pCNL algorithm for BM(1). Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates.
-
-\begin{figure}[htpb]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/BM1_L2"}
-\caption{Error analysis for the PAIS-pCNL and pCNL algorithms.
-The solid blue line, and dashed red line compare the algorithms with fixed optimal scaling parameters, and the blue crossed and magenta crossed lines compare the adaptive algorithms.
-The setup is as described in Section~\ref{sec:Implementation P2} (2,3).}
-\label{fig:BM1_L2}
-\end{center}
-\end{figure}
-
-We can see that the adaptive algorithms compare closely with the respective nonadaptive algorithms and the improvement PAIS offers remains significant.
-The PAIS-pCNL algorithm requires 45\% fewer iterations than the standard pCNL algorithm.
-
-For BM(2) the algorithms are run with the global optimal value of $\delta^*$, and with the local optimal value of $\delta^*$.
-Figure~\ref{fig:BM2_L2} shows that the algorithms using the globally optimal $\delta^*$ convergence diagnostics converge slowly towards the true posterior after a long burn-in period, whereas the algorithms using the local optimal $\delta^*$ converge quickly, but get stuck in one mode meaning that the convergence rate flattens out.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/BM2_L2"}
-\caption{pCNL and PAIS-pCNL convergence statistics using locally and globally optimal $\delta$. The setup is as described in Section~\ref{sec:Implementation P2} (2).}
-\label{fig:BM2_L2}
-\end{center}
-\end{figure}
-
-The adaptive algorithm as stated in Section~\ref{sec:adapt} is one way of combining the two regimes.
-Another method uses a small number of `scout' chains with a large $\delta$ to continually search out new modes.
-Other methods of mode searches are described in \cite{lan2013wormhole}, and the regeneration method is applicable\cite{nummelin1984general}.
-
-Figure~\ref{fig:BM2_AL2} shows the success of the (A)PAIS-pCNL algorithms in converging to the posterior, compared to the (A)pCNL algorithms.
-We can see that using the pCNL algorithm for this problem would be infeasible.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/BM2_AL2"}
-\caption{Convergence graphs for Problem BM(2), the (A)pCNL and (A)PAIS-pCNL algorithms have been run with optimal $\delta^*$ for the L2 error and the adaptive algorithm described in Section~\ref{sec:adapt}.
-The setup is as described in Section~\ref{sec:Implementation P2} (3).}
-\label{fig:BM2_AL2}
-\end{center}
-\end{figure}
-
-\subsubsection{A useful property of the PAIS algorithm for multimodal distributions}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/BM2_suction"}
-\caption{This figure demonstrates the redistribution property of the PAIS algorithm. Initially there is one chain in the positive mode, and 49 chains in the negative mode.}
-\label{fig:BM2_suction}
-\end{center}
-\end{figure}
-
-The biggest issue for the Metropolis-Hastings algorithms when sampling from a posterior such as the one in BM(2) is that it is unlikely that the same number of chains will occur in each of the modes, and since there is no interaction between the chains, there is no way to remedy this problem.
-The PAIS algorithm tackles this problem with its resampling step.
-The algorithm uses its dynamic kernel to build up an approximation of the posterior at each iteration, and then compares this to the posterior distribution via the weights function.
-Any large discrepancy in the approximation will result in a large or small weight being assigned to the relevant chain, meaning the chain will either pull other chains towards it or be sucked towards a chain with a larger weight.
-In this way, the algorithm allows chains to `teleport' to regions of the posterior which are in need of more weight.
-Figure~\ref{fig:BM2_suction} shows Problem BM(2) with initially 1 chain in the positive mode, and 49 chains in the negative mode.
-It takes only a handful of iterations for the algorithm to balance out the chains into 25 chains in each mode.
-The chains switch modes without having to climb the energy gradient in the middle.
-
-\section{Concluding Remarks}
-
-We have explored the application of parallelised MCMC algorithms in low dimensional Inverse problems.
-We have demonstrated numerically that these algorithms converge faster than the analogous Metropolis-Hastings algorithms, further experimentation with the Random Walk Metropolis-Hastings (RWMH), Metropolis Adjusted Langevin Algorithm (MALA) and preconditioned Crank-Nicolson (pCN) proposals has yielded similar results.
-
-One exception where the proposal distribution makes a difference to
-the convergence of the PAIS algorithm is where the KL divergence is
-large e.g. the PAIS-pCN algorithm struggles to sample from the
-distribution in Section~\ref{sec:problem 2}, while the pCN manages to
-produce a sample, albeit with a slow burn-in period. Additionally the
-PAIS-RWMH algorithm has no problems sampling from this distribution.
-
-\section{More conclusions to add}
-
-Incomplete PAIS (i.e. each processor only needs last reported value of
-the other chains)
-
-Using less than optimal resampler (lower cost, Schefzik, Roman,
-Thordis L. Thorarinsdottir, and Tilmann Gneiting. "Uncertainty
-quantification in complex simulation models using ensemble copula
-coupling." Statistical Science 28.4 (2013): 616-640.) COLIN TO DO THIS
-
-Use of other MCMC proposals (Riemann, HMC, etc)
-
-Importantly, we have compared the efficiency of our parallel scheme
-with a naive parallelisation of serial methods. Thus our increase in
-efficiency is over and above an $N$-fold increase, where $N$ is the
-number of cores or processors at our disposal. Our approach
-demonstrates a better-than-linear speed-up with the number
-processors/cores used. Thus, our approach is not only embarrasingly
-parallel (as coined by Cleve Moler in \cite{moler1986matrix}), but
-humiliatingly so.
-
-%It is commonly agreed that is preferable to run one MCMC long chain
-%than to run many shorter ones - this is largely due to the burn-in
-%period that is required.
-
-\bibliographystyle{siam}
-\bibliography{refs}
+\end{appendix}
 

### 2015-08-14 PAIS.tex
Cleared out old figures
Added section on calculating speed up

@@
 \newcommand{\E}{{\mathbb E}}
+
 \newcommand{\picturesAB}[6]{
@@
 Figure~\ref{fig:MH1 L2} shows that the PAIS-pCNL algorithm converges to the posterior distribution faster than the standard function space pCNL algorithm, in both L2 error and relative error in the moments.
-To reach a desired tolerance, the PAIS-pCNL algorithm requires 40\% fewer iterations than the pCNL algorithm.
+A description of the speed up attained by this algorithm is given in Section~\ref{sec:calc_saving}.
 
@@
 \centering
-\includegraphics[width=0.45\textwidth]{"figures/pCNL2 opt delta"}
-\includegraphics[width=0.45\textwidth]{"figures/PAIS pCNL2 opt delta"}
+\includegraphics[width=0.45\textwidth]{"figures/pCNL2o"}
+\includegraphics[width=0.45\textwidth]{"figures/PAISpCNL2o"}
 \caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 2}.
@@
 The adaptive algorithms both struggle with the first moment for the first million iterations, but produce better estimates after 10 million iterations.
-We see that to obtain a given tolerance level, the PAIS-pCNL algorithm requires 35\% fewer iterations than the pCNL algorithm.
 
@@
 respective nonadaptive algorithms and the improvement PAIS offers
-remains significant. The PAIS-pCNL algorithm requires 45\% fewer
-iterations than the naively parallelised pCNL algorithm.
+remains significant.
 
@@
 
-\subsubsection{A useful property of the PAIS algorithm for multimodal distributions}
+\subsubsection{Calculating the Speed Up in Convergence}\label{sec:calc_saving}
 
 \begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{"figures/calc_saving"}
+\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2}$ as a percentage of pCNL iterations.
+The relative L2 error graphs are from the problem in Section~\ref{sec:problem 1}.}
+\label{fig:calc_saving}
+\end{center}
+\end{figure}
+
+The graphs in the previous section clearly show that the PAIS-pCNL algorithm converges faster than the pCNL algorithm.
+We can calculate the number of iterations required to achieve a particular tolerance level in our solution for each algorithm and compare these to calculate a percentage saving.
+In Figure~\ref{fig:calc_saving} we demonstrate our calculation of the savings.
+We regress through the relative L2 error curve for both algorithms fixing the exponent at $-1/2$, i.e.
+look for the constants $c_1$ and $c_2$ such that
+\[
+	\text{pCNL L2 error} = c_1x^{-1/2}, \quad \text{and} \quad \text{PAIS-pCNL L2 error} = c_2x^{-1/2}.
+\]
+Then, for a chosen tolerance, $c_1x_1^{-1/2} = c_2x_2^{-1/2}$, where $x_1$ and $x_2$ are the number of iterations it takes the pCNL and PAIS-pCNL algorithms to reach a particular tolerance, given by the vertical lines on Figure~\ref{fig:calc_saving}.
+To find a percentage of iterations required by the PAIS-pCNL algorithm to achieve the same tolerance as the pCNL algorithm, we then use the expression
+\[
+	\frac{x_2}{x_1} = \left(\frac{c_2}{c_1}\right)^2.
+\]
+A summary of the percentage of iterations required using the PAIS algorithm compared with the respective Metropolis-Hastings algorithms is given in Table~\ref{table:calc_savings}.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{|l|r|r|}
+\hline
+		& Gaussian (low KL div) & Gaussian (high KL div) \\ \hline
+	RWMH & 14\% & 14\% \\
+	pCN & 33\% & - \\
+	MALA & 41\% & 42\% \\
+	pCNL & 62\% & 66\% \\ \hline
+		& Bimodal (low KL div) & Bimodal (high KL div) \\ \hline
+	RWMH & 40\% & 44\% \\
+	pCN & 32\% & 4\% \\
+	MALA & 36\% & 40\% \\
+	pCNL & 56\% & 0.02\% \\ \hline
+\end{tabular}
+\caption{Iterations for the PAIS algorithms required to achieve a desired tolerance as a percentage of the number of iterations required by the respective Metropolis algorithms.}
+\label{table:calc_savings}
+\end{table}
+
+\subsubsection{A Useful Property of the PAIS Algorithm for Multimodal Distributions}
+
+\begin{figure}[!h]
 \begin{center}
@@
 \section{The adaptive PAIS algorithm}\label{Sec:App}
-
-\begin{table}
+\begin{table}[!h]
 \begin{mdframed}
 \begin{algorithmic}
-\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
-\STATE Choose $\delta^{(0)} \in (0, 2]$, $\delta_L^{(0)} = 0.99\delta^{(0)}$, $\delta_U^{(0)} = 1.01\delta^{(0)} \wedge 2$.
-\FOR{$i=0,1,2, \ldots, N$}
-\STATE Sample $y_j^{(i)}~\sim~\nu(\cdot;x_j^{(i)},\delta_L^{(i)})$ for $j=1, \ldots,M/2$.
-\STATE Sample $y_j^{(i)}~\sim~\nu(\cdot;x_j^{(i)},\delta_U^{(i)})$ for $j=M/2+1, \ldots, M$.
-\STATE $\chi(y;\X^{(i)}) = M^{-1}
-\left(\sum_{j=1}^{M/2} \nu(y;x_j^{(i)},\delta_L^{(i)})+\sum_{j=M/2+1}^{M} \nu(y;x_j^{(i)},\delta_U^{(i)})\right)$.
-\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
-\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
+\STATE $x_j^{(0)} \sim \mu_0$, for $j \in L\cup U$, where $L = \{j\}_{j=1}^{M/2},\ U = \{j\}_{j=M/2+1}^M$.
+\STATE Choose $\delta^{(1)} \in (0, 2]$. Set $\delta_{L,U}^{(1)} = (1\pm 0.01)\delta^{(1)}\wedge 2$.
+\FOR{$i=1,2, \ldots, N$}
+\STATE $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\delta_L^{(i)})$ for $j \in L$, and $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\delta_U^{(i)})$ for $j \in U$.
+\STATE Calculate 
+\[
+	w^{(i)}_j = \frac{\pi(y_j^{(i)})}{\nu(y_j^{(i)};X^{(i-1)})},
+\]
+where
+\[
+	\nu(y;X) = \frac{1}{M}\sum_{j\in L} q(y;x_j,\delta_L)+\frac{1}{M}\sum_{j\in U} q(y;x_j,\delta_U).
+\]
+
 \IF{$i$ is in $\{n_k\}_{k=1}$}
-	\STATE For $w_{kj} = w_j^{(k)}$, $N$ iterations since last update,
-	\STATE $T_L = NM(\sum_{k=i-N}^N\sum_{j=1}^{M/2} w_{kj})^2/(\sum_{k=i-N}^N\sum_{j=1}^{M/2} w_{kj}^2)$.
-	\STATE $T_U = NM(\sum_{k=i-N}^N\sum_{j=M/2+1}^M w_{kj})^2/(\sum_{k=i-N}^N\sum_{j=M/2+1}^M w_{kj}^2)$.
-	\STATE $\delta^{(i+1)} = \delta^{(i)} - \Delta\D T$.
+	\STATE For $w_{kj} = w_j^{(k)}$, $S = n_k - n_{k-1}$,
+	\STATE $T_L = SM(\sum_{k=i-S}^S\sum_{j\in L} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in L} w_{kj}^2)$.
+	\STATE $T_U = SM(\sum_{k=i-S}^S\sum_{j\in U} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in U} w_{kj}^2)$.
+	\STATE $\delta^{(i+1)} = \delta^{(i)} - \Delta t \displaystyle\frac{T_U - T_L}{\delta_U^{(i)}-\delta_L^{(i)}}$.
 \ELSE
@@
 \ENDIF
-\IF{$i < K$}
-	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=1}^{M/2} \rightarrow (\frac{1}{M/2}, x_j^{(i+1)})_{j=1}^{M/2}$
-	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=M/2+1}^M \rightarrow (\frac{1}{M/2}, x_j^{(i+1)})_{j=M/2+1}^M$
+
+\IF{$i < N_\text{stop}$}
+	\IF{$i < N_\text{join}$}
+		\STATE Resample 
+	\[
+		(w_j^{(i)},y_j^{(i)})_{j\in L} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in L}, \quad (w_j^{(i)},y_j^{(i)})_{j\in U} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in U}.
+\]
 	\ELSE
-	\STATE Resample: $(w_j^{(i)},y_j^{(i)})_{j=1}^M \rightarrow (\frac{1}{M}, x_j^{(i+1)})_{j=1}^M$
+		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
+	\ENDIF
+	\STATE $\delta_{L,U}^{(i)} = \delta^{(i)} \pm 2\sqrt{2}\delta^{(i)}/\sqrt{NM} \wedge 2$.
+\ELSE
+		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
+		\STATE $\delta_L^{(i+1)} = \delta_U^{(i+1)} = \delta^{(i+1)}$.
 \ENDIF

### 2015-08-14 PAIS.tex
updated adaptive algorithm description

@@
 large, then the method may become unstable and inefficient.
+We also choose two iteration numbers at which we change our adaptive strategy.
+At iteration $N_\text{join}$ we resample using the full ensemble size instead of using to subsets of the full ensemble.
+This allows us to fine-tune the scaling parameter using the full resampler, but slows down the rate of convergence of the scaling parameter.
+Finally we choose a time, $N_\text{stop}$, at which we stop updating our parameter to guarantee ergodicity of the MCMC algorithm.
 

### 2015-08-17 PAIS.tex
a few suggested edits.

@@
 \begin{abstract}
-  Markov chain Monte Carlo methods are commonly used and powerful
+  Markov chain Monte Carlo methods are a powerful and commonly used
   family of numerical methods for sampling from complex probability
-  distributions. As the applications that these methods are being
-  applied to increase in size and complexity, the need for efficient
+  distributions. As applications of these methods increase in size and complexity, the need for efficient
   methods which can exploit the parallel architectures which are
@@
 methods were first developed in the 70s\cite{hastings1970monte}, and with the development of
-faster more powerful computers, have become ever more important in a
-whole range of fields in statistics, science and engineering. However,
-despite the advances in hardware that have made it possible to fully
-characteries the posterior distributions of many problems, it remains
-unfeasible to use MCMC methods for many more. In particular, when
+to use MCMC methods for many more. In particular, when
 considering Bayesian inverse problems, each MCMC step may involve the
@@
 
-Many advances have been made in the field of MCMC, to design ever more
-complex methods, which more intelligently propose moves which leads to
-faster converging methods. Function space versions of standard methods
+Many advances have been made in the field of MCMC to design ever more
+complex methods that propose moves more intelligently, leading to
+rapidly converging methods. Function space versions of standard methods
 such as the random walk Metropolis-Hastings (RWMH) algorithm or the
@@
 even higher order gradient information\cite{bui2014solving}.  These
-methods allow us to more fully explore the posterior distribution, at
-the cost of fewer iterations of the method.
+methods allow us explore the posterior distribution more fully 
+with fewer iterations.
 
@@
 development of computing hardware. Moore's law, which predicted that
-the number of transistors that it is possible to fit on a single
-microchip will double every two years, has been largely followed
-since the early 70s\cite{moore1998cramming}. However, the main revolution with regards to
-scientific computing in recent decades has been the increasing focus on parallel
-architectures. The efficient exploitation of these facilities is the key to
-solving many of the computational challenges that we currently face.
+the number of transistors that can to fit on a single microchip will
+double every two years, has been largely followed since the early
+70s\cite{moore1998cramming}. In recent times, it has become necessary
+to use parallel architectures to exploit Moore's law. The efficient
+exploitation of these facilities is the key to solving many of the
+computational challenges that we currently face.
 
@@
 In this paper, we present a framework for parallelisation of
-importance sampling, which allows us to use any of the current
+importance sampling, which can be built around any of the current
 Metropolis-based methodologies in order to create an efficient target
 proposal from the current state of all of the chains in the
-ensemble. In Section \ref{Sec:Prelim} we outline some preliminaries,
+ensemble. The idea is to consider the parallel chains as an ensemble,
+and to resample using a transformation based on optimal transport.
+Samplers based on optimal transport have also been considered in
+\cite{el2012bayesian}.
+
+In Section \ref{Sec:Prelim} we outline some preliminaries,
 including the general set up for Bayesian inverse problems, the
@@
 automatically tuned to provide optimal convergence. In
-Section \ref{Sec:Num}, we present some numerical experiments which
-demonstrate the savings that can be made by employing this approach as
+Section \ref{Sec:Num}, we present some numerical experiments that
+demonstrate the savings available by employing this approach as
 opposed to a naive/trivial parallelisation of existing MCMC
-methods. In Section \ref{Sec:Conc}, we will summarise our results and
+methods. In Section \ref{Sec:Conc}, we summarise our results and
 suggest some areas for future investigation.
@@
 \subsection{Bayesian inverse problems}
-In this paper, we particularly focus on the use of MCMC methods in characterising
+In this paper, we focus on the use of MCMC methods for characterising
 posterior probability distributions in Bayesian inverse problems. We
@@
 The parameter $u$ is observed
-through the observation operator $\mathcal{G}:X \to
-\mathbb{R}^d$. Since observations are rarely, if ever, perfect, we
+through the observation operator $\mathcal{G}:X \to\mathbb{R}^d$. 
+Since observations are never perfect, we
 assume that these measurements $\data$ are subject to Gaussian noise, so that
@@
 Given that a trivial coupling $T^t$ always exists in the space of transference plans, $\Pi(\mu_{X}, \mu_{Y})$, we can find a coupling $T^*$ which maximises the correlation between $X$ and $Y$ \cite{cotter2012ensemble}.
-This coupling is the solution to a linear programming problem in $M^2$ variables with $2M-1$ constraints as defined in \cite{reich2013nonparametric}.
+This coupling is the solution to a linear programming problem in $M^2$ variables with $2M-1$ constraints.
 Maximising the correlation ensures that the new sample is as much like the original sample as possible with the additional property that the sample is evenly weighted.
 
-As proposed in \cite{reich2013nonparametric} a Monte Carlo algorithm can be implemented to resample from a weighted ensemble.
+A Monte Carlo algorithm can be implemented to resample from a weighted ensemble.
 We create a weighted sample, then solve the optimal transport problem which produces the coupling described above, we can draw a new sample from the evenly weighted distribution.
@@
 \frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
-\STATE Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
+\STATE Resample using ETMC: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
 \ENDFOR 
@@
 $\chi$ is a close approximation to the posterior density $\pi$.
-If the proposal distribution is too over-dispersed, then the algorithm will propose states a long way from the current state but there will be a relatively low acceptance rate, impacting the quality of inferences on the posterior.
+If the proposal distribution is over-dispersed, then the algorithm will propose states a long way from the current state at the cost of a relatively low acceptance rate, impacting the quality of inferences on the posterior.
 Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the space so the mixing rate, and hence convergence rate, will be slow.
@@
 MCMC algorithms can be assessed by comparing their approximation of the posterior to the analytic distribution.
-To assess this, a distance metric must be decided on, such as the relative error between the sample moments and the true moments, or the relative L2 error between the true density, $\pi(x|D)$, and the constructed histogram.
+To assess this, a distance metric must be chosen.
+Examples are the relative error between the sample moments and the true moments, or the relative L2 error between the true density, $\pi(x|D)$, and the constructed histogram.
 The relative error in the $m$-th moment is
@@
 \end{equation}
-where the points $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional regions the histogram is defined over, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of a bin, and $B_i$ is the value of the $i$th bin.
+where the points $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional histogram bins, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of a bin, and $B_i$ is the value of the $i$th bin.
 
@@
 the last expression we see that when the variance of the weights is
-zero, $\neff = M$, which is our ideal scenario.
-In a neighbourhood around $\delta_{\text{var}}^*$, $\neff$ decreases, when the variance increases.
+zero, $\neff = M$; this is our ideal scenario.
+In a neighbourhood around $\delta_{\text{var}}^*$, $\neff$ decreases when the variance increases.
 So if $v(w(y))$ can be equal to zero for a particular problem and proposal distribution, maximising the effective sample size is equivalent to minimising the variance of the weights.
@@
 struggle with multimodal problems, particularly where switches between
-the modes are rare, resulting incorrectly proportioned modes in the
-histograms produced. With the PAIS algorithm, we see that the
-resampling step allows redistribute chains to new modes as they are
+the modes are rare, resulting in incorrectly proportioned modes in the
+histograms, for example. With the PAIS algorithm, we see that the
+resampling step redistributes chains to new modes as they are
 found. This means that we expect the number of chains in a mode to be
@@
 We have explored the application of parallelised MCMC algorithms in
-low dimensional Inverse problems. We have demonstrated numerically
+low dimensional inverse problems. We have demonstrated numerically
 that these algorithms converge faster than the analogous naively parallelised

### 2015-08-25 PAIS.tex
removed OT dimension scaling and some regression detail

@@
 \usepackage{pifont}
-\usepackage{bbm,algorithmic,mdframed,placeins}
+\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs}
 \setlength{\parindent}{0in}
@@
 
-Ideally the dimension of particles in a filter will not affect the convergence of the method.
+%Ideally the dimension of particles in a filter will not affect the convergence of the method.
 Unfortunately, the approximation of continuous density functions is computationally infeasible in high dimensions and the required ensemble size scales exponentially with the dimension of the problem\cite{silverman1986density,snyder2008obstacles}.
 
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/ET_dim_scaling}
-\caption{Demonstration of the ETMC algorithm rapidly failing as the dimension increases. The relative error shown is that between the sampled mean and the true mean.}
-\label{fig:ET_dim_scaling}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm
-loses its ability to accurately preserve the mean of the posterior
-distribution as the dimension increases. Clearly for a simple Gaussian
-problem, we can not resample in much more than 10 dimensions and hope
-to recover a representative sample, without hugely increasing the
-number of particles in the ensemble. As the cost of the ETMC is
-$\mathcal{O}(n^2)$, this does limit the dimension for which this
-approach can be used.
+%\begin{figure}[htb]
+%\begin{center}
+%\includegraphics[width=\textwidth]{figures/ET_dim_scaling}
+%\caption{Demonstration of the ETMC algorithm rapidly failing as the dimension increases. The relative error shown is that between the sampled mean and the true mean.}
+%\label{fig:ET_dim_scaling}
+%\end{center}
+%\end{figure}
+
+%Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm loses its ability to accurately preserve the mean of the posterior distribution as the dimension increases.
+Clearly for a simple Gaussian problem, we can not resample in much more than 10 dimensions and hope to recover a representative sample, without hugely increasing the number of particles in the ensemble.
+As the cost of the ETMC is $\mathcal{O}(n^2)$, this does limit the dimension for which this approach can be used.
 
@@
 
-The graphs in the previous section clearly show that the PAIS-pCNL algorithm converges faster than the pCNL algorithm.
+The graphs in the previous section clearly show that the PAIS-pCNL algorithm converges faster than the pCNL algorithm when both are parallelised with the same number of threads.
 We can calculate the number of iterations required to achieve a particular tolerance level in our solution for each algorithm and compare these to calculate a percentage saving.
 In Figure~\ref{fig:calc_saving} we demonstrate our calculation of the savings.
-We regress through the relative L2 error curve for both algorithms fixing the exponent at $-1/2$, i.e.
-look for the constants $c_1$ and $c_2$ such that
-\[
-	\text{pCNL L2 error} = c_1x^{-1/2}, \quad \text{and} \quad \text{PAIS-pCNL L2 error} = c_2x^{-1/2}.
-\]
-Then, for a chosen tolerance, $c_1x_1^{-1/2} = c_2x_2^{-1/2}$, where $x_1$ and $x_2$ are the number of iterations it takes the pCNL and PAIS-pCNL algorithms to reach a particular tolerance, given by the vertical lines on Figure~\ref{fig:calc_saving}.
-To find a percentage of iterations required by the PAIS-pCNL algorithm to achieve the same tolerance as the pCNL algorithm, we then use the expression
-\[
-	\frac{x_2}{x_1} = \left(\frac{c_2}{c_1}\right)^2.
-\]
+The constants $c_1$ and $c_2$ are found by regressing through the data with a fixed exponent of $-1/2$ excluding the intial data points where the graph has not finished burning in.
+
 A summary of the percentage of iterations required using the PAIS algorithm compared with the respective Metropolis-Hastings algorithms is given in Table~\ref{table:calc_savings}.
+The blank entries correspond to occasions when the MH algorithms haven't converged to the posterior distribution.
 
@@
 \centering
-\begin{tabular}{|l|r|r|}
-\hline
-		& Gaussian (low KL div) & Gaussian (high KL div) \\ \hline
-	RWMH & 14\% & 14\% \\
-	pCN & 33\% & - \\
-	MALA & 41\% & 42\% \\
-	pCNL & 62\% & 66\% \\ \hline
-		& Bimodal (low KL div) & Bimodal (high KL div) \\ \hline
-	RWMH & 40\% & 44\% \\
-	pCN & 32\% & 4\% \\
-	MALA & 36\% & 40\% \\
-	pCNL & 56\% & 0.02\% \\ \hline
+\begin{tabular}{|l|l|r|r|}
+\hline
+		& & Low KL div & High KL div \\ \hline
+	\multirow{4}{*}{Gaussian}	&RWMH & 14\% & 14\% \\
+	&pCN & 33\% & - \\
+	&MALA & 41\% & 42\% \\
+	&pCNL & 62\% & 66\% \\ \hline
+	\multirow{4}{*}{Bimodal}	&RWMH & 40\% & 44\% \\
+	&pCN & 32\% & - \\
+	&MALA & 36\% & 40\% \\
+	&pCNL & 56\% & - \\ \hline
 \end{tabular}

### 2015-08-25 PAIS.tex
some final edits in the conclusions mostly.

@@
 that allow us to sample from complex probability distributions. MCMC
-methods were first developed in the 70s\cite{hastings1970monte}, and with the development of
-to use MCMC methods for many more. In particular, when
+methods were first developed in the 70s\cite{hastings1970monte}, and
+with the development of and with the development of faster more
+powerful computers, have become ever more important in a whole range
+of fields in statistics, science and engineering.  In particular, when
 considering Bayesian inverse problems, each MCMC step may involve the
@@
 \ref{sec:filters}, and the pCNL proposal distribution described in
-Section \ref{Sec:pCNL}
+Section \ref{Sec:pCNL}.
 
@@
 
-Important samplers can be a very efficient method for sampling from a
-probability distribution. A proposal density is chosen from which we
-can draw samples, and each sample is assigned a weight given by the
+Important sampling can be a very efficient method for sampling from a
+probability distribution. A proposal density is chosen, from which we
+can draw samples. Each sample is assigned a weight given by the
 ratio of the target density and the proposal density at that
@@
 parallel (as coined by Cleve Moler in \cite{moler1986matrix}), but
-humiliatingly so.
+humiliatingly so, provided that an optimal transport algorithm can be
+efficiently implemented in parallel and so does not dominate the
+communication costs.
 
@@
 
-The PAIS approach is limited by the curse of dimensionality, since the
-cost of the resampling algorithm becomes prohibitively large. The cost
-of the resampler is $\mathcal{O}(n^2)$, where $n$ is the number of
-particles in the ensemble. Other resamplers exist which are not as
-costly, but for which there is some sacrifice in terms of optimality
-of the output. For example, the resampler proposed in
-\cite{schefzik2013uncertainty}  has computational complexity
-$\mathcal{O}(n)$. This might allow the PAIS to be used in higher
-dimensional problems, and is an avenue for future investigation.
-
 Another strength of the PAIS is that it can also be used with any MCMC
 proposal. There are a growing number of increasing sophisticated MCMC
-algorithms (HMC, Riemann manifold MCMC etc) which could be incorporated into this framework, leading to
-even more efficient algorithms, and this is another opportunity for
-future work.
+algorithms (HMC, Riemann manifold MCMC etc) which could be
+incorporated into this framework, leading to even more efficient
+algorithms, and this is another opportunity for future work.
 
@@
 to more efficient use of the computer architecture.
+
+One limitation of the PAIS approach as described above is that a
+direct solver (such as FastEMD \cite{FastEMD}) has computational cost
+$\mathcal{O}(n^3\log n)$, where $n$ is the number of particles in the
+ensemble.  The use of approximate solutions from iterative solvers
+might allow the PAIS to be used in higher dimensional problems, and is
+an avenue for future investigation. A second limitation is associated
+with the problem of accurately approximating measures with empirical
+measures in high dimensional phase space. This is a well-known issue
+in filtering, where one possible solution is to ``localise'' the
+impact of observations, as rigorously analysed in
+\cite{rebeschini2013can}. An established technique in Ensemble Kalman
+filters, localisation has also been incorporated into the Ensemble
+Transform Particle Filter \cite{ChRe2015}. This issue could be
+addressed in the PAIS by using localisation when computing the
+transform, but keeping the unlocalised weights in the importance
+sampling for consistency.
 

### 2015-08-26 PAIS.tex
Spellcheck and cover letter
Added a cover letter and did a spell check.

@@
 In Figure~\ref{fig:calc_saving} we demonstrate our calculation of the savings.
-The constants $c_1$ and $c_2$ are found by regressing through the data with a fixed exponent of $-1/2$ excluding the intial data points where the graph has not finished burning in.
+The constants $c_1$ and $c_2$ are found by regressing through the data with a fixed exponent of $-1/2$ excluding the initial data points where the graph has not finished burning in.
 

### 2015-08-26 coverletter.tex
Spellcheck and cover letter
Added a cover letter and did a spell check.

\documentclass{letter}
\usepackage{txfonts}

\setlength{\textheight}{10.2in} 
\setlength{\topmargin}{-0.9in}

\signature{Simon Cotter}
\address{School of Mathematics \\ Alan Turing Building \\ Oxford
  Road\\Manchester\\ M13 9PL \\ UK \\\textit{email:} simon.cotter@manchester.ac.uk}

\begin{document}

\date{}
\begin{letter}{ }

\opening{To whom it may concern,}

Please find attached a manuscript that we would like you to consider for publication in the SIAM Journal of  Scientific Computing, entitled ``Parallel Adaptive Importance Sampling". We look forward to hearing from you in due course.

\closing{Yours sincerely,}
\end{letter}

\end{document}

### 2016-03-22 PAIS.tex
updated effective sample size

@@
 the last expression we see that when the variance of the weights is
-zero, $\neff = M$; this is our ideal scenario.
-In a neighbourhood around $\delta_{\text{var}}^*$, $\neff$ decreases when the variance increases.
-So if $v(w(y))$ can be equal to zero for a particular problem and proposal distribution, maximising the effective sample size is equivalent to minimising the variance of the weights.
-In the PAIS algorithm we have a dynamic proposal which is perturbed at every iteration.
-
-The statistic $\neff$ converges faster than the variance of the weights, and so is preferable as a means of tuning $\delta$.
-While $\delta_{\text{var}}^*$ and the optimal value found using the effective sample size, $\delta_{\text{eff}}^*$, coincide when $v(w(y)) = 0$, this is not usually possible, so the methods will find different optimal values of $\delta^*$.
-
-The $\neff$ statistic also has another useful property; if we imagine the algorithm in the burn-in phase, for example, we have $M$ processes in the tail of a Gaussian curve searching for the area of high density.
+zero, $\neff = M$; this is our ideal scenario. Maximising the effective sample size is equivalent to minimising the variance of the weights.
+
+The statistic $\neff$ is better scaled than the variance of the weights, and is easier to interpret, so is preferable as a means of tuning $\delta$.
+In the following numerics, we use the effective sample size calculated at each iteration and averaged across the entire run to tune the scaling parameters.
+We do this because this is the method we use to tune the scaling parameters on the fly, comparison to the optimal scaling parameter found using the global optimum of the variance of the weights is also included for comparison.
+
+The effective sample size also has another useful property; if we imagine the algorithm in the burn-in phase, for example, we have $M$ processes in the tail of a Gaussian curve searching for the area of high density.
 If the processes are evenly spaced, then the particle closest to the mean will have an exponentially higher weight assigned to it.
@@
 Here we look at how the statistic behaves as we vary the ensemble size, $M$.
-Figure~\ref{fig:neff-M} shows results for the second problem discussed in \ref{sec:problem 2} when using the PAIS-pCNL algorithm to explore the posterior distribution.
+Figure~\ref{fig:neff-M} shows results for the Gaussian posterior discussed in \ref{sec:problem 2} when using the PAIS-MALA algorithm.
 The left part of the figure shows that as the ensemble size increases, the value of $\delta$ which gives the optimal effective sample size ratio decreases.

### 2016-03-22 PAIS.tex
removed G1

@@
 \usepackage{pifont}
-\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs}
+\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
 \setlength{\parindent}{0in}
@@
 In this example we look at how PAIS compares to naively parallelised
-MCMC. We compare the pCNL algorithm \cite{cotter2013mcmc} with its
-PAIS variant, the PAIS-pCNL algorithm. We consider several statistics
+MH. We compare the RWMH algorithm \cite{cotter2013mcmc} with its
+PAIS variant, the PAIS-RW algorithm. We consider several statistics
 for measuring the efficiency of the PAIS algorithm compared to
@@
 
-Since we are comparing against trivially parallelised MCMC algorithms,
+Since we are comparing against naively parallelised MCMC algorithms,
 we also need to decide which statistic $T(\delta)$ to optimise for
 these approaches.
-In the examples which follow, we have optimised the naively parallelised pCNL algorithm using the optimal acceptance rate $\hat{\alpha} = 0.75$.
-This value was found by using the L2 error minimums to calculate $\delta^*$.
+In the examples which follow, we have optimised the naively parallelised RWMH algorithm using the optimal acceptance rate $\hat{\alpha} = 0.5$.
+This value differs from the theoretical symptotic value of 0.234 which applies in higher dimensions, this higher value is commonly used for one dimensional Gaussian posteriors.
 This means that we are minimising the statistic
@@
 Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions.
-Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the resulting form
+Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
 \begin{equation}\label{eqn:Gaussian posterior}
@@
 numerics which follow, we choose $\Sigma = \sigma^2I_N$ and
-$\mathcal{T} = \tau^2I_d$ with $\tau^2 =2$ and $\sigma^2 = 0.1$, and
-we observe $u=-2.5$ noisily such that
+$\mathcal{T} = \tau^2I_d$ with $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
+we observe $u=4$ noisily such that
 \[
@@
 These values result in a posterior density in which the vast majority
-of the density is contained inside the high density region of the
-prior. This means that it should be straightforward for the algorithm
-to find the stationary distribution. The Kullback-Leibler (KL)
+of the density is out in the tails of the
+prior distribution. The Kullback-Leibler (KL)
 divergence, which gives us a measure of how different the prior and
-posterior are, is $D_{KL}(\mu_Y || \mu_0) = 2.67$ for this problem.
+posterior are, is $D_{KL}(\mu_Y || \mu_0) = 4.67$ for this problem.
 
@@
 We run the PAIS-pCNL algorithm for 10,000 iterations and pCNL for 100,000 iterations, each with $M=50$ processes.
-We took 32 repeats of both algorithms and then used the medians of the statistics to find the optimal parameters.
+We took 32 repeats of both algorithms and then used the geometric means of the sample statistics to find the optimal parameters.
 
@@
 The algorithms are run using the optimal parameters found in (1).
-The relative L2 error (Equation~\ref{eqn:L2_error}) is used as a measure of accuracy.
+The relative L2 error, given in Equation~\ref{eqn:L2_error}, is used as a measure of accuracy.
 The simulation for each algorithm was repeated 24 times.
@@
 \centering
-\includegraphics[width=0.45\textwidth]{"figures/pCNL1o"}
-\includegraphics[width=0.45\textwidth]{"figures/PAISpCNL1o"}
-\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 1}.
-The setup is as in Section~\ref{sec:Implementation P1}.}
+\subfigure[RWMH algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/pCNL2o"}}
+\subfigure[PAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/PAISpCNL2o"}}
+\caption{Finding optimal values of $\delta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}.}
 \label{fig:P1 opt delta}
@@
     \end{minipage}
+	\vspace{1mm}
 	\caption{Optimal values of $\delta$ summarised from Figure~\ref{fig:P1 opt delta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
@@
 
-Figure~\ref{fig:P1 opt delta} (left) shows the two values of $\delta$ which may be optimal for the pCNL algorithm, found at the turning points.
+Figure~\ref{fig:P1 opt delta} (a) shows the two values of $\delta$ which may be optimal for the pCNL algorithm, found at the turning points.
 The first estimate comes from the relative L2 error, and the second comes from the variance of the estimate of the mean.
-The optimal acceptance rates of the function space algorithms are expected to be slightly higher than their finite dimensional versions.
-Since the L2 error estimate of the histogram gives an acceptance rate which is near to 100\%, we say that the optimal acceptance rate is that as given by the variance of the mean, roughly 75\%.
-The results in Figure~\ref{fig:P1 opt delta} are summarised in Table~\ref{table:prob1 opt delta} (left).
-
-Figure~\ref{fig:P1 opt delta} (right) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
+The results in Figure~\ref{fig:P1 opt delta} are summarised in Table~\ref{table:prob1 opt delta}.
+
+Figure~\ref{fig:P1 opt delta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
 Although the L2 error graph is noisy, the maximum in the effective sample size is close to the minimum in the L2 error.
 The minimum in the variance of the weights however is a long way away.
-We choose the effective sample size as the best estimator of the optimal scaling parameter because of this, and because the effective sample size statistic converges to a smooth graph faster than either the variance of the weights or the L2 error.
+Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter.
+In general this indicator underestimates the estimate found by using the entire sample.
 
@@
 \centering
-\includegraphics[width=0.45\textwidth]{"figures/pCNL1l"}
-\includegraphics[width=0.45\textwidth]{"figures/pCNL1m"}
-\caption{Relative error in the first moment (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iterations for problem~\ref{sec:problem 1}.
-The setup is as in Section~\ref{sec:Implementation P1} (2, 3).}
+\subfigure[Relative L2 error.]{\includegraphics[width=0.45\textwidth]{"figures/pCNL2l"}}
+\subfigure[Relative error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/pCNL2m"}}
+\caption{Error analysis for the (A)RWMH and (A)PAIS-RW algorithms. The setup is as in Section~\ref{sec:Implementation P1} (2, 3).}
 \label{fig:MH1 L2}
@@
 MCMC should give us added value over and above that provided by naive parallelism.
-
-\subsection{Sampling from a one dimensional Gaussian with a higher KL divergence}\label{sec:problem 2}
-
-In this example we use the same setup used in Section~\ref{sec:problem 1}.
-We choose a posterior distribution which has most of its mass far out in the tails of the prior distribution.
-The KL divergence is $D_{KL}(\mu_Y||\mu_0) = 4.670$.
-This means that the algorithm will have to ``work harder'' to find the area of high probability in the posterior density.
-
-\subsubsection{Target distribution}
-
-As in the previous example, we use the identity observation operator
-$\G(u)=u$, which results in the Gaussian posteriors in
-Equation~\ref{eqn:Gaussian posterior}. However, this time we choose
-$\sigma^2=0.01$ and $\tau^2=0.01$ so the posterior distribution is
-$\N(D/(1+\sigma^2/\tau^2), \tau^2\sigma^2/(\sigma^2+\tau^2)) = \N(D/2,
-0.005)$ which for large $D$ is a long way out in the tail of the prior
-with a very small variance. For the simulations which follow we
-observe a reading of $u=4$, with observational noise drawn from $\N(0, \sigma^2)$.
-
-\subsubsection{Optimal values of $\delta$}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.45\textwidth]{"figures/pCNL2o"}
-\includegraphics[width=0.45\textwidth]{"figures/PAISpCNL2o"}
-\caption{Finding optimal values of $\delta$ for the pCNL (left) algorithm and PAIS-pCNL (right) algorithm for the problem in Section~\ref{sec:problem 2}.
-The setup is as in Section~\ref{sec:Implementation P1}.}
-\label{fig:P2 opt delta}
-\end{figure}
-
-\begin{table}[!htb]
-    \begin{minipage}{.5\linewidth}
-      \centering
-        \begin{tabular}{|c|r|}
-	\hline
-	Statistic											& pCNL \\ \hline
-	$\delta_{\text{L2}}^*$								& 8.6e-2 \\
-	$\delta_{\text{var}(\hat{\mu})}^*$					& 9.1e-1 \\
-	Acceptance Rate ($\delta_{\text{L2}}^*$	)			& 9.9e-1 \\
-	Acceptance Rate ($\delta_{\text{var}(\hat{\mu})}^*$)	& 8.1e-1 \\
-	\hline
-	\end{tabular}
-    \end{minipage}%
-    \begin{minipage}{.5\linewidth}
-      \centering
-        \begin{tabular}{|l|r|r|}
-	\hline
-	Statistic							& PAIS-pCNL \\ \hline
-	$\delta_{\text{eff}}^*$				& 2.6e-1 \\
-	$\delta_{\text{var}(w(y))}^*$		& 2.6e-1 \\
-	$\delta_{\text{L2}}^*$				& 2.8e-1 \\
-	\hline
-	\end{tabular}
-    \end{minipage}
-	\caption{Optimal values of $\delta$ summarised from Figure~\ref{fig:P2 opt delta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
-	\label{table:P2 opt delta}
-\end{table}
-
-We find the optimal values of $\delta$ using the same methods described in Sections~\ref{sec:Implementation P1} and \ref{sec:Optimal values P1}.
-The results are displayed in Figure~\ref{fig:P2 opt delta} and Table~\ref{table:P2 opt delta}.
-There is a huge difference between the two error estimates of $\delta$ for pCNL, although the corresponding variance graph is very flat making it sensitive to Monte Carlo error.
-The variance of the mean estimate has an 81\% acceptance rate, which is larger than the pCNL acceptance rate found previously.
-
-For the PAIS-pCNL algorithm, it is again clear that the effective sample size ratio is a useful statistic for judging the optimal value of $\delta$.
-The relative L2 error estimate of $\delta^*$, shown in Table~\ref{table:P2 opt delta}, is slightly higher than the other two estimates, but from the graph there again seems to be a fairly flat wide minimum which is in the same region as the optimal effective sample size ratio.
-The variance of the weights becomes so small in the critical region that we get numerical zeroes, which means that we could not use this method to tune the scaling parameter adaptively.
-
-\subsubsection{Convergence of pCNL vs PAIS-pCNL}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.45\textwidth]{"figures/pCNL2l"}
-\includegraphics[width=0.45\textwidth]{"figures/pCNL2m"}
-\caption{Relative error in the first moment (right) and histograms (left) produced by the (A)pCNL and (A)PAIS-pCNL algorithms against iterations for problem~\ref{sec:problem 2}.
-The setup is as described in Section~\ref{sec:Implementation P1} (2,3).}
-\label{fig:MH2 L2}
-\end{figure}
-
-Figure~\ref{fig:MH2 L2} shows that the PAIS-pCNL algorithm, converges to the posterior distribution faster than the pCNL algorithm.
-The adaptive algorithms both struggle with the first moment for the first million iterations, but produce better estimates after 10 million iterations.
 

### 2016-03-24 PAIS.tex
new RW numerics for G2

@@
 In this example we look at how PAIS compares to naively parallelised
-MH. We compare the RWMH algorithm \cite{cotter2013mcmc} with its
+MH. We compare the RWMH algorithm with its
 PAIS variant, the PAIS-RW algorithm. We consider several statistics
@@
 In the examples which follow, we have optimised the naively parallelised RWMH algorithm using the optimal acceptance rate $\hat{\alpha} = 0.5$.
-This value differs from the theoretical symptotic value of 0.234 which applies in higher dimensions, this higher value is commonly used for one dimensional Gaussian posteriors.
-This means that we are minimising the statistic
+This value differs from the theoretical asymptotic value of 0.234 which applies in higher dimensions, this higher acceptance rate is commonly used for one dimensional Gaussian posteriors.
+To optimise the algorithm we minimise the statistic
 \[
@@
 First we calculate the optimal value of $\delta$ by optimising the statistics described in Section~\ref{sec:statistics}.
-Once we have these parameters, we run the algorithms and compare the convergence speeds of the algorithms.
-Finally, we implement the adaptive algorithms described in Section~\ref{sec:adapt} and compare the convergence of these algorithms with the nonadaptive algorithms.
+We then run the algorithms with optimal parameters to calculate and compare the convergence rates.
+Finally, we implement the adaptive algorithms described in Section~\ref{sec:adapt} and compare the convergence rates of these algorithms with the nonadaptive algorithms.
 
@@
 parameters we choose 32 values of $\delta$ evenly spaced on a log scale
-between $[10^{-5}, 2]$.
-We run the PAIS-pCNL algorithm for 10,000 iterations and pCNL for 100,000 iterations, each with $M=50$ processes.
+in he interval $[10^{-5}, 2]$.
+We run the PAIS-RW algorithm for 100,000 iterations and RWMH for 1,000,000 iterations, each with $M=50$ processes.
 We took 32 repeats of both algorithms and then used the geometric means of the sample statistics to find the optimal parameters.
 
-{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} and Section~\ref{sec:problem 2} for 10 million iterations, again with 50 processes.
-The algorithms are run using the optimal parameters found in (1).
-The relative L2 error, given in Equation~\ref{eqn:L2_error}, is used as a measure of accuracy.
-The simulation for each algorithm was repeated 24 times.
+{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for 1 million iterations, again with 50 processes.
+The simulations are repeated 32 times using the optimal parameters found in (1).
+The performance of the algorithms is judged by the convergence of the relative L2 error statistic in Equation~\ref{eqn:L2_error}.
 
@@
 \centering
-\subfigure[RWMH algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/pCNL2o"}}
-\subfigure[PAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/PAISpCNL2o"}}
+\subfigure[RWMH algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_rwmho"}}
+\subfigure[PAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
 \caption{Finding optimal values of $\delta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}.}
@@
 	Statistic											& pCNL \\ \hline
-	$\delta_{\text{L2}}^*$								& 3.7e-3 \\
-	$\delta_{\text{var}(\hat{\mu})}^*$					& 5.8e-2 \\
-	Acceptance Rate ($\delta_{\text{L2}}^*$)				& 9.9e-1 \\
-	Acceptance Rate ($\delta_{\text{var}(\hat{\mu})}^*$)	& 7.4e-1 \\
+	$\delta_{\text{L2}}^*$								& 2.3e-4 \\
+	$\delta_{\%}^*$									& 1.2e-1 \\
+	Acceptance Rate ($\delta_{\text{L2}}^*$)				& 9.0e-1 \\
+	Acceptance Rate ($\delta_{\%}^*$)					& 5.0e-1 \\
 	\hline
@@
 	Statistic							& PAIS-pCNL \\ \hline
-	$\delta_{\text{eff}}^*$				& 1.5e-2 \\
-	$\delta_{\text{var}(w(y))}^*$		& 6.4e-2 \\
-	$\delta_{\text{L2}}^*$				& 1.7e-2 \\
+	$\delta_{\text{eff}}^*$				& 1.1e-3 \\
+	$\delta_{\text{var}(w(y))}^*$			& 1.7e-3 \\
+	$\delta_{\text{L2}}^*$				& 7.6e-4 \\
 	\hline
@@
 
-Figure~\ref{fig:P1 opt delta} (a) shows the two values of $\delta$ which may be optimal for the pCNL algorithm, found at the turning points.
-The first estimate comes from the relative L2 error, and the second comes from the variance of the estimate of the mean.
+Figure~\ref{fig:P1 opt delta} (a) shows the two values of $\delta$ which are optimal according to the acceptance rate and relative L2 error criteria for the RWMH algorithm.
+The smaller estimate comes from the relative L2 error, and the larger from the acceptance rate.
 The results in Figure~\ref{fig:P1 opt delta} are summarised in Table~\ref{table:prob1 opt delta}.
+Since in general we cannot calculate the relative L2 error, we must optimise the algorithm using the acceptance rate.
+From the relative L2 error curve we can see that the minimum is very wide and despite the optimal values being very different there is not a large difference in the convergence rate.
 
 Figure~\ref{fig:P1 opt delta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
-Although the L2 error graph is noisy, the maximum in the effective sample size is close to the minimum in the L2 error.
-The minimum in the variance of the weights however is a long way away.
+The relative L2 error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative L2 error.
 Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter.
@@
 
-\subsubsection{Convergence of pCNL vs PAIS-pCNL}
+\subsubsection{Convergence of RWMH vs PAIS-RW}
 
@@
 \centering
-\subfigure[Relative L2 error.]{\includegraphics[width=0.45\textwidth]{"figures/pCNL2l"}}
-\subfigure[Relative error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/pCNL2m"}}
+\subfigure[Relative L2 error.]{\includegraphics[width=0.45\textwidth]{"figures/G2_l2"}}
+\subfigure[Relative error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/G2_moments"}}
 \caption{Error analysis for the (A)RWMH and (A)PAIS-RW algorithms. The setup is as in Section~\ref{sec:Implementation P1} (2, 3).}
@@
 
-Figure~\ref{fig:MH1 L2} shows that the PAIS-pCNL algorithm converges to the posterior distribution faster than the standard function space pCNL algorithm, in both L2 error and relative error in the moments.
+Figure~\ref{fig:MH1 L2} shows that the PAIS-RW algorithm converges to the posterior distribution significantly faster than the RWMH algorithm, in both L2 error and relative error in the moments.
 A description of the speed up attained by this algorithm is given in Section~\ref{sec:calc_saving}.
 
-Both adaptive algorithms are run with initial values of $\delta_0=0.1$.
-From Figure~\ref{fig:MH1 L2} we can see that the APAIS-pCNL algorithm converges at least as quickly as the PAIS-pCNL algorithm.
-The ApCNL algorithm initially has trouble converging to the posterior; during the initial burn-in phase, some chains find themselves a long way out in the tails where due to the high gradient they will overshoot the high density region and reject almost all proposed values.
+Both adaptive algorithms are run with initial values of $\delta_0=2.0$.
+Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance.
+The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth making it difficult to find the optimal parameter.
 
@@
 \label{fig:problem 3 posteriors}
-\end{center}
+\end{center}repo
 \end{figure}
@@
 	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$ \\ \hline
-	pCNL								& 1.9e-1					& - \\
-	PAIS-pCNL							& -						& 3.9e-2\\
+	RWMH								& 1.2e-2					& - \\
+	PAIS-RW								& -						& 3.9e-2\\
 	\hline

### 2016-04-08 PAIS.tex
Added ensemble size analysis and improved some numerics discussion

@@
 $\chi$ is a close approximation to the posterior density $\pi$.
-If the proposal distribution is over-dispersed, then the algorithm will propose states a long way from the current state at the cost of a relatively low acceptance rate, impacting the quality of inferences on the posterior.
-Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the space so the mixing rate, and hence convergence rate, will be slow.
+If the proposal distribution is widely over-dispersed, then the algorithm will propose states a long way from the current state resulting in a relatively low acceptance rate, impacting the quality of inferences on the posterior.
+Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the parameter space and so the mixing rate, and hence convergence rate, will be slow.
 It is therefore necessary to find a proposal distribution which is slightly over-dispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.
@@
 algorithm. Algorithms which use this method to find optimal proposal
-distributions are known as adaptive MCMC algorithms. Adaptive MCMC
+distributions are known as adaptive MCMC algorithms. Adaptive MH
 algorithms will converge to the stationary distribution $\pi(\cdot)$,
@@
 
-The variance of the proposal distributions that are used within the
-proposal mixture distribution plays a key role in how well the
-proposal distribution represents the target distribution, and
-therefore how quickly the algorithm converges. A large over-dispersed
-proposal distribution will sample a lot from the tails of the
-distribution, but may also find other regions of high density with
-respect to the target distribution. A very small variance in the
-proposal distributions will lead to a very rough proposal mixture
-distribution. In practise, we wish to find the happy medium in which
-the proposal distribution is slightly more dispersed than the target
-density, whilst also being a good representation of the regions which
-have high density with respect to the target distribution.
-
-Adaptively choosing the variance of the proposal distributions with a
+Algorithms which use mixture proposals, e.g. PAIS, must tune the variance of the individual kernels within the proposal mixture. Adaptively choosing the variance of the kernels starting with a
 large initial guess allows us to first explore the state space
-globally, searching for multiple modes. Reducing the proposal variances to an
+globally, searching for the locations of the modes. Reducing the proposal variances to an
 optimal value then allows us to explore each region efficiently. The
@@
 effectively what the optimal variance of the proposal distributions
-should be.
+should be. This is discussed in Section~\ref{sec:adapt}.
 
@@
 Unlike Metropolis-Hastings algorithms, the PAIS algorithm does not accept or reject proposed values, so we need another method of measuring the optimality of $\delta$.
-Section~\ref{sec:statistics} gives some possible methods of tuning $\delta$.
+Section~\ref{sec:statistics} gives some possible methods for tuning $\delta$.
 
@@
 
-\subsubsection{Determining optimal scaling parameter $\delta$ using error analysis}
+\subsubsection{Determining optimal scaling parameter using error analysis}
 
@@
 \]
-where $\{x_i\}_{i=1}^N$ is a sample of size $N$ produced by the algorithm. The relative L2 error between a continuous function to a piecewise constant function, $e$, is
+where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative L2 error between a continuous function to a piecewise constant function, $e$, is
 \begin{equation}\label{eqn:L2_error}
@@
 \end{equation}
-where the points $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional histogram bins, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of a bin, and $B_i$ is the value of the $i$th bin.
-
-These statistics are not practical for finding optimal values of
+where the regions $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional histogram bins, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of each bin, and $B_i$ is the value of the $i$th bin.
+
+These statistics cannot be used in practise to find optimal values of
 $\delta$ since they require knowledge of the analytic solution, and
-require that the algorithm be run for a long time to build up a
+the algorithm must be run for a long time to build up a
 sufficiently large sample. However they can be used to assess the
-ability of other statistical measures to find the optimal proposal
+ability of other indicators to find the optimal proposal
 variances in a controlled setting.
 
+%%% remove this paragraph?
 Another related statistic, is the variance of the sample estimate of the first moment $\hat{\mu}$.
@@
 
-The following statistics can be used for importance sampling algorithms.
-
-\subsubsection{Determining optimal $\delta$ using the variance of the weights}
-
-Importance samplers assign a weight to each sample they produce, based on a ratio of the posterior to the proposal at that point.
-This type of sampler is most efficient when the posterior is proportional to the proposal distribution, in this case the weights are constant, and so the variance of the weights, $var(w(y))$, is zero.
-Hence, the optimal value of $\delta$, is
+The following statistics can be used specifically for importance sampling algorithms.
+
+\subsubsection{The variance of the weights}
+
+Importance samplers assign a weight to each sample they produce based on a ratio of the posterior to the proposal at that point.
+This method is most efficient when the target is proportional to the proposal distribution, in this case the weights are constant and so the variance of the weights, $var(w(y))$, is zero.
+Hence, we would like to choose the value of $\delta$ which minimises the variance of the weights,
 \[
@@
 \]
-The mean of the estimator $\text{var}(w(y))$ is a smooth function, and so it is used to tune $\delta$ during the burn-in phase of MCMC algorithms.
-The variance of the estimator can be large, especially far away from the optimal value, so it can take a large number of samples to calculate descent directions.
-
-\subsubsection{Determining optimal $\delta$ using the effective sample size}\label{sec:ess}
-
-The effective sample size, $\neff$, can be used to assess the efficiency of importance samplers.
+The mean of the estimator $\text{var}(w(y))$ is smooth enough that it can be used to tune $\delta$ during the burn-in phase of MCMC algorithms.
+However the variance of the estimator can be large, especially far away from the optimal value, so it can take a large number of iterations to calculate descent directions.
+
+\subsubsection{The effective sample size}\label{sec:ess}
+
+The effective sample size, $\neff$, can also be used to assess the efficiency of importance samplers.
 Ideally, in each iteration, we would like all $M$ of our samples to provide us with new information about the posterior distribution.
-In practise, we are unlikely to achieve a ratio of exactly 1.
+In practise, we cannot achieve a perfect effective sample size of $M$.
 
@@
 
-\begin{figure}[htpb]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/neff-M}
-\caption{Left: A contour plot showing how $\delta_{\text{eff}}$ varies with the ensemble size.
-The black line highlights the optimal value of $\delta$.
-Right: The value of the effective sample size ratio for the optimal $\delta$ at each ensemble size.
-i.e.
-the value of the effective sample size ratio along the black line in the left figure.
-Set up for this problem is given in Section~\ref{sec:problem 2}.}
+\begin{figure}[htb]
+\centering
+\subfigure[Contours showing optimal ranges of the scaling parameter.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM_contour"}}
+\subfigure[The highest effective sample size achieved for each ensemble size.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM"}}
+\caption{The behaviour of the effective sample size as the ensemble size increases. This analysis is from the Gaussian posterior in Section~\ref{sec:problem 1} using the PAIS-RW algorithm.}
 \label{fig:neff-M}
-\end{center}
 \end{figure}
@@
 in different situations.
-In Section \ref{Sec:Num} we see how it behaves for different observation operators.
 Here we look at how the statistic behaves as we vary the ensemble size, $M$.
-Figure~\ref{fig:neff-M} shows results for the Gaussian posterior discussed in \ref{sec:problem 2} when using the PAIS-MALA algorithm.
-The left part of the figure shows that as the ensemble size increases, the value of $\delta$ which gives the optimal effective sample size ratio decreases.
+Figure~\ref{fig:neff-M} shows results for the Gaussian posterior discussed in Section~\ref{sec:problem 1} when using the PAIS-RW algorithm.
+Figure~\ref{fig:neff-M} (a) shows that as the ensemble size increases, the value of $\delta$ which gives the optimal effective sample size ratio decreases.
 This is to be expected since we are trying to fit more kernels into the posterior distribution, so the variance of each kernel needs to be decreased.
+
+Figure~\ref{fig:neff-M} (b) shows that as the ensemble size increases, the efficiency of the sampler also increases.
 
@@
 uninformative prior for the scaling parameter.
-Alternatively, the parameter may be randomly sampled from some interval at various points during the chain, a benefit being that it allows some exploration of the state space but never converges to an optimal value of the scaling parameter.
-We choose to use a divide and conquer scheme which optimises the effective sample size.
+Alternatively, the parameter may be randomly sampled from some interval at various points during the chain.
+This results in some iterations which explore the state space, and some which make local moves.
+Algorithms of this type do not converges to an optimal value of the scaling parameter.
+
+We choose to use a divide and conquer scheme which optimises the effective sample size (or any other diagnostic).
 Some more sophisticated examples are described in \cite{roberts2009examples} and \cite{Ji2013adaptive}.
-The proposed algorithm, is not presented as an optimal strategy but as an example of the benefits of tuning the algorithm using the effective sample size statistic.
-
-From hereon in, we will assume that we are using the proposal from the
-pCNL algorithm given in Section \ref{Sec:pCNL}, which is given by:
-\begin{equation}Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
-2\delta\mathcal{C}\nabla \Phi(u)+
-\sqrt{8\delta} W\right], \qquad W \sim \mu_0,
+Algorithm~\ref{tab:adapt} in Appedix~\ref{Sec:App}, is not presented as an optimal strategy but as a demonstration that the PAIS algorithm can be adaptively tuned effectively.
+
+From here on in, we will use the random walk proposal,
+\begin{equation}Y = X_{i-1}+\beta W_{i-1}, \qquad W_{i-1} \sim \mathcal{N}(0,1),
 \end{equation}
-where $X$ is our current state and $Y$ is our proposed state. The
-scaling parameter $\delta$ dictates the variance in the proposal distribution.
-
- In the adaptive algorithm described in Table~\ref{tab:adapt} in
- Appendix \ref{Sec:App}, we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm.
-We must choose some sequence of iterations at which to update $\delta^{(j)}$, $\{n_k\}_{k=1}$, which can be decided using a sequence in which the terms grow exponentially further apart.
-The algorithm is given for the PAIS method but a similar method is used to adaptively calculate $\delta^*$ for the pCNL algorithm.
+where $X_{i-1}$ is our current state, $Y$ is our proposed state, and $\beta = \sqrt{8\delta}/(2+\delta)$. In the numerics section we will refer to $\delta$ as the scaling parameter.
+
+ In Algorithm~\ref{tab:adapt}, we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges roughly to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm.
+We must choose some sequence of iterations, $\{n_k\}_{k=1}$, at which to update $\delta$.
+These $n_k$ can be constructed by a sequence in which the terms grow exponentially further apart.
+The algorithm is given for the PAIS method but a similar method is used to adaptively calculate the optimal scaling parameter, $\delta^*$, for the MH algorithm.
 
 When implemented, there are a number of parameters which must be
-chosen to allow efficient tuning of $\delta$. Firstly, the initial
-value of $\delta$ should be chosen so that the chains spread out
-quickly across the state space. However, if it is chosen to be too
-large, then the method may become unstable and inefficient.
+chosen to ensure efficient tuning of $\delta$. Firstly, the initial
+value of $\delta$ should be chosen so that the chains quickly explore the parameter space. However, if $\delta$ is chosen to be too
+large, then the method may become unstable.
 We also choose two iteration numbers at which we change our adaptive strategy.
-At iteration $N_\text{join}$ we resample using the full ensemble size instead of using to subsets of the full ensemble.
-This allows us to fine-tune the scaling parameter using the full resampler, but slows down the rate of convergence of the scaling parameter.
-Finally we choose a time, $N_\text{stop}$, at which we stop updating our parameter to guarantee ergodicity of the MCMC algorithm.
+At iteration $N_\text{join}$ we resample using the full ensemble instead of using two subsets of the full ensemble.
+When using two subsets we are training the scaling parameter using the wrong ensemble size, this allows faster convergence but leads to a sub-optimal scaling parameter.
+Resampling with the full ensemble towards the end of the tuning allows fast convergence to the right order of magnitude, followed by fine tuning to the optimal scaling parameter.
+Finally we choose a time, $N_\text{stop}$, at which we stop updating our parameter.
 
@@
 
-In this example we look at how PAIS compares to naively parallelised
-MH. We compare the RWMH algorithm with its
-PAIS variant, the PAIS-RW algorithm. We consider several statistics
-for measuring the efficiency of the PAIS algorithm compared to
-existing Metropolis-Hastings algorithms when applied to a simple one
-dimensional Gaussian posterior.
-
-Since we are comparing against naively parallelised MCMC algorithms,
+In this example we compare the naively parallelised RWMH algorithm with its
+PAIS variant, the PAIS-RW algorithm. We perform this comparison using the relative L2 error defined in Equation~\ref{eqn:L2_error}, as well as the relative error in the first moment.
+
+Since we are comparing against naively parallelised MH algorithms,
 we also need to decide which statistic $T(\delta)$ to optimise for
@@
 \begin{equation}\label{eqn:Gaussian posterior}
-	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_\Sigma - \frac{1}{2}\big\|u\big\|^2_{\mathcal{T}}\right),
+	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
 \end{equation}
-where $\Sigma$ and $\mathcal{T}$ are the covariances of the
+where $\sigma^2$ and $\tau^2$ are the variances of the
 observational noise and prior distributions respectively. In the
-numerics which follow, we choose $\Sigma = \sigma^2I_N$ and
-$\mathcal{T} = \tau^2I_d$ with $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
+numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
 we observe $u=4$ noisily such that
 \[
-D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\Sigma).
+D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2).
 \]
@@
 parameters we choose 32 values of $\delta$ evenly spaced on a log scale
-in he interval $[10^{-5}, 2]$.
-We run the PAIS-RW algorithm for 100,000 iterations and RWMH for 1,000,000 iterations, each with $M=50$ processes.
+in the interval $[10^{-5}, 2]$.
+We run the PAIS-RW and RWMH algorithms for one million iterations, each with an ensemble size of $M=50$.
 We took 32 repeats of both algorithms and then used the geometric means of the sample statistics to find the optimal parameters.
 
-{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for 1 million iterations, again with 50 processes.
+{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for one million iterations, again with $M=50$.
 The simulations are repeated 32 times using the optimal parameters found in (1).
@@
 algorithms, and again use the relative L2 error to compare
-efficiency. The initial value of $\delta$ is given in the discussion
-of each simulation.
+efficiency. The adaptive algorithms are initialised with $\delta^{(1)}$ = 2.
 
@@
 Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance.
-The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth making it difficult to find the optimal parameter.
+The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth at the optimal value making it difficult to minimise.
 
@@
 
-Throughout this paper, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales if we were to increase the ensemble size, and if there is some limit below which the algorithm fails.
-We implement the problem in Section~\ref{sec:problem 1}, using the pCNL algorithm with ensemble sizes ranging from $M=2$ up to $M=494$.
+Throughout this paper, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales when we increase the ensemble size, and if there is some limit below which the algorithm fails.
+We implement the problem in Section~\ref{sec:problem 1}, using the RWMH and PAIS-RW algorithms with ensemble sizes in the inteveral $M \in [1, 160]$.
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{"figures/PAIS_saving"}
-\caption{Ratio of PAIS-pCNL samples required to reach the same tolerance as the pCNL algorithm.}
+\includegraphics[width=0.7\textwidth]{"figures/PAIS_saving"}
+\caption{Ratio of PAIS-RW samples required to reach the same tolerance as the RWMH algorithm.}
 \label{fig:PAIS_saving}
@@
 Figure~\ref{fig:PAIS_saving} was produced using the method of finding
-optimal $\delta$ described in Section~\ref{sec:Implementation P1}(1),
+optimal $\delta$ described in Section~\ref{sec:Implementation P1} (1),
 then running 32 repeats at each ensemble size. The convergence rates
@@
 noisy but demonstrates that increasing the ensemble size continues to
-reduce the number of iterations required in comparison with naive
-MCMC. When $M<8$, the algorithm takes a long time to reach
-stationarity. This indicates superlinear improvement of PAIS with
+reduce the number of iterations required in comparison with naively parallelised MH. When $M<8$, the algorithm takes a long time to reach
+stationarity. The decreaseing trend indicates superlinear improvement of PAIS with
 respect to ensemble size, in terms of the number of iterations
 required, which is a demonstration of our belief that parallelism of
-MCMC should give us added value over and above that provided by naive parallelism.
+MCMC should give us added value over and above that provided by naive parallelism. This decrease is linked to the increasing effective sample size shown in Figure~\ref{fig:neff-M} (b).
 
@@
 In this section we investigate the behaviour of the PAIS algorithm
-when applied to bimodal problems. Metropolis-Hastings methods can
+when applied to bimodal problems. MH methods can
 struggle with multimodal problems, particularly where switches between
 the modes are rare, resulting in incorrectly proportioned modes in the
-histograms, for example. With the PAIS algorithm, we see that the
-resampling step redistributes chains to new modes as they are
-found. This means that we expect the number of chains in a mode to be
-approximately proportional to the probability mass in that mode.
-As a result, reconstructed posteriors with disproportional modes, as is familiar with the Metropolis-Hastings algorithms, are not produced.
-We again look at an `easy' problem, BM(1), which has a KL divergence of 0.880, and a `harder' problem, BM(2), which has a KL divergence of 3.647.
-Problem BM(1) has two modes which are separated by a smaller energy barrier.
-In BM(2) we increase the distance between the two modes which has the effect of increasing the required energy to jump between modes.
+histograms. This example demonstrates that the PAIS algorithm redistributes 
+chains to new modes as they are found. This means that we expect the number of chains in a mode to be
+approximately proportional to the probability density in that mode. As a result, reconstructed posteriors with disproportional modes, as is familiar with the MH algorithms, are not produced.
+
+We look at an `easy' problem, $B_1$, which has a KL divergence of 0.880, and a `harder' problem, $B_2$, which has a KL divergence of 3.647.
+Problem $B_1$ has two modes which are separated by a small energy barrier.
+In $B_2$ we increase the distance between the two modes which has the effect of increasing the required energy to jump between modes.
 These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.
@@
 The following setup is the same for both problems.
-We consider an observation operator $\G(u) = u^2$, and assign the prior $u \sim \mu_0 = \N(0, \tau^2=0.25)$.
+We consider a nonlinear observation operator $\G(u) = u^2$, and assign the prior $u \sim \mu_0 = \N(0, \tau^2=0.25)$.
 We assume that a noisy reading, $D$, is taken according to $D = \G(u) + \varepsilon$, where $\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$.
@@
 \includegraphics[width=\textwidth]{"figures/posteriors3"}
-\caption{Posterior distributions for problem BM(1) and BM(2).
-Problem BM(1) has a noisy data value of  0.921312 which results in an energy barrier which is relatively easy to cross, as well as their common prior distribution.}
+\caption{Posterior distributions for problem $B_1$ and $B_2$.
+Problem $B_1$ has a noisy data value of  0.921312 which results in an energy barrier which is relatively easy to cross, as well as their common prior distribution.}
 \label{fig:problem 3 posteriors}
-\end{center}repo
+\end{center}
 \end{figure}
@@
 
-The numerical implementation for most of the following simulations follow the same setup as described in Section~\ref{sec:Implementation P1}.
-The only exception is that the convergence plots for both adaptive and nonadaptive algorithms are run for $10^6$ iterations instead of $10^7$.
+The numerical implementation for the following simulations follow the same setup as described in Section~\ref{sec:Implementation P1}.
 
@@
 
-Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for pCNL and we use the effective sample size to find the optimal values for PAIS-pCNL.
+Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for RWMH and we use the effective sample size to find the optimal values for PAIS-RW.
 Table~\ref{table:BM_opt_delta} gives the optimal values of $\delta$ for both problems.
@@
 	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$ \\ \hline
-	RWMH								& 1.2e-2					& - \\
-	PAIS-RW								& -						& 3.9e-2\\
+	RWMH								& 1.3e-1					& - \\
+	PAIS-RW								& -						& 5.4e-3\\
 	\hline
@@
 	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$	& $\delta^*_{\text{L2}}$ \\ \hline
-	pCNL								& 5.8e-2					& - 						& 9.1e-1\\
-	PAIS-pCNL							& -						& 2.6e-2 					& 2.6e-2\\
+	RWMH								& 2.6e-2					& - 						& 9.1e-1\\
+	PAIS-RW								& -						& 1.1e-3 					& 8.1e-3\\
+	MALA								& 2.6e-2					& - 						& 1.3e-1\\
+	PAIS-MALA							& -						& 3.7e-3 					& 8.1e-3\\
 	\hline
@@
     \end{minipage}
-	\caption{Optimal values of $\delta$ for BM(1) (left) and BM(2) (right).}
+	\vspace{1mm}
+	\caption{Optimal values of $\delta$ for $B_1$ (left) and $B_2$ (right).}
 	\label{table:BM_opt_delta}
@@
 
-In problem BM(1), the pCNL algorithm has the
-higher value of $\delta$, which corresponds to a close to independence
-sampling, effectively sampling from the prior. This is because the
-regions with the majority of the target density are well covered by
-the prior. The PAIS-pCNL algorithm samples more efficiently from its
-mixture $\chi$, and so a lower value of $\delta$ is more efficient.
-
-Problem BM(2) is much harder than BM(1); transitions between the modes
+Problem $B_2$ is much harder than $B_1$; transitions between the modes
 are extremely unlikely for the standard pCNL algorithm. This means that we need to consider the
@@
 
-To get correctly proportioned modes with the pCNL algorithm it is
+To get correctly proportioned modes with the RWMH algorithm it is
 important that the chains can transition between the modes, which
 means that $\delta$ must be large. However, this leads to a lower
-acceptance rate, and so we sacrifice convergence locally. The prior distribution $\mu_0$ is
-not a good approximation of the posterior distribution, and so too
-large a value of $\delta$ (which leads to an independence sampler
-using the prior as a proposal distribution) leads to an inefficient
-method sampling. For these reasons, the pCNL algorithm is very slow to
+acceptance rate, and so we sacrifice convergence locally. For this reason, the RWMH algorithm is very slow to
 converge for problems of this type.
 
-We can achieve these two regimes in pCNL by tuning $\delta$ using the acceptance rate for local convergence, and by L2 error for global convergence.
-Similarly in PAIS-pCNL we can use the effective sample size for local convergence, and the L2 error for global convergence.
-
-From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime, meaning that both will result in inefficient sampling.
-The PAIS-pCNL algorithm manages to sample the detail and the large scale behaviour with the same value of $\delta^*$: a clear advantage to using this algorithm for this problem.
-
-\subsubsection{Convergence of pCNL vs PAIS-pCNL}
-
-We see a significant speed with the PAIS-pCNL algorithm for BM(1). Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates.
+We can achieve these two regimes in RWMH by tuning $\delta$ using the acceptance rate for local convergence, and by L2 error for global convergence.
+Similarly in PAIS-RW we can use the effective sample size for local convergence, and the L2 error for global convergence.
+
+From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime in both RWMH and MALA, meaning that both will result in inefficient sampling.
+The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\delta^*$: a clear advantage to using this algorithm for this problem.
+
+\subsubsection{Convergence of RWMH vs PAIS-RW}
 
@@
 \caption{Error analysis for the PAIS-pCNL and pCNL algorithms for
-  problem BM(1).
+  problem $B_1$.
 The solid blue line, and dashed red line compare the algorithms with fixed optimal scaling parameters, and the blue crossed and magenta crossed lines compare the adaptive algorithms.
@@
 
+As in the Gaussian example we see a significant speed up with the PAIS-RW algorithm for $B_1$.
+Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates.
 We can see that the adaptive algorithms compare closely with the
@@
 
-For BM(2) the algorithms are run with the global optimal value of $\delta^*$, and with the local optimal value of $\delta^*$.
+For $B_2$ the algorithms are run with the global optimal value of $\delta^*$, and with the local optimal value of $\delta^*$.
 Figure~\ref{fig:BM2_L2} shows that the algorithms using the globally optimal $\delta^*$ convergence diagnostics converge slowly towards the true posterior after a long burn-in period, whereas the algorithms using the local optimal $\delta^*$ converge quickly, but get stuck in one mode meaning that the convergence rate flattens out.
@@
 \caption{pCNL and PAIS-pCNL convergence statistics using locally and
-  globally optimal $\delta$, for problem BM(2). The setup is as described in Section~\ref{sec:Implementation P2} (2).}
+  globally optimal $\delta$, for problem $B_2$. The setup is as described in Section~\ref{sec:Implementation P2} (2).}
 \label{fig:BM2_L2}
@@
 \includegraphics[width=\textwidth]{"figures/BM2_AL2"}
-\caption{Convergence graphs for problem BM(2), the (A)pCNL and (A)PAIS-pCNL algorithms have been run with optimal $\delta^*$ for the L2 error and the adaptive algorithm described in Section~\ref{sec:adapt}.
+\caption{Convergence graphs for problem $B_2$, the (A)pCNL and (A)PAIS-pCNL algorithms have been run with optimal $\delta^*$ for the L2 error and the adaptive algorithm described in Section~\ref{sec:adapt}.
 The setup is as described in Section~\ref{sec:Implementation P2} (3).}
@@
 The biggest issue for the Metropolis-Hastings algorithms when sampling
-from a posterior such as the one in BM(2) is that it is unlikely that
+from a posterior such as the one in $B_2$ is that it is unlikely that
 the correct ratio of chains will occur in each of the modes, and since there is no interaction between the chains, there is no way to remedy this problem.
@@
 In this way, the algorithm allows chains to `teleport' to regions of the posterior which are in need of more exploration.
-Figure~\ref{fig:BM2_suction} shows Problem BM(2) with initially 1 chain in the positive mode, and 49 chains in the negative mode.
+Figure~\ref{fig:BM2_suction} shows Problem $B_2$ with initially 1 chain in the positive mode, and 49 chains in the negative mode.
 It takes only a handful of iterations for the algorithm to balance out the chains into 25 chains in each mode.

### 2016-04-12 PAIS.tex
Updated problem B2

@@
 
-%Ideally the dimension of particles in a filter will not affect the convergence of the method.
-Unfortunately, the approximation of continuous density functions is computationally infeasible in high dimensions and the required ensemble size scales exponentially with the dimension of the problem\cite{silverman1986density,snyder2008obstacles}.
-
-%\begin{figure}[htb]
-%\begin{center}
-%\includegraphics[width=\textwidth]{figures/ET_dim_scaling}
-%\caption{Demonstration of the ETMC algorithm rapidly failing as the dimension increases. The relative error shown is that between the sampled mean and the true mean.}
-%\label{fig:ET_dim_scaling}
-%\end{center}
-%\end{figure}
-
-%Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm loses its ability to accurately preserve the mean of the posterior distribution as the dimension increases.
-Clearly for a simple Gaussian problem, we can not resample in much more than 10 dimensions and hope to recover a representative sample, without hugely increasing the number of particles in the ensemble.
-As the cost of the ETMC is $\mathcal{O}(n^2)$, this does limit the dimension for which this approach can be used.
-
 \subsection{Deficiencies of Metropolis-type MCMC schemes}
@@
 
-Important sampling can be a very efficient method for sampling from a
+Importance sampling can be a very efficient method for sampling from a
 probability distribution. A proposal density is chosen, from which we
@@
 variances in a controlled setting.
-
-%%% remove this paragraph?
-Another related statistic, is the variance of the sample estimate of the first moment $\hat{\mu}$.
-Assuming that the algorithm is converging to the invariant distribution, convergence will be fastest when the variance of the estimate is minimised.
-This fact follows from the convergence rate of Monte Carlo algorithms, $\sigma/\sqrt{n}$.
-The proposal distribution which minimises the variance, $\sigma^2$, of $\hat{\mu}$, is the most desirable.
-This variance statistic often converges faster than the moment itself, and does not require any knowledge of the true value of the moment.
 
@@
 
-\subsubsection{Numerical Implementation}\label{sec:Implementation P2}
-
-The numerical implementation for the following simulations follow the same setup as described in Section~\ref{sec:Implementation P1}.
-
 \subsubsection{Calculating values of Optimal $\delta^*$}\label{sec:BM1_opt_delta}
@@
 	RWMH								& 2.6e-2					& - 						& 9.1e-1\\
-	PAIS-RW								& -						& 1.1e-3 					& 8.1e-3\\
-	MALA								& 2.6e-2					& - 						& 1.3e-1\\
-	PAIS-MALA							& -						& 3.7e-3 					& 8.1e-3\\
+	PAIS-RW								& -						& 1.3e-3 					& 8.1e-3\\
 	\hline
@@
 
-Problem $B_2$ is much harder than $B_1$; transitions between the modes
-are extremely unlikely for the standard pCNL algorithm. This means that we need to consider the
+It is relatively simple to find optimal scaling parameters for problem $B_1$.
+These values are given in Table~\ref{table:BM_opt_delta} (left).
+However problem $B_2$ is much harder; transitions between the modes are extremely unlikely for the standard RWMH algorithm.
+This means that we need to consider the
 convergence on two levels; we should consider the algorithm's ability
-to find all the modes, and to sample them thoroughly and in the correct proportions.
+to find all the modes, and also whether it can sample them in the correct proportions.
 
@@
 
-From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime in both RWMH and MALA, meaning that both will result in inefficient sampling.
+From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime in RWMH, and so will result in inefficient sampling.
 The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\delta^*$: a clear advantage to using this algorithm for this problem.
@@
 
-\begin{figure}[htpb]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/BM1_L2"}
-\caption{Error analysis for the PAIS-pCNL and pCNL algorithms for
-  problem $B_1$.
-The solid blue line, and dashed red line compare the algorithms with fixed optimal scaling parameters, and the blue crossed and magenta crossed lines compare the adaptive algorithms.
-The setup is as described in Section~\ref{sec:Implementation P2} (2,3).}
+\begin{figure}[htb]
+\centering
+\subfigure[Relative L2 error.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_L2"}}
+\subfigure[Absolute error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_moments"}}
+\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_1$. Set up described in Section~\ref{sec:Implementation P1}.}
 \label{fig:BM1_L2}
-\end{center}
 \end{figure}
@@
 We can see that the adaptive algorithms compare closely with the
-respective nonadaptive algorithms and the improvement PAIS offers
-remains significant.
+respective nonadaptive algorithms and the improvement PAIS offers remains significant.
+
+\begin{figure}[htb]
+\centering
+\subfigure[Relative L2 error using the optimal scaling parameters.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
+\subfigure[Comparing of the relative L2 error of the algorithms optimised by L2 minimisation with locally optimised PAIS-RW with a scout chain.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
+\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_2$. Set up described in Section~\ref{sec:Implementation P1}.}
+\label{fig:BM2_L2}
+\end{figure}
 
 For $B_2$ the algorithms are run with the global optimal value of $\delta^*$, and with the local optimal value of $\delta^*$.
-Figure~\ref{fig:BM2_L2} shows that the algorithms using the globally optimal $\delta^*$ convergence diagnostics converge slowly towards the true posterior after a long burn-in period, whereas the algorithms using the local optimal $\delta^*$ converge quickly, but get stuck in one mode meaning that the convergence rate flattens out.
-
-\begin{figure}[h]
+Figure~\ref{fig:BM2_L2} (a) shows that the algorithms using the globally optimal $\delta^*$ convergence at the desired rate, whereas the algorithms optimised using the acceptance rate and effective sample size initially converge faster but at some point forget the location of one of the modes, causing the convergence to flatten out.
+
+Using the PAIS algorithm we can minimise the impact of forgotten modes by constantly allowing the algorithm to search them out.
+Since we have parallel chains, we can run PAIS-RW with the majority of chains using the locally optimal value of $\delta$, and one or two chains with a larger scaling parameter.
+These chains with larger scaling parameters act both as `scouts' for new modes, and act to aid in the overdispersal of the proposal distribution.
+Figure~\ref{fig:BM2_L2} (b) shows the results of using 49 chains with the local optimal scaling parameter, and one chain with ten times the local optimal scaling parameter.
+We see that modes are not forgotten and the algorithm converges with the improvement we see from PAIS-RW in the other problems.
+Other methods of mode searches are described in \cite{lan2013wormhole}.
+
+The adaptive algorithm can just as easily be applied to the PAIS algorithm with the `scout' chains described in the previous paragraph.
+Since we need two equally sized sub-ensembles, we will use two groups of 24 ensembles with the same proposal distributions, and each group will also have a `scout' chain with a scaling parameter ten times that of the rest of the group.
+
+\begin{figure}[htb]
 \begin{center}
-\includegraphics[width=\textwidth]{"figures/BM2_L2"}
-\caption{pCNL and PAIS-pCNL convergence statistics using locally and
-  globally optimal $\delta$, for problem $B_2$. The setup is as described in Section~\ref{sec:Implementation P2} (2).}
-\label{fig:BM2_L2}
-\end{center}
-\end{figure}
-
-The adaptive algorithm as stated in Section~\ref{sec:adapt} is one way of combining the two regimes.
-Another method uses a small number of `scout' chains with a large $\delta$ to continually search out new modes.
-Other methods of mode searches are described in \cite{lan2013wormhole}, and the regeneration method is applicable\cite{nummelin1984general}.
-
-Figure~\ref{fig:BM2_AL2} shows the success of the (A)PAIS-pCNL algorithms in converging to the posterior, compared to the (A)pCNL algorithms.
-We can see that using the pCNL algorithm for this problem would be infeasible.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/BM2_AL2"}
-\caption{Convergence graphs for problem $B_2$, the (A)pCNL and (A)PAIS-pCNL algorithms have been run with optimal $\delta^*$ for the L2 error and the adaptive algorithm described in Section~\ref{sec:adapt}.
-The setup is as described in Section~\ref{sec:Implementation P2} (3).}
+\includegraphics[width=0.7\textwidth]{"figures/BM2_AL2"}
+\caption{Convergence of the relative L2 error for problem $B_2$, comparing the globally optimised nonadaptive algorithms with the locally optimised adaptive algorithms.
+The setup is as described in Section~\ref{sec:Implementation P1} (3).}
 \label{fig:BM2_AL2}
@@
 
+Comparing the convergence of the adaptive algorithms against the nonadaptive algorithms in Figure~\ref{fig:BM2_AL2} shows that the algorithms behave as expected.
+The adaptive RWMH algorithm tuned using the acceptance rate converges at the same rate as the L2 optimised algorithm until the scaling parameter gets small where the algorithm forgets one of the modes.
+The adaptive PAIS-RW with scout chains tuned to the effective sample size converges at about the same rate as the locally optimised nonadaptive algorithm also with scouts.
+
 \subsubsection{Calculating the Speed Up in Convergence}\label{sec:calc_saving}
@@
 \includegraphics[width=\textwidth]{"figures/calc_saving"}
-\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2}$ as a percentage of pCNL iterations.
+\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2.5}$ as a percentage of pCNL iterations.
 The relative L2 error graphs are from the problem in Section~\ref{sec:problem 1}.}
@@
 
-The graphs in the previous section clearly show that the PAIS-pCNL algorithm converges faster than the pCNL algorithm when both are parallelised with the same number of threads.
+The graphs in the previous section clearly show that the PAIS-RW algorithm converges faster than the RWMH algorithm when both are parallelised with the same ensemble size.
 We can calculate the number of iterations required to achieve a particular tolerance level in our solution for each algorithm and compare these to calculate a percentage saving.
@@
 A summary of the percentage of iterations required using the PAIS algorithm compared with the respective Metropolis-Hastings algorithms is given in Table~\ref{table:calc_savings}.
-The blank entries correspond to occasions when the MH algorithms haven't converged to the posterior distribution.
+The blank entries correspond to occasions when either the MH algorithm or PAIS algorithm hasn't converged to the posterior distribution.
 
@@
 \centering
-\begin{tabular}{|l|l|r|r|}
+\begin{tabular}{|l|r|r|r|}
 \hline
-		& & Low KL div & High KL div \\ \hline
-	\multirow{4}{*}{Gaussian}	&RWMH & 14\% & 14\% \\
-	&pCN & 33\% & - \\
-	&MALA & 41\% & 42\% \\
-	&pCNL & 62\% & 66\% \\ \hline
-	\multirow{4}{*}{Bimodal}	&RWMH & 40\% & 44\% \\
-	&pCN & 32\% & - \\
-	&MALA & 36\% & 40\% \\
-	&pCNL & 56\% & - \\ \hline
+		& Gaussian & $B_1$ & $B_2$ \\ \hline
+	RWMH & 10\% & 32\% & 12.6\% (scout) \\
+	pCN & - & 32\% & - \\
+	MALA & 42\% & 36\% & 40\% \\
+	pCNL & 66\% & 56\% & - \\ \hline
 \end{tabular}
-\caption{Iterations for the PAIS algorithms required to achieve a desired tolerance as a percentage of the number of iterations required by the respective Metropolis algorithms.}
+\caption{Iterations for the PAIS algorithms required to achieve a desired tolerance as a percentage of the number of iterations required by the respective MH algorithms.
+The pCN and pCNL proposal distributions are taken from~\cite{cotter2013mcmc}.}
 \label{table:calc_savings}

### 2016-04-13 PAIS.tex
missed a bit

@@
 \includegraphics[width=\textwidth]{"figures/calc_saving"}
-\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2.5}$ as a percentage of pCNL iterations.
-The relative L2 error graphs are from the problem in Section~\ref{sec:problem 1}.}
+\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2.5}$ as a percentage of MH iterations.
+The relative L2 error graphs are from the Gaussian problem in Section~\ref{sec:problem 1}.}
 \label{fig:calc_saving}

### 2016-05-30 PAIS.tex
Chemical system added

@@
 \setlength{\parindent}{0in}
-\usepackage{amsmath,amsfonts,amsbsy}
+\usepackage{amsmath,amsfonts,amsbsy,amssymb}
 \newcommand{\RARR}[3]{#1
@@
 }}
+
+\makeatletter  
+\newcommand{\xleftrightarrows}[2][]{\mathrel{%  
+ \raise.40ex\hbox{$  
+       \ext@arrow 3095\leftarrowfill@{\phantom{#1}}{#2}$}%  
+ \setbox0=\hbox{$\ext@arrow 0359\rightarrowfill@{#1}{\phantom{#2}}$}%  
+ \kern-\wd0 \lower.4ex\box0}}
+
+\newcommand{\xrightleftarrows}[2][]{\mathrel{%  
+ \raise.40ex\hbox{$\ext@arrow 3095\rightarrowfill@{\phantom{#1}}{#2}$}%  
+ \setbox0=\hbox{$\ext@arrow 0359\leftarrowfill@{#1}{\phantom{#2}}$}%  
+ \kern-\wd0 \lower.4ex\box0}}
+
+\def\leftrightarrowfill@{%
+ \arrowfill@\leftarrow\relbar\rightarrow%
+ }  
+\makeatother
+
 \author{Colin Cotter\thanks{Department of Mathematics, Imperial
@@
 
+\subsection{Sampling from non-Gaussian bivariate distributions}\label{sec:chem}
+%***some introduction to chemical systems and QSSA
+In this section we apply the PAIS algorithm to a more complicated posterior distribution.
+The field of biochemical kinetics gives rise to multiscale stochastic problems which remain a challenge both theoretically and computationally.
+Biochemical reactions occur in single cells between a number of chemical populations and the rates of these reactions can vary on vastly different timescales.
+It is these reaction rates which we are interested in finding descriptions for.
+
+It is often possible to isolate subsystems of fast and slow reactions.
+The quasi-steady-state assumption (QSSA) is that these fast subsystems converge in distribution much faster than the slow subsystem.
+Approximating the fast subsystem as a steady state results in a simple analysis for the fast reaction rates which we can use to simulate the full system and obtain distributions for each of the reaction rates.
+
+We introduce the simple chemical system,
+\begin{equation}\label{eqn:full_system}
+	\emptyset \xrightarrow{k_1} X \xleftrightarrows[k_3]{k_2} Y \xrightarrow{k_4} \emptyset,
+\end{equation}
+where reactions $X\rightarrow Y$ and $Y\rightarrow X$ happen much faster than the creation and destruction of $X$ and $Y$.
+Using the QSSA we can combine the chemical species $X$ and $Y$ into a single species $S=X+Y$, and represent our system as,
+\begin{equation}\label{eqn:QSSA_system}
+	\emptyset \xrightarrow{k_1} S \xrightarrow{\hat{k}_4} \emptyset.
+\end{equation}
+Here, the new reaction rate $\hat{k}_4$ is found to be
+\[
+	\hat{k}_4(s) = \mathbb{E}\left[k_4Y|S=s\right] = \frac{k_2k_4S}{k_2+k_3}, \quad \text{since} \quad Y \sim \text{Bin}\left(S, \frac{k_2}{k_2+k_3}\right).
+\]
+By observing the system in Equation~\ref{eqn:QSSA_system}, we indirectly observe the rates $k_2$ and $k_3$ through $\hat{k}_4$, resulting in infinitely many possibilities for $k_2$ and $k_3$.
+Along with the intrinsic randomness and errors in observations, which we assume to follow a Gamma distribution with some variance $\sigma^2$, this results in a long and thin posterior distribution when put into the framework of a Bayesian inverse problem.
+
+% solution to QSSA ode and Gamma noise assumption
+\subsubsection{Target Distribution}
+
+We now formalise the posterior of interest.
+We look for a distribution over the parameter $\mathbf{k} = (k_2, k_3)^T$, given that we know $k_1 = 100$ and $k_4=1$.
+We generate our data by making ten observations of the system at $t_i = 2, 4, \dots, 20$, here simulated by solving the full system, Equation~\ref{eqn:full_system}, governed by the differential equations
+\begin{align*}
+	\frac{\text{d}X}{\text{d}t} &= k_1 - k_2X(t)+k_3Y(t), \\
+	\frac{\text{d}Y}{\text{d}t} &= k_2X(t) - (k_3+k_4)Y(t),
+\end{align*}
+with initial conditions $X(0)=Y(0) = 0$, and parameter values $\mathbf{k} = (50, 100)^T$.
+We then add noise taken from a Gamma distribution with variance $\sigma^2 = 225$ and centred at each $S(t_i)=X(t_i)+Y(t_i)$.
+
+In our modelling we use the QSSA to simplify this system of differential equations into the one parameter system based on Equation~\ref{eqn:QSSA_system},
+\[
+	\frac{\text{d}S}{\text{d}t} = k_1 - \frac{k_2k_4}{k_2+k_3}S(t).
+\]
+The observation operator, $\mathcal{G}: (\mathbf{k},t) \mapsto S(t)$, maps from parameter space onto population space at a time $t$. This means that for the $i$th observation we assume,
+\[
+	 D_i \sim \text{Gamma}(\alpha_i, \beta_i),
+\]
+where
+\[
+	 \alpha_i = \frac{\mathcal{G}(\mathbf{k}, t_i)^2}{\sigma^2},\  \beta_i = \frac{\mathcal{G}(\mathbf{k}, t_i)}{\sigma^2}, \quad i = 1, \dots, 10.
+\]
+
+We assign Gamma priors to $\mathbf{k}$ with mean $\alpha_0/\beta_0 = 75$ and variance $\alpha_0/\beta_0^2 = 100$, resulting in the posterior
+\[
+	\pi(\mathbf{k}|\mathbf{D}) \propto \left[\prod\limits_{i=1}^{10} \text{Gamma}(D_i; \alpha_i, \beta_i)\right]\text{Gamma}(k_2; \alpha_0, \beta_0)\text{Gamma}(k_3; \alpha_0, \beta_0).
+\]
+
+\subsubsection{Implementation}\label{sec:chem_implementation}
+% explanation of l2 convergence measure
+For this problem there is no analytic form for the normalisation constant, and numerical methods implemented in MATLAB and Mathematica have proven to be unreliable.
+To demonstrate the convergence of the algorithm, we first run a MH algorithm for much longer than we normally would ($8\times 10^{10}$ samples), and consider the histogram produced to be the truth.
+We then perform our usual simulations with the PAIS and MH algorithms to compare the rate at which the histograms converge to the `true' histogram over the same grid.
+% modifications to algorithm + spikes
+In this problem we modify our PAIS algorithm to use a mixture of Gamma distributions in the proposal distribution instead of a Gaussian mixture.
+Similarly we use a Gamma proposal in the MH algorithm.
+This ensures that we only propose positive parameters.
+Having a heavier right tail in the proposal distribution also means that the posterior is absolutely continuous with respect to the proposal.
+When this is not the case, the weight function can tend to zero or infinity when $k_i \rightarrow \infty$.
+Sampling under these conditions can result in very spiky approximations of the tails of the posterior distribution resulting in slow convergence.
+%*** AMR and increase in sample size
+
+%***numerics.
+\subsubsection{Convergence of PAIS-RW and PAIS-LA vs MH with Gamma proposals}\label{sec:chem_conv}
+
+In the numerics which follow we have used the AMR algorithm to resample with an ensemble size of $M=2500$.
+The method otherwise remains the same as in previous sections.
+We perform test runs to find the optimal scaling parameters for both RW proposals and MALA-type proposals.
+We then calculate the convergence rates of the algorithms by producing 50 million samples from the posterior with each algorithm, and repeat the simulation 32 times.
+
+\begin{figure}[htb]
+\centering
+\subfigure[Optimal values of the scaling parameter.]{\includegraphics[width=0.45\textwidth]{"figures/C1_ESS"}}
+\subfigure[Relative L2 error against number of iterations.]{\includegraphics[width=0.45\textwidth]{"figures/C1_L2"}}
+\caption{Convergence of the PAIS-RW and PAIS-LA algorithms for the chemical system problem described in Section~\ref{sec:chem}.
+Implementation described in Sections~\ref{sec:chem_implementation} and \ref{sec:chem_conv}.}
+\label{fig:C1_ESS}
+\end{figure}
+
+\begin{table}[!htb]
+      \centering
+        \begin{tabular}{|l|r|r|r|}
+	\hline
+	Algorithm	& MH-Gam & PAIS-RW & PAIS-LA \\ \hline
+	$\delta^*$	& 7.2e-0     & 8.9e-1    & 1.0e-0 \\
+	\hline
+	\end{tabular}
+	\vspace{1mm}
+	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithms are optimised using the effective sample size.}
+	\label{table:C1_opt_delta}
+\end{table}
+
+In Figure~\ref{fig:C1_ESS} (a) we see that the two PAIS algorithms have similar optimal values of the scaling parameter, also shown in Table~\ref{table:C1_opt_delta}, however the MALA-type proposal achieves a higher effective sample size with the same ensemble size.
+Convergence after the first 50 million samples is shown in Figure~\ref{fig:C1_ESS} (b) and demonstrates that the PAIS algorithm converges faster than the MH algorithm.
+The slightly unstable convergence in the PAIS-RW algorithm is the effect of spikes appearing in the tails of the posterior distribution.
+This is also what has caused the slightly lower effective sample size.
+
 \section{Discussion and Conclusions}\label{Sec:Conc}
@@
 Algorithm (MALA) and preconditioned Crank-Nicolson (pCN) proposals has
-yielded similar results\cite{Paul}.%  Importantly, we have demonstrated
-% that the convergence rate improves superlinearly with respect to the
-% number of processors used.
+yielded similar results\cite{Paul}.
 

### 2016-05-31 PAIS.tex
converted deltas to betas throughout

@@
 for near Gaussian targets we have 23.4\% for RWMH and 57.4\% for MALA\cite{roberts2001optimal}.
-Unlike Metropolis-Hastings algorithms, the PAIS algorithm does not accept or reject proposed values, so we need another method of measuring the optimality of $\delta$.
-Section~\ref{sec:statistics} gives some possible methods for tuning $\delta$.
+Unlike Metropolis-Hastings algorithms, the PAIS algorithm does not accept or reject proposed values, so we need another method of measuring the optimality of $\beta$.
+Section~\ref{sec:statistics} gives some possible methods for tuning $\beta$.
 
@@
 These statistics cannot be used in practise to find optimal values of
-$\delta$ since they require knowledge of the analytic solution, and
+$\beta$ since they require knowledge of the analytic solution, and
 the algorithm must be run for a long time to build up a
@@
 This method is most efficient when the target is proportional to the proposal distribution, in this case the weights are constant and so the variance of the weights, $var(w(y))$, is zero.
-Hence, we would like to choose the value of $\delta$ which minimises the variance of the weights,
-\[
-	\delta^*_{\text{var}} = \arg\min_{\delta} \text{var}(w(y)).
-\]
-The mean of the estimator $\text{var}(w(y))$ is smooth enough that it can be used to tune $\delta$ during the burn-in phase of MCMC algorithms.
+Hence, we would like to choose the value of $\beta$ which minimises the variance of the weights,
+\[
+	\beta^*_{\text{var}} = \arg\min_{\beta} \text{var}(w(y)).
+\]
+The mean of the estimator $\text{var}(w(y))$ is smooth enough that it can be used to tune $\beta$ during the burn-in phase of MCMC algorithms.
 However the variance of the estimator can be large, especially far away from the optimal value, so it can take a large number of iterations to calculate descent directions.
@@
 
-The statistic $\neff$ is better scaled than the variance of the weights, and is easier to interpret, so is preferable as a means of tuning $\delta$.
+The statistic $\neff$ is better scaled than the variance of the weights, and is easier to interpret, so is preferable as a means of tuning the scaling parameter.
 In the following numerics, we use the effective sample size calculated at each iteration and averaged across the entire run to tune the scaling parameters.
@@
 Figure~\ref{fig:neff-M} shows results for the Gaussian posterior discussed in Section~\ref{sec:problem 1} when using the PAIS-RW algorithm.
-Figure~\ref{fig:neff-M} (a) shows that as the ensemble size increases, the value of $\delta$ which gives the optimal effective sample size ratio decreases.
+Figure~\ref{fig:neff-M} (a) shows that as the ensemble size increases, the scaling parameter which gives the optimal effective sample size ratio decreases.
 This is to be expected since we are trying to fit more kernels into the posterior distribution, so the variance of each kernel needs to be decreased.
@@
 Some more sophisticated examples are described in \cite{roberts2009examples} and \cite{Ji2013adaptive}.
-Algorithm~\ref{tab:adapt} in Appedix~\ref{Sec:App}, is not presented as an optimal strategy but as a demonstration that the PAIS algorithm can be adaptively tuned effectively.
+Algorithm~\ref{tab:adapt} in Appendix~\ref{Sec:App}, is not presented as an optimal strategy but as a demonstration that the PAIS algorithm can be adaptively tuned effectively.
 
@@
 \end{equation}
-where $X_{i-1}$ is our current state, $Y$ is our proposed state, and $\beta = \sqrt{8\delta}/(2+\delta)$. In the numerics section we will refer to $\delta$ as the scaling parameter.
-
- In Algorithm~\ref{tab:adapt}, we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges roughly to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm.
-We must choose some sequence of iterations, $\{n_k\}_{k=1}$, at which to update $\delta$.
+where $X_{i-1}$ is our current state and $Y$ is our proposed state. In the numerics section we will refer to $\beta$ as the scaling parameter.
+
+ In Algorithm~\ref{tab:adapt}, we calculate a sequence $\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal scaling parameter $\beta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm.
+We must choose some sequence of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$.
 These $n_k$ can be constructed by a sequence in which the terms grow exponentially further apart.
-The algorithm is given for the PAIS method but a similar method is used to adaptively calculate the optimal scaling parameter, $\delta^*$, for the MH algorithm.
+The algorithm is given for the PAIS method but a similar method is used to adaptively calculate the optimal scaling parameter, $\beta^*$, for the MH algorithm.
 
 When implemented, there are a number of parameters which must be
-chosen to ensure efficient tuning of $\delta$. Firstly, the initial
-value of $\delta$ should be chosen so that the chains quickly explore the parameter space. However, if $\delta$ is chosen to be too
+chosen to ensure efficient tuning of $\beta$. Firstly, the initial
+value of $\beta$ should be chosen so that the chains quickly explore the parameter space. However, if $\beta$ is chosen to be too
 large, then the method may become unstable.
@@
 Since we are comparing against naively parallelised MH algorithms,
-we also need to decide which statistic $T(\delta)$ to optimise for
+we also need to decide which statistic $T(\beta)$ to optimise for
 these approaches.
@@
 \[
-	T_{\text{MH}}(\delta) = \left| \frac{N_{\text{acc}}(\delta)}{N_{\text{total}}} - \hat{\alpha} \right|,
-\]
-where $N_{\text{acc}}(\delta)$ is the number of accepted moves and
+	T_{\text{MH}}(\beta) = \left| \frac{N_{\text{acc}}(\beta)}{N_{\text{total}}} - \hat{\alpha} \right|,
+\]
+where $N_{\text{acc}}(\beta)$ is the number of accepted moves and
 $N_{\text{total}}$ is the total number of samples produced. For the
 PAIS algorithm, we are maximising the effective sample size as
-discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\delta) =
-\neff(\delta)$.
+discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\beta) =
+\neff(\beta)$.
 
@@
 In each of the following simulations, we perform three tasks.
-First we calculate the optimal value of $\delta$ by optimising the statistics described in Section~\ref{sec:statistics}.
+First we calculate the optimal value of $\beta$ by optimising the statistics described in Section~\ref{sec:statistics}.
 We then run the algorithms with optimal parameters to calculate and compare the convergence rates.
@@
 {\bf (1) Finding the optimal parameters}: To find the optimal
-parameters we choose 32 values of $\delta$ evenly spaced on a log scale
+parameters we choose 32 values of $\beta$ evenly spaced on a log scale
 in the interval $[10^{-5}, 2]$.
@@
 algorithms, and again use the relative L2 error to compare
-efficiency. The adaptive algorithms are initialised with $\delta^{(1)}$ = 2.
-
-\subsubsection{Optimal values of $\delta$}\label{sec:Optimal values P1}
+efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 2.
+
+\subsubsection{Optimal values of $\beta$}\label{sec:Optimal values P1}
 
@@
 \subfigure[PAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
-\caption{Finding optimal values of $\delta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}.}
-\label{fig:P1 opt delta}
+\caption{Finding optimal values of $\beta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}.}
+\label{fig:P1 opt beta}
 \end{figure}
@@
 	\hline
-	Statistic											& pCNL \\ \hline
-	$\delta_{\text{L2}}^*$								& 2.3e-4 \\
-	$\delta_{\%}^*$									& 1.2e-1 \\
-	Acceptance Rate ($\delta_{\text{L2}}^*$)				& 9.0e-1 \\
-	Acceptance Rate ($\delta_{\%}^*$)					& 5.0e-1 \\
+	Statistic											& RWMH \\ \hline
+	$\beta_{\text{L2}}^*$								& 2.1e-2 \\
+	$\beta_{\%}^*$									& 1.5e-1 \\
+	Acceptance Rate ($\beta_{\text{L2}}^*$)				& 9.0e-1 \\
+	Acceptance Rate ($\beta_{\%}^*$)					& 5.0e-1 \\
 	\hline
@@
 	\hline
-	Statistic							& PAIS-pCNL \\ \hline
-	$\delta_{\text{eff}}^*$				& 1.1e-3 \\
-	$\delta_{\text{var}(w(y))}^*$			& 1.7e-3 \\
-	$\delta_{\text{L2}}^*$				& 7.6e-4 \\
+	Statistic							& PAIS-RW \\ \hline
+	$\beta_{\text{eff}}^*$				& 4.7e-2 \\
+	$\beta_{\text{var}(w(y))}^*$		& 5.8e-2 \\
+	$\beta_{\text{L2}}^*$				& 3.9e-2 \\
 	\hline
@@
 	\vspace{1mm}
-	\caption{Optimal values of $\delta$ summarised from Figure~\ref{fig:P1 opt delta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
-	\label{table:prob1 opt delta}
+	\caption{Optimal values of $\beta$ summarised from Figure~\ref{fig:P1 opt beta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
+	\label{table:prob1 opt beta}
 \end{table}
 
-Figure~\ref{fig:P1 opt delta} (a) shows the two values of $\delta$ which are optimal according to the acceptance rate and relative L2 error criteria for the RWMH algorithm.
+Figure~\ref{fig:P1 opt beta} (a) shows the two values of $\beta$ which are optimal according to the acceptance rate and relative L2 error criteria for the RWMH algorithm.
 The smaller estimate comes from the relative L2 error, and the larger from the acceptance rate.
-The results in Figure~\ref{fig:P1 opt delta} are summarised in Table~\ref{table:prob1 opt delta}.
+The results in Figure~\ref{fig:P1 opt beta} are summarised in Table~\ref{table:prob1 opt beta}.
 Since in general we cannot calculate the relative L2 error, we must optimise the algorithm using the acceptance rate.
@@
 
-Figure~\ref{fig:P1 opt delta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
+Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
 The relative L2 error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative L2 error.
@@
 
-Both adaptive algorithms are run with initial values of $\delta_0=2.0$.
+Both adaptive algorithms are run with initial values of $\beta_0=2.0$.
 Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance.
@@
 Figure~\ref{fig:PAIS_saving} was produced using the method of finding
-optimal $\delta$ described in Section~\ref{sec:Implementation P1} (1),
+optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
 then running 32 repeats at each ensemble size. The convergence rates
@@
 
-\subsubsection{Calculating values of Optimal $\delta^*$}\label{sec:BM1_opt_delta}
+\subsubsection{Calculating values of Optimal $\beta^*$}\label{sec:BM1_opt_beta}
 
 Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for RWMH and we use the effective sample size to find the optimal values for PAIS-RW.
-Table~\ref{table:BM_opt_delta} gives the optimal values of $\delta$ for both problems.
+Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both problems.
 
@@
 	\hline
-	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$ \\ \hline
-	RWMH								& 1.3e-1					& - \\
-	PAIS-RW								& -						& 5.4e-3\\
+	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$ \\ \hline
+	RWMH								& 4.8e-1					& - \\
+	PAIS-RW								& -						& 1.0e-1\\
 	\hline
@@
 	\hline
-	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$	& $\delta^*_{\text{L2}}$ \\ \hline
-	RWMH								& 2.6e-2					& - 						& 9.1e-1\\
-	PAIS-RW								& -						& 1.3e-3 					& 8.1e-3\\
+	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$	& $\beta^*_{\text{L2}}$ \\ \hline
+	RWMH								& 2.3e-1					& - 						& 9.3e-1\\
+	PAIS-RW								& -						& 5.1e-2 					& 1.3e-1\\
 	\hline
@@
 	\vspace{1mm}
-	\caption{Optimal values of $\delta$ for $B_1$ (left) and $B_2$ (right).}
-	\label{table:BM_opt_delta}
+	\caption{Optimal values of $\beta$ for $B_1$ (left) and $B_2$ (right).}
+	\label{table:BM_opt_beta}
 \end{table}
@@
 It is relatively simple to find optimal scaling parameters for problem $B_1$.
-These values are given in Table~\ref{table:BM_opt_delta} (left).
+These values are given in Table~\ref{table:BM_opt_beta} (left).
 However problem $B_2$ is much harder; transitions between the modes are extremely unlikely for the standard RWMH algorithm.
@@
 important that the chains can transition between the modes, which
-means that $\delta$ must be large. However, this leads to a lower
+means that $\beta$ must be large. However, this leads to a lower
 acceptance rate, and so we sacrifice convergence locally. For this reason, the RWMH algorithm is very slow to
@@
 
-We can achieve these two regimes in RWMH by tuning $\delta$ using the acceptance rate for local convergence, and by L2 error for global convergence.
+We can achieve these two regimes in RWMH by tuning $\beta$ using the acceptance rate for local convergence, and by L2 error for global convergence.
 Similarly in PAIS-RW we can use the effective sample size for local convergence, and the L2 error for global convergence.
 
-From Table~\ref{table:BM_opt_delta} (right) we see that there is a large difference between the optimal value of $\delta$ for each regime in RWMH, and so will result in inefficient sampling.
-The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\delta^*$: a clear advantage to using this algorithm for this problem.
+From Table~\ref{table:BM_opt_beta} (right) we see that there is a large difference between the optimal value of $\beta$ for each regime in RWMH, and so will result in inefficient sampling.
+The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\beta^*$: a clear advantage to using this algorithm for this problem.
 
@@
 
-For $B_2$ the algorithms are run with the global optimal value of $\delta^*$, and with the local optimal value of $\delta^*$.
-Figure~\ref{fig:BM2_L2} (a) shows that the algorithms using the globally optimal $\delta^*$ convergence at the desired rate, whereas the algorithms optimised using the acceptance rate and effective sample size initially converge faster but at some point forget the location of one of the modes, causing the convergence to flatten out.
+For $B_2$ the algorithms are run with the global optimal value of $\beta^*$, and with the local optimal value of $\beta^*$.
+Figure~\ref{fig:BM2_L2} (a) shows that the algorithms using the globally optimal $\beta^*$ convergence at the desired rate, whereas the algorithms optimised using the acceptance rate and effective sample size initially converge faster but at some point forget the location of one of the modes, causing the convergence to flatten out.
 
 Using the PAIS algorithm we can minimise the impact of forgotten modes by constantly allowing the algorithm to search them out.
-Since we have parallel chains, we can run PAIS-RW with the majority of chains using the locally optimal value of $\delta$, and one or two chains with a larger scaling parameter.
+Since we have parallel chains, we can run PAIS-RW with the majority of chains using the locally optimal value of $\beta$, and one or two chains with a larger scaling parameter.
 These chains with larger scaling parameters act both as `scouts' for new modes, and act to aid in the overdispersal of the proposal distribution.
@@
 	Algorithm	& MH-Gam & PAIS-RW & PAIS-LA \\ \hline
-	$\delta^*$	& 7.2e-0     & 8.9e-1    & 1.0e-0 \\
+	$\beta^*$	& 7.2e-0     & 8.9e-1    & 1.0e-0 \\
 	\hline
@@
 	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithms are optimised using the effective sample size.}
-	\label{table:C1_opt_delta}
+	\label{table:C1_opt_beta}
 \end{table}
 
-In Figure~\ref{fig:C1_ESS} (a) we see that the two PAIS algorithms have similar optimal values of the scaling parameter, also shown in Table~\ref{table:C1_opt_delta}, however the MALA-type proposal achieves a higher effective sample size with the same ensemble size.
+In Figure~\ref{fig:C1_ESS} (a) we see that the two PAIS algorithms have similar optimal values of the scaling parameter, also shown in Table~\ref{table:C1_opt_beta}, however the MALA-type proposal achieves a higher effective sample size with the same ensemble size.
 Convergence after the first 50 million samples is shown in Figure~\ref{fig:C1_ESS} (b) and demonstrates that the PAIS algorithm converges faster than the MH algorithm.
@@
 \STATE $x_j^{(0)} \sim \mu_0$, for $j \in L\cup U$, where $L = \{j\}_{j=1}^{M/2},\ U = \{j\}_{j=M/2+1}^M$.
-\STATE Choose $\delta^{(1)} \in (0, 2]$. Set $\delta_{L,U}^{(1)} = (1\pm 0.01)\delta^{(1)}\wedge 2$.
+\STATE Choose $\beta^{(1)} \in (0, 2]$. Set $\beta_{L,U}^{(1)} = (1\pm 0.01)\beta^{(1)}\wedge 2$.
 \FOR{$i=1,2, \ldots, N$}
-\STATE $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\delta_L^{(i)})$ for $j \in L$, and $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\delta_U^{(i)})$ for $j \in U$.
+\STATE $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\beta_L^{(i)})$ for $j \in L$, and $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\beta_U^{(i)})$ for $j \in U$.
 \STATE Calculate 
@@
 \[
-	\nu(y;X) = \frac{1}{M}\sum_{j\in L} q(y;x_j,\delta_L)+\frac{1}{M}\sum_{j\in U} q(y;x_j,\delta_U).
+	\nu(y;X) = \frac{1}{M}\sum_{j\in L} q(y;x_j,\beta_L)+\frac{1}{M}\sum_{j\in U} q(y;x_j,\beta_U).
 \]
@@
 	\STATE $T_U = SM(\sum_{k=i-S}^S\sum_{j\in U} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in U} w_{kj}^2)$.
-	\STATE $\delta^{(i+1)} = \delta^{(i)} - \Delta t \displaystyle\frac{T_U - T_L}{\delta_U^{(i)}-\delta_L^{(i)}}$.
+	\STATE $\beta^{(i+1)} = \beta^{(i)} - \Delta t \displaystyle\frac{T_U - T_L}{\beta_U^{(i)}-\beta_L^{(i)}}$.
 \ELSE
-	\STATE $\delta^{(i+1)} = \delta^{(i)}$.
+	\STATE $\beta^{(i+1)} = \beta^{(i)}$.
 \ENDIF
@@
 	\ENDIF
-	\STATE $\delta_{L,U}^{(i)} = \delta^{(i)} \pm 2\sqrt{2}\delta^{(i)}/\sqrt{NM} \wedge 2$.
+	\STATE $\beta_{L,U}^{(i)} = \beta^{(i)} \pm 2\sqrt{2}\beta^{(i)}/\sqrt{NM} \wedge 2$.
 \ELSE
 		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
-		\STATE $\delta_L^{(i+1)} = \delta_U^{(i+1)} = \delta^{(i+1)}$.
+		\STATE $\beta_L^{(i+1)} = \beta_U^{(i+1)} = \beta^{(i+1)}$.
 \ENDIF

### 2016-05-31 PAIS.tex
added reference for 1d Gaussian 50% + added note about \beta_L2^*

@@
 In the examples which follow, we have optimised the naively parallelised RWMH algorithm using the optimal acceptance rate $\hat{\alpha} = 0.5$.
-This value differs from the theoretical asymptotic value of 0.234 which applies in higher dimensions, this higher acceptance rate is commonly used for one dimensional Gaussian posteriors.
+This value differs from the theoretical asymptotic value of 0.234 which applies in higher dimensions, this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}.
 To optimise the algorithm we minimise the statistic
@@
 	\vspace{1mm}
-	\caption{Optimal values of $\beta$ summarised from Figure~\ref{fig:P1 opt beta}. Statistics calculated as described in Section~\ref{sec:statistics}.}
+	\caption{Optimal values of $\beta$ summarised from Figure~\ref{fig:P1 opt beta}.
+Statistics calculated as described in Section~\ref{sec:statistics}.
+The values $\beta^*_{\text{L2}}$ and $\beta^*_{\%}$ are the optimal scaling parameters found by optimising the relative L2 errors and acceptance rate respectively.
+Similarly $\beta_{\text{eff}}^*$ and $\beta_{\text{var}(w(y))}^*$ optimise the effective sample size and variance of the weights statistics.}
 	\label{table:prob1 opt beta}

### 2016-06-01 PAIS.tex
Added AMR section
Added AMR section

@@
 \newcommand{\E}{{\mathbb E}}
+\renewcommand{\b}[1]{{\bf #1}}
+\DeclareMathOperator*{\argmin}{arg\,min}
+\DeclareMathOperator*{\argmax}{arg\,max}
 
@@
 this set of weighted samples. This is achieved by implementing the
-ETMC algorithm. Once we once again have a set of evenly weighted
+ETPF algorithm. Once we once again have a set of evenly weighted
 samples that we believe represents the target distribution well, we
@@
 \frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
-\STATE Resample using ETMC: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
+\STATE Resample using ETPF: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
 \ENDFOR 
@@
 value of $\beta$ should be chosen so that the chains quickly explore the parameter space. However, if $\beta$ is chosen to be too
-large, then the method may become unstable.
-We also choose two iteration numbers at which we change our adaptive strategy.
-At iteration $N_\text{join}$ we resample using the full ensemble instead of using two subsets of the full ensemble.
-When using two subsets we are training the scaling parameter using the wrong ensemble size, this allows faster convergence but leads to a sub-optimal scaling parameter.
-Resampling with the full ensemble towards the end of the tuning allows fast convergence to the right order of magnitude, followed by fine tuning to the optimal scaling parameter.
-Finally we choose a time, $N_\text{stop}$, at which we stop updating our parameter.
+large, then the method may become unstable. We also choose two
+iteration numbers at which we change our adaptive strategy. At
+iteration $N_\text{join}$ we resample using the full ensemble instead
+of using two subsets of the full ensemble. When using two subsets we
+are training the scaling parameter using the wrong ensemble size, this
+allows faster convergence but leads to a sub-optimal scaling
+parameter. Resampling with the full ensemble towards the end of the
+tuning allows fast convergence to the right order of magnitude,
+followed by fine tuning to the optimal scaling parameter. Finally we
+choose a time, $N_\text{stop}$, at which we stop updating our
+parameter.
+
+\section{Approximate Multinomial Resampling}
+Although the ETPF is optimal in terms of preserving statistics of the
+sample, it can also become quite costly as the number of ensemble members
+is increased. It is arguable that in the context of PAIS, we do not
+require this degree of accuracy, and that a faster more approximate
+method for resampling could be employed. One approach would be to use
+the bootstrap resampler, which simply takes the $N$ ensemble members'
+weights and constructs a multinomial distribution, from which $N$
+samples are drawn. This is essentially the cheapest resampling
+algorithm that one could construct. However it too has some
+drawbacks. The algorithm is random, and as such it is possible for all
+of the ensemble members in a particular region not to be sampled. This
+could be particularly problematic when attempting to sample from a
+multimodal distribution, where it might take a long time to find one
+of the modes again. The bootstrap filter also does not guarantee to
+preserve the mean of the weighted sample as the ETPF.
+
+Ideally, we would like to use a resampling algorithm which is not
+prohibitively costly for moderately or large sized ensembles, but
+which also preserves the mean of the samples, and also makes it much
+harder for the new samples to have missed a significant region in the
+density. This motivates the following algorithm, which we refer to as
+approximate multinomial resampling (AMR).
+
+Instead of sampling $N$ times from an $N$-dimensional multinomial
+distribution as is the case with the bootstrap algorithm, we sample
+once each from $N$ different multinomials. Suppose that we have $N$
+samples $y_n$ with weights $w_n$. The multinomial sampled from in the
+bootstrap filter has a vector of probabilities given by:
+\begin{equation*}
+\frac{1}{\sum w_n} [w_1,w_2,\ldots,w_N] = \bar{\bf w},
+\end{equation*}
+with associated states $y_n$.
+We wish to find $N$ vectors $\{{\bf p}_1,{\bf p}_2,\ldots,{\bf p}_N\}
+\subset \mathbb{R}^N_{\ge 0}$
+such that  $\frac{1}{N} \sum {\bf p}_i = \bar{\bf w}$. The AMR is then
+given by a sample from each of the multinomials defined by the vectors
+${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,N}]$ with associated states ${\bf x}_i$. Alternatively, as with the ETPF, a deterministic sample
+can be chosen by picking each sample to be equal to the mean value of
+each of these multinomial distributions, i.e. each new sample
+$\hat{x}_i$ is given by:
+\begin{equation}
+\hat{x}_i = \sum p_{i,j} x_j, \qquad i \in \{1,2,\ldots,N.
+\end{equation}
+
+The resulting sample has several properties which are advantageous in
+the context of being used with the PAIS algorithm. Firstly, we have
+effectively chopped up the multinomial distribution into used in the
+bootstrap filter into $N$ pieces, and we can guarantee that exactly
+one sample will be taken from each section. This leads to a much
+smaller chance of losing entire modes in the density. Secondly, if we do not make a random sample for
+each multinomial with probability vector $\bf p_i$ but instead take
+the mean of the multinomial to be the sample, this algorithm preserves
+the mean of the sample exactly. Lastly, as we will see shortly, this
+algorithm is significantly less computationally intensive than the
+ETPF.
+
+There are of course infinitely many different ways that one could use
+to split the original multinomial up into $N$ parts, some of which
+will be far from optimal. The method that we have arrived at is based
+on the idea that we want to group together states which are
+geographically close to each other, preferentially doing this 
+for the states with the biggest weights.
+
+\begin{table}
+\begin{mdframed}
+\begin{algorithmic}
+\STATE $\b{z} = N\bar{\b{w}}$
+\FOR{$i=1,2,\ldots,N$}
+\STATE $J = \argmax_j z_j$
+\STATE $p_{i,J} = \min\{1,z_J\}$
+\STATE $z_J = z_J - p_{i,J}$
+\WHILE{$\sum_j p_{i,j} <1$}
+\STATE $K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$
+\STATE $p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$
+\STATE $z_K = z_K - p_{i,K}$
+\ENDWHILE
+\STATE $x_i = \sum_k p_{i,k}y_k$.
+\ENDFOR
+\end{algorithmic}
+\end{mdframed}
+\caption{The approximate multinomial resampler (AMR) algorithm.\label{tab:AMR}}
+\end{table}
+
+Table \ref{tab:AMR} describes the basis of the algorithm with
+deterministic resampling, using the means of each of the
+sub-multinomials. This resampler was designed with the aims of being
+more efficient than the ETPF, and more accurate than straight multinomial
+resampling. Therefore we now present numerical examples which
+demonstrate this.
+
+\begin{figure}[htb]
+\centering
+\subfigure[Relative error in $\mathbb{E}(X)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX"}}
+\subfigure[Relative error in
+$\mathbb{E}(X^2)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX2"}}\\
+\subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX3"}}
+\subfigure[Cost of resamplers per iteration]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
+\caption{Finding optimal values of $\beta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}.}
+\label{fig:AMR}
+\end{figure}
+
+To test the accuracy and speed of the three resamplers (ETPF,
+bootstrap and AMR), we drew a sample of size $N$ from the proposal distribution
+$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
+target distribution of $\mathcal{N}(2,3)$. The statistics of the
+resampled outputes were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
+three moments of the samples changes with $N$ for the three different
+samplers. As expected, the AMR lies somewhere between the high
+accuracy of the ETPF and the less accurate bootstrap
+resampling. Note that only the error for the bootstrap multinomial
+sampler is presented for the first moment since both the ETPF and the
+AMR preserve the mean up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
+measured in seconds, scales with the ensemble size for the three
+different methods. These results demonstrate that the AMR behaves how we wish, and
+importantly ensures that exactly one sample of the output will lie in
+each region with weights up to $\frac{1}{N}$ of the total.
+
+We will use the AMR in the numerics in Section \ref{sec:chem_conv}
+where we have chosen to use a larger ensemble size. We do not claim
+that the AMR is the optimal choice within PAIS, but it does have
+favourable features, and demonstrates how different choices of
+resampler can affect the speed and accuracy of the PAIS algorithm.
 

### 2016-06-01 PAIS.tex
adjusted the notation in the AMR section and linked it to the numerics

@@
 
-\section{Approximate Multinomial Resampling}
+\section{Approximate Multinomial Resampling}\label{sec:AMR}
 Although the ETPF is optimal in terms of preserving statistics of the
@@
 method for resampling could be employed. One approach would be to use
-the bootstrap resampler, which simply takes the $N$ ensemble members'
-weights and constructs a multinomial distribution, from which $N$
+the bootstrap resampler, which simply takes the $M$ ensemble members'
+weights and constructs a multinomial distribution, from which $M$
 samples are drawn. This is essentially the cheapest resampling
@@
 multimodal distribution, where it might take a long time to find one
-of the modes again. The bootstrap filter also does not guarantee to
-preserve the mean of the weighted sample as the ETPF.
+of the modes again. The bootstrap filter is also not guaranteed to
+preserve the mean of the weighted sample, unlike the ETPF.
 
 Ideally, we would like to use a resampling algorithm which is not
-prohibitively costly for moderately or large sized ensembles, but
-which also preserves the mean of the samples, and also makes it much
-harder for the new samples to have missed a significant region in the
+prohibitively costly for moderately or large sized ensembles,
+which preserves the mean of the samples, and which makes it much
+harder for the new samples to forget a significant region in the
 density. This motivates the following algorithm, which we refer to as
@@
 
-Instead of sampling $N$ times from an $N$-dimensional multinomial
+Instead of sampling $M$ times from an $M$-dimensional multinomial
 distribution as is the case with the bootstrap algorithm, we sample
-once each from $N$ different multinomials. Suppose that we have $N$
+once each from $M$ different multinomials. Suppose that we have $M$
 samples $y_n$ with weights $w_n$. The multinomial sampled from in the
@@
 \begin{equation*}
-\frac{1}{\sum w_n} [w_1,w_2,\ldots,w_N] = \bar{\bf w},
+\frac{1}{\sum w_n} [w_1,w_2,\ldots,w_M] = \bar{\bf w},
 \end{equation*}
 with associated states $y_n$.
-We wish to find $N$ vectors $\{{\bf p}_1,{\bf p}_2,\ldots,{\bf p}_N\}
-\subset \mathbb{R}^N_{\ge 0}$
-such that  $\frac{1}{N} \sum {\bf p}_i = \bar{\bf w}$. The AMR is then
+We wish to find $M$ vectors $\{{\bf p}_1,{\bf p}_2,\ldots,{\bf p}_M\}
+\subset \mathbb{R}^M_{\ge 0}$
+such that  $\frac{1}{M} \sum {\bf p}_i = \bar{\bf w}$. The AMR is then
 given by a sample from each of the multinomials defined by the vectors
-${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,N}]$ with associated states ${\bf x}_i$. Alternatively, as with the ETPF, a deterministic sample
+${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,M}]$ with associated states ${\bf x}_i$. Alternatively, as with the ETPF, a deterministic sample
 can be chosen by picking each sample to be equal to the mean value of
@@
 \begin{equation}
-\hat{x}_i = \sum p_{i,j} x_j, \qquad i \in \{1,2,\ldots,N.
+\hat{x}_i = \sum p_{i,j} x_j, \qquad i \in \{1,2,\ldots,M\}.
 \end{equation}
@@
 effectively chopped up the multinomial distribution into used in the
-bootstrap filter into $N$ pieces, and we can guarantee that exactly
+bootstrap filter into $M$ pieces, and we can guarantee that exactly
 one sample will be taken from each section. This leads to a much
@@
 There are of course infinitely many different ways that one could use
-to split the original multinomial up into $N$ parts, some of which
+to split the original multinomial up into $M$ parts, some of which
 will be far from optimal. The method that we have arrived at is based
@@
 \begin{algorithmic}
-\STATE $\b{z} = N\bar{\b{w}}$
-\FOR{$i=1,2,\ldots,N$}
+\STATE $\b{z} = M\bar{\b{w}}$
+\FOR{$i=1,2,\ldots,M$}
 \STATE $J = \argmax_j z_j$
@@
 To test the accuracy and speed of the three resamplers (ETPF,
-bootstrap and AMR), we drew a sample of size $N$ from the proposal distribution
+bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
 $\mathcal{N}(1,2)$. Importance weights were assigned, based on a
@@
 resampled outputes were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
-three moments of the samples changes with $N$ for the three different
+three moments of the samples changes with $M$ for the three different
 samplers. As expected, the AMR lies somewhere between the high
@@
 importantly ensures that exactly one sample of the output will lie in
-each region with weights up to $\frac{1}{N}$ of the total.
+each region with weights up to $\frac{1}{M}$ of the total.
 
@@
 We then perform our usual simulations with the PAIS and MH algorithms to compare the rate at which the histograms converge to the `true' histogram over the same grid.
+
 % modifications to algorithm + spikes
@@
 Sampling under these conditions can result in very spiky approximations of the tails of the posterior distribution resulting in slow convergence.
-%*** AMR and increase in sample size
-
-%***numerics.
+
+% AMR and increase in sample size
+We now significantly increase the ensemble size we are using from $M=50$ to $M=2500$.
+This allows us to build a better approximation of the posterior distribution for our proposals.
+Following the discussion in Section~\ref{sec:AMR}, it is clear that to keep the runtime of the resampler negligible compared with the calculation of the posterior we must switch from the ETPF resampler to the AMR algorithm.
+
+% numerics.
 \subsubsection{Convergence of PAIS-RW and PAIS-LA vs MH with Gamma proposals}\label{sec:chem_conv}

### 2016-06-01 PAIS.tex
added description of PAIS-LA and sorted out the names

@@
 In this problem we modify our PAIS algorithm to use a mixture of Gamma distributions in the proposal distribution instead of a Gaussian mixture.
+The PAIS-LA algorithm uses the Langevin proposal distribution from the standard MALA algorithm with a pertubation taken from a Gamma distribution with mean $X+\frac{1}{2}\beta^2\nabla \log\pi(\cdot)$, and  variance $\beta^2$.
 Similarly we use a Gamma proposal in the MH algorithm.
@@
 % numerics.
-\subsubsection{Convergence of PAIS-RW and PAIS-LA vs MH with Gamma proposals}\label{sec:chem_conv}
+\subsubsection{Convergence of PAIS-Gamma and PAIS-LA vs MH with Gamma proposals}\label{sec:chem_conv}
 
@@
 The method otherwise remains the same as in previous sections.
-We perform test runs to find the optimal scaling parameters for both RW proposals and MALA-type proposals.
+We perform test runs to find the optimal scaling parameters for both Gamma proposals and MALA-type proposals.
 We then calculate the convergence rates of the algorithms by producing 50 million samples from the posterior with each algorithm, and repeat the simulation 32 times.
@@
 \subfigure[Relative L2 error against number of iterations.]{\includegraphics[width=0.45\textwidth]{"figures/C1_L2"}}
-\caption{Convergence of the PAIS-RW and PAIS-LA algorithms for the chemical system problem described in Section~\ref{sec:chem}.
+\caption{Convergence of the PAIS-Gamma and PAIS-LA algorithms for the chemical system problem described in Section~\ref{sec:chem}.
 Implementation described in Sections~\ref{sec:chem_implementation} and \ref{sec:chem_conv}.}
@@
 	\hline
-	Algorithm	& MH-Gam & PAIS-RW & PAIS-LA \\ \hline
-	$\beta^*$	& 7.2e-0     & 8.9e-1    & 1.0e-0 \\
+	Algorithm	& MH-Gamma & PAIS-Gamma & PAIS-LA \\ \hline
+	$\beta^*$	& 2.7e-0     & 9.4e-1    & 1.0e-0 \\
 	\hline
@@
 Convergence after the first 50 million samples is shown in Figure~\ref{fig:C1_ESS} (b) and demonstrates that the PAIS algorithm converges faster than the MH algorithm.
-The slightly unstable convergence in the PAIS-RW algorithm is the effect of spikes appearing in the tails of the posterior distribution.
+The slightly unstable convergence in the PAIS-Gamma algorithm is the effect of spikes appearing in the tails of the posterior distribution.
 This is also what has caused the slightly lower effective sample size.

### 2016-06-02 PAIS.tex
Abstract/Intro/Prelims
Updates to the first two sections and the abstract to reflect changes
in the rest of the document.

@@
   better than linear improvements in convergence rates with respect to
-  the number of processors used.
+  the number of processors used. We also introduce a new resampling
+  strategy, approximate multinomial resampling (AMR), which while not
+  as accurate as other schemes is substantially costly for large
+  ensemble sizes, which can then be used in conjunction with PAIS for
+  complex problems.
 \end{abstract}
@@
 methods were first developed in the 70s\cite{hastings1970monte}, and
-with the development of and with the development of faster more
+with the development of faster, more
 powerful computers, have become ever more important in a whole range
@@
 In this paper, we present a framework for parallelisation of
-importance sampling, which can be built around any of the current
+importance sampling, which can be built around many of the current
 Metropolis-based methodologies in order to create an efficient target
@@
 
-In Section \ref{Sec:Prelim} we outline some preliminaries,
-including the general set up for Bayesian inverse problems, the
-preconditioned Crank-Nicolson Langevin (pCNL) algorithm and a
-brief review of the optimal transport resampler,  both of which we will be
-employing within the algorithm. We describe the Parallel Adaptive
-Importance Sampler (PAIS) in Section \ref{Sec:PAIS}, and in Section
-\ref{Sec:adapt} we describe how algorithmic parameters can be
-automatically tuned to provide optimal convergence. In
-Section \ref{Sec:Num}, we present some numerical experiments that
-demonstrate the savings available by employing this approach as
-opposed to a naive/trivial parallelisation of existing MCMC
-methods. In Section \ref{Sec:Conc}, we summarise our results and
-suggest some areas for future investigation.
+***SECTIONS SUMMARY NEEDS REWRITING***
 
@@
 
-\subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}\label{Sec:pCNL}
-In recent years, work has been carried out to frame MCMC proposal
-distributions on function space\cite{cotter2013mcmc}. These new
-discretisations perform comparably with the original versions in low
-dimensions. If gradient information regarding the observation operator is
-available, then a range of MCMC methods are available which exploit this information to improve mixing rates. One
-example of such an algorithm is MALA. In \cite{cotter2013mcmc}, a function space version of this
-method was presented, the pCNL algorithm, and is described in full in Table
-\ref{tab:pCNL}. The proposal used in this method comes about through
-a Crank-Nicolson approximation of the Langevin SDE, whose invariant
-measure is the posterior measure $\mu_Y$.
-
-\begin{table}
-\begin{mdframed}
-\begin{algorithmic}
-\STATE $X = x_0$
-\FOR{$i=1,2,3,\ldots$}
-\STATE $Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
-2\delta\mathcal{C}\nabla \Phi(u)+
-\sqrt{8\delta} W\right] $, $W \sim \mu_0$
-\STATE $a(X_{i-1},Y) = \min \left \{ 1,  \exp(\Phi(X_{i-1}) - \Phi(Y) ) \right \}$.
-\STATE $u \sim U([0,1])$
-\IF{$u<a(X_{i-1},Y)$}
-\STATE $X_i = Y$
-\ELSE 
-\STATE $X_i = X_{i-1}$
-\ENDIF
-\ENDFOR
-\end{algorithmic}
-\end{mdframed}\caption{A pseudo-code representation of the preconditioned Crank-Nicolson Langevin
-   (pCNL) algorithm. $\delta \in (0,2]$ is a step size parameter.}
-\label{tab:pCNL}
-\end{table}
+% \subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}\label{Sec:pCNL}
+% In recent years, work has been carried out to frame MCMC proposal
+% distributions on function space\cite{cotter2013mcmc}. These new
+% discretisations perform comparably with the original versions in low
+% dimensions. If gradient information regarding the observation operator is
+% available, then a range of MCMC methods are available which exploit this information to improve mixing rates. One
+% example of such an algorithm is MALA. In \cite{cotter2013mcmc}, a function space version of this
+% method was presented, the pCNL algorithm, and is described in full in Table
+% \ref{tab:pCNL}. The proposal used in this method comes about through
+% a Crank-Nicolson approximation of the Langevin SDE, whose invariant
+% measure is the posterior measure $\mu_Y$.
+
+% \begin{table}
+% \begin{mdframed}
+% \begin{algorithmic}
+% \STATE $X = x_0$
+% \FOR{$i=1,2,3,\ldots$}
+% \STATE $Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
+% 2\delta\mathcal{C}\nabla \Phi(u)+
+% \sqrt{8\delta} W\right] $, $W \sim \mu_0$
+% \STATE $a(X_{i-1},Y) = \min \left \{ 1,  \exp(\Phi(X_{i-1}) - \Phi(Y) ) \right \}$.
+% \STATE $u \sim U([0,1])$
+% \IF{$u<a(X_{i-1},Y)$}
+% \STATE $X_i = Y$
+% \ELSE 
+% \STATE $X_i = X_{i-1}$
+% \ENDIF
+% \ENDFOR
+% \end{algorithmic}
+% \end{mdframed}\caption{A pseudo-code representation of the preconditioned Crank-Nicolson Langevin
+%    (pCNL) algorithm. $\delta \in (0,2]$ is a step size parameter.}
+% \label{tab:pCNL}
+% \end{table}
 
@@
 One advantage of the particle filter is that there are convergence
-results for this method as the number of particles is increased. The
-downside is that, the effective sample size decreases at each iteration. One way to tackle
+results for this method as the number of particles is increased. One
+downside is that the effective sample size decreases at each
+iteration, resulting in degeneration of the approximation of the
+posterior. One way to tackle
 this is to employ a resampling scheme. The aim of a successful
 resample is to take your unevenly weighted ensemble and return a new
-ensemble of particles with even weights which is highly correlated to the original samples.
+ensemble of particles with even weights which is highly correlated to
+the original samples. Another downside is that the required ensemble
+size increases quickly with dimension, making it difficult to use in
+high-dimensional problems.
 
@@
 
-A Monte Carlo algorithm can be implemented to resample from a weighted ensemble.
-We create a weighted sample, then solve the optimal transport problem which produces the coupling described above, we can draw a new sample from the evenly weighted distribution.
-Reich suggests using the mean of the evenly weighted distribution to produce a consistent estimator.
-
-Analysis of this method shows that as the ensemble size increases, the statistics of the evenly weighted sample approach those of the posterior distribution, at least well enough for a proposal distribution as described in Section~\ref{Sec:PAIS}.
-The histogram of the evenly weighted sample exhibits small oscillations in the tails of the posterior, and also struggles to deal with discontinuities.
+A Monte Carlo algorithm can be implemented to resample from a weighted
+ensemble. We create a weighted sample, then solve the optimal
+transport problem which produces the coupling described above, we can
+draw a new sample from the evenly weighted distribution. Reich
+suggests using the mean of the evenly weighted distribution to produce
+a consistent estimator. Analysis of this method shows that as the
+ensemble size increases, the statistics of the evenly weighted sample
+approach those of the posterior distribution.
 
@@
 All MCMC methods are trivially parallelisable. One can take a method
-and simply implement it simultaneously over a set of processes. All
+and simply implement it simultaneously over a set of processors. All
 of the states of all of the processes can be recorded, and in the time
@@
 
-However, we argue that this is a far from optimal
-scenario.
-First of all, unless we have a lot of information about the posterior, we will begin the algorithm a long way from statistical equilibrium.
-Some initial iterations then are not samples from the posterior, and must be thrown away.
-This process is known as the burn-in.
-In a trivially parallelised scenario, each process must perform this process independently.
+However, we argue that this is not an optimal
+scenario. First of all, unless we have a lot of information about the
+posterior, we will intialise the algorithm's initial state in the
+tails of the distribution. The samples that are initially made as the
+algorithm finds its way to the mode(s) of the distribution cannot be
+considered to samples from the target distribution, and must be thrown
+away. This process is known as the burn-in. In a trivially
+parallelised scenario, each process must perform this process
+independently, and therefore mass parallelisation makes no inroads to
+cutting this cost.
 
 Moreover, many MCMC algorithms suffer from poor mixing, especially in
-multimodal systems. The time for an MCMC trajectory to switch between modes can be
-large and given that a large number of switches are required before we have a
+multimodal systems. The amount of samples that it takes for an MCMC trajectory to switch between modes can be
+large, and given that a large number of switches are required before we have a
 good idea of the relative probability densities of these different
@@
 (PAIS). In the next section we will introduce the method in its most
-general form. We will then introduce the version that we have
-implemented, which utilises the resampler recently suggested by
-Reich\cite{reich2013nonparametric} described in Section
-\ref{sec:filters}, and the pCNL proposal distribution described in
-Section \ref{Sec:pCNL}.
+general form.
 

### 2016-06-02 PAIS.tex
Updated sections 3+4

@@
 this is not the case. The aim of the PAIS is to use the states of a
-set of parallel MCMC chains to construct a proposal distribution which
-will be as close as possible to the target density. Given enough
-processors, the states of all of these chains at one point
+ensemble ofl MCMC chains to construct a proposal distribution which
+will be as close as possible to the target density. Given a big enough
+ensemble, the states of all of these chains at one point
 in time may be reasonably representative sample of the target density.
@@
 The proposal distribution could be constructed in many different ways,
-but we choose to use a mixture distribution, made up of MCMC proposals
-(in this paper, specifically the pCNL proposal from Section \ref{Sec:pCNL}). Once the proposal is
+but we choose to use a mixture distribution, made up of a sum of MCMC
+proposal distributions for each of the members of the ensemble. Once the proposal is
 constructed, we can sample a new set of states from the proposal
@@
 need to create a set of evenly weighted samples which best represent
-this set of weighted samples. This is achieved by implementing the
-ETPF algorithm. Once we once again have a set of evenly weighted
-samples that we believe represents the target distribution well, we
-can iterate the process once again. The algorithm is given in more detail in Table \ref{tab:PAIS}.
+this set of weighted samples. This is achieved by implementing a
+resampling algorithm. Initially we will use the
+ETPF algorithm\cite{reich2013nonparametric}, although we will suggest an alternative strategy in
+Section \ref{sec:AMR}. The output of the resampling algorithm gives us
+a set of evenly weighted
+samples that we believe represents the target distribution well, and
+from this point we
+can iterate the process once again. The algorithm is summarised in Table \ref{tab:PAIS}.
 
 We wish to sample states $x \in X$ from a posterior
-probability distribution $\mu_Y$. Since we have $M$ processes, we
+probability distribution $\mu_D$, where $D$ represents our data which
+we wish to use to infer $x$. Since we have $M$ processes, we
 represent the current state of all of the Markov chains as a vector
@@
 $\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
-the pCNL proposal presented in Section \ref{Sec:pCNL}.
+the random walk Metropolis-Hastings proposal density $\nu(\cdot,x) ~
+\mathcal{N}(x,\beta^2)$, where $\beta^2\in \mathbb{R}$ defines the
+variance of the proposal.
 
@@
 \frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
-\STATE Resample using ETPF: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
+\STATE Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
 \ENDFOR 
@@
 importance samplers, the weighted samples
-$(\W^{(i)},\Y^{(i)})_{i=1}^N$, or their statistics, are stored.
-
-The key here is to choose a suitable transition kernel $\nu$ such that
-if $X^{(i)}$ is a decent representative sample of the posterior,
-the mixture density $\chi(\cdot;\X^{(i)})$ is a reasonable
+$(\W^{(i)},\Y^{(i)})_{i=1}^N$ are the samples from the posterior that
+we will analyse.
+
+The key is to choose a suitable transition kernel $\nu$ such that
+if $X^{(i)}$ is a good representative sample of the posterior,
+then the mixture density $\chi(\cdot;\X^{(i)})$ is a good
 approximation of the posterior distribution. If this is the case,
-the newly proposed states $\Y^{(i)}$ will also be a good (and relatively
-independent) sample of the posterior with low variance in the
+the newly proposed states $\Y^{(i)}$ will also be a good sample of the posterior with low variance in the
 weights $\W^{(i)}$.
@@
 In Section \ref{Sec:Num}, we will demonstrate how the algorithm
-performs, using pCNL algorithm. We do not claim that this choice is optimal, but
+performs, primarily using RWMH proposals. We do not claim that this choice is optimal, but
 is simply chosen as an example to show that sharing information across
@@
 the likelihood from a very large data set this could lead to a
-large saving of computational cost.
+large saving of computational cost. We have observed that
+using more complex (and hence more expensive) kernels $\nu$, does not
+significantly improve the speed of convergence of the algorithm.
+
+Care needs to be taken when choosing the proposal distribution to
+ensure that the proposals are absolutely continuous with respect to
+the posterior distribution. In Section \ref{sec:chem} we will consider
+an inverse problem with Gamma priors. Since the posterior distribution
+in this context is heavy-tailed, if we use a lighter tailed
+distribution for $\nu$, such as a Gaussian kernel, then importance
+weights in the tails will be unbounded, which will hamper convergence
+of the algorithm.
 
@@
 
-Efficient selection of scaling parameters in MCMC algorithms is critical to achieving optimal mixing rates and hence achieving fast convergence to the target density.
-One significant difficulty is finding an appropriate $\nu$ such that
-$\chi$ is a close approximation to the posterior density $\pi$.
-If the proposal distribution is widely over-dispersed, then the algorithm will propose states a long way from the current state resulting in a relatively low acceptance rate, impacting the quality of inferences on the posterior.
-Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the parameter space and so the mixing rate, and hence convergence rate, will be slow.
+Efficient selection of scaling parameters in MCMC algorithms is
+critical to achieving optimal mixing rates and hence achieving fast
+convergence to the target density. One aspect worthy of  consideration
+with the PAIS, is finding an appropriate $\nu$ such that
+$\chi$ is a close approximation to the posterior density $\pi$. If the
+proposal distribution is over-dispersed, then the algorithm will often
+propose states in the tails of the distribution, resulting in larger
+variance of the weights, and therefore slower converegence to the
+posterior distribution. Similarly, if the proposal distribution is
+under-dispersed, the proposals will be highly correlated with the
+previous states, and the algorithm will take a long time to fully explore the parameter spac.
 It is therefore necessary to find a proposal distribution which is slightly over-dispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.
 
-Proposal distributions which are slightly over-dispersed as described
-above, can be found by tuning the variance of the proposal
-distribution $\nu$ during the burn-in phase of the
+Most commonly used MCMC proposals have parametric dependence which
+allows the user to control their variance. For example, in the RWMH
+proposal $y = x + \beta \eta$, the parameter $\beta$ controls how
+correlated the proposal state $y$ is to the current state
+$x$. Therefore the proposal distributions can be tuned such that they are are slightly over-dispersed, as described
+above. This tuning can take place during the burn-in phase of the
 algorithm. Algorithms which use this method to find optimal proposal
-distributions are known as adaptive MCMC algorithms. Adaptive MH
-algorithms will converge to the stationary distribution $\pi(\cdot)$,
-irrespective of whether the adaptive parameter itself converges to an
-optimal value, under the following
-conditions\cite{roberts2007coupling,roberts2009examples}: 
-\begin{enumerate}
-\item The adaptive parameter exhibits diminishing adaptation, which says that the amount of movement in the adaptive parameter decreases with the length of the chain. 
-\item The smallest number of steps for which the kernel has sufficiently converged, from an initial state $x$, is finite, by bounded convergence.
-\end{enumerate}
-
-Algorithms which use mixture proposals, e.g. PAIS, must tune the variance of the individual kernels within the proposal mixture. Adaptively choosing the variance of the kernels starting with a
-large initial guess allows us to first explore the state space
-globally, searching for the locations of the modes. Reducing the proposal variances to an
+distributions are known as adaptive MCMC algorithms, and have been
+shown to be convergent provided that they satisfy certain conditions\cite{roberts2007coupling,roberts2009examples}.
+
+Algorithms which use mixture proposals, e.g. PAIS, must tune the
+variance of the individual kernels within the proposal mixture. This
+adaptivity during the burn-in has some added benefits over and above
+finding an optimal parameter regime for the algorithm. If the initial
+value of the proposal variances is chosen to be very large, then
+proposed moves will be made far and wide, expediting the early
+mode-finding stages of the algorithm. Adaptively reducing the proposal variances to an
 optimal value then allows us to explore each region efficiently. The
-fact that we have multiple chains allows us to assess quickly and
+fact that we have an ensemble of chains allows us to assess quickly and
 effectively what the optimal variance of the proposal distributions
-should be. This is discussed in Section~\ref{sec:adapt}.
+should be.
 
@@
 
-MCMC algorithms can be assessed by comparing their approximation of the posterior to the analytic distribution.
-To assess this, a distance metric must be chosen.
-Examples are the relative error between the sample moments and the true moments, or the relative L2 error between the true density, $\pi(x|D)$, and the constructed histogram.
-The relative error in the $m$-th moment is
+MCMC algorithms can be assessed by comparing their approximation of
+the posterior to the analytic distribution, in cases where the
+posterior distribution can be computed using alternative methods. To
+assess this, a distance metric on distributions must be chosen. Examples are the
+relative error between the sample moments and the posterior's moments, or the
+relative $L^2$ error between the true density, $\pi(x|D)$, and the
+constructed histogram. The relative error in the $m$-th moment is
+given by:
 \[
@@
 \]
-where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative L2 error between a continuous function to a piecewise constant function, $e$, is
+where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative $L^2$
+error between a continuous function to a piecewise constant function,
+$e$, can be given by considering the difference in mass between the
+normalised histogram of the samples and the posterior distribution
+over a set of disjoint sets or ``bins'':
 \begin{equation}\label{eqn:L2_error}
@@
 
-These statistics cannot be used in practise to find optimal values of
+These statistics cannot be used in general to find optimal values of
 $\beta$ since they require knowledge of the analytic solution, and
@@
 ability of other indicators to find the optimal proposal
-variances in a controlled setting.
-
-The following statistics can be used specifically for importance sampling algorithms.
+variances in a controlled setting. The following statistics can be used specifically for importance sampling algorithms.
 
@@
 
-Importance samplers assign a weight to each sample they produce based on a ratio of the posterior to the proposal at that point.
-This method is most efficient when the target is proportional to the proposal distribution, in this case the weights are constant and so the variance of the weights, $var(w(y))$, is zero.
+Importance samplers assign a weight to each sample they produce based
+on a ratio of the posterior to the proposal at that point. Importance
+samplers are most efficient when the target is proportional to the
+proposal distribution.
+In this case the weights are all equal, and so the variance of the weights, $var(w(y))$, is zero.
 Hence, we would like to choose the value of $\beta$ which minimises the variance of the weights,
 \[
-	\beta^*_{\text{var}} = \arg\min_{\beta} \text{var}(w(y)).
+	\beta^*_{\text{var}} = \argmin_{\beta} \text{var}(w(y)).
 \]
-The mean of the estimator $\text{var}(w(y))$ is smooth enough that it can be used to tune $\beta$ during the burn-in phase of MCMC algorithms.
-However the variance of the estimator can be large, especially far away from the optimal value, so it can take a large number of iterations to calculate descent directions.
+In our experience, the mean of the estimator
+$\text{var}(w(y))$ is a smooth enough function of $\beta$ that it can
+be used to tune the proposal variance during the burn-in phase of MCMC
+algorithms. However the variance of the estimator of the variance can be large, especially far away from the optimal value, so it can take a large number of iterations to calculate descent directions.
 
@@
 
-The statistic $\neff$ is better scaled than the variance of the weights, and is easier to interpret, so is preferable as a means of tuning the scaling parameter.
-In the following numerics, we use the effective sample size calculated at each iteration and averaged across the entire run to tune the scaling parameters.
-We do this because this is the method we use to tune the scaling parameters on the fly, comparison to the optimal scaling parameter found using the global optimum of the variance of the weights is also included for comparison.
-
-The effective sample size also has another useful property; if we imagine the algorithm in the burn-in phase, for example, we have $M$ processes in the tail of a Gaussian curve searching for the area of high density.
-If the processes are evenly spaced, then the particle closest to the mean will have an exponentially higher weight assigned to it.
-The effective sample size in this scenario will be close to 1.
-As the algorithm burns in, the processes find the flatter area near the mean so the number of samples contributing information to the posterior will increase.
+The statistic $\neff$ is easier to deal with than the variance of the
+weights, as it varies between $1$ and $M$ as opposed to the variance
+which can vary over many orders of magnitude.
+Therefore it is preferable as a means of tuning the scaling
+parameter. In the following numerics, we use the effective sample size
+calculated at each iteration and averaged across the entire run to
+tune the scaling parameters. We do this because when we tune this
+parameter on the fly, we use the single-iteration average to estimate
+optimality. The optimal scaling parameter found using the global
+optimum of the variance of the weights is also included for
+comparison. Note that for the majority of iterations the value of
+$\neff$ is an overestimate of this statistic over a larger number of
+samples. Rare events which result in a large importance weight bring
+this statistic down, and care must be made not to overfit the proposal
+variance to $\neff$ on an iteration by iteration basis. This can be
+accounted for by increasing the variance slightly once the adaptive
+algorithm has arrived on a value using single iteration values of $\neff$.
+
+The effective sample size also has another useful property; if we
+imagine the algorithm in the burn-in phase, for example, we have $M$
+processes in the tail of a Gaussian curve searching for the area of
+high density. If the processes are evenly spaced, then the particle
+closest to the mean will have an exponentially higher weight assigned
+to it. The effective sample size in this scenario will be close to
+1. As the algorithm burns in, the ensemble populates the regions where
+the majority of the probability density lies, and the proposal
+distributions better represent the posterior distribution. This leads
+to smaller variance in the weights, and a bigger effective sample size.
 By this argument we can see that rising $\neff$ signals the end of the burn-in period.
-
-\subsubsection{Behaviour of the effective sample size for varying ensemble size}
 
@@
 how to tune the proposal variance, we need to look at how it behaves
-in different situations.
-Here we look at how the statistic behaves as we vary the ensemble size, $M$.
-Figure~\ref{fig:neff-M} shows results for the Gaussian posterior discussed in Section~\ref{sec:problem 1} when using the PAIS-RW algorithm.
-Figure~\ref{fig:neff-M} (a) shows that as the ensemble size increases, the scaling parameter which gives the optimal effective sample size ratio decreases.
-This is to be expected since we are trying to fit more kernels into the posterior distribution, so the variance of each kernel needs to be decreased.
-
-Figure~\ref{fig:neff-M} (b) shows that as the ensemble size increases, the efficiency of the sampler also increases.
-
-\subsection{The Adaptive Algorithm}\label{sec:adapt}
+in different situations. Here we look at how the statistic behaves as
+we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results
+for the Gaussian posterior discussed in Section~\ref{sec:problem 1}
+when using the PAIS algorithm with RWMH proposals, each with variance
+$\beta^2 \in \mathbb{R}_{>0}$. Figure~\ref{fig:neff-M} (a) shows that
+as the ensemble size increases, the scaling parameter which gives the
+optimal effective sample size ratio decreases. This is to be expected
+since if we are trying to approximate the posterior with a mixture
+distribution of a small number of Gaussians, the optimal variance will
+naturally be larger so that the whole of the significant regions are
+covered by the proposal distribution. As the number increases, the
+optimal variances decrease so that finer details in the posterior can
+be better represented in the proposal. Figure~\ref{fig:neff-M} (b) shows that as the ensemble size increases, the efficiency of the sampler also increases.
+
+\subsection{Adaptive PAIS}\label{sec:adapt}
 
@@
 optimal value, and we may need an
-uninformative prior for the scaling parameter.
-Alternatively, the parameter may be randomly sampled from some interval at various points during the chain.
-This results in some iterations which explore the state space, and some which make local moves.
-Algorithms of this type do not converges to an optimal value of the scaling parameter.
-
-We choose to use a divide and conquer scheme which optimises the effective sample size (or any other diagnostic).
-Some more sophisticated examples are described in \cite{roberts2009examples} and \cite{Ji2013adaptive}.
-Algorithm~\ref{tab:adapt} in Appendix~\ref{Sec:App}, is not presented as an optimal strategy but as a demonstration that the PAIS algorithm can be adaptively tuned effectively.
+uninformative prior for the scaling parameter. Alternatively, the
+parameter may be randomly sampled at various points during the
+evolution of the chain. This results in some iterations which make
+larger global moves in the state space between modes in the target
+distribution, and some which make local moves. Algorithms of this type
+do not converge to an optimal value of the scaling parameter. We choose to use a divide and conquer scheme which optimises the
+effective sample size (or any other diagnostic). Some more
+sophisticated examples are described in \cite{roberts2009examples} and
+\cite{Ji2013adaptive}. 
+% Algorithm~\ref{tab:adapt} in Appendix~\ref{Sec:App}, is not presented as an optimal strategy but as a demonstration that the PAIS algorithm can be adaptively tuned effectively.
 
 From here on in, we will use the random walk proposal,
-\begin{equation}Y = X_{i-1}+\beta W_{i-1}, \qquad W_{i-1} \sim \mathcal{N}(0,1),
+\begin{equation}y = x_{i-1}+\beta \omega_{i-1}, \qquad \omega_{i-1}
+  \sim \mathcal{N}(0,1) \qquad {\rm i.i.d.}
 \end{equation}
-where $X_{i-1}$ is our current state and $Y$ is our proposed state. In the numerics section we will refer to $\beta$ as the scaling parameter.
-
- In Algorithm~\ref{tab:adapt}, we calculate a sequence $\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal scaling parameter $\beta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm.
-We must choose some sequence of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$.
-These $n_k$ can be constructed by a sequence in which the terms grow exponentially further apart.
-The algorithm is given for the PAIS method but a similar method is used to adaptively calculate the optimal scaling parameter, $\beta^*$, for the MH algorithm.
-
-When implemented, there are a number of parameters which must be
-chosen to ensure efficient tuning of $\beta$. Firstly, the initial
-value of $\beta$ should be chosen so that the chains quickly explore the parameter space. However, if $\beta$ is chosen to be too
-large, then the method may become unstable. We also choose two
-iteration numbers at which we change our adaptive strategy. At
-iteration $N_\text{join}$ we resample using the full ensemble instead
-of using two subsets of the full ensemble. When using two subsets we
-are training the scaling parameter using the wrong ensemble size, this
-allows faster convergence but leads to a sub-optimal scaling
-parameter. Resampling with the full ensemble towards the end of the
-tuning allows fast convergence to the right order of magnitude,
-followed by fine tuning to the optimal scaling parameter. Finally we
-choose a time, $N_\text{stop}$, at which we stop updating our
-parameter.
+where $x_{i-1}$ is our current state and $y$ is our proposed state. In the numerics section we will refer to $\beta$ as the scaling parameter.
+
+ Using an adaptive strategy, we calculate a sequence
+ $\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal
+ scaling parameter $\beta^*$, resulting in the optimal transition
+ density $\chi$ for our MCMC algorithm. We must choose some sequence
+ of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$, and due
+ to the constraints on adaptive MCMC
+ algorithms\cite{roberts2007coupling,roberts2009example}, these $n_k$
+ must grow exponentially further apart. This same adaptive approach
+ can also be applied to the trivially parallelised MCMC algorithms to adaptively calculate their optimal scaling parameter $\beta^*$.
+
+% When implemented, there are a number of parameters which must be
+% chosen to ensure efficient tuning of $\beta$. Firstly, the initial
+% value of $\beta$ should be chosen so that the chains quickly explore the parameter space. However, if $\beta$ is chosen to be too
+% large, then the method may become unstable. We also choose two
+% iteration numbers at which we change our adaptive strategy. At
+% iteration $N_\text{join}$ we resample using the full ensemble instead
+% of using two subsets of the full ensemble. When using two subsets we
+% are training the scaling parameter using the wrong ensemble size, this
+% allows faster convergence but leads to a sub-optimal scaling
+% parameter. Resampling with the full ensemble towards the end of the
+% tuning allows fast convergence to the right order of magnitude,
+% followed by fine tuning to the optimal scaling parameter. Finally we
+% choose a time, $N_\text{stop}$, at which we stop updating our
+% parameter.
 
@@
 In this example we compare the naively parallelised RWMH algorithm with its
-PAIS variant, the PAIS-RW algorithm. We perform this comparison using the relative L2 error defined in Equation~\ref{eqn:L2_error}, as well as the relative error in the first moment.
+PAIS variant, the PAIS-RW algorithm. We perform this comparison using the relative $L^2$ error defined in Equation~\ref{eqn:L2_error}, as well as the relative error in the first moment.
 
@@
 \begin{equation}\label{eqn:Gaussian posterior}
-	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
+	\text{law}(\mu_D) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
 \end{equation}
@@
 divergence, which gives us a measure of how different the prior and
-posterior are, is $D_{KL}(\mu_Y || \mu_0) = 4.67$ for this problem.
+posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this problem.
 
@@
 The simulations are repeated 32 times using the optimal parameters found in (1).
-The performance of the algorithms is judged by the convergence of the relative L2 error statistic in Equation~\ref{eqn:L2_error}.
+The performance of the algorithms is judged by the convergence of the relative $L^2$ error statistic in Equation~\ref{eqn:L2_error}.
 
@@
 adaptive algorithms under the same conditions as the nonadaptive
-algorithms, and again use the relative L2 error to compare
+algorithms, and again use the relative $L^2$ error to compare
 efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 2.
@@
 Statistics calculated as described in Section~\ref{sec:statistics}.
-The values $\beta^*_{\text{L2}}$ and $\beta^*_{\%}$ are the optimal scaling parameters found by optimising the relative L2 errors and acceptance rate respectively.
+The values $\beta^*_{\text{L2}}$ and $\beta^*_{\%}$ are the optimal scaling parameters found by optimising the relative $L^2$ errors and acceptance rate respectively.
 Similarly $\beta_{\text{eff}}^*$ and $\beta_{\text{var}(w(y))}^*$ optimise the effective sample size and variance of the weights statistics.}
@@
 
-Figure~\ref{fig:P1 opt beta} (a) shows the two values of $\beta$ which are optimal according to the acceptance rate and relative L2 error criteria for the RWMH algorithm.
-The smaller estimate comes from the relative L2 error, and the larger from the acceptance rate.
+Figure~\ref{fig:P1 opt beta} (a) shows the two values of $\beta$ which are optimal according to the acceptance rate and relative $L^2$ error criteria for the RWMH algorithm.
+The smaller estimate comes from the relative $L^2$ error, and the larger from the acceptance rate.
 The results in Figure~\ref{fig:P1 opt beta} are summarised in Table~\ref{table:prob1 opt beta}.
-Since in general we cannot calculate the relative L2 error, we must optimise the algorithm using the acceptance rate.
-From the relative L2 error curve we can see that the minimum is very wide and despite the optimal values being very different there is not a large difference in the convergence rate.
+Since in general we cannot calculate the relative $L^2$ error, we must optimise the algorithm using the acceptance rate.
+From the relative $L^2$ error curve we can see that the minimum is very wide and despite the optimal values being very different there is not a large difference in the convergence rate.
 
 Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
-The relative L2 error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative L2 error.
+The relative $L^2$ error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative $L^2$ error.
 Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter.
@@
 \centering
-\subfigure[Relative L2 error.]{\includegraphics[width=0.45\textwidth]{"figures/G2_l2"}}
+\subfigure[Relative $L^2$ error.]{\includegraphics[width=0.45\textwidth]{"figures/G2_l2"}}
 \subfigure[Relative error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/G2_moments"}}
@@
 
-Figure~\ref{fig:MH1 L2} shows that the PAIS-RW algorithm converges to the posterior distribution significantly faster than the RWMH algorithm, in both L2 error and relative error in the moments.
+Figure~\ref{fig:MH1 L2} shows that the PAIS-RW algorithm converges to the posterior distribution significantly faster than the RWMH algorithm, in both $L^2$ error and relative error in the moments.
 A description of the speed up attained by this algorithm is given in Section~\ref{sec:calc_saving}.
@@
 
-We can achieve these two regimes in RWMH by tuning $\beta$ using the acceptance rate for local convergence, and by L2 error for global convergence.
-Similarly in PAIS-RW we can use the effective sample size for local convergence, and the L2 error for global convergence.
+We can achieve these two regimes in RWMH by tuning $\beta$ using the acceptance rate for local convergence, and by $L^2$ error for global convergence.
+Similarly in PAIS-RW we can use the effective sample size for local convergence, and the $L^2$ error for global convergence.
 
@@
 \centering
-\subfigure[Relative L2 error.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_L2"}}
+\subfigure[Relative $L^2$ error.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_L2"}}
 \subfigure[Absolute error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_moments"}}
@@
 \centering
-\subfigure[Relative L2 error using the optimal scaling parameters.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
-\subfigure[Comparing of the relative L2 error of the algorithms optimised by L2 minimisation with locally optimised PAIS-RW with a scout chain.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
+\subfigure[Relative $L^2$ error using the optimal scaling parameters.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
+\subfigure[Comparing of the relative $L^2$ error of the algorithms optimised by $L^2$ minimisation with locally optimised PAIS-RW with a scout chain.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
 \caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_2$. Set up described in Section~\ref{sec:Implementation P1}.}
@@
 \includegraphics[width=0.7\textwidth]{"figures/BM2_AL2"}
-\caption{Convergence of the relative L2 error for problem $B_2$, comparing the globally optimised nonadaptive algorithms with the locally optimised adaptive algorithms.
+\caption{Convergence of the relative $L^2$ error for problem $B_2$, comparing the globally optimised nonadaptive algorithms with the locally optimised adaptive algorithms.
 The setup is as described in Section~\ref{sec:Implementation P1} (3).}
@@
 Comparing the convergence of the adaptive algorithms against the nonadaptive algorithms in Figure~\ref{fig:BM2_AL2} shows that the algorithms behave as expected.
-The adaptive RWMH algorithm tuned using the acceptance rate converges at the same rate as the L2 optimised algorithm until the scaling parameter gets small where the algorithm forgets one of the modes.
+The adaptive RWMH algorithm tuned using the acceptance rate converges at the same rate as the $L^2$ optimised algorithm until the scaling parameter gets small where the algorithm forgets one of the modes.
 The adaptive PAIS-RW with scout chains tuned to the effective sample size converges at about the same rate as the locally optimised nonadaptive algorithm also with scouts.
@@
 \caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2.5}$ as a percentage of MH iterations.
-The relative L2 error graphs are from the Gaussian problem in Section~\ref{sec:problem 1}.}
+The relative $L^2$ error graphs are from the Gaussian problem in Section~\ref{sec:problem 1}.}
 \label{fig:calc_saving}
@@
 \subfigure[Optimal values of the scaling parameter.]{\includegraphics[width=0.45\textwidth]{"figures/C1_ESS"}}
-\subfigure[Relative L2 error against number of iterations.]{\includegraphics[width=0.45\textwidth]{"figures/C1_L2"}}
+\subfigure[Relative $L^2$ error against number of iterations.]{\includegraphics[width=0.45\textwidth]{"figures/C1_L2"}}
 \caption{Convergence of the PAIS-Gamma and PAIS-LA algorithms for the chemical system problem described in Section~\ref{sec:chem}.
@@
 
-\begin{appendix}
-\section{The adaptive PAIS algorithm}\label{Sec:App}
-\begin{table}[!h]
-\begin{mdframed}
-\begin{algorithmic}
-\STATE $x_j^{(0)} \sim \mu_0$, for $j \in L\cup U$, where $L = \{j\}_{j=1}^{M/2},\ U = \{j\}_{j=M/2+1}^M$.
-\STATE Choose $\beta^{(1)} \in (0, 2]$. Set $\beta_{L,U}^{(1)} = (1\pm 0.01)\beta^{(1)}\wedge 2$.
-\FOR{$i=1,2, \ldots, N$}
-\STATE $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\beta_L^{(i)})$ for $j \in L$, and $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\beta_U^{(i)})$ for $j \in U$.
-\STATE Calculate 
-\[
-	w^{(i)}_j = \frac{\pi(y_j^{(i)})}{\nu(y_j^{(i)};X^{(i-1)})},
-\]
-where
-\[
-	\nu(y;X) = \frac{1}{M}\sum_{j\in L} q(y;x_j,\beta_L)+\frac{1}{M}\sum_{j\in U} q(y;x_j,\beta_U).
-\]
-
-\IF{$i$ is in $\{n_k\}_{k=1}$}
-	\STATE For $w_{kj} = w_j^{(k)}$, $S = n_k - n_{k-1}$,
-	\STATE $T_L = SM(\sum_{k=i-S}^S\sum_{j\in L} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in L} w_{kj}^2)$.
-	\STATE $T_U = SM(\sum_{k=i-S}^S\sum_{j\in U} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in U} w_{kj}^2)$.
-	\STATE $\beta^{(i+1)} = \beta^{(i)} - \Delta t \displaystyle\frac{T_U - T_L}{\beta_U^{(i)}-\beta_L^{(i)}}$.
-\ELSE
-	\STATE $\beta^{(i+1)} = \beta^{(i)}$.
-\ENDIF
-
-\IF{$i < N_\text{stop}$}
-	\IF{$i < N_\text{join}$}
-		\STATE Resample 
-	\[
-		(w_j^{(i)},y_j^{(i)})_{j\in L} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in L}, \quad (w_j^{(i)},y_j^{(i)})_{j\in U} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in U}.
-\]
-	\ELSE
-		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
-	\ENDIF
-	\STATE $\beta_{L,U}^{(i)} = \beta^{(i)} \pm 2\sqrt{2}\beta^{(i)}/\sqrt{NM} \wedge 2$.
-\ELSE
-		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
-		\STATE $\beta_L^{(i+1)} = \beta_U^{(i+1)} = \beta^{(i+1)}$.
-\ENDIF
-\ENDFOR
-\end{algorithmic}
-\end{mdframed}\caption{A pseudo-code representation of the adaptive
-  PAIS algorithm.}
-\label{tab:adapt}
-\end{table}
-\end{appendix}
+% \begin{appendix}
+% \section{The adaptive PAIS algorithm}\label{Sec:App}
+% \begin{table}[!h]
+% \begin{mdframed}
+% \begin{algorithmic}
+% \STATE $x_j^{(0)} \sim \mu_0$, for $j \in L\cup U$, where $L = \{j\}_{j=1}^{M/2},\ U = \{j\}_{j=M/2+1}^M$.
+% \STATE Choose $\beta^{(1)} \in (0, 2]$. Set $\beta_{L,U}^{(1)} = (1\pm 0.01)\beta^{(1)}\wedge 2$.
+% \FOR{$i=1,2, \ldots, N$}
+% \STATE $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\beta_L^{(i)})$ for $j \in L$, and $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\beta_U^{(i)})$ for $j \in U$.
+% \STATE Calculate 
+% \[
+% 	w^{(i)}_j = \frac{\pi(y_j^{(i)})}{\nu(y_j^{(i)};X^{(i-1)})},
+% \]
+% where
+% \[
+% 	\nu(y;X) = \frac{1}{M}\sum_{j\in L} q(y;x_j,\beta_L)+\frac{1}{M}\sum_{j\in U} q(y;x_j,\beta_U).
+% \]
+
+% \IF{$i$ is in $\{n_k\}_{k=1}$}
+% 	\STATE For $w_{kj} = w_j^{(k)}$, $S = n_k - n_{k-1}$,
+% 	\STATE $T_L = SM(\sum_{k=i-S}^S\sum_{j\in L} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in L} w_{kj}^2)$.
+% 	\STATE $T_U = SM(\sum_{k=i-S}^S\sum_{j\in U} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in U} w_{kj}^2)$.
+% 	\STATE $\beta^{(i+1)} = \beta^{(i)} - \Delta t \displaystyle\frac{T_U - T_L}{\beta_U^{(i)}-\beta_L^{(i)}}$.
+% \ELSE
+% 	\STATE $\beta^{(i+1)} = \beta^{(i)}$.
+% \ENDIF
+
+% \IF{$i < N_\text{stop}$}
+% 	\IF{$i < N_\text{join}$}
+% 		\STATE Resample 
+% 	\[
+% 		(w_j^{(i)},y_j^{(i)})_{j\in L} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in L}, \quad (w_j^{(i)},y_j^{(i)})_{j\in U} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in U}.
+% \]
+% 	\ELSE
+% 		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
+% 	\ENDIF
+% 	\STATE $\beta_{L,U}^{(i)} = \beta^{(i)} \pm 2\sqrt{2}\beta^{(i)}/\sqrt{NM} \wedge 2$.
+% \ELSE
+% 		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
+% 		\STATE $\beta_L^{(i+1)} = \beta_U^{(i+1)} = \beta^{(i+1)}$.
+% \ENDIF
+% \ENDFOR
+% \end{algorithmic}
+% \end{mdframed}\caption{A pseudo-code representation of the adaptive
+%   PAIS algorithm.}
+% \label{tab:adapt}
+% \end{table}
+% \end{appendix}
 

### 2016-06-02 PAIS.tex
Updated AMR section
Just some small tweaks.

@@
   naive parallelisation of serial MCMC methods using the same number
-  of processors, for low dimensional problems, and in fact shows
+  of ensemble members, for low dimensional problems, and in fact shows
   better than linear improvements in convergence rates with respect to
-  the number of processors used. We also introduce a new resampling
+  the number of ensemble members. We also introduce a new resampling
   strategy, approximate multinomial resampling (AMR), which while not
@@
 All MCMC methods are trivially parallelisable. One can take a method
-and simply implement it simultaneously over a set of processors. All
-of the states of all of the processes can be recorded, and in the time
-that it takes one process to draw $N$ samples, $M$ processes can draw
+and simply implement it simultaneously over a set of processors in an ensemble. All
+of the states of all of the ensemble member can be recorded, and in the time
+that it takes one MCMC chain to draw $N$ samples, $M$ ensemble members
+can draw
 $NM$ samples.
@@
 away. This process is known as the burn-in. In a trivially
-parallelised scenario, each process must perform this process
+parallelised scenario, each ensemble member must perform this process
 independently, and therefore mass parallelisation makes no inroads to
@@
 Moreover, a trivially parallelised MCMC scheme is exactly that -
-trivial. Intuition suggests that we can gain some speed up by sharing information across the processes and that is
-exactly what we wish to demonstrate in this paper.
+trivial. Intuition suggests that we can gain some speed up by sharing
+information across the ensemble members, and this is
+what we wish to demonstrate in this paper.
 
@@
 probability distribution $\mu_D$, where $D$ represents our data which
-we wish to use to infer $x$. Since we have $M$ processes, we
+we wish to use to infer $x$. Since we have $M$ ensemble members, we
 represent the current state of all of the Markov chains as a vector
@@
 is simply chosen as an example to show that sharing information across
-processes can improve on the original MCMC algorithm and lead to
+ensemble members can improve on the original MCMC algorithm and lead to
 convergence in fewer evaluations of $\G$. This is important since if
@@
 imagine the algorithm in the burn-in phase, for example, we have $M$
-processes in the tail of a Gaussian curve searching for the area of
-high density. If the processes are evenly spaced, then the particle
+ensemble members in the tail of a Gaussian curve searching for the area of
+high density. If the ensemble members are evenly spaced, then the particle
 closest to the mean will have an exponentially higher weight assigned
@@
 given by a sample from each of the multinomials defined by the vectors
-${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,M}]$ with associated states ${\bf x}_i$. Alternatively, as with the ETPF, a deterministic sample
+${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,M}]$ with associated states ${\bf y}_i$. Alternatively, as with the ETPF, a deterministic sample
 can be chosen by picking each sample to be equal to the mean value of
@@
 the context of being used with the PAIS algorithm. Firstly, we have
-effectively chopped up the multinomial distribution into used in the
+effectively chopped up the multinomial distribution used in the
 bootstrap filter into $M$ pieces, and we can guarantee that exactly
 one sample will be taken from each section. This leads to a much
-smaller chance of losing entire modes in the density. Secondly, if we do not make a random sample for
+smaller chance of losing entire modes in the density, if each of the
+sub-multinomials is picked in an appropriate fashion. Secondly, if we do not make a random sample for
 each multinomial with probability vector $\bf p_i$ but instead take
@@
 deterministic resampling, using the means of each of the
-sub-multinomials. This resampler was designed with the aims of being
+sub-multinomials as the new samples. This resampler was designed with the aims of being
 more efficient than the ETPF, and more accurate than straight multinomial
@@
 target distribution of $\mathcal{N}(2,3)$. The statistics of the
-resampled outputes were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
+resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
 three moments of the samples changes with $M$ for the three different
@@
 sampler is presented for the first moment since both the ETPF and the
-AMR preserve the mean up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
+AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
 measured in seconds, scales with the ensemble size for the three
@@
 
+***REWRITE***
+
 We have explored the application of parallelised MCMC algorithms in
@@
 with a naive parallelisation of serial methods. Thus our increase in
-efficiency is over and above an $N$-fold increase, where $N$ is the
-number of cores or processors at our disposal. Our approach
-demonstrates a better-than-linear speed-up with the number
-processors/cores used. Thus, our approach is not only embarrassingly
-parallel (as coined by Cleve Moler in \cite{moler1986matrix}), but
-humiliatingly so, provided that an optimal transport algorithm can be
-efficiently implemented in parallel and so does not dominate the
-communication costs.
+efficiency is over and above an $M$-fold increase, where $M$ is the
+number of ensemble members used. Our approach
+demonstrates a better-than-linear speed-up with the number of ensemble
+members used.
 
@@
 algorithm's ability to redistribute, through the resampling regime,
-the chains to regions which require more exploration. This allows the
+the ensemble members to regions which require more exploration. This allows the
 method to be used to sample from complex multimodal distribution.
@@
 
-One disadvantage of parallelised algorithms is that often different
-processors will complete their tasks in different amount of times, for
-a number of reasons, often due to communication between the
-processors. One approach which could be applied to the PAIS to avoid
-this is for each processor to immediately start a new iteration, only
-using the last values of the other processors that were communicated
-to it. This incomplete approach would still be valid, and could lead
-to more efficient use of the computer architecture.
+% One disadvantage of parallelised algorithms is that often different
+% processors will complete their tasks in different amount of times, for
+% a number of reasons, often due to communication between the
+% processors. One approach which could be applied to the PAIS to avoid
+% this is for each processor to immediately start a new iteration, only
+% using the last values of the other processors that were communicated
+% to it. This incomplete approach would still be valid, and could lead
+% to more efficient use of the computer architecture.
 

### 2016-06-02 PAIS.tex

@@
 given by a sample from each of the multinomials defined by the vectors
-${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,M}]$ with associated states ${\bf y}_i$. Alternatively, as with the ETPF, a deterministic sample
+${\bf p}_i = [p_{i,1},p_{i,2},\ldots,p_{i,M}]$ with associated states ${\bf x}_i$. Alternatively, as with the ETPF, a deterministic sample
 can be chosen by picking each sample to be equal to the mean value of
@@
 the context of being used with the PAIS algorithm. Firstly, we have
-effectively chopped up the multinomial distribution used in the
+effectively chopped up the multinomial distribution into used in the
 bootstrap filter into $M$ pieces, and we can guarantee that exactly
 one sample will be taken from each section. This leads to a much
-smaller chance of losing entire modes in the density, if each of the
-sub-multinomials is picked in an appropriate fashion. Secondly, if we do not make a random sample for
+smaller chance of losing entire modes in the density. Secondly, if we do not make a random sample for
 each multinomial with probability vector $\bf p_i$ but instead take
@@
 deterministic resampling, using the means of each of the
-sub-multinomials as the new samples. This resampler was designed with the aims of being
+sub-multinomials. This resampler was designed with the aims of being
 more efficient than the ETPF, and more accurate than straight multinomial
@@
 target distribution of $\mathcal{N}(2,3)$. The statistics of the
-resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
+resampled outputes were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
 three moments of the samples changes with $M$ for the three different
@@
 sampler is presented for the first moment since both the ETPF and the
-AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
+AMR preserve the mean up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
 measured in seconds, scales with the ensemble size for the three

### 2016-06-02 PAIS.tex
Some changes
Not very informative really!

--- 2016-06-02 PAIS.tex
@@
  to the constraints on adaptive MCMC
- algorithms\cite{roberts2007coupling,roberts2009example}, these $n_k$
+ algorithms\cite{roberts2007coupling,roberts2009examples}, these $n_k$
  must grow exponentially further apart. This same adaptive approach
@@
 we also need to decide which statistic $T(\beta)$ to optimise for
-these approaches.
-In the examples which follow, we have optimised the naively parallelised RWMH algorithm using the optimal acceptance rate $\hat{\alpha} = 0.5$.
-This value differs from the theoretical asymptotic value of 0.234 which applies in higher dimensions, this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}.
+these approaches. In the examples which follow, we have optimised the
+naively parallelised RWMH algorithm using the optimal acceptance rate
+$\hat{\alpha} = 0.5$. This value differs from the theoretical
+asymptotic value of 0.234 which applies in higher dimensions, but this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}.
 To optimise the algorithm we minimise the statistic
@@
 PAIS algorithm, we are maximising the effective sample size as
-discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\beta) =
-\neff(\beta)$.
+discussed in Section~\ref{sec:ess}.
 
@@
 \[
-D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2).
+D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2) = \mathcal{N}(4,0.01).
 \]
@@
 divergence, which gives us a measure of how different the prior and
-posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this problem.
+posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
+problem. A KL divergence of zero indicates that two distributions are
+identical almost everywhere.
 
@@
 Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter.
-In general this indicator underestimates the estimate found by using the entire sample.
+In general this indicator underestimates the value found by using the entire sample.
 
@@
 
-Both adaptive algorithms are run with initial values of $\beta_0=2.0$.
+Both adaptive algorithms are run with initial values of $\beta=2.0$.
 Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance.
@@
 
-Throughout this paper, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales when we increase the ensemble size, and if there is some limit below which the algorithm fails.
+Throughout this example, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales when we increase the ensemble size, and if there is some limit below which the algorithm fails.
 We implement the problem in Section~\ref{sec:problem 1}, using the RWMH and PAIS-RW algorithms with ensemble sizes in the inteveral $M \in [1, 160]$.
@@
 reduce the number of iterations required in comparison with naively parallelised MH. When $M<8$, the algorithm takes a long time to reach
-stationarity. The decreaseing trend indicates superlinear improvement of PAIS with
+stationarity. The decreasing trend indicates superlinear improvement of PAIS with
 respect to ensemble size, in terms of the number of iterations
@@
 
-We look at an `easy' problem, $B_1$, which has a KL divergence of 0.880, and a `harder' problem, $B_2$, which has a KL divergence of 3.647.
-Problem $B_1$ has two modes which are separated by a small energy barrier.
-In $B_2$ we increase the distance between the two modes which has the effect of increasing the required energy to jump between modes.
-These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.
-
-\subsubsection{Target Distribution}
+We look at an `easy' problem, $B_1$, which has a KL divergence of
+0.880, and a `harder' problem, $B_2$, which has a KL divergence of
+3.647. Problem $B_1$ has two modes which are not too far apart. In
+$B_2$ we increase the distance between the two modes which has the
+effect of increasing the expected number of iterations that it takes
+for a MCMC chain to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.
+
+\subsubsection{Target Distribution} \label{sec:tar}
 
@@
 \includegraphics[width=\textwidth]{"figures/posteriors3"}
-\caption{Posterior distributions for problem $B_1$ and $B_2$.
-Problem $B_1$ has a noisy data value of  0.921312 which results in an energy barrier which is relatively easy to cross, as well as their common prior distribution.}
+\caption{Posterior distributions for problems $B_1$ and $B_2$ as
+  described in Section \ref{sec:tar}.}
 \label{fig:problem 3 posteriors}
@@
 
-Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for RWMH and we use the effective sample size to find the optimal values for PAIS-RW.
+Calculating the optimal values of the scaling parameters for this
+problem is similar to the previous example; we check only the acceptance rate to find the optimal values for RWMH and we use the effective sample size to find the optimal values for PAIS-RW.
 Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both problems.
@@
 These values are given in Table~\ref{table:BM_opt_beta} (left).
-However problem $B_2$ is much harder; transitions between the modes are extremely unlikely for the standard RWMH algorithm.
+However problem $B_2$ is much harderas transitions between the modes are extremely unlikely for the standard RWMH algorithm.
 This means that we need to consider the
@@
 From Table~\ref{table:BM_opt_beta} (right) we see that there is a large difference between the optimal value of $\beta$ for each regime in RWMH, and so will result in inefficient sampling.
-The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\beta^*$: a clear advantage to using this algorithm for this problem.
+The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\beta^*$; a clear advantage to using this algorithm for this problem.
 
@@
 
-As in the Gaussian example we see a significant speed up with the PAIS-RW algorithm for $B_1$.
-Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates.
-We can see that the adaptive algorithms compare closely with the
+As in the Gaussian example we see a significant speed up with the
+PAIS-RW algorithm for problem $B_1$. Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates. We can see that the adaptive algorithms compare closely with the
 respective nonadaptive algorithms and the improvement PAIS offers remains significant.
@@
 
-Comparing the convergence of the adaptive algorithms against the nonadaptive algorithms in Figure~\ref{fig:BM2_AL2} shows that the algorithms behave as expected.
-The adaptive RWMH algorithm tuned using the acceptance rate converges at the same rate as the $L^2$ optimised algorithm until the scaling parameter gets small where the algorithm forgets one of the modes.
-The adaptive PAIS-RW with scout chains tuned to the effective sample size converges at about the same rate as the locally optimised nonadaptive algorithm also with scouts.
+Comparing the convergence of the adaptive algorithms against the
+nonadaptive algorithms in Figure~\ref{fig:BM2_AL2} shows that the
+algorithms behave as expected. The adaptive RWMH algorithm tuned using
+the acceptance rate converges at the same rate as the $L^2$ optimised
+algorithm until the scaling parameter gets small, and therefore
+switches between the modes are rare, and the relative heights of the
+modes are decided by the arbitrary proportion of chains which are in
+each mode at this point. The adaptive PAIS-RW with scout chains tuned to the effective sample size converges at about the same rate as the locally optimised nonadaptive algorithm also with scouts.
 
@@
 \includegraphics[width=\textwidth]{"figures/calc_saving"}
-\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $10^{-2.5}$ as a percentage of MH iterations.
+\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $~10^{-2.85}$ as a percentage of MH iterations.
 The relative $L^2$ error graphs are from the Gaussian problem in Section~\ref{sec:problem 1}.}

### 2016-06-02 PAIS.tex
Numerics sections checked

@@
 \includegraphics[width=\textwidth]{"figures/calc_saving"}
-\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $~10^{-2.85}$ as a percentage of MH iterations.
+\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $~10^{-2.8}$ as a percentage of MH iterations.
 The relative $L^2$ error graphs are from the Gaussian problem in Section~\ref{sec:problem 1}.}
@@
 
-It is often possible to isolate subsystems of fast and slow reactions.
-The quasi-steady-state assumption (QSSA) is that these fast subsystems converge in distribution much faster than the slow subsystem.
-Approximating the fast subsystem as a steady state results in a simple analysis for the fast reaction rates which we can use to simulate the full system and obtain distributions for each of the reaction rates.
-
-We introduce the simple chemical system,
+It is often possible to isolate which reactions are occurring more
+frequently (the fast reactions) and which are occurring less
+frequently (the slow reactions). The quasi-steady-state assumption
+(QSSA), which assumes that the fast reactions converge in distribution
+on a timescale which is negligible with respect to the rate of
+occurrence of the slow reactions. This assumption allows us to
+approximate the dynamics of the slowly changing quantities in the
+system by assuming that the fast quantities are in equillibrium with
+respect to the fast reactions in isolation. This kind of model
+reduction can be used to approximate the likelihood in an inverse
+problem where we wish to recover the reaction parameters in the
+system.
+
+Let us consider a simple example by introducing the following simple
+chemical system involving two chemical species $S_1$ and $S_2$:
 \begin{equation}\label{eqn:full_system}
-	\emptyset \xrightarrow{k_1} X \xleftrightarrows[k_3]{k_2} Y \xrightarrow{k_4} \emptyset,
+	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
 \end{equation}
-where reactions $X\rightarrow Y$ and $Y\rightarrow X$ happen much faster than the creation and destruction of $X$ and $Y$.
-Using the QSSA we can combine the chemical species $X$ and $Y$ into a single species $S=X+Y$, and represent our system as,
+Each arrow represents a reaction from a reactant to a product, with
+some rate constant $k_i$, and where the rates of the reactions are
+assumed to follow mass action kinetics. We assume that we are in a
+parameter regime such that the reactions $S_1\rightarrow S_2$ and $S_2\rightarrow S_1$ occur
+much much more frequently than the other reactions. Notice that both
+chemical species are involved in fast reactions. However, the quantity
+$\mathcal{S} = S_1 + S_2$ is conserved by both of the fast reactions,
+and as such, this is the slowly changing quantity in this system.
+The effective dynamics of $\mathcal{S}$ can be represented as follows.
 \begin{equation}\label{eqn:QSSA_system}
@@
 \end{equation}
-Here, the new reaction rate $\hat{k}_4$ is found to be
+Here, the new reaction rate $\hat{k}_4$ is approximated through
+application of the QSSA to be
 \[
-	\hat{k}_4(s) = \mathbb{E}\left[k_4Y|S=s\right] = \frac{k_2k_4S}{k_2+k_3}, \quad \text{since} \quad Y \sim \text{Bin}\left(S, \frac{k_2}{k_2+k_3}\right).
+	\hat{k}_4(s) = \mathbb{E}\left[k_4S_2|\mathcal{S}=s\right] = \frac{k_2k_4\mathcal{S}}{k_2+k_3}.
 \]
-By observing the system in Equation~\ref{eqn:QSSA_system}, we indirectly observe the rates $k_2$ and $k_3$ through $\hat{k}_4$, resulting in infinitely many possibilities for $k_2$ and $k_3$.
-Along with the intrinsic randomness and errors in observations, which we assume to follow a Gamma distribution with some variance $\sigma^2$, this results in a long and thin posterior distribution when put into the framework of a Bayesian inverse problem.
+The value of $\mathbb{E}\left[k_4S_2|\mathcal{S}=s\right]$ is
+approximated by finding the steady state of the ODE representing the fast subsystem of
+reactions:
+\[S_1 \xleftrightarrows[k_3]{k_2} S_2, \qquad S_1 + S_2 = s\]
+
+If we assume that we know the rate constants $k_1$ and $k_4$, and
+observe the system in Equation~\ref{eqn:QSSA_system}, we indirectly
+observe the rates $k_2$ and $k_3$ through the effective rate
+$\hat{k}_4$ of the degradation of $\mathcal{S}$. Our
+observations are uninformative about these reaction rates, as there
+are surfaces in parameter space along which the effective rate
+$\hat{k}_4$ is invariant, leading to a highly illposed inverse
+problem. Making the assumption that the errors in our observations of
+the value of $\mathcal{S}$ are Gamma distributed with some variance
+$\sigma^2$, this results in a long and thin posterior distribution on
+$k_2$ and $k_3$. This type of problem is notoriously difficult to
+sample from using standard MH algorithms, as the algorithms quickly
+find a point on this manifold on which $\hat{k}_4$ is invariant, but
+exploration along its length is slow.
 
@@
 \begin{align*}
-	\frac{\text{d}X}{\text{d}t} &= k_1 - k_2X(t)+k_3Y(t), \\
-	\frac{\text{d}Y}{\text{d}t} &= k_2X(t) - (k_3+k_4)Y(t),
+	\frac{\text{d}S_1}{\text{d}t} &= k_1 - k_2S_1(t)+k_3S_2(t), \\
+	\frac{\text{d}S_2}{\text{d}t} &= k_2S_1(t) - (k_3+k_4)S_2(t),
 \end{align*}
-with initial conditions $X(0)=Y(0) = 0$, and parameter values $\mathbf{k} = (50, 100)^T$.
-We then add noise taken from a Gamma distribution with variance $\sigma^2 = 225$ and centred at each $S(t_i)=X(t_i)+Y(t_i)$.
+with initial conditions $S_1(0)=S_2(0) = 0$, and parameter values $\mathbf{k} = (50, 100)^T$.
+We then add noise taken from a Gamma distribution with variance $\sigma^2 = 225$ and centred at each $\mathcal{S}(t_i)=S_1(t_i)+S_2(t_i)$.
 
@@
 
-We assign Gamma priors to $\mathbf{k}$ with mean $\alpha_0/\beta_0 = 75$ and variance $\alpha_0/\beta_0^2 = 100$, resulting in the posterior
+We assign Gamma priors to $\mathbf{k}$ with mean $\alpha_0/\beta_0 =
+75$ and variance $\alpha_0/\beta_0^2 = 100$ in both coordinates, resulting in the posterior
 \[
@@
 % explanation of l2 convergence measure
-For this problem there is no analytic form for the normalisation constant, and numerical methods implemented in MATLAB and Mathematica have proven to be unreliable.
-To demonstrate the convergence of the algorithm, we first run a MH algorithm for much longer than we normally would ($8\times 10^{10}$ samples), and consider the histogram produced to be the truth.
-We then perform our usual simulations with the PAIS and MH algorithms to compare the rate at which the histograms converge to the `true' histogram over the same grid.
+For this problem there is no analytic form for the normalisation
+constant, and numerical methods implemented in MATLAB and Mathematica
+have proven to be unreliable. To demonstrate the convergence of the
+algorithm, we first run a MH algorithm for much longer than we
+normally would ($8\times 10^{10}$ samples), and consider the histogram
+produced to be a well-converged approximation of the true posterior
+distribution. We then perform our usual simulations with the PAIS and
+MH algorithms to compare the rate at which the histograms converge to
+the posterior over the same histogram bins.
 
 % modifications to algorithm + spikes
-In this problem we modify our PAIS algorithm to use a mixture of Gamma distributions in the proposal distribution instead of a Gaussian mixture.
-The PAIS-LA algorithm uses the Langevin proposal distribution from the standard MALA algorithm with a pertubation taken from a Gamma distribution with mean $X+\frac{1}{2}\beta^2\nabla \log\pi(\cdot)$, and  variance $\beta^2$.
-Similarly we use a Gamma proposal in the MH algorithm.
-This ensures that we only propose positive parameters.
-Having a heavier right tail in the proposal distribution also means that the posterior is absolutely continuous with respect to the proposal.
-When this is not the case, the weight function can tend to zero or infinity when $k_i \rightarrow \infty$.
-Sampling under these conditions can result in very spiky approximations of the tails of the posterior distribution resulting in slow convergence.
+In this problem we modify our PAIS algorithm to use a mixture of Gamma
+distributions in the proposal distribution instead of a Gaussian
+mixture. The PAIS-LA algorithm uses the Langevin proposal distribution
+from the standard MALA algorithm with a pertubation taken from a Gamma
+distribution with mean $X+\frac{1}{2}\beta^2\nabla \log\pi(\cdot)$,
+and  variance $\beta^2$. Similarly we use a Gamma proposal in the MH
+algorithm. This ensures that we only propose positive
+parameters. Having a heavier right tail in the proposal distribution
+also means that the posterior is absolutely continuous with respect to
+the proposal. When this is not the case, the weight function can tend
+to zero or infinity when $k_i \rightarrow \infty$, which results in
+poor, very spiky approximations of the tails of the posterior distribution leading to slow convergence.
 
@@
 
-In Figure~\ref{fig:C1_ESS} (a) we see that the two PAIS algorithms have similar optimal values of the scaling parameter, also shown in Table~\ref{table:C1_opt_beta}, however the MALA-type proposal achieves a higher effective sample size with the same ensemble size.
+In Figure~\ref{fig:C1_ESS} (a) we see that the two PAIS algorithms have similar optimal values of the scaling parameter, also shown in Table~\ref{table:C1_opt_beta}.
+However, the MALA-type proposal achieves a higher effective sample size with the same ensemble size.
 Convergence after the first 50 million samples is shown in Figure~\ref{fig:C1_ESS} (b) and demonstrates that the PAIS algorithm converges faster than the MH algorithm.

### 2016-06-02 PAIS.tex
Conclusions

@@
 
-***REWRITE***
-
 We have explored the application of parallelised MCMC algorithms in
@@
 Random Walk Metropolis-Hastings (RWMH), Metropolis Adjusted Langevin
-Algorithm (MALA) and preconditioned Crank-Nicolson (pCN) proposals has
+Algorithm (MALA), preconditioned Crank-Nicolson (pCN) and Hamiltonian
+Monte Carlo (HMC) proposals has
 yielded similar results\cite{Paul}.
@@
 One limitation of the PAIS approach as described above is that a
-direct solver (such as FastEMD \cite{FastEMD}) has computational cost
-$\mathcal{O}(n^3\log n)$, where $n$ is the number of particles in the
-ensemble.  The use of approximate solutions from iterative solvers
-might allow the PAIS to be used in higher dimensional problems, and is
-an avenue for future investigation. A second limitation is associated
-with the problem of accurately approximating measures with empirical
-measures in high dimensional phase space. This is a well-known issue
-in filtering, where one possible solution is to ``localise'' the
-impact of observations, as rigorously analysed in
-\cite{rebeschini2013can}. An established technique in Ensemble Kalman
-filters, localisation has also been incorporated into the Ensemble
-Transform Particle Filter \cite{ChRe2015}. This issue could be
-addressed in the PAIS by using localisation when computing the
-transform, but keeping the unlocalised weights in the importance
-sampling for consistency.
+direct solver of the ETPF problem (such as FastEMD \cite{FastEMD}) has computational cost
+$\mathcal{O}(M^3\log M)$, where $M$ is the number of particles in the
+ensemble. As such, we introduced a more approximate resampler the
+approximate multinomial resampler, which allows us to push the
+approach to the limit
+with much larger ensemble sizes. The PAIS framework is very flexible
+in terms of being able to use any combination of proposal
+distributions and resampling algorithms that one wishes.
+
+%  A second limitation is associated
+% with the problem of accurately approximating measures with empirical
+% measures in high dimensional phase space. This is a well-known issue
+% in filtering, where one possible solution is to ``localise'' the
+% impact of observations, as rigorously analysed in
+% \cite{rebeschini2013can}. An established technique in Ensemble Kalman
+% filters, localisation has also been incorporated into the Ensemble
+% Transform Particle Filter \cite{ChRe2015}. This issue could be
+% addressed in the PAIS by using localisation when computing the
+% transform, but keeping the unlocalised weights in the importance
+% sampling for consistency.
 

### 2016-06-02 PAIS.tex
Sections summary updated

@@
 
-***SECTIONS SUMMARY NEEDS REWRITING***
+In Section \ref{Sec:Prelim} we introduce some mathematical
+preliminaries upon which we will later rely. In Section \ref{Sec:PAIS}
+we present the general framework of the PAIS algorithm. In Section
+\ref{Sec:adapt} we consider adaptive versions of PAIS which
+automatically tune algorithmic parameters concerned with the proposal
+distributions. In Section \ref{sec:AMR} we introduce the approximate
+multinomial resampling (AMR) algorithm which is a less accurate but
+faster alternative to the ETPF resampler. In Section \ref{Sec:Num} we
+present some numerical examples, before a brief conclusion and
+discussion in Section \ref{Sec:Conc}.
 

### 2016-06-02 PAIS.tex
further elaboration on proposal distribution in section 6.3.2

--- 2016-06-02 PAIS.tex
@@
 distributions in the proposal distribution instead of a Gaussian
-mixture. The PAIS-LA algorithm uses the Langevin proposal distribution
+mixture. The PAIS-Gamma and MH-Gamma algorithms use a Gamma
+proposal distribution with mean centred at the previous state,
+\[
+	Y \sim \text{Gamma}(\cdot;\alpha^*, \beta^*) \quad \text{where} \quad \frac{\alpha^*}{\beta^*} = X \ \text{and} \ \frac{\alpha^*}{(\beta^*)^2} = \beta^2.
+\]
+
+Similarly the PAIS-LA algorithm uses the Langevin proposal distribution
 from the standard MALA algorithm with a pertubation taken from a Gamma
-distribution with mean $X+\frac{1}{2}\beta^2\nabla \log\pi(\cdot)$,
-and  variance $\beta^2$. Similarly we use a Gamma proposal in the MH
-algorithm. This ensures that we only propose positive
+distribution,
+\[
+	Y \sim \text{Gamma}(\cdot; \alpha^*, \beta^*), \quad \text{where} \quad \frac{\alpha^*}{\beta^*} = X+\frac{1}{2}\beta^2\nabla\log\pi(X) \ \text{and} \ \frac{\alpha^*}{(\beta^*)^2} = \beta^2.
+\]
+This ensures that we only propose positive
 parameters. Having a heavier right tail in the proposal distribution

### 2016-06-02 PAIS.tex
S -> \mathcal{S}
Small changes to chemical systems section

@@
 
-In our modelling we use the QSSA to simplify this system of differential equations into the one parameter system based on Equation~\ref{eqn:QSSA_system},
+In our modelling we use the QSSA to simplify this system of differential equations into the one dimensional system based on Equation~\ref{eqn:QSSA_system},
 \[
-	\frac{\text{d}S}{\text{d}t} = k_1 - \frac{k_2k_4}{k_2+k_3}S(t).
+	\frac{\text{d}\mathcal{S}}{\text{d}t} = k_1 - \frac{k_2k_4}{k_2+k_3}\mathcal{S}(t).
 \]
@@
 \[
-	Y \sim \text{Gamma}(\cdot;\alpha^*, \beta^*) \quad \text{where} \quad \frac{\alpha^*}{\beta^*} = X \ \text{and} \ \frac{\alpha^*}{(\beta^*)^2} = \beta^2.
+	y \sim \text{Gamma}(\cdot;\alpha^*, \beta^*) \quad \text{where} \quad \frac{\alpha^*}{\beta^*} = x \ \text{and} \ \frac{\alpha^*}{(\beta^*)^2} = \beta^2.
 \]
@@
 \[
-	Y \sim \text{Gamma}(\cdot; \alpha^*, \beta^*), \quad \text{where} \quad \frac{\alpha^*}{\beta^*} = X+\frac{1}{2}\beta^2\nabla\log\pi(X) \ \text{and} \ \frac{\alpha^*}{(\beta^*)^2} = \beta^2.
+	y \sim \text{Gamma}(\cdot; \alpha^*, \beta^*), \quad \text{where} \quad \frac{\alpha^*}{\beta^*} = x+\frac{1}{2}\beta^2\nabla\log\pi(X) \ \text{and} \ \frac{\alpha^*}{(\beta^*)^2} = \beta^2.
 \]

### 2016-06-05 PAIS.tex
added missed S
added some evidence that high ESS = end of burnin

@@
 closest to the mean will have an exponentially higher weight assigned
-to it. The effective sample size in this scenario will be close to
+to it. The effective sample size ratio in this scenario will be close to
 1. As the algorithm burns in, the ensemble populates the regions where
@@
 By this argument we can see that rising $\neff$ signals the end of the burn-in period.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.65\textwidth]{"figures/C1_burnin"}
+\caption{The effective sample size ratio and relative L2 error between the proposal and posterior distributions at each of the first 30 iterations.
+These numerics are taken from a simulation of the problem in Section~\ref{sec:chem} using the PAIS-Gamma algorithm.}
+\label{fig:neff-burnin}
+\end{figure}
+
+Figure~\ref{fig:neff-burnin} shows that the effective sample size reaches stationarity at the same time as the relative L2 error between the posterior distribution and the proposal distribution reaches its minimum.
+Detecting this stationarity allows us to automatically determine the end of the burn-in phase of the algorithm.
 
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth]{"figures/PAIS_saving"}
+\includegraphics[width=0.6\textwidth]{"figures/PAIS_saving"}
 \caption{Ratio of PAIS-RW samples required to reach the same tolerance as the RWMH algorithm.}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth]{"figures/BM2_AL2"}
+\includegraphics[width=0.6\textwidth]{"figures/BM2_AL2"}
 \caption{Convergence of the relative $L^2$ error for problem $B_2$, comparing the globally optimised nonadaptive algorithms with the locally optimised adaptive algorithms.
@@
 \begin{center}
-\includegraphics[width=\textwidth]{"figures/calc_saving"}
+\includegraphics[width=0.8\textwidth]{"figures/calc_saving"}
 \caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $~10^{-2.8}$ as a percentage of MH iterations.
@@
 \begin{equation}\label{eqn:QSSA_system}
-	\emptyset \xrightarrow{k_1} S \xrightarrow{\hat{k}_4} \emptyset.
+	\emptyset \xrightarrow{k_1} \mathcal{S} \xrightarrow{\hat{k}_4} \emptyset.
 \end{equation}

### 2016-06-14 PAIS.tex
fixed some typos in the introduction+preliminaries

@@
 even higher order gradient information\cite{bui2014solving}.  These
-methods allow us explore the posterior distribution more fully 
+methods allow us to explore the posterior distribution more fully 
 with fewer iterations.
@@
 As such, the development of efficient parallel MCMC algorithms is an
-important area for research. Since MCMC methods can be trivially
+important area for research. Since MCMC methods can be naively
 parallelised by simply running many independent chains in parallel,
@@
 proposal from the current state of all of the chains in the
-ensemble. The idea is to consider the parallel chains as an ensemble,
-and to resample using a transformation based on optimal transport.
-Samplers based on optimal transport have also been considered in
-\cite{el2012bayesian}.
+ensemble. The idea is to consider the current state on each of a set of
+parallel chains as an ensemble, and to resample using a transformation
+based on optimal transport. Samplers based on optimal transport
+have also been considered in \cite{el2012bayesian}.
 
@@
 multinomial resampling (AMR) algorithm which is a less accurate but
-faster alternative to the ETPF resampler. In Section \ref{Sec:Num} we
-present some numerical examples, before a brief conclusion and
-discussion in Section \ref{Sec:Conc}.
+faster alternative to resamplers which solve the optimal transport
+problem exactly. In Section \ref{Sec:Num} we present some numerical
+examples, before a brief conclusion and discussion in Section
+\ref{Sec:Conc}.
 
@@
 Since observations are never perfect, we
-assume that these measurements $\data$ are subject to Gaussian noise, so that
+assume that these measurements $\data$ are subject to Gaussian noise,
+so that
 \begin{equation}\label{eqn:obs}
@@
 \end{equation}
-For example, if $u$ are the rates of reactions in a chemical system, $\G$ might return the times at which each reaction occurs, or some summary of this information.
+For example, if $u$ are the rates of reactions in a chemical system, $\G$ might
+return the quantities of each chemical species at a particular time, or some
+summary of this information.
 
@@
 likelihood of observing the data $\data$ given the parameter $u =
-u^*$. Rearranging \eqref{eqn:obs} and using the distribution of $\varepsilon$, we get:
+u^*$. Rearranging \eqref{eqn:obs} and using the distribution of
+$\varepsilon$, we get:
 \begin{equation}\label{eqn:like}
@@
 
-The Ensemble Transform (ET) method proposed by Reich \cite{reich2013nonparametric} makes use of optimal transportation as described in \cite{villani2003topics,villani2008optimal}.
-The transform takes a sample of weighted particles $\{y_i\}_{i=1}^M$ from $\mu_Y$ and converts it into a sample of evenly weighted particles $\{x_i\}_{i=1}^M$ from $\mu_X$, by means of defining a coupling $T^*$ between $Y$ and $X$.
-Given that a trivial coupling $T^t$ always exists in the space of transference plans, $\Pi(\mu_{X}, \mu_{Y})$, we can find a coupling $T^*$ which maximises the correlation between $X$ and $Y$ \cite{cotter2012ensemble}.
-This coupling is the solution to a linear programming problem in $M^2$ variables with $2M-1$ constraints.
-Maximising the correlation ensures that the new sample is as much like the original sample as possible with the additional property that the sample is evenly weighted.
+The ensemble transform particle filter (ETPF) proposed by Reich
+\cite{reich2013nonparametric} makes use of optimal transportation as
+described in \cite{villani2003topics,villani2008optimal}. The transform
+takes a sample of weighted particles $\{y_i\}_{i=1}^M$ from $\mu_Y$
+and converts it into a sample of evenly weighted particles $\{x_i\}_{i=1}^M$
+from $\mu_X$, by means of defining a coupling $T^*$ between $Y$ and
+$X$. Given that a trivial coupling $T^t$ always exists in the space of
+transference plans, $\Pi(\mu_{X}, \mu_{Y})$, we can find a coupling
+$T^*$ which maximises the correlation between $X$ and $Y$
+\cite{cotter2012ensemble}. This coupling is the solution to a linear
+programming problem in $M^2$ variables with $2M-1$ constraints.
+Maximising the correlation ensures that the new sample is as much like
+the original sample as possible with the additional property that the sample
+is evenly weighted.
 
@@
 \subsection{Deficiencies of Metropolis-type MCMC schemes}
-All MCMC methods are trivially parallelisable. One can take a method
+All MCMC methods are naively parallelisable. One can take a method
 and simply implement it simultaneously over a set of processors in an ensemble. All
@@
 that it takes one MCMC chain to draw $N$ samples, $M$ ensemble members
-can draw
-$NM$ samples.
+can draw $NM$ samples.
 
@@
 scenario. First of all, unless we have a lot of information about the
-posterior, we will intialise the algorithm's initial state in the
+posterior, we will initialise the algorithm's initial state in the
 tails of the distribution. The samples that are initially made as the
 algorithm finds its way to the mode(s) of the distribution cannot be
-considered to samples from the target distribution, and must be thrown
-away. This process is known as the burn-in. In a trivially
-parallelised scenario, each ensemble member must perform this process
-independently, and therefore mass parallelisation makes no inroads to
-cutting this cost.
+considered to be samples from the target distribution, and must be thrown
+away. This process is known as the burn-in. In a trivially parallelised
+scenario, each ensemble member must perform this process independently,
+and therefore mass parallelisation makes no inroads to cutting this cost.
 
 Moreover, many MCMC algorithms suffer from poor mixing, especially in
-multimodal systems. The amount of samples that it takes for an MCMC trajectory to switch between modes can be
-large, and given that a large number of switches are required before we have a
-good idea of the relative probability densities of these different
-regions, it can be prohibitively expensive.
+multimodal systems. The amount of samples that it takes for an MCMC
+trajectory to switch between modes can be large, and given that a large
+number of switches are required before we have a good idea of the relative
+probability densities of these different regions, it can be prohibitively
+expensive.
 

### 2016-06-19 PAIS.tex
typos + corrections in sections 3+4

@@
 able to make direct or indirect noisy observations. For now
-we say that $u$ is a member of a Hilbert
+we say that $x$ is a member of a Hilbert
 space $X$.
 
-The parameter $u$ is observed
+The parameter $x$ is observed
 through the observation operator $\mathcal{G}:X \to\mathbb{R}^d$. 
@@
 \begin{equation}\label{eqn:obs}
-	\data = \mathcal{G}(u) + \varepsilon, \qquad \varepsilon \sim \mu_{\varepsilon} = \mathcal{N}(0,\Sigma).
+	\data = \mathcal{G}(x) + \varepsilon, \qquad \varepsilon \sim \mu_{\varepsilon} = \mathcal{N}(0,\Sigma).
 \end{equation}
-For example, if $u$ are the rates of reactions in a chemical system, $\G$ might
+For example, if $x$ are the rates of reactions in a chemical system, $\G$ might
 return the quantities of each chemical species at a particular time, or some
@@
 These modelling assumptions allow us to construct the 
-likelihood of observing the data $\data$ given the parameter $u =
-u^*$. Rearranging \eqref{eqn:obs} and using the distribution of
+likelihood of observing the data $\data$ given the parameter $x =
+x^*$. Rearranging \eqref{eqn:obs} and using the distribution of
 $\varepsilon$, we get:
 \begin{equation}\label{eqn:like}
-	\mathbb{P}(\data|u=u^*) \propto \exp \left ( -\frac{1}{2} \|\mathcal{G}(u^*)
-	  - \data\|_\Sigma^2 \right ) = \exp\left(-\Phi(u^*)\right),
+	\mathbb{P}(\data|x=x^*) \propto \exp \left ( -\frac{1}{2} \|\mathcal{G}(x^*)
+	  - \data\|_\Sigma^2 \right ) = \exp\left(-\Phi(x^*)\right),
 \end{equation}
-where $\| x - y \|_\Sigma$ is the Mahalanobis distance between $x$ and $y$.
+where $\| y_1 - y_2 \|_\Sigma$ is the Mahalanobis distance between $y_1$ and $y_2 \in \mathbb{R}^d$.
 
@@
 \begin{equation}\label{eqn:RND}
-	\frac{d\mu_Y}{d\mu_0} \propto \exp \left ( -\Phi(u^*) \right ).
+	\frac{d\mu_Y}{d\mu_0} \propto \exp \left ( -\Phi(x^*) \right ).
 \end{equation}
@@
 similar areas to the target density, and incredibly inefficient when
-this is not the case. The aim of the PAIS is to use the states of a
-ensemble ofl MCMC chains to construct a proposal distribution which
-will be as close as possible to the target density. Given a big enough
-ensemble, the states of all of these chains at one point
-in time may be reasonably representative sample of the target density.
+this is not the case. The aim of the PAIS is to use an ensemble of states,
+each coming from a particular time point in an MCMC chain, to construct a proposal
+distribution which will be as close as possible to the target density. If this
+ensemble is large enough, the distribution of states will be representative
+of the target density.
 
@@
 $\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
-the random walk Metropolis-Hastings proposal density $\nu(\cdot,x) ~
+the random walk Metropolis-Hastings proposal density $\nu(\cdot,x) \sim
 \mathcal{N}(x,\beta^2)$, where $\beta^2\in \mathbb{R}$ defines the
@@
 $\chi$ is a close approximation to the posterior density $\pi$. If the
-proposal distribution is over-dispersed, then the algorithm will often
+proposal distribution is too over-dispersed, then the algorithm will often
 propose states in the tails of the distribution, resulting in larger
-variance of the weights, and therefore slower converegence to the
+variance of the weights, and therefore slower convergence to the
 posterior distribution. Similarly, if the proposal distribution is
 under-dispersed, the proposals will be highly correlated with the
-previous states, and the algorithm will take a long time to fully explore the parameter spac.
-It is therefore necessary to find a proposal distribution which is slightly over-dispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.
+previous states, and the algorithm will take a long time to fully explore
+the parameter space. It is therefore necessary to find a proposal
+distribution which is slightly over-dispersed to ensure the entire
+posterior is explored\cite{gelman1992inference}, but is as close to the
+posterior as possible.
 
@@
 should be.
+
+The alternative to using adaptive procedures to tune the scaling
+parameters is to perform exploratory simulations to find the optimal
+regimes by trial and error. This can be very costly and the optimal
+parameters cannot realistically be found to more than a couple of
+significant figures. It is therefore important that MCMC algorithms
+provide a feasible adaptive strategy to bypass this exploratory stage.
 
@@
 \end{equation}
-where the regions $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional histogram bins, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of each bin, and $B_i$ is the value of the $i$th bin.
+where the regions $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional
+histogram bins, so that $\bigcup_i R_i \subseteq X$ and
+$R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the
+volume of each bin, and $B_i$ is the value of the $i$th bin. This
+metric converges to the standard definition of the relative $L^2$
+error as $v\rightarrow 0$.
 
@@
 \includegraphics[width=0.65\textwidth]{"figures/C1_burnin"}
-\caption{The effective sample size ratio and relative L2 error between the proposal and posterior distributions at each of the first 30 iterations.
+\caption{The effective sample size ratio and relative $L^2$ error between the proposal and posterior distributions at each of the first 30 iterations.
 These numerics are taken from a simulation of the problem in Section~\ref{sec:chem} using the PAIS-Gamma algorithm.}
@@
 
-Figure~\ref{fig:neff-burnin} shows that the effective sample size reaches stationarity at the same time as the relative L2 error between the posterior distribution and the proposal distribution reaches its minimum.
+Figure~\ref{fig:neff-burnin} shows that the effective sample size reaches stationarity at the same time as the relative $L^2$ error between the posterior distribution and the proposal distribution reaches its minimum.
 Detecting this stationarity allows us to automatically determine the end of the burn-in phase of the algorithm.
@@
 for the Gaussian posterior discussed in Section~\ref{sec:problem 1}
-when using the PAIS algorithm with RWMH proposals, each with variance
+when using the PAIS algorithm with RW proposals, each with variance
 $\beta^2 \in \mathbb{R}_{>0}$. Figure~\ref{fig:neff-M} (a) shows that
@@
  scaling parameter $\beta^*$, resulting in the optimal transition
- density $\chi$ for our MCMC algorithm. We must choose some sequence
+ density $\chi$ for our MCMC algorithm. This optimal value will differ
+ depending on the criterion we are optimising. We must choose some sequence
  of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$, and due
@@
 \subfigure[Relative $L^2$ error using the optimal scaling parameters.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
-\subfigure[Comparing of the relative $L^2$ error of the algorithms optimised by $L^2$ minimisation with locally optimised PAIS-RW with a scout chain.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
+\subfigure[Comparison of the relative $L^2$ error of the algorithms optimised by $L^2$ minimisation with locally optimised PAIS-RW with a scout chain.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
 \caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_2$. Set up described in Section~\ref{sec:Implementation P1}.}

### 2016-06-19 PAIS.tex
corrections in the rest of the sections

@@
 In this example we compare the naively parallelised RWMH algorithm with its
-PAIS variant, the PAIS-RW algorithm. We perform this comparison using the relative $L^2$ error defined in Equation~\ref{eqn:L2_error}, as well as the relative error in the first moment.
+PAIS variant, the PAIS-RW algorithm. We asses the performance of the PAIS algorithm
+using the relative $L^2$ error defined in Equation~\ref{eqn:L2_error}, as well as the
+relative error in the first moment.
 
 Since we are comparing against naively parallelised MH algorithms,
-we also need to decide which statistic $T(\beta)$ to optimise for
-these approaches. In the examples which follow, we have optimised the
-naively parallelised RWMH algorithm using the optimal acceptance rate
-$\hat{\alpha} = 0.5$. This value differs from the theoretical
-asymptotic value of 0.234 which applies in higher dimensions, but this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}.
-To optimise the algorithm we minimise the statistic
+we also need to decide which statistics $T(\beta)$ provide the best criterions
+for obtaining the optimal scaling parameters. In the examples which follow, we have
+optimised the naively parallelised RWMH algorithm using the optimal
+acceptance rate $\hat{\alpha} = 0.5$. This value differs from the theoretical
+asymptotic value of 0.234 which applies in higher dimensions, but this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}. To find the optimal
+scaling parameter we minimise the statistic
 \[
@@
 $N_{\text{total}}$ is the total number of samples produced. For the
-PAIS algorithm, we are maximising the effective sample size as
+PAIS algorithm, we maximise the effective sample size as
 discussed in Section~\ref{sec:ess}.
@@
 
-Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions.
+Consider the simple case of a linear observation operator $\G(x) = x$, where the prior on $x$ and the observational noise follow Gaussian distributions.
 Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
 \begin{equation}\label{eqn:Gaussian posterior}
-	\text{law}(\mu_D) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
+	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right),
 \end{equation}
@@
 numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
-we observe $u=4$ noisily such that
+we observe $x_\text{ref}=4$ noisily such that
 \[
-D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2) = \mathcal{N}(4,0.01).
+D = \mathcal{G}(x) + \eta \sim \mathcal{N}(\mathcal{G}(x_\text{ref}),\sigma^2) = \mathcal{N}(4,0.01).
 \]
@@
 algorithms, and again use the relative $L^2$ error to compare
-efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 2.
+efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 1.
 
@@
 
-Both adaptive algorithms are run with initial values of $\beta=2.0$.
+Both adaptive algorithms are run with initial values of $\beta=1$.
 Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance.
@@
 noisy but demonstrates that increasing the ensemble size continues to
-reduce the number of iterations required in comparison with naively parallelised MH. When $M<8$, the algorithm takes a long time to reach
-stationarity. The decreasing trend indicates superlinear improvement of PAIS with
+reduce the number of iterations required in comparison with naively
+parallelised MH. The decreasing trend indicates superlinear improvement of PAIS with
 respect to ensemble size, in terms of the number of iterations
@@
 
-The following setup is the same for both problems.
-We consider a nonlinear observation operator $\G(u) = u^2$, and assign the prior $u \sim \mu_0 = \N(0, \tau^2=0.25)$.
-We assume that a noisy reading, $D$, is taken according to $D = \G(u) + \varepsilon$, where $\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$.
-This results in the non-Gaussian posterior
+The following setup is the same for both problems. We consider a
+non-linear observation operator $\G(x) = x^2$, and assign the prior
+$x \sim \mu_0 = \N(0, \tau^2=0.25)$. We assume that a noisy reading,
+$D$, is taken according to $D = \G(x_\text{ref}) + \varepsilon$, where
+$\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$. This results
+in the non-Gaussian posterior
 \[
-	\pi(u|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|u^2 - D\|^2 - \frac{1}{2\tau^2}\|u\|^2\right).
+	\pi(x|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|x^2 - D\|^2 - \frac{1}{2\tau^2}\|x\|^2\right).
 \]
-To create the `easy' problem we say that the true value of $\G(u) = 0.75$, and the `hard' problem is generated using $\G(u) = 2$.
-In the numerics which follow we draw noise from $\mu_\varepsilon$ to generate our data point.
+To create the `easy' problem we say that the true value of
+$\G(x_\text{ref}) = 0.75$, and the `hard' problem is generated using
+$\G(x_\text{ref}) = 2$. In the numerics which follow we draw noise from
+$\mu_\varepsilon$ to generate our data point.
 
@@
 Calculating the optimal values of the scaling parameters for this
-problem is similar to the previous example; we check only the acceptance rate to find the optimal values for RWMH and we use the effective sample size to find the optimal values for PAIS-RW.
-Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both problems.
+problem is similar to the previous example; we check only the acceptance
+rate to find the optimal values for RWMH and we use the effective
+sample size to find the optimal values for PAIS-RW.
+Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both
+problems. The subscript on $\beta$ refers to the criterion which has been optimised.
 
@@
 It is relatively simple to find optimal scaling parameters for problem $B_1$.
-These values are given in Table~\ref{table:BM_opt_beta} (left).
-However problem $B_2$ is much harderas transitions between the modes are extremely unlikely for the standard RWMH algorithm.
-This means that we need to consider the
+These values are given in Table~\ref{table:BM_opt_beta} (left). However
+problem $B_2$ is much harder as transitions between the modes are extremely
+unlikely for the standard RWMH algorithm. This means that we need to consider the
 convergence on two levels; we should consider the algorithm's ability
-to find all the modes, and also whether it can sample them in the correct proportions.
+to find both the modes, and also whether it can sample them in the correct proportions.
 
 To get correctly proportioned modes with the RWMH algorithm it is
-important that the chains can transition between the modes, which
+important that the chains can transition between the modes frequently, which
 means that $\beta$ must be large. However, this leads to a lower
@@
 are surfaces in parameter space along which the effective rate
-$\hat{k}_4$ is invariant, leading to a highly illposed inverse
+$\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
 problem. Making the assumption that the errors in our observations of
@@
 MH algorithms to compare the rate at which the histograms converge to
-the posterior over the same histogram bins.
+the posterior over the same mesh.
 

### 2016-06-20 PAIS.tex
Final? edits.

@@
 considered to be samples from the target distribution, and must be thrown
-away. This process is known as the burn-in. In a trivially parallelised
+away. This process is known as the burn-in. In a naively parallelised
 scenario, each ensemble member must perform this process independently,
@@
 
-Moreover, a trivially parallelised MCMC scheme is exactly that -
-trivial. Intuition suggests that we can gain some speed up by sharing
+Moreover, a naively parallelised MCMC scheme is exactly that -
+naive. Intuition suggests that we can gain some speed up by sharing
 information across the ensemble members, and this is
@@
 
-These deficiencies of the trivial method of parallelising MCMC methods
+These deficiencies of the naive method of parallelising MCMC methods
 motivated the development of the Parallel Adaptive Importance Sampler
@@
 \includegraphics[width=0.65\textwidth]{"figures/C1_burnin"}
-\caption{The effective sample size ratio and relative $L^2$ error between the proposal and posterior distributions at each of the first 30 iterations.
-These numerics are taken from a simulation of the problem in Section~\ref{sec:chem} using the PAIS-Gamma algorithm.}
+\caption{The effective sample size ratio and relative $L^2$ difference
+between the proposal and posterior distributions at each of the first 30
+iterations. These numerics are taken from a simulation of the problem in
+Section~\ref{sec:chem} using the PAIS-Gamma algorithm.}
 \label{fig:neff-burnin}
@@
 
-Figure~\ref{fig:neff-burnin} shows that the effective sample size reaches stationarity at the same time as the relative $L^2$ error between the posterior distribution and the proposal distribution reaches its minimum.
-Detecting this stationarity allows us to automatically determine the end of the burn-in phase of the algorithm.
+Figure~\ref{fig:neff-burnin} shows that the effective sample size flattens
+out at the same time as the relative $L^2$ difference between
+the posterior distribution and the proposal distribution stabilises close to
+its minimum. Detecting this stationarity allows us to automatically
+determine the end of the burn-in phase of the algorithm.
 
@@
 to split the original multinomial up into $M$ parts, some of which
-will be far from optimal. The method that we have arrived at is based
-on the idea that we want to group together states which are
-geographically close to each other, preferentially doing this 
-for the states with the biggest weights.
-
-\begin{table}
+will be far from optimal. The method that we have chosen is loosely
+based on the idea of optimal transport. We search out states with the
+largest weights, and choose a cluster around these points based on
+the closest states geographically. This method is not optimal since
+once most of the clusters have been selected the remaining states
+may be spread across the parameter space.
+
+\begin{table}[htb]
 \begin{mdframed}
@@
 sub-multinomials as the new samples. This resampler was designed with the aims of being
-more efficient than the ETPF, and more accurate than straight multinomial
+numerically cheaper than the ETPF, and more accurate than straight multinomial
 resampling. Therefore we now present numerical examples which
@@
 that these algorithms converge faster than the analogous naively parallelised
-Metropolis-Hastings algorithms. Further experimentation with the
-Random Walk Metropolis-Hastings (RWMH), Metropolis Adjusted Langevin
-Algorithm (MALA), preconditioned Crank-Nicolson (pCN) and Hamiltonian
-Monte Carlo (HMC) proposals has
-yielded similar results\cite{Paul}.
+Metropolis-Hastings algorithms. Further experimentation with the Metropolis
+Adjusted Langevin Algorithm (MALA), preconditioned Crank-Nicolson (pCN),
+preconditioned Crank-Nicolson Langevin (pCNL) and Hamiltonian
+Monte Carlo (HMC) proposals has yielded similar results\cite{Paul}.
 

### 2016-06-24 PAIS.tex
Colin's suggestions

@@
 In many MCMC algorithms such as the Random Walk Metropolis-Hastings (RWMH) algorithm, the optimal scaling parameter can be found by searching for the parameter value which gives an optimal acceptance rate, e.g.
-for near Gaussian targets we have 23.4\% for RWMH and 57.4\% for MALA\cite{roberts2001optimal}.
+for near Gaussian targets the optimal rates are 23.4\% for RWMH and 57.4\% for MALA\cite{roberts2001optimal}.
 Unlike Metropolis-Hastings algorithms, the PAIS algorithm does not accept or reject proposed values, so we need another method of measuring the optimality of $\beta$.
@@
 In this example we compare the naively parallelised RWMH algorithm with its
-PAIS variant, the PAIS-RW algorithm. We asses the performance of the PAIS algorithm
+PAIS variant, the PAIS-RW algorithm. The PAIS algorithm is implemented using the
+ETPF to perform the resampling step. We asses the performance of the PAIS algorithm
 using the relative $L^2$ error defined in Equation~\ref{eqn:L2_error}, as well as the
@@
 parameters we choose 32 values of $\beta$ evenly spaced on a log scale
-in the interval $[10^{-5}, 2]$.
-We run the PAIS-RW and RWMH algorithms for one million iterations, each with an ensemble size of $M=50$.
-We took 32 repeats of both algorithms and then used the geometric means of the sample statistics to find the optimal parameters.
+in the interval $[10^{-5}, 2]$. We run the PAIS-RW and RWMH algorithms
+for one million iterations, each with an ensemble size of $M=50$. We took
+32 repeats of both algorithms and then used the geometric means of the
+sample statistics to find the optimal parameters.
 
@@
 \subfigure[PAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
-\caption{Finding optimal values of $\beta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}.}
+\caption{Finding optimal values of $\beta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
 \label{fig:P1 opt beta}
@@
 \subfigure[Relative error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/G2_moments"}}
-\caption{Error analysis for the (A)RWMH and (A)PAIS-RW algorithms. The setup is as in Section~\ref{sec:Implementation P1} (2, 3).}
+\caption{Error analysis for the (A)RWMH and (A)PAIS-RW algorithms. The setup is as in Section~\ref{sec:Implementation P1} (2, 3). Resampling is performed using the ETPF.}
 \label{fig:MH1 L2}
@@
 \subfigure[Absolute error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_moments"}}
-\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_1$. Set up described in Section~\ref{sec:Implementation P1}.}
+\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_1$. Set up described in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
 \label{fig:BM1_L2}
@@
 \subfigure[Comparison of the relative $L^2$ error of the algorithms optimised by $L^2$ minimisation with locally optimised PAIS-RW with a scout chain.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
-\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_2$. Set up described in Section~\ref{sec:Implementation P1}.}
+\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_2$. Set up described in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
 \label{fig:BM2_L2}
@@
 \caption{Convergence of the relative $L^2$ error for problem $B_2$, comparing the globally optimised nonadaptive algorithms with the locally optimised adaptive algorithms.
-The setup is as described in Section~\ref{sec:Implementation P1} (3).}
+The setup is as described in Section~\ref{sec:Implementation P1} (3).
+Resampling is performed using the ETPF.}
 \label{fig:BM2_AL2}
@@
 \caption{Convergence of the PAIS-Gamma and PAIS-LA algorithms for the chemical system problem described in Section~\ref{sec:chem}.
-Implementation described in Sections~\ref{sec:chem_implementation} and \ref{sec:chem_conv}.}
+Implementation described in Sections~\ref{sec:chem_implementation} and \ref{sec:chem_conv}.
+Resampling is performed using the AMR scheme.}
 \label{fig:C1_ESS}

### 2016-06-28 PAIS.tex
Final minor corrections
Some minor corrections and typos. References cleaned up.

@@
   strategy, approximate multinomial resampling (AMR), which while not
-  as accurate as other schemes is substantially costly for large
+  as accurate as other schemes is substantially less costly for large
   ensemble sizes, which can then be used in conjunction with PAIS for
@@
 required before Monte Carlo error is reduced to acceptable levels,
-these types of problem remain frustratingly out of our grasp.
+the application of MCMC methods to these types of problem remain frustratingly out of our grasp.
 
@@
 complex methods that propose moves more intelligently, leading to
-rapidly converging methods. Function space versions of standard methods
+rapidly converging approximations. Function space versions of standard methods
 such as the random walk Metropolis-Hastings (RWMH) algorithm or the
@@
 development of computing hardware. Moore's law, which predicted that
-the number of transistors that can to fit on a single microchip will
+the number of transistors that can fit onto a single microchip will
 double every two years, has been largely followed since the early
 70s\cite{moore1998cramming}. In recent times, it has become necessary
-to use parallel architectures to exploit Moore's law. The efficient
-exploitation of these facilities is the key to solving many of the
+to use parallel architectures in order for this trend to continue. The efficient
+exploitation of these architectures is the key to solving many of the
 computational challenges that we currently face.
@@
 In this paper, we focus on the use of MCMC methods for characterising
-posterior probability distributions in Bayesian inverse problems. We
+posterior probability distributions arising from Bayesian inverse problems. We
 wish to learn about a particular unknown quantity $u$, of which we are
@@
 One advantage of the particle filter is that there are convergence
-results for this method as the number of particles is increased. One
+results for this method as the number of particles is increased.  One downside is that the required ensemble
+size increases quickly with dimension, making it difficult to use in
+high-dimensional problems. Another
 downside is that the effective sample size decreases at each
@@
 this is to employ a resampling scheme. The aim of a successful
-resample is to take your unevenly weighted ensemble and return a new
+resample is to take the unevenly weighted ensemble and return a new
 ensemble of particles with even weights which is highly correlated to
-the original samples. Another downside is that the required ensemble
-size increases quickly with dimension, making it difficult to use in
-high-dimensional problems.
+the original samples.
 
@@
 this is not the case. The aim of the PAIS is to use an ensemble of states,
-each coming from a particular time point in an MCMC chain, to construct a proposal
+each coming from a MCMC chain, to construct a proposal
 distribution which will be as close as possible to the target density. If this
@@
 using more complex (and hence more expensive) kernels $\nu$, does not
-significantly improve the speed of convergence of the algorithm.
+significantly improve the speed of convergence of the algorithm for
+simple examples\cite{Paul}.
 
@@
 convergence to the target density. One aspect worthy of  consideration
-with the PAIS, is finding an appropriate $\nu$ such that
+with the PAIS, is finding an appropriate proposal kernel $\nu$ such
+that the mixture distribution
 $\chi$ is a close approximation to the posterior density $\pi$. If the
@@
 previous states, and the algorithm will take a long time to fully explore
-the parameter space. It is therefore necessary to find a proposal
+the parameter space, and worse, will lead regularly to states proposed
+in the tails of the proposal distribution with very large weights. It is therefore necessary to find a proposal
 distribution which is slightly over-dispersed to ensure the entire
@@
 significant figures. It is therefore important that MCMC algorithms
-provide a feasible adaptive strategy to bypass this exploratory stage.
+provide a feasible adaptive strategy.
 
@@
 given by:
-\[
+\begin{equation}\label{eq:34567}
 	\left|\frac{N^{-1}\sum_{i=1}^N \! x_i^m - \mathbb{E}[X^m]}{\mathbb{E}[X^m]}\right|,
-\]
+\end{equation}
 where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative $L^2$
@@
 Therefore it is preferable as a means of tuning the scaling
-parameter. In the following numerics, we use the effective sample size
+parameter. In all of the numerics which follow, we use the effective sample size
 calculated at each iteration and averaged across the entire run to
@@
 \begin{equation}y = x_{i-1}+\beta \omega_{i-1}, \qquad \omega_{i-1}
-  \sim \mathcal{N}(0,1) \qquad {\rm i.i.d.}
+  \sim \mathcal{N}(0,\Sigma) \qquad {\rm i.i.d.}
 \end{equation}
-where $x_{i-1}$ is our current state and $y$ is our proposed state. In the numerics section we will refer to $\beta$ as the scaling parameter.
+where $x_{i-1}$ is our current state, $y$ is our proposed state and
+$\Sigma$ is a covariance operator, which could come from the prior distribution. In the numerics section we will refer to $\beta$ as the scaling parameter.
 
@@
 resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
-three moments of the samples changes with $M$ for the three different
+three moments of the samples changes with ensemble size $M$ for the three different
 samplers. As expected, the AMR lies somewhere between the high
@@
 PAIS variant, the PAIS-RW algorithm. The PAIS algorithm is implemented using the
-ETPF to perform the resampling step. We asses the performance of the PAIS algorithm
-using the relative $L^2$ error defined in Equation~\ref{eqn:L2_error}, as well as the
-relative error in the first moment.
+ETPF to perform the resampling step. We assess the performance of the PAIS algorithm
+using the relative $L^2$ error defined in \eqref{eqn:L2_error}, as well as the
+relative error in the first moment \eqref{eq:34567}.
 
@@
 Consider the simple case of a linear observation operator $\G(x) = x$, where the prior on $x$ and the observational noise follow Gaussian distributions.
-Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
+Then, following \eqref{eqn:like}, the Gaussian posterior has the form
 \begin{equation}\label{eqn:Gaussian posterior}
@@
 The simulations are repeated 32 times using the optimal parameters found in (1).
-The performance of the algorithms is judged by the convergence of the relative $L^2$ error statistic in Equation~\ref{eqn:L2_error}.
+The performance of the algorithms is judged by the convergence of the relative $L^2$ error statistic in \eqref{eqn:L2_error}.
 
@@
 
-Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights.
-The relative $L^2$ error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative $L^2$ error.
-Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter.
-In general this indicator underestimates the value found by using the entire sample.
+Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio
+compared to the error analysis and the variance of the weights. The
+relative $L^2$ error graph is noisy, but it is clear that the maximum
+in the effective sample size and the minimum in the variance of the
+weights are both close to the minimum in the relative $L^2$ error. Due
+to this we say that the estimate of the effective sample size found by
+averaging the statistic over each iteration is a good indicator for
+the optimal scaling parameter. In general this indicator overestimates
+the value of $\neff$ found by using the entire sample.
 
@@
 
-Both adaptive algorithms are run with initial values of $\beta=1$.
-Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance.
-The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth at the optimal value making it difficult to minimise.
+Both adaptive algorithms are run with initial values of
+$\beta=1$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
+period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
+by the end of the simulation window is matching its performance. The
+ARWMH algorithm does not perform quite as well, this is possibly due
+to the fact that the acceptance rate cost function is not particularly
+smooth at the optimal value making it difficult to minimise.
 
@@
 
+\subsubsection{Target Distribution} \label{sec:tar}
 We look at an `easy' problem, $B_1$, which has a KL divergence of
@@
 for a MCMC chain to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.
-
-\subsubsection{Target Distribution} \label{sec:tar}
 
@@
 from a posterior such as the one in $B_2$ is that it is unlikely that
-the correct ratio of chains will occur in each of the modes, and since there is no interaction between the chains, there is no way to remedy this problem.
+the correct ratio of chains will be maintained in each of the modes, and since there is no interaction between the chains, there is no way to remedy this problem.
 The PAIS algorithm tackles this problem with its resampling step.
@@
 frequently (the slow reactions). The quasi-steady-state assumption
-(QSSA), which assumes that the fast reactions converge in distribution
+(QSSA) is the assumption that the fast reactions converge in distribution
 on a timescale which is negligible with respect to the rate of
@@
 some rate constant $k_i$, and where the rates of the reactions are
-assumed to follow mass action kinetics. We assume that we are in a
+assumed to follow mass action kinetics. We denote the concentration of
+species $S_i$ by $X_i$. We assume that we are in a
 parameter regime such that the reactions $S_1\rightarrow S_2$ and $S_2\rightarrow S_1$ occur
@@
 chemical species are involved in fast reactions. However, the quantity
-$\mathcal{S} = S_1 + S_2$ is conserved by both of the fast reactions,
+$\mathcal{S} = X_1 + X_2$ is conserved by both of the fast reactions,
 and as such, this is the slowly changing quantity in this system.
@@
 \[
-	\hat{k}_4(s) = \mathbb{E}\left[k_4S_2|\mathcal{S}=s\right] = \frac{k_2k_4\mathcal{S}}{k_2+k_3}.
+	\hat{k}_4(s) = \mathbb{E}\left[k_4X_2|\mathcal{S}=s\right] = \frac{k_2k_4\mathcal{S}}{k_2+k_3}.
 \]
-The value of $\mathbb{E}\left[k_4S_2|\mathcal{S}=s\right]$ is
+The value of $\mathbb{E}\left[k_4X_2|\mathcal{S}=s\right]$ is
 approximated by finding the steady state of the ODE representing the fast subsystem of
 reactions:
-\[S_1 \xleftrightarrows[k_3]{k_2} S_2, \qquad S_1 + S_2 = s\]
+\[S_1 \xleftrightarrows[k_3]{k_2} S_2, \qquad X_1 + X_2 = s\]
 
 If we assume that we know the rate constants $k_1$ and $k_4$, and
-observe the system in Equation~\ref{eqn:QSSA_system}, we indirectly
+observe the system in \eqref{eqn:QSSA_system}, we indirectly
 observe the rates $k_2$ and $k_3$ through the effective rate
@@
 We look for a distribution over the parameter $\mathbf{k} = (k_2, k_3)^T$, given that we know $k_1 = 100$ and $k_4=1$.
-We generate our data by making ten observations of the system at $t_i = 2, 4, \dots, 20$, here simulated by solving the full system, Equation~\ref{eqn:full_system}, governed by the differential equations
+We generate our data by making ten observations of the system at $t_i = 2, 4, \dots, 20$, here simulated by solving the full system, \eqref{eqn:full_system}, governed by the differential equations
 \begin{align*}
@@
 
-In our modelling we use the QSSA to simplify this system of differential equations into the one dimensional system based on Equation~\ref{eqn:QSSA_system},
+In our modelling we use the QSSA to simplify this system of differential equations into the one dimensional system based on \eqref{eqn:QSSA_system},
 \[
@@
 \]
-
+These are parameters required to give us a distribution with mean
+$\mathcal{G}(\mathbf{k}$ and variance $\sigma^2$.
 We assign Gamma priors to $\mathbf{k}$ with mean $\alpha_0/\beta_0 =

### 2016-06-28 PAIS.tex
Final changes + final figure
Final couple of typos and added the posterior distribution figure for
the chemical system section.

@@
 % solution to QSSA ode and Gamma noise assumption
-\subsubsection{Target Distribution}
+\subsubsection{Target Distribution}\label{sec:345}
 
@@
 \begin{align*}
-	\frac{\text{d}S_1}{\text{d}t} &= k_1 - k_2S_1(t)+k_3S_2(t), \\
-	\frac{\text{d}S_2}{\text{d}t} &= k_2S_1(t) - (k_3+k_4)S_2(t),
+	\frac{\text{d}S_1}{\text{d}t} &= k_1 - k_2X_1(t)+k_3X_2(t), \\
+	\frac{\text{d}S_2}{\text{d}t} &= k_2X_1(t) - (k_3+k_4)X_2(t),
 \end{align*}
-with initial conditions $S_1(0)=S_2(0) = 0$, and parameter values $\mathbf{k} = (50, 100)^T$.
-We then add noise taken from a Gamma distribution with variance $\sigma^2 = 225$ and centred at each $\mathcal{S}(t_i)=S_1(t_i)+S_2(t_i)$.
+with initial conditions $X_1(0)=X_2(0) = 0$, and parameter values $\mathbf{k} = (50, 100)^T$.
+We then add noise taken from a Gamma distribution with variance $\sigma^2 = 225$ and centred at each $\mathcal{S}(t_i)=X_1(t_i)+X_2(t_i)$.
 
@@
 \]
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.65\textwidth]{"figures/C1_posterior"}
+\caption{The posterior distribution on the parameters $k_2$ and $k_3$
+  given the data and priors described in Section \ref{sec:345}.}
+\label{fig:345}
+\end{figure}
+
+Figure \ref{fig:345} presents a visualisation of the posterior
+distribution for this problem, found by exhaustive MH simulation.
 

### 2016-06-28 PAIS.tex
Added EPSRC acknowledgement
Crucial!
Coverletter and spellings

@@
     College, London, UK} \and Simon Cotter\thanks{School of
-    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
+    Mathematics, University of Manchester, Manchester, UK. e:
+    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
+    grant award EP/L023393/1} \and Paul Russell\thanks{School of
     Mathematics, University of Manchester, Manchester, UK}}
@@
   family of numerical methods for sampling from complex probability
-  distributions. As applications of these methods increase in size and complexity, the need for efficient
-  methods which can exploit the parallel architectures which are
-  prevalent in high performance computing increases. In this paper, we
-  aim to develop a framework for scalable parallel MCMC algorithms. At
-  each iteration, an importance sampling proposal distribution is
-  formed using the current states of all of the chains within an
-  ensemble. Once weighted samples have been produced from this, a
-  state-of-the-art resampling method is then used to create an evenly
-  weighted sample ready for the next iteration. We demonstrate that
-  this parallel adaptive importance sampling (PAIS) method outperforms
-  naive parallelisation of serial MCMC methods using the same number
-  of ensemble members, for low dimensional problems, and in fact shows
+  distributions. As applications of these methods increase in size and
+  complexity, the need for efficient methods which can exploit the
+  parallel architectures which are prevalent in high performance
+  computing increases. In this paper, we aim to develop a framework
+  for scalable parallel MCMC algorithms. At each iteration, an
+  importance sampling proposal distribution is formed using the
+  current states of all of the chains within an ensemble. Once
+  weighted samples have been produced from this, a state-of-the-art
+  resampling method is then used to create an evenly weighted sample
+  ready for the next iteration. We demonstrate that this parallel
+  adaptive importance sampling (PAIS) method outperforms naive
+  parallelisation of serial MCMC methods using the same number of
+  ensemble members, for low dimensional problems, and in fact shows
   better than linear improvements in convergence rates with respect to
@@
 Throughout this example, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales when we increase the ensemble size, and if there is some limit below which the algorithm fails.
-We implement the problem in Section~\ref{sec:problem 1}, using the RWMH and PAIS-RW algorithms with ensemble sizes in the inteveral $M \in [1, 160]$.
+We implement the problem in Section~\ref{sec:problem 1}, using the RWMH and PAIS-RW algorithms with ensemble sizes in the interval $M \in [1, 160]$.
 
@@
 Since we have parallel chains, we can run PAIS-RW with the majority of chains using the locally optimal value of $\beta$, and one or two chains with a larger scaling parameter.
-These chains with larger scaling parameters act both as `scouts' for new modes, and act to aid in the overdispersal of the proposal distribution.
+These chains with larger scaling parameters act both as `scouts' for new modes, and act to aid in the over-dispersal of the proposal distribution.
 Figure~\ref{fig:BM2_L2} (b) shows the results of using 49 chains with the local optimal scaling parameter, and one chain with ten times the local optimal scaling parameter.
@@
 approximate the dynamics of the slowly changing quantities in the
-system by assuming that the fast quantities are in equillibrium with
+system by assuming that the fast quantities are in equilibrium with
 respect to the fast reactions in isolation. This kind of model
@@
 Similarly the PAIS-LA algorithm uses the Langevin proposal distribution
-from the standard MALA algorithm with a pertubation taken from a Gamma
+from the standard MALA algorithm with a perturbationperturbation taken from a Gamma
 distribution,

### 2016-06-28 coverletter.tex
Coverletter and spellings

@@
 
-Please find attached a manuscript that we would like you to consider for publication in the SIAM Journal of  Scientific Computing, entitled ``Parallel Adaptive Importance Sampling".
+Please find attached a manuscript that we would like you to consider for publication in the SIAM Journal on  Uncertainty Quantification, entitled ``Parallel Adaptive Importance Sampling".
 We look forward to hearing from you in due course.

### 2016-11-21 PAIS.tex
Added mixture model

@@
 
-\subsection{Sampling from Bimodal Distributions}
+\subsection{Sampling from Bimodal Distributions}\label{sec:bimodal}
 
@@
 \subsubsection{Target Distribution} \label{sec:tar}
+
 We look at an `easy' problem, $B_1$, which has a KL divergence of
@@
 
+\subsection{A Mixture Model}\label{sec:mixture}
+
+The technique of mixture modelling employs well known parametric families to construct an approximating nonparametric distribution which may have a complex structure.
+Most commonly, Gaussian kernels are used since underlying properties in the data can often be assumed to follow a Gaussian distribution.
+An example would be if a practitioner were to measure the heights of one hundred adults, but failed to record their gender.
+The data could be considered as one population with two sub populations, male and female.
+The problem then might be to find the average height of adult males from the data.
+In this case, since height is often considered to follow a Gaussian distribution, it makes sense to model the population as a mixture of two univariate Gaussian distributions.
+
+A well known problem in the Bayesian treatment of mixture modelling is that of identification, sometimes referred to as the label-switching phenomenon.
+The likelihood distribution for mixture models is invariant under permutations of the mixture labels.
+If a mixture has $n$ means and the point $(\mu_1, \dots, \mu_n)$ maximises the likelihood, then the likelihood will also be maximised by $(\mu_{\varphi(1)}, \dots, \mu_{\varphi(n)})$ for all permutations $\varphi(\cdot)$.
+This means that the number of modes in the posterior distribution is of order $\mathcal{O}(n!)$.
+As we have seen it can be hard for standard MCMC algorithms to obtain reliable inference for posterior distributions with a large number of modes, or even a small number of modes which are separated by a large distance.
+
+\subsubsection{Target Distribution}\label{sec:mixture_target}
+
+In particular we look at a Gaussian mixture where we assume that there are two subpopulations within the overall population.
+Since both subpopulations are assumed to follow a Gaussian distribution we have five parameters which we need to be estimated, two means $\mu_{1,2}$, two variances $\sigma^2_{1,2}$, and the probability, $p$, that an individual observation comes from one of the subpopulations.
+We have 100 data points, $D_i$, which we assume to be distributed according to
+\[
+	D_i \sim p\mathcal{N}(\mu_1, \sigma^2_1) + (1-p)\mathcal{N}(\mu_2, \sigma^2_2), \quad i = 1,\dots,100,
+\]
+where $p \in [0, 1]$, $\mu_{1,2} \in \mathbb{R}$ and $\sigma^2_{1,2} \in \mathbb{R}^+$. Due to the domains of these parameters, we assign the priors
+\[
+	p \sim \text{Beta}(1,1), \quad \mu_{1,2} \sim \mathcal{N}(0, 4) \quad \text{and} \quad \sigma^2_{1,2} \sim \text{Gamma}(\alpha=2, \beta=1).
+\]
+If we collect these parameters in the vector $\theta = (p, \mu_1, \sigma^2_1, \mu_2, \sigma^2_2)^\top$, the resulting posterior distribution is
+\begin{equation}\label{eq:mixture_posterior}
+	\pi(\theta|D) \propto \prod\limits_{i=1}^{100} (p\mathcal{N}(D_i;\mu_1, \sigma^2_1) + (1-p)\mathcal{N}(D_i;\mu_2, \sigma^2_2))\prod\limits_{i=1}^5 \pi_0^i(\theta_i),
+\end{equation}
+where $\pi_0^i(\cdot)$ is the prior density function corresponding to $\theta_i$.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=\textwidth]{"figures/PAIS_best_posterior"}
+\caption{The posterior distribution of $\theta$ as given in Equation~\eqref{eq:mixture_posterior}.
+The main diagonal contains the marginal distributions of each $\theta_i$, and the lower triangular contours represents the correlation between pairs of parameters.}
+\label{fig:mixture_posterior}
+\end{figure}
+
+Figure \ref{fig:mixture_posterior} presents a visualisation of the posterior distribution for this problem, created from a single PAIS simulation.
+
+\subsubsection{Implementation}\label{sec:mixture_implementation}
+
+We have no analytic form for the posterior distribution for this model, and MH cannot reliably sample from this type of bimodal distribution.
+We do however know that the probability density should be evenly divided between the two modes.
+This is due to the symmetric prior for $p$, and and the same priors being assigned to $\mu_1$ and $\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$.
+To decide which mode a sample belongs to we define a plane which bisects the posterior so that each point on this plane lies exactly halfway between the two true solutions to the inverse problem i.e.
+the value of $\theta$ used to generate the data, and also the $\theta$ obtained by a relabelling of the parameters.
+Now that we can assign a sample to a particular mode, we can calculate the density in each mode by summing the weights associated to all samples in that mode,
+\[
+	\bar{w}_k = \sum\limits_{i=1}^N w_i\text{I}(X_i \in \text{Mode $k$}), \quad k = 1, 2,
+\]
+and the relative error in the amount of density in each mode is then
+\[
+	w_\text{error} = 2\left|\frac{\bar{w}_1}{\bar{w}_1+\bar{w}_2} - \frac{1}{2}\right|.
+\]
+
+Since the probability $p$ is constrained to lie in the interval $[0,1]$, and the variances must be positive, it can be wasteful to use Gaussian proposal distributions.
+For this example we make proposals by matching the proposal distribution to the prior, scaling the variance and centring the proposal at the previous value.
+So
+\[
+	q_p \sim \text{Beta}(\delta^{-2}p, \delta^{-2}(1-p)), \quad q_{\mu_{1,2}} \sim \mathcal{N}(\mu_{1,2}, 4\delta^2) \quad \text{and} \quad q_{\sigma^2_{1,2}} \sim \text{Gamma}(\alpha^*, \beta^*),
+\]
+where $\alpha^* = \sigma^2_{1,2}\beta^*$, $\beta^* = \sigma^2_{1,2}/2\delta^2$ and $\delta$ is a scaling parameter to be tuned.
+This means that our proposal distributions will not be a mixture of multivariate Gaussians, but independent mixtures of univariate Beta, Gamma and Gaussian distributions.
+
+In the numerics which follow we have increased the ensemble size from $M=50$ to $M=500$ to compensate for the increase in dimension.
+We also use the AMR algorithm for the resampling step because of the reduced computational cost.
+The method otherwise remains the same as in previous examples.
+We perform test runs to find the optimal scaling parameters considering convergence to modes with equal density.
+We then calculate the convergence rates of the algorithms by producing 50 million samples from the posterior with each algorithm, and repeat the simulation 32 times.
+
+% numerics.
+\subsubsection{Convergence of MH vs PAIS}\label{sec:mixture_conv}
+
+\begin{table}[!htb]
+      \centering
+        \begin{tabular}{|l|r|r|}
+	\hline
+	Algorithm	& MH & PAIS \\ \hline
+	$\delta^*$	& 1.3e-1     & 2.3e-1 \\
+	\hline
+	\end{tabular}
+	\vspace{1mm}
+	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithm is optimised using the effective sample size.}
+	\label{table:mixture_opt_beta}
+\end{table}
+
+The optimal scaling parameters for the MH and PAIS algorithms with the proposal distributions described in Section~\ref{sec:mixture_implementation} are given in Table~\ref{table:mixture_opt_beta}.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.8\textwidth]{"figures/Mode_proportions"}
+\caption{Convergence of the PAIS algorithm for the mixture model described in Section~\ref{sec:mixture}.
+Implementation described in Sections~\ref{sec:mixture_implementation} and \ref{sec:mixture_conv}.
+Resampling is performed using the AMR scheme.}
+\label{fig:mixture_modes}
+\end{figure}
+
+Convergence of the relative error for the two algorithms is displayed in Figure~\ref{fig:mixture_modes}.
+PAIS converges at the expected $\mathcal{O}(1/\sqrt{N})$ rate, whereas the MH algorithm converges to locally smooth histograms but with the wrong proportion of samples in each mode.
+The relatively low value of the error for the MH example is due to the priors covering the sample space evenly, however since transitions are near impossible with a small value of the scaling parameter, this error will take a very long time to reduce.
+This problem was discussed in Section~\ref{sec:bimodal}.
+
 \section{Discussion and Conclusions}\label{Sec:Conc}

### 2016-11-23 PAIS.tex
some small edits to the mixture section

@@
 In particular we look at a Gaussian mixture where we assume that there are two subpopulations within the overall population.
-Since both subpopulations are assumed to follow a Gaussian distribution we have five parameters which we need to be estimated, two means $\mu_{1,2}$, two variances $\sigma^2_{1,2}$, and the probability, $p$, that an individual observation comes from one of the subpopulations.
+Since both subpopulations will be approximated by Gaussian distributions we have five parameters which we need to be estimated, two means $\mu_{1,2}$, two variances $\sigma^2_{1,2}$, and the probability, $p$, that an individual observation belongs to the first subpopulation.
 We have 100 data points, $D_i$, which we assume to be distributed according to
@@
 \]
-where $p \in [0, 1]$, $\mu_{1,2} \in \mathbb{R}$ and $\sigma^2_{1,2} \in \mathbb{R}^+$. Due to the domains of these parameters, we assign the priors
+where $p \in [0, 1]$, $\mu_{1,2} \in \mathbb{R}$ and $\sigma^2_{1,2} \in \mathbb{R}^+$. Due to the domains of these parameters and also some prior knowledge, we assign the priors
 \[
@@
 \includegraphics[width=\textwidth]{"figures/PAIS_best_posterior"}
-\caption{The posterior distribution of $\theta$ as given in Equation~\eqref{eq:mixture_posterior}.
+\caption{The posterior distribution of $\theta$ as given in Equation~\eqref{eq:mixture_posterior} as found from 10 million samples from the PAIS algorithm.
 The main diagonal contains the marginal distributions of each $\theta_i$, and the lower triangular contours represents the correlation between pairs of parameters.}
@@
 We perform test runs to find the optimal scaling parameters considering convergence to modes with equal density.
-We then calculate the convergence rates of the algorithms by producing 50 million samples from the posterior with each algorithm, and repeat the simulation 32 times.
+We then calculate the convergence rates of the algorithms by producing 10 million samples from the posterior with each algorithm, and repeat the simulation 32 times.
 

### 2016-12-04 PAIS.tex
Added section on consistency

@@
 
+\section{Consistency of PAIS}
+
+As outlined in~\cite{martino2015adaptive} consistency of population AIS algorithms can be considered in two different cases.
+In the first case we fix the number of iterations $N < \infty$, but allow the population size to grow to inifinity $M\rightarrow\infty$.
+In the second case we hold the population size fixed $M<\infty$, and allow infinitely many iterations $N\rightarrow\infty$.
+
+In case one, from standard importance sampling results we know that for an iteration $n$, as $M\rightarrow\infty$, we obtain a consistent estimator for any statistic of interest,
+\[
+	\hat{\phi}(\Theta) \approx \frac{1}{\hat{Z}}\sum\limits_{i=1}^M \! \frac{1}{M}w_i\phi(\theta_i) \rightarrow \phi(\Theta),
+\]
+where the normalisation constant, $\hat{Z} = \frac{1}{M}\sum_{i=1}^M \! w_i$, also converges to the true normalisation constant $Z$~\cite{robert2013monte}.
+
+Case two is slightly more involved.
+Estimation of the normalisation constant $Z$ is biased, and so estimates of statistics are sums of independent but biased estimators.
+Since the estimators are independent, the proof of consistency of PAIS in this second limit can be approached in the same way as the PMC algorithm, where it has been shown that $\hat{Z}\rightarrow Z$~\cite{robert2013monte}.
+Since the normalisation constant is consistent, sums of the independent estimators are also consistent.
+
 \section{Numerical Examples}\label{Sec:Num}

### 2016-12-04 PAIS.tex
Edited adaptive PAIS and algorithm optimisation sections

@@
 \usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
+\usepackage[ruled,vlined,linesnumbered,algosection]{algorithm2e}
+
 \setlength{\parindent}{0in}
@@
 \STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
-\FOR{$i=0,1,2, ..., N$}
+\FOR{$i=0,1,2, ..., N-1$}
 \STATE $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^T, \quad y_j^{(i)} \sim
@@
 critical to achieving optimal mixing rates and hence achieving fast
-convergence to the target density. One aspect worthy of  consideration
+convergence to the target density. It is well known that a scaling
+parameter which is either too large or too small results in a Markov
+chain with high autocorrelation. One aspect worthy of consideration
 with the PAIS, is finding an appropriate proposal kernel $\nu$ such
-that the mixture distribution
-$\chi$ is a close approximation to the posterior density $\pi$. If the
-proposal distribution is too over-dispersed, then the algorithm will often
-propose states in the tails of the distribution, resulting in larger
-variance of the weights, and therefore slower convergence to the
-posterior distribution. Similarly, if the proposal distribution is
-under-dispersed, the proposals will be highly correlated with the
-previous states, and the algorithm will take a long time to fully explore
-the parameter space, and worse, will lead regularly to states proposed
-in the tails of the proposal distribution with very large weights. It is therefore necessary to find a proposal
-distribution which is slightly over-dispersed to ensure the entire
-posterior is explored\cite{gelman1992inference}, but is as close to the
-posterior as possible.
+that the mixture distribution $\chi$ is a close approximation to the
+posterior density $\pi$.
+
+%If the
+%proposal distribution is too over-dispersed, then the algorithm will often
+%propose states in the tails of the distribution, resulting in larger
+%variance of the weights, and therefore slower convergence to the
+%posterior distribution. Similarly, if the proposal distribution is
+%under-dispersed, the proposals will be highly correlated with the
+%previous states, and the algorithm will take a long time to fully explore
+%the parameter space, and worse, will lead regularly to states proposed
+%in the tails of the proposal distribution with very large weights. It is therefore necessary to find a proposal
+%distribution which is slightly over-dispersed to ensure the entire
+%posterior is explored\cite{gelman1992inference}, but is as close to the
+%posterior as possible.
 
@@
 allows the user to control their variance. For example, in the RWMH
-proposal $y = x + \beta \eta$, the parameter $\beta$ controls how
-correlated the proposal state $y$ is to the current state
-$x$. Therefore the proposal distributions can be tuned such that they are are slightly over-dispersed, as described
-above. This tuning can take place during the burn-in phase of the
-algorithm. Algorithms which use this method to find optimal proposal
-distributions are known as adaptive MCMC algorithms, and have been
-shown to be convergent provided that they satisfy certain conditions\cite{roberts2007coupling,roberts2009examples}.
+proposal $y = x + \beta \eta$, the parameter $\beta$ is the variance
+of the proposal distribution. Therefore the proposal distributions can
+be tuned such that they are are slightly over-dispersed. This tuning
+can take place during the burn-in phase of the algorithm. Algorithms
+which use this method to find optimal proposal distributions are known
+as adaptive MCMC algorithms, and have been shown to be convergent
+provided that they satisfy certain
+conditions\cite{roberts2007coupling,roberts2009examples}.
+Alternatively, a costly trial and error scheme with short runs of the
+MCMC algorithm can be used to find an acceptable value of $\beta$.
 
@@
 variance of the individual kernels within the proposal mixture. This
-adaptivity during the burn-in has some added benefits over and above
+adaptivity during early iterations has some added benefits over and above
 finding an optimal parameter regime for the algorithm. If the initial
-value of the proposal variances is chosen to be very large, then
-proposed moves will be made far and wide, expediting the early
-mode-finding stages of the algorithm. Adaptively reducing the proposal variances to an
-optimal value then allows us to explore each region efficiently. The
-fact that we have an ensemble of chains allows us to assess quickly and
-effectively what the optimal variance of the proposal distributions
-should be.
-
-The alternative to using adaptive procedures to tune the scaling
-parameters is to perform exploratory simulations to find the optimal
-regimes by trial and error. This can be very costly and the optimal
-parameters cannot realistically be found to more than a couple of
-significant figures. It is therefore important that MCMC algorithms
-provide a feasible adaptive strategy.
-
-In many MCMC algorithms such as the Random Walk Metropolis-Hastings (RWMH) algorithm, the optimal scaling parameter can be found by searching for the parameter value which gives an optimal acceptance rate, e.g.
-for near Gaussian targets the optimal rates are 23.4\% for RWMH and 57.4\% for MALA\cite{roberts2001optimal}.
-Unlike Metropolis-Hastings algorithms, the PAIS algorithm does not accept or reject proposed values, so we need another method of measuring the optimality of $\beta$.
-Section~\ref{sec:statistics} gives some possible methods for tuning $\beta$.
+value of the proposal variance is chosen to be very large, then the early
+mode-finding stages of the algorithm are expedited. Adaptively
+reducing the proposal variances to an optimal value then allows us to
+explore each region efficiently. Using an ensemble of chains allows
+quick and effective assessment of the value of the optimal scaling
+parameter.
+
+%The alternative to using adaptive procedures to tune the scaling
+%parameters is to perform exploratory simulations to find the optimal
+%regimes by trial and error. This can be very costly and the optimal
+%parameters cannot realistically be found to more than a couple of
+%significant figures. It is therefore important that MCMC algorithms
+%provide a feasible adaptive strategy.
+
+In many MCMC algorithms such as the Random Walk Metropolis-Hastings
+(RWMH) algorithm, the optimal scaling parameter can be found by
+searching for the parameter value which gives an optimal acceptance
+rate, e.g. for near Gaussian targets the optimal rates are 23.4\% for
+RWMH and 57.4\% for MALA\cite{roberts2001optimal}. This method
+is not applicable to PAIS so we must use other statistics to optimise
+the scaling parameter. Section~\ref{sec:statistics} gives some
+possible methods for tuning $\beta$.
 
@@
 
-MCMC algorithms can be assessed by comparing their approximation of
-the posterior to the analytic distribution, in cases where the
-posterior distribution can be computed using alternative methods. To
-assess this, a distance metric on distributions must be chosen. Examples are the
-relative error between the sample moments and the posterior's moments, or the
-relative $L^2$ error between the true density, $\pi(x|D)$, and the
-constructed histogram. The relative error in the $m$-th moment is
-given by:
+%Sampling algorithms can be assessed by comparing their approximation of
+%the posterior to the analytic distribution, in cases where the
+%posterior distribution can be computed using alternative methods.
+
+When available, an analytic form for the target distribution allows us
+to assess the convergence of sampling algorithms to the target
+distribution. Common metrics for this task include the relative error
+between the sample moments and the target's moments, or the
+relative $L^2$ error between the sample histogram and the target
+density, $\pi(x|D)$. The relative error in the $n$-th moment,
+$\hat{m}_n$, is given by:
 \begin{equation}\label{eq:34567}
-	\left|\frac{N^{-1}\sum_{i=1}^N \! x_i^m - \mathbb{E}[X^m]}{\mathbb{E}[X^m]}\right|,
+	e = \left|\frac{\hat{m}_n - \mathbb{E}[X^n]}{\mathbb{E}[X^n]}\right|, \quad \text{where} \quad \hat{m}_n = \frac{1}{N}\sum_{i=1}^N \! x_i^n,
 \end{equation}
-where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative $L^2$
-error between a continuous function to a piecewise constant function,
-$e$, can be given by considering the difference in mass between the
-normalised histogram of the samples and the posterior distribution
-over a set of disjoint sets or ``bins'':
+where $\{x_i\}_{i=1}^N$ is a sample of size $N$.
+
+The relative $L^2$ error, $E$, between a continuous function to a
+piecewise constant function, can be given by considering the
+difference in mass between the self-normalised histogram of the
+samples and the posterior distribution over a set of disjoint sets or
+``bins'':
 \begin{equation}\label{eqn:L2_error}
-	e^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a\right]^2,
+	E^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(s|D) \, \mbox{d}s - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(s|D) \, \mbox{d}s\right]^2,
 \end{equation}
@@
 sufficiently large sample. However they can be used to assess the
-ability of other indicators to find the optimal proposal
-variances in a controlled setting. The following statistics can be used specifically for importance sampling algorithms.
-
-\subsubsection{The variance of the weights}
-
-Importance samplers assign a weight to each sample they produce based
-on a ratio of the posterior to the proposal at that point. Importance
-samplers are most efficient when the target is proportional to the
-proposal distribution.
+ability of other indicators to find the optimal scaling parameters in
+a controlled setting.
+
+%\subsubsection{The variance of the weights}
+
+%Importance samplers assign a weight to each sample they produce based
+%on a ratio of the posterior to the proposal at that point. Importance
+%samplers are most efficient when the target is proportional to the
+%proposal distribution.
 In this case the weights are all equal, and so the variance of the weights, $var(w(y))$, is zero.
-Hence, we would like to choose the value of $\beta$ which minimises the variance of the weights,
-\[
-	\beta^*_{\text{var}} = \argmin_{\beta} \text{var}(w(y)).
-\]
-In our experience, the mean of the estimator
-$\text{var}(w(y))$ is a smooth enough function of $\beta$ that it can
-be used to tune the proposal variance during the burn-in phase of MCMC
-algorithms. However the variance of the estimator of the variance can be large, especially far away from the optimal value, so it can take a large number of iterations to calculate descent directions.
+Hence, we would like to %choose the value of $\beta$ which minimises the variance of the weights,
+%\[
+%	\beta^*_{\text{var}} = \argmin_{\beta} \text{var}(w(y)).
+%\]
+%In our experience, the mean of the estimator
+%$\text{var}(w(y))$ is a smooth enough function of $\beta$ that it can
+%be used to tune the proposal variance during the burn-in phase of MCMC
+%algorithms.
+However the variance of the estimator of the variance can be large, especially far away from the optimal value, so it can take a %large number of iterations to calculate descent directions.
 
@@
 
-The effective sample size, $\neff$, can also be used to assess the efficiency of importance samplers.
-Ideally, in each iteration, we would like all $M$ of our samples to provide us with new information about the posterior distribution.
-In practise, we cannot achieve a perfect effective sample size of $M$.
-
-The effective sample size can be defined in the following way:
+The effective sample size, $\neff$, can be used to assess the
+efficiency of importance samplers. Ideally, in each iteration, we
+would like all $M$ of our samples to provide us with new information
+about the posterior distribution. In practise, we cannot achieve a
+perfect effective sample size of $M$.
+
+The effective sample size of a weighted sample is defined in the
+following way:
 \[
@@
 The second two expressions are true when $M\rightarrow\infty$. From
-the last expression we see that when the variance of the weights is
-zero, $\neff = M$; this is our ideal scenario. Maximising the effective sample size is equivalent to minimising the variance of the weights.
+the last expression, if the variance of the weights is zero then
+$\neff = M$; this is our ideal scenario. Maximising the effective
+sample size is equivalent to minimising the variance of the weights.
 
 The statistic $\neff$ is easier to deal with than the variance of the
-weights, as it varies between $1$ and $M$ as opposed to the variance
-which can vary over many orders of magnitude.
-Therefore it is preferable as a means of tuning the scaling
-parameter. In all of the numerics which follow, we use the effective sample size
-calculated at each iteration and averaged across the entire run to
-tune the scaling parameters. We do this because when we tune this
-parameter on the fly, we use the single-iteration average to estimate
-optimality. The optimal scaling parameter found using the global
-optimum of the variance of the weights is also included for
-comparison. Note that for the majority of iterations the value of
-$\neff$ is an overestimate of this statistic over a larger number of
-samples. Rare events which result in a large importance weight bring
-this statistic down, and care must be made not to overfit the proposal
-variance to $\neff$ on an iteration by iteration basis. This can be
-accounted for by increasing the variance slightly once the adaptive
-algorithm has arrived on a value using single iteration values of $\neff$.
+weights, as it takes values on $[1, M]$ while the variance of the
+weights takes values on $\mathbb{R}_\geq$. Since the variance can vary
+over many orders of magnitude bringing numerical issues, the effective
+sample size is more desirable.
+
+%In the numerics which follow, we use the effective sample size
+%calculated at each iteration and averaged across the entire run to
+%tune the scaling parameters. We do this because when we tune this
+%parameter on the fly, we use the single-iteration average to estimate
+%optimality. The optimal scaling parameter found using the global
+%optimum of the variance of the weights is also included for
+%comparison. Note that for the majority of iterations the value of
+%$\neff$ is an overestimate of this statistic over a larger number of
+%samples. Rare events which result in a large importance weight bring
+%this statistic down, and care must be made not to overfit the proposal
+%variance to $\neff$ on an iteration by iteration basis. This can be
+%accounted for by increasing the variance slightly once the adaptive
+%algorithm has arrived on a value using single iteration values of $\neff$.
+
+In this paper we tune our simulations using the effective sample size
+statistic. We calculate this statistic using a sample size of $Mn_k$,
+where $1 \leq n_k\leq N$ is sufficiently large enough to obtain a
+reasonable estimate of $\neff$ with scaling parameter $\delta_k$.
+Calculating $\neff$ over these subsets of the simulations tends to
+underestimate the optimal value of the scaling parameter due to the
+possibility of missing unlikely proposals with extreme weights.
 
 The effective sample size also has another useful property; if we
-imagine the algorithm in the burn-in phase, for example, we have $M$
-ensemble members in the tail of a Gaussian curve searching for the area of
-high density. If the ensemble members are evenly spaced, then the particle
-closest to the mean will have an exponentially higher weight assigned
-to it. The effective sample size ratio in this scenario will be close to
-1. As the algorithm burns in, the ensemble populates the regions where
-the majority of the probability density lies, and the proposal
-distributions better represent the posterior distribution. This leads
-to smaller variance in the weights, and a bigger effective sample size.
-By this argument we can see that rising $\neff$ signals the end of the burn-in period.
+consider the algorithm in the early stages, for example, we have all
+$M$ particles in the tails of the target searching for a mode. The
+particle closest to the mode will have an exponentially higher weight
+and the effective sample size will be close to 1. In later stages the
+ensemble populates high density regions, and better represents the
+posterior distribution. This leads to smaller variation in the
+weights, and a higher effective sample size. By looking for
+approximations of the effective sample size which look like a
+stationary distribution, we can tell when the algorithm is working
+efficiently.
 
@@
 \caption{The effective sample size ratio and relative $L^2$ difference
-between the proposal and posterior distributions at each of the first 30
-iterations. These numerics are taken from a simulation of the problem in
-Section~\ref{sec:chem} using the PAIS-Gamma algorithm.}
+$E$ during the first 30 iterations. These numerics are taken from the
+example in Section~\ref{sec:chem} using the PAIS-Gamma algorithm.}
 \label{fig:neff-burnin}
@@
 
-Figure~\ref{fig:neff-burnin} shows that the effective sample size flattens
-out at the same time as the relative $L^2$ difference between
-the posterior distribution and the proposal distribution stabilises close to
-its minimum. Detecting this stationarity allows us to automatically
-determine the end of the burn-in phase of the algorithm.
+Figure~\ref{fig:neff-burnin} demonstrates that the effective sample
+size flattens out as the relative $L^2$ difference between the
+posterior distribution and the proposal distribution stabilises close
+to its minimum.
 
@@
 \subfigure[The highest effective sample size achieved for each ensemble size.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM"}}
-\caption{The behaviour of the effective sample size as the ensemble size increases. This analysis is from the Gaussian posterior in Section~\ref{sec:problem 1} using the PAIS-RW algorithm.}
+\caption{The behaviour of the effective sample size as the ensemble size increases. This analysis is of the example in Section~\ref{sec:problem 1} using the PAIS-RW algorithm.}
 \label{fig:neff-M}
@@
 
-If we are to use the effective sample size ratio as an indicator of
-how to tune the proposal variance, we need to look at how it behaves
-in different situations. Here we look at how the statistic behaves as
-we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results
-for the Gaussian posterior discussed in Section~\ref{sec:problem 1}
-when using the PAIS algorithm with RW proposals, each with variance
-$\beta^2 \in \mathbb{R}_{>0}$. Figure~\ref{fig:neff-M} (a) shows that
-as the ensemble size increases, the scaling parameter which gives the
-optimal effective sample size ratio decreases. This is to be expected
-since if we are trying to approximate the posterior with a mixture
-distribution of a small number of Gaussians, the optimal variance will
-naturally be larger so that the whole of the significant regions are
-covered by the proposal distribution. As the number increases, the
-optimal variances decrease so that finer details in the posterior can
-be better represented in the proposal. Figure~\ref{fig:neff-M} (b) shows that as the ensemble size increases, the efficiency of the sampler also increases.
+Figure~\ref{fig:neff-M} shows how the effective sample size used in
+PAIS is affected by the ensemble size. We see from subfigure (a) that
+as the ensemble size increases, the optimal scaling parameter
+decreases. This is expected since the scaling parameter must be
+reduced if we are to fit more kernels into the same target. We also
+see that the algorithm becomes less sensitive to changes in the
+scaling parameter as the ensemble size increases. Subfigure (b) shows
+that as the ensemble size increases, the algorithm becomes more
+efficient.
+
+%If we are to use the effective sample size ratio as an indicator of
+%how to tune the proposal variance, we need to look at how it behaves
+%in different situations. Here we look at how the statistic behaves as
+%we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results
+%for the Gaussian posterior discussed in Section~\ref{sec:problem 1}
+%when using the PAIS algorithm with RW proposals, each with variance
+%$\beta^2 \in \mathbb{R}_{>0}$. Figure~\ref{fig:neff-M} (a) shows that
+%as the ensemble size increases, the scaling parameter which gives the
+%optimal effective sample size ratio decreases. This is to be expected
+%since if we are trying to approximate the posterior with a mixture
+%distribution of a small number of Gaussians, the optimal variance will
+%naturally be larger so that the whole of the significant regions are
+%covered by the proposal distribution. As the number increases, the
+%optimal variances decrease so that finer details in the posterior can
+%be better represented in the proposal. Figure~\ref{fig:neff-M} (b) shows that as the ensemble size increases, the efficiency of the sampler also increases.
 
@@
 
-A popular approach for adaptive MCMC algorithms is to view the scaling
-parameter as a random variable which we can sample during the course
-of the MCMC iterations. However, it can be slow to converge to the
-optimal value, and we may need an
-uninformative prior for the scaling parameter. Alternatively, the
-parameter may be randomly sampled at various points during the
-evolution of the chain. This results in some iterations which make
-larger global moves in the state space between modes in the target
-distribution, and some which make local moves. Algorithms of this type
-do not converge to an optimal value of the scaling parameter. We choose to use a divide and conquer scheme which optimises the
-effective sample size (or any other diagnostic). Some more
-sophisticated examples are described in \cite{roberts2009examples} and
-\cite{Ji2013adaptive}. 
-% Algorithm~\ref{tab:adapt} in Appendix~\ref{Sec:App}, is not presented as an optimal strategy but as a demonstration that the PAIS algorithm can be adaptively tuned effectively.
-
-From here on in, we will use the random walk proposal,
-\begin{equation}y = x_{i-1}+\beta \omega_{i-1}, \qquad \omega_{i-1}
-  \sim \mathcal{N}(0,\Sigma) \qquad {\rm i.i.d.}
-\end{equation}
-where $x_{i-1}$ is our current state, $y$ is our proposed state and
-$\Sigma$ is a covariance operator, which could come from the prior distribution. In the numerics section we will refer to $\beta$ as the scaling parameter.
-
- Using an adaptive strategy, we calculate a sequence
- $\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal
- scaling parameter $\beta^*$, resulting in the optimal transition
- density $\chi$ for our MCMC algorithm. This optimal value will differ
- depending on the criterion we are optimising. We must choose some sequence
- of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$, and due
- to the constraints on adaptive MCMC
- algorithms\cite{roberts2007coupling,roberts2009examples}, these $n_k$
- must grow exponentially further apart. This same adaptive approach
- can also be applied to the trivially parallelised MCMC algorithms to adaptively calculate their optimal scaling parameter $\beta^*$.
+%A popular approach for adaptive MCMC algorithms is to view the scaling
+%parameter as a random variable which we can sample during the course
+%of the MCMC iterations. However, it can be slow to converge to the
+%optimal value, and we may need an
+%uninformative prior for the scaling parameter. Alternatively, the
+%parameter may be randomly sampled at various points during the
+%evolution of the chain. This results in some iterations which make
+%larger global moves in the state space between modes in the target
+%distribution, and some which make local moves. Algorithms of this type
+%do not converge to an optimal value of the scaling parameter.
+
+To adapt the scaling parameter $\beta$, we use a version of the
+gradient ascent method modified for a stochastic function. Some more
+sophisticated examples are described in
+\cite{roberts2009examples,Ji2013adaptive,andrieu2006ergodicity}.
+%Algorithm~\ref{tab:adapt} in Appendix~\ref{Sec:App}, is not
+%presented as an optimal strategy but as a demonstration that the PAIS
+%algorithm can be adaptively tuned effectively.
+
+%From here on in, we use the random walk proposal unless otherwise stated,
+%\begin{equation}y = x_{i-1}+\beta \omega_{i-1}, \qquad \omega_{i-1}
+%  \sim \mathcal{N}(0,\Sigma) \qquad {\rm i.i.d.}
+%\end{equation}
+%where $x_{i-1}$ is our current state, $y$ is our proposed state and
+%$\Sigma$ is a covariance operator which might come from the prior
+%distribution. In the numerics section we will refer to $\beta$ as the
+%scaling parameter.
+
+%Using an adaptive strategy, we calculate a sequence
+%$\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal
+%scaling parameter $\beta^*$, resulting in the optimal transition
+%density $\chi$ for our MCMC algorithm. This optimal value will differ
+%depending on the criterion we are optimising. We must choose some sequence
+%of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$, and due
+%to the constraints on adaptive MCMC
+%algorithms\cite{roberts2007coupling,roberts2009examples}, these $n_k$
+%must grow exponentially further apart. This same adaptive approach
+%can also be applied to the trivially parallelised MCMC algorithms to adaptively calculate their optimal scaling parameter $\beta^*$.
+
+\begin{table}[!h]
+\centering
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+	Define update times $\{n_k\}_k$.\;
+	\For{$n=1,\dots,N$}{
+		Compute normal PAIS steps.\;
+		\If {$n \in \{n_k\}$}
+		{
+			Divide ensemble into two halves. Use these halves to estimate the gradient in $\neff$ at $\delta_k$.\;
+			Update $\delta_k$ using gradient ascent,
+				\[
+					\delta_{k+1} = \delta_k + \gamma\nabla\neff.
+				\]
+			\label{algline:gradient_ascent}
+		}
+	}
+\caption{The Adaptive PAIS Algorithm.\label{alg:adaptPAIS}}
+\end{algorithm}
+\end{table}
+Our adaptive algorithm is given in Algorithm~\ref{alg:adaptPAIS}.
+We choose update times which, as suggested in Section~\ref{sec:ess} allow for a reasonable estimate of the effective sample size, but do not waste too many iterations.
+In Step~\ref{algline:gradient_ascent}, the gradient ascent parameter $\gamma$ may decrease over time, e.g.
+as a function of $n_k$.
 

### 2016-12-04 PAIS.tex
changed algorithm style

@@
 \usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
-\usepackage[ruled,vlined,linesnumbered,algosection]{algorithm2e}
+\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
 
@@
 from this point we
-can iterate the process once again. The algorithm is summarised in Table \ref{tab:PAIS}.
+can iterate the process once again. The algorithm is summarised in Algorithm \ref{alg:PAIS}.
 
@@
 
-\begin{table}
-\begin{mdframed}
-\begin{algorithmic}
-\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
-\FOR{$i=0,1,2, ..., N-1$}
-\STATE $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^T, \quad y_j^{(i)} \sim
-\nu(\cdot;x_j^{(i)})$
-\STATE $\chi(y;\X^{(i)}) = \frac{1}{M}
-\sum_{j=1}^M \nu(y;x_j^{(i)})$.
-\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
-\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
-\STATE Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
-\ENDFOR 
-\end{algorithmic}
-\end{mdframed}
-
-\caption{A pseudo-code representation of the Parallel Adaptive
-  Importance Sampler (PAIS).}
-\label{tab:PAIS}
+%\begin{table}
+%\begin{mdframed}
+%\begin{algorithmic}
+%\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
+%\FOR{$i=0,1,2, ..., N-1$}
+%\STATE $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^T, \quad y_j^{(i)} \sim
+%\nu(\cdot;x_j^{(i)})$
+%\STATE $\chi(y;\X^{(i)}) = \frac{1}{M}
+%\sum_{j=1}^M \nu(y;x_j^{(i)})$.
+%\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
+%\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
+%\STATE Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
+%\ENDFOR 
+%\end{algorithmic}
+%\end{mdframed}
+%\caption{A pseudo-code representation of the Parallel Adaptive
+%  Importance Sampler (PAIS).}
+%\label{tab:PAIS}
+%\end{table}
+
+\begin{table}[!h]
+\centering
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+	Set $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^\top$.\;
+	\For {$i = 1, \dots, N$}
+	{
+		Sample $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^\top, \quad y_j^{(i)} \sim
+\nu(\cdot;x_j^{(i-1)})$.\;
+		Calculate $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^\top,$ \quad $w^{(i)}_j =
+\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i-1)})}$, where
+		\[
+			\chi(y;\X^{(i-1)}) = \frac{1}{M}\sum_{j=1}^M \nu(y;x_j^{(i-1)}).
+		\]
+
+		Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i)})$.
+	}
+\caption{The parallel adaptive importance sampler (PAIS).\label{alg:PAIS}}
+\end{algorithm}
 \end{table}
@@
 
-\begin{table}[htb]
-\begin{mdframed}
-\begin{algorithmic}
-\STATE $\b{z} = M\bar{\b{w}}$
-\FOR{$i=1,2,\ldots,M$}
-\STATE $J = \argmax_j z_j$
-\STATE $p_{i,J} = \min\{1,z_J\}$
-\STATE $z_J = z_J - p_{i,J}$
-\WHILE{$\sum_j p_{i,j} <1$}
-\STATE $K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$
-\STATE $p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$
-\STATE $z_K = z_K - p_{i,K}$
-\ENDWHILE
-\STATE $x_i = \sum_k p_{i,k}y_k$.
-\ENDFOR
-\end{algorithmic}
-\end{mdframed}
-\caption{The approximate multinomial resampler (AMR) algorithm.\label{tab:AMR}}
+\begin{table}[!h]
+\centering
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+	$\b{z} = M\bar{\b{w}}$.\;
+	\For {$i = 1,\dots, M$}
+	{
+		$J = \argmax_j z_j$.\;
+		$p_{i,J} = \min\{1,z_J\}$.\;
+		$z_J = z_J - p_{i,J}$.\;
+		\While {$\sum_j p_{i,j} <1$}
+		{
+			$K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$.\;
+			$p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$.\;
+			$z_K = z_K - p_{i,K}$.\;
+		}
+		$x_i = \sum_k p_{i,k}y_k$.\;
+	}
+\caption{The approximate multinomial resampler (AMR).\label{alg:AMR}}
+\end{algorithm}
 \end{table}
 
-Table \ref{tab:AMR} describes the basis of the algorithm with
+Algorithm~\ref{alg:AMR} describes the basis of the algorithm with
 deterministic resampling, using the means of each of the

### 2016-12-04 PAIS.tex
added section on AIS

@@
 
+\subsection{Adaptive Importance Sampling}
+
+Bayesian philosophy has become wide\-spread in statistical inference
+largely due to the development of efficient methods, such as MCMC and
+particle filtering, for calculating previously intractable integrals.
+A related class of algorithms, adaptive importance sampling (AIS)
+\cite{liu2008monte} reviewed in\cite{bugallo2015adaptive}, had
+received less attention until their practical applicability was
+demonstrated in the
+mid-2000s\cite{celeux2006iterated,cappepopulation,isard1998condensation,bink2008bayesian}.
+AIS methods produce a sequence of approximating distributions,
+constructed from standard distributions, from which samples can be
+easily drawn. At each iteration the samples are weighted, often using
+standard importance sampling methods. The weighted samples are used to
+train the adapting sequence of distributions so that samples are drawn
+more efficiently as the iterations progress. The weighted samples form
+a sample from the posterior distribution under some mild
+conditions~\cite{robert2013monte,martino2015adaptive}.
+
+The AIS algorithm is Markovian, conditioned on the method of adaption,
+but differs from MCMC methods in that the sequence does not have the
+posterior as its limiting distribution. Convergence is ensured by the
+weighting of samples, which does not rely on algorithm asymptotics.
+This means that after any iteration the algorithm can be stopped and
+the sample obtained is a sample from the posterior distribution, with
+no consideration for burn-in times~\cite{douc2007convergence}.
+
+A popular AIS algorithm is the population monte carlo (PMC) sampling
+method~\cite{cappe2012population}. This algorithm uses a population of
+samples to build a mixture or kernel density estimate (KDE) of the
+posterior distribution. This KDE is optimal in the sense that it
+minimises the Kullback-Leibler divergence between the mixture and the
+posterior distribution~\cite{douc2007convergence}. The efficiency of
+this optimisation is restricted by the component kernel(s) chosen, and
+the quality of the current sample from the posterior.
+
+The PMC algorithm is often implemented as a $D$-kernel
+method~\cite{douc2007convergence}. In this method one choses $D$
+different measures, and a number of samples $N$ such that the total
+population size $N = MD$. At each iteration and for each mixture
+component, one of these $D$ kernels is selected via some optimisation
+criterion and $M$ samples are produced from this kernel. The $M$
+samples from each kernel attempt to stabilise the resampling scheme so
+that information about a particular region is not lost.
+
+In practice, the PMC algorithm can suffer from spiky posteriors due to
+poor approximation properties in some parts of the posterior. The
+adaptive multiple importance sampling algorithm
+(AMIS)~\cite{cornuet2012adaptive} aims to reduce the impact of these
+poor approximations. AMIS considers the entire sample at time $n$ to
+be a draw from all approximations up to time $n$. This means that the
+importance weights for the whole sample must be recalculated at each
+iteration. These weights are known as deterministic mixture
+weights~\cite{owen2000safe,veach1995optimally}, where the denominator
+is the sum of all kernels, $q_t(\cdot), 0\leq t\leq n$,
+\[
+	w_n = \pi(Y_n) \Big/ \sum\limits_{t=0}^n \! q_t(Y_n).
+\]
+The advantage of this system is that at a time $n_1$ a region might be
+badly approximated, it might be better approximated at a later time
+$n_2$, so the weight for the sample $Y_{n_1}$ will be more
+representative of the density in that region, if it is considered as a
+draw from a mixture of samples up to time $n_2$. These weights are
+however biased, consistency was proved in~\cite{marin2014consistency}.
+AMIS has been successfully implemented for problems in population
+genetics~\cite{siren2011reconstructing}.
+
+A combination of AMIS and PMC results in a stable PMC algorithm known
+as adaptive population importance sampling
+(APIS)~\cite{martino2015adaptive}. This algorithm utilises the
+deterministic mixture weights to stabilise the algorithm within a
+single population, then `restarts' the kernel approximation in the
+next iteration. This memoryless property makes the algorithm much more
+flexible to adaptation. A combination with the $D$-kernel PMC method
+allows for adaptive learning of kernel, kernel mean and kernel
+covariances.
+
 \section{The Parallel Adaptive Importance Sampler \allowbreak (PAIS)}\label{Sec:PAIS}

### 2016-12-04 PAIS.tex
added consistency section to introduction contents

@@
 faster alternative to resamplers which solve the optimal transport
-problem exactly. In Section \ref{Sec:Num} we present some numerical
-examples, before a brief conclusion and discussion in Section
-\ref{Sec:Conc}.
+problem exactly. In Section~\ref{sec:consistency} we consider
+consistency of the PAIS algorithm. In Section \ref{Sec:Num} we present
+some numerical examples, before a brief conclusion and discussion in
+Section \ref{Sec:Conc}.
 
@@
 
-\section{Consistency of PAIS}
+\section{Consistency of PAIS}\label{sec:consistency}
 

### 2016-12-04 PAIS.tex
Made it clear that we save (w,Y) as the sample not X.

@@
 this set of weighted samples. This is achieved by implementing a
-resampling algorithm. Initially we will use the
-ETPF algorithm\cite{reich2013nonparametric}, although we will suggest an alternative strategy in
-Section \ref{sec:AMR}. The output of the resampling algorithm gives us
-a set of evenly weighted
-samples that we believe represents the target distribution well, and
-from this point we
-can iterate the process once again. The algorithm is summarised in Algorithm \ref{alg:PAIS}.
+resampling algorithm. Initially we will use the ETPF
+algorithm\cite{reich2013nonparametric}, although we will suggest an
+alternative strategy in Section \ref{sec:AMR}. The resampling
+algorithm gives us a set of evenly weighted samples which represents
+the target distribution well, which we can use to iterate the process
+again. The algorithm is summarised in Algorithm \ref{alg:PAIS}.
 
@@
 	}
+	Output $(\W, \Y)$.\;
 \caption{The parallel adaptive importance sampler (PAIS).\label{alg:PAIS}}
@@
 The key is to choose a suitable transition kernel $\nu$ such that
-if $X^{(i)}$ is a good representative sample of the posterior,
+if $X^{(i)}$ is a representative sample of the posterior,
 then the mixture density $\chi(\cdot;\X^{(i)})$ is a good

### 2016-12-05 PAIS.tex
references fixed
added 'cost per sig fig' comparison to AMR section

@@
 draw from a mixture of samples up to time $n_2$. These weights are
-however biased, consistency was proved in~\cite{marin2014consistency}.
+however biased, consistency was proved in~\cite{marin2012consistency}.
 AMIS has been successfully implemented for problems in population
@@
 \subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX3"}}
-\subfigure[Cost of resamplers per iteration]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
-\caption{Finding optimal values of $\beta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}.}
+\subfigure[Cost of resamplers per significant figure]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
+\caption{Comparison of the performance between different resampling schemes. The example in Section~\ref{sec:problem 1} is implemented for this demonstration.}
 \label{fig:AMR}

### 2017-01-16 old_intro.tex
New intro
New intro written.

--- 2016-12-05 PAIS.tex
@@
 
-\subsection{Sampling from a one dimensional Gaussian distribution}\label{sec:problem 1}
-
-In this example we compare the naively parallelised RWMH algorithm with its
-PAIS variant, the PAIS-RW algorithm. The PAIS algorithm is implemented using the
-ETPF to perform the resampling step. We assess the performance of the PAIS algorithm
-using the relative $L^2$ error defined in \eqref{eqn:L2_error}, as well as the
-relative error in the first moment \eqref{eq:34567}.
-
-Since we are comparing against naively parallelised MH algorithms,
-we also need to decide which statistics $T(\beta)$ provide the best criterions
-for obtaining the optimal scaling parameters. In the examples which follow, we have
-optimised the naively parallelised RWMH algorithm using the optimal
-acceptance rate $\hat{\alpha} = 0.5$. This value differs from the theoretical
-asymptotic value of 0.234 which applies in higher dimensions, but this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}. To find the optimal
-scaling parameter we minimise the statistic
+In this section we demonstrate convergence of the PAIS algorithm. We
+also investigate the convergence properties of PAIS compared with
+naively parallelised Metropolis-Hastings samplers. We assess the
+performance of the PAIS algorithm using the relative $L^2$ error
+defined in \eqref{eqn:L2_error}, as well as the relative error in the
+first moment \eqref{eq:34567}.
+
+In the one-dimensional examples which follow, we have optimised the
+naively parallelised RWMH algorithm using the optimal acceptance rate
+$\hat{\alpha} = 0.5$. This value differs from the theoretical
+asymptotic value of 0.234 which applies in higher dimensions, but this
+higher acceptance rate is commonly used for one dimensional Gaussian
+posteriors~\cite{rosenthal2011optimal}. To find the optimal scaling
+parameter we minimise the statistic
 \[
@@
 
+\subsection{Sampling from a one dimensional Gaussian distribution}\label{sec:problem 1}
+
+This first example shows how PAIS handles searching for the mode of a
+Gaussian distribution when the initial state is a long way out in the
+tails.
+
 \subsubsection{Target distribution}
 
-Consider the simple case of a linear observation operator $\G(x) = x$, where the prior on $x$ and the observational noise follow Gaussian distributions.
-Then, following \eqref{eqn:like}, the Gaussian posterior has the form
+Consider the simple case of a linear observation operator $\G(x) = x$,
+where the prior on $x$ and the observational noise follow centred
+Gaussian distributions. Then, following \eqref{eqn:like}, the Gaussian
+posterior has the form
 \begin{equation}\label{eqn:Gaussian posterior}
@@
 observational noise and prior distributions respectively. In the
-numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
-we observe $x_\text{ref}=4$ noisily such that
+numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$,
+and we observe $x_\text{ref}=4$ noisily such that
 \[
@@
 
-In each of the following simulations, we perform three tasks.
-First we calculate the optimal value of $\beta$ by optimising the statistics described in Section~\ref{sec:statistics}.
-We then run the algorithms with optimal parameters to calculate and compare the convergence rates.
-Finally, we implement the adaptive algorithms described in Section~\ref{sec:adapt} and compare the convergence rates of these algorithms with the nonadaptive algorithms.
+In each of the following simulations, we perform three tasks. First we
+calculate the optimal value of $\beta$ by optimising the statistics
+described in Section~\ref{sec:statistics}. We then run the algorithms
+with optimal parameters to calculate and compare the convergence
+rates. Finally, we implement the adaptive algorithms described in
+Section~\ref{sec:adapt} and compare the convergence rates of these
+algorithms with the nonadaptive algorithms.
 
@@
 
-{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for one million iterations, again with $M=50$.
-The simulations are repeated 32 times using the optimal parameters found in (1).
-The performance of the algorithms is judged by the convergence of the relative $L^2$ error statistic in \eqref{eqn:L2_error}.
+{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the
+algorithms in Section~\ref{sec:problem 1} for one million iterations,
+again with $M=50$. The simulations are repeated 32 times using the
+optimal parameters found in (1). The performance of the algorithms is
+judged by the convergence of the relative $L^2$ error statistic in
+\eqref{eqn:L2_error}.
 
@@
 \subfigure[PAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
-\caption{Finding optimal values of $\beta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
+\caption{Finding optimal values of $\beta$ for the example in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
 \label{fig:P1 opt beta}
@@
 
-Figure~\ref{fig:P1 opt beta} (a) shows the two values of $\beta$ which are optimal according to the acceptance rate and relative $L^2$ error criteria for the RWMH algorithm.
-The smaller estimate comes from the relative $L^2$ error, and the larger from the acceptance rate.
-The results in Figure~\ref{fig:P1 opt beta} are summarised in Table~\ref{table:prob1 opt beta}.
-Since in general we cannot calculate the relative $L^2$ error, we must optimise the algorithm using the acceptance rate.
-From the relative $L^2$ error curve we can see that the minimum is very wide and despite the optimal values being very different there is not a large difference in the convergence rate.
+Figure~\ref{fig:P1 opt beta} (a) shows the two values of $\beta$ which
+are optimal according to the acceptance rate and relative $L^2$ error
+criteria for the RWMH algorithm. The smaller estimate comes from the
+relative $L^2$ error, and the larger from the acceptance rate. The
+results in Figure~\ref{fig:P1 opt beta} are summarised in
+Table~\ref{table:prob1 opt beta}. Since in general we cannot calculate
+the relative $L^2$ error, we must optimise the algorithm using the
+acceptance rate. From the relative $L^2$ error curve we can see that
+the minimum is very wide and despite the optimal values being very
+different there is not a large difference in the convergence rate.
 
@@
 averaging the statistic over each iteration is a good indicator for
-the optimal scaling parameter. In general this indicator overestimates
-the value of $\neff$ found by using the entire sample.
+the optimal scaling parameter.
 
@@
 
-Figure~\ref{fig:MH1 L2} shows that the PAIS-RW algorithm converges to the posterior distribution significantly faster than the RWMH algorithm, in both $L^2$ error and relative error in the moments.
-A description of the speed up attained by this algorithm is given in Section~\ref{sec:calc_saving}.
+Figure~\ref{fig:MH1 L2} shows that the PAIS-RW algorithm converges to
+the posterior distribution significantly faster than the RWMH
+algorithm, in both $L^2$ error and relative error in the moments. A
+description of the speed up attained by this algorithm is given in
+Section~\ref{sec:calc_saving}.
 
@@
 ARWMH algorithm does not perform quite as well, this is possibly due
-to the fact that the acceptance rate cost function is not particularly
-smooth at the optimal value making it difficult to minimise.
+to the fact that $T_{\text{MH}}(\beta)$ is not smooth at the optimal
+value making it difficult to minimise.
 
@@
 
-Throughout this example, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales when we increase the ensemble size, and if there is some limit below which the algorithm fails.
-We implement the problem in Section~\ref{sec:problem 1}, using the RWMH and PAIS-RW algorithms with ensemble sizes in the interval $M \in [1, 160]$.
+Throughout this example, we use an ensemble size $M=50$, but it is
+interesting to see how the PAIS algorithm scales when we increase the
+ensemble size, and if there is some limit below which the algorithm
+fails. We implement the problem in Section~\ref{sec:problem 1}, using
+the RWMH and PAIS-RW algorithms with ensemble sizes in the interval $M
+\in [1, 160]$.
 
@@
 reduce the number of iterations required in comparison with naively
-parallelised MH. The decreasing trend indicates superlinear improvement of PAIS with
-respect to ensemble size, in terms of the number of iterations
-required, which is a demonstration of our belief that parallelism of
-MCMC should give us added value over and above that provided by naive parallelism. This decrease is linked to the increasing effective sample size shown in Figure~\ref{fig:neff-M} (b).
+parallelised MH. The decreasing trend shows superlinear improvement of
+PAIS with respect to ensemble size, in terms of the number of
+iterations required, which is a demonstration of our belief that
+parallelism of MCMC should give us added value over and above that
+provided by naive parallelism. This decrease is due to the increasing
+effective sample size shown in Figure~\ref{fig:neff-M} (b).
 

### 2017-01-16 PAIS.tex
Intro update
Small change to intro.

@@
   ensemble sizes, which can then be used in conjunction with PAIS for
-  complex problems.
+  complex problems. In particular, we demonstrate this methodology's
+  superior sampling for multimodal problems, such as those arising
+  from inference for mixture models.
 \end{abstract}
@@
 \section{Introduction}
-%\subsection{Lit. review}
-%\subsection{Optimal transport resampler}
-%\subsection{Function Space MCMC}
-Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
-that allow us to sample from complex probability distributions. MCMC
-methods were first developed in the 70s\cite{hastings1970monte}, and
-with the development of faster, more
-powerful computers, have become ever more important in a whole range
-of fields in statistics, science and engineering.  In particular, when
-considering Bayesian inverse problems, each MCMC step may involve the
-numerical solution of one or more PDE. As many samples are usually
-required before Monte Carlo error is reduced to acceptable levels,
-the application of MCMC methods to these types of problem remain frustratingly out of our grasp.
-
-Many advances have been made in the field of MCMC to design ever more
-complex methods that propose moves more intelligently, leading to
-rapidly converging approximations. Function space versions of standard methods
-such as the random walk Metropolis-Hastings (RWMH) algorithm or the
-Metropolis adjusted Langevin algorithm (MALA), whose convergence rates
-are independent of dimension have been
-developed\cite{cotter2013mcmc}. The hybrid (or Hamiltonian) Monte
-Carlo (HMC) method uses Hamiltonian dynamics in order to propose and
-accept moves to states which are a long way away from the current
-position\cite{sexton1992hamiltonian}, and function space analogues of
-this have also been proposed\cite{beskos2011hybrid}. Riemann
+Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
+importance and interest in the last 20 years or so. They allow us to
+sample from complex probability distributions which we would not be
+able to sample from directly. In particular, these methods have
+revolutionised the way in which inverse problems can be tackled,
+allowing full posterior sampling when using a Bayesian framework.
+
+However, this often comes at a very high cost, with a very large
+number of iterations required in order for the empirical approximation
+of the posterior to be considered good enough. As the cost of
+computing likelihoods can be extremely large, this means that many
+problems of interest are simply computationally intractable.
+
+This problem has been tackled in a variety of different ways. One
+approach is to constuct increasingly complex MCMC methods which are
+able to use the structure of the posterior to make more intelligent
+proposals, leading to more thorough exploration of the posterior with
+fewer iterations. For example, the Hamiltonian or Hybrid Monte Carlo
+(HMC) algorithm uses gradient information and simplectic integrators
+in order to make very large moves in state with relatively high
+acceptance probability\cite{sexton1992hamiltonian}. Non-reversible
+methods are also becoming quite popular as they can improve
+mixing\cite{bierkens2015non}. Riemann
 manifold Monte Carlo methods exploit the Riemann geometry of the
@@
 successfully applied to MALA-type proposals and methods which exploit
-even higher order gradient information\cite{bui2014solving}.  These
-methods allow us to explore the posterior distribution more fully 
-with fewer iterations.
-
-Simultaneously, great strides are continually being made in the
-development of computing hardware. Moore's law, which predicted that
-the number of transistors that can fit onto a single microchip will
-double every two years, has been largely followed since the early
-70s\cite{moore1998cramming}. In recent times, it has become necessary
-to use parallel architectures in order for this trend to continue. The efficient
-exploitation of these architectures is the key to solving many of the
-computational challenges that we currently face.
-
-As such, the development of efficient parallel MCMC algorithms is an
-important area for research. Since MCMC methods can be naively
-parallelised by simply running many independent chains in parallel,
-the focus needs to be on the development of methods which gain some
-added benefit through parallelism. One class of parallel MCMC method
-uses multiple proposals, with only one of these proposals being
+even higher order gradient information\cite{bui2014solving}.
+
+Since the clock speed of an individual processor is no longer
+following Moore's law\cite{moore1998cramming}, improvements in
+computational power are largely coming from the parallelisation of
+multiple cores. As such, the area of parallel MCMC methods is becoming
+increasingly of interest. One class of parallel MCMC method uses
+multiple proposals, with only one of these proposals being
 accepted. Examples of this approach include multiple try
@@
 parallelisation of MCMC methods was presented, which demonstrated
-speed ups of up to two orders of magnitude when compared with serial methods.
+speed ups of up to two orders of magnitude when compared with serial
+methods.
+
+Other methods are also designed for particular scenarios. For
+instance, the challenge of sampling from high/infinite-dimensional
+posterior distributions is a big one. The majority of
+Metropolis-Hastings algorithms suffer from the curse of
+dimensionality, requiring more samples for a given degree of accuracy
+as the dimension of problem is increased. However some
+dimension-independent methods have been developed, based on
+Crank-Nicolson discretisations of certain stochastic differential
+equations\cite{cotter2013mcmc}. Other ideas such as chain
+adaptation\cite{haario2005componentwise} and early rejection of
+samples can also aid reduction of the computational
+workload\cite{solonen2012efficient}.
+
+However, high dimensionality is not the only challenge that we may
+face. Complex structure in low
+dimensions can cause significant issues. These issues may arise due to
+large correlations between parameters in the posterior, leading to
+long thin curved structures which many standard methods can struggle
+with. These features are common, for example, in inverse problems
+related to epidemiology and
+other biological applications\cite{house2016bayesian}. Multimodality of the posterior can also lead to
+incredibly slow convergence in many methods. Many methods allow for
+good exploration of the current mode, but the waiting time to the next
+switch of the chain to another mode may be large. Since many switches
+are required in order for the correct weighting to be given to each
+mode, and for all of the modes to be explored fully, this presents a
+significant challenge.
+
+One application where this is an ever-present problem is that of
+mixture models. Given a dataset, where we know that the data is from
+two or more different distributions, we wish to be able to identify
+the parameters, e.g. the mean and variance and relative weight, of
+each part of the mixture\cite{marin2005bayesian}. The resulting posterior
+distribution is invariably a multimodal distribution, since it is
+usually invariant to permutations. Metropolis-Hastings algorithms, for
+example, will often fail to converge in a reasonable time frame for
+problems such as this. Since the posterior may be multimodal,
+independent of this label switching, it is still important to be able
+to efficiently sample from the whole posterior.
 
@@
 proposal from the current state of all of the chains in the
-ensemble. The idea is to consider the current state on each of a set of
-parallel chains as an ensemble, and to resample using a transformation
-based on optimal transport. Samplers based on optimal transport
-have also been considered in \cite{el2012bayesian}.
+ensemble. The method makes use of a resampler based on optimal
+transport which has been used in the context of particle
+filters\cite{reich2013nonparametric}. In particular we demonstrate the
+advantages of this method when attempting to sample from multimodal
+posterior distributions, such as those arising from inference for
+mixture models.
 
@@
 
-In this section we investigate the behaviour of the PAIS algorithm
+In this second example we investigate the behaviour of the PAIS algorithm
 when applied to bimodal problems. MH methods can

### 2017-01-16 PAIS.tex
Cut AIS section and added paragraphs to introduction

@@
 
+Importance sampling methods are another class of methods which allow
+us to sample from complex probability distribution. A related class of algorithms, adaptive importance sampling (AIS)
+\cite{liu2008monte} reviewed in\cite{bugallo2015adaptive}, had
+received less attention until their practical applicability was
+demonstrated in the
+mid-2000s\cite{celeux2006iterated,cappepopulation,isard1998condensation,bink2008bayesian}.
+AIS methods produce a sequence of approximating distributions,
+constructed from standard distributions, from which samples can be
+easily drawn. At each iteration the samples are weighted, often using
+standard importance sampling methods. The weighted samples are used to
+train the adapting sequence of distributions so that samples are drawn
+more efficiently as the iterations progress. The weighted samples form
+a sample from the posterior distribution under some mild
+conditions~\cite{robert2013monte,martino2015adaptive}.
+
+Ensemble importance sampling schemes also exist, such as population Monte Carlo (PMC)~\cite{cappe2012population}. This algorithm uses a population of
+samples to build a mixture or kernel density estimate (KDE) of the
+posterior distribution. The efficiency of
+this optimisation is restricted by the component kernel(s) chosen, and
+the quality of the current sample from the posterior. Extensions, such
+as adaptive multiple importance sampling algorithm
+(AMIS)~\cite{cornuet2012adaptive}, and adaptive population importance sampling
+(APIS)~\cite{martino2015adaptive} have allowed these methods to
+applied to various applications, including population
+genetics~\cite{siren2011reconstructing}.
+
 In this paper, we present a framework for parallelisation of
@@
 general form.
-
-\subsection{Adaptive Importance Sampling}
-
-Bayesian philosophy has become wide\-spread in statistical inference
-largely due to the development of efficient methods, such as MCMC and
-particle filtering, for calculating previously intractable integrals.
-A related class of algorithms, adaptive importance sampling (AIS)
-\cite{liu2008monte} reviewed in\cite{bugallo2015adaptive}, had
-received less attention until their practical applicability was
-demonstrated in the
-mid-2000s\cite{celeux2006iterated,cappepopulation,isard1998condensation,bink2008bayesian}.
-AIS methods produce a sequence of approximating distributions,
-constructed from standard distributions, from which samples can be
-easily drawn. At each iteration the samples are weighted, often using
-standard importance sampling methods. The weighted samples are used to
-train the adapting sequence of distributions so that samples are drawn
-more efficiently as the iterations progress. The weighted samples form
-a sample from the posterior distribution under some mild
-conditions~\cite{robert2013monte,martino2015adaptive}.
-
-The AIS algorithm is Markovian, conditioned on the method of adaption,
-but differs from MCMC methods in that the sequence does not have the
-posterior as its limiting distribution. Convergence is ensured by the
-weighting of samples, which does not rely on algorithm asymptotics.
-This means that after any iteration the algorithm can be stopped and
-the sample obtained is a sample from the posterior distribution, with
-no consideration for burn-in times~\cite{douc2007convergence}.
-
-A popular AIS algorithm is the population monte carlo (PMC) sampling
-method~\cite{cappe2012population}. This algorithm uses a population of
-samples to build a mixture or kernel density estimate (KDE) of the
-posterior distribution. This KDE is optimal in the sense that it
-minimises the Kullback-Leibler divergence between the mixture and the
-posterior distribution~\cite{douc2007convergence}. The efficiency of
-this optimisation is restricted by the component kernel(s) chosen, and
-the quality of the current sample from the posterior.
-
-The PMC algorithm is often implemented as a $D$-kernel
-method~\cite{douc2007convergence}. In this method one choses $D$
-different measures, and a number of samples $N$ such that the total
-population size $N = MD$. At each iteration and for each mixture
-component, one of these $D$ kernels is selected via some optimisation
-criterion and $M$ samples are produced from this kernel. The $M$
-samples from each kernel attempt to stabilise the resampling scheme so
-that information about a particular region is not lost.
-
-In practice, the PMC algorithm can suffer from spiky posteriors due to
-poor approximation properties in some parts of the posterior. The
-adaptive multiple importance sampling algorithm
-(AMIS)~\cite{cornuet2012adaptive} aims to reduce the impact of these
-poor approximations. AMIS considers the entire sample at time $n$ to
-be a draw from all approximations up to time $n$. This means that the
-importance weights for the whole sample must be recalculated at each
-iteration. These weights are known as deterministic mixture
-weights~\cite{owen2000safe,veach1995optimally}, where the denominator
-is the sum of all kernels, $q_t(\cdot), 0\leq t\leq n$,
-\[
-	w_n = \pi(Y_n) \Big/ \sum\limits_{t=0}^n \! q_t(Y_n).
-\]
-The advantage of this system is that at a time $n_1$ a region might be
-badly approximated, it might be better approximated at a later time
-$n_2$, so the weight for the sample $Y_{n_1}$ will be more
-representative of the density in that region, if it is considered as a
-draw from a mixture of samples up to time $n_2$. These weights are
-however biased, consistency was proved in~\cite{marin2012consistency}.
-AMIS has been successfully implemented for problems in population
-genetics~\cite{siren2011reconstructing}.
-
-A combination of AMIS and PMC results in a stable PMC algorithm known
-as adaptive population importance sampling
-(APIS)~\cite{martino2015adaptive}. This algorithm utilises the
-deterministic mixture weights to stabilise the algorithm within a
-single population, then `restarts' the kernel approximation in the
-next iteration. This memoryless property makes the algorithm much more
-flexible to adaptation. A combination with the $D$-kernel PMC method
-allows for adaptive learning of kernel, kernel mean and kernel
-covariances.
 

### 2017-01-16 cover_letter.tex
Cut AIS section and added paragraphs to introduction

\documentclass{letter}
\usepackage{txfonts}

\setlength{\textheight}{10.2in} 
\setlength{\topmargin}{-0.9in}

\signature{Dr Simon Cotter (corresponding author), Dr Colin Cotter and Paul Russell}
\address{School of Mathematics \\ Alan Turing Building \\ Oxford
  Road\\Manchester\\ M13 9PL \\ UK \\\textit{email:} simon.cotter@manchester.ac.uk}

\begin{document}

\date{}
\begin{letter}{ }

\opening{Dear Prof. Higdon,}

I am writing with regards to our previous submission to the SIAM/ASA Journal of Uncertainty Quantification, entitled ``Parallel Adaptive Importance Sampling''. Please forgive us for the length of this cover letter, but we felt strongly that we needed to address a few issues from our previous submission.

This manuscript had previously been submitted to the SIAM Journal for Scientific Computing, back in August 2015. We received 2 pretty positive reviews, although one review criticised our lack of a high dimensional example. The introduction at the time was possibly slightly misleading as application areas were mentioned (meteorology, oceanography, oil well diagnostic problems) where the unknowns in the inverse problem are high dimensional. We used these examples as these are the areas which are commonly used to motivate particle filters. Since our method uses a state-of-the-art resampler from a particle filter method, this felt like an appropriate setting for our method also. The particle filter methodology has a large literature, all of which uses these motivations, when in fact particle filters, like all importance sampling-based algorithms, suffer badly from the curse of dimensionality. These papers often demonstrate their algorithm's efficacy by applying them to synthetic data sets arising from chaotic and low-dimensional dynamical systems, such as various iterations of the Lorenz equations. The criticisms that were levelled at our paper could just as well be applied to this entire thriving area of mathematical enquiry.

Despite this, the editor stated that our paper could only be accepted if we could demonstrate that the method works for a inverse problem with around $10^9$ unknowns. This request was insurmountable, unreasonable, and not one that anybody who is aware of the dimensionality restrictions of importance sampling/kernel estimation problems would ask. Besides this, even if we did have such an algorithm, the request would require us to have access to the sort of size of computing infrastructure that is available to the MET office. As such, we didn't feel that it would be worth our time in resubmitting to this particular journal.

As a result of this, we undertook a thorough rewrite of the paper, addressing as many of the other concerns of the referees as we could. We then considered which other journals would be a good fit for this work, and felt that SIAM UQ would be perfect. At this stage, we did make an error, in that when submitting, we failed to disclose the previous submission to SISC. In particular, we should have submitted a detailed description of the changes we had made to the manuscipt in response to the referee's comments. This is entirely our own error. However, SIAM's own policy dictates that any resubmissions to a different SIAM journal will be treated as a new submission. Unfortunately, our experience with this second submission has not reflected this.

In the response letter, the associate editor stated that: \emph{"You will see that both referees reports are quite short -- that is my doing. I was aware (through the SIAM grapevine) that you had previously submitted this paper to SISC and that lengthy reviews had been obtained. So, in the interest of not using too much of people's time, I asked for indicative reviews."} This clearly indicates that this submission was not treated as a new submission, and the referees were instructed not to give a full review. Moreover, their reviews have in our view been unfairly coloured by the previous reviews from the SISC submission which were provided to them. In particular, they criticise us due to \emph{"the computed example provides no information about whether or not the proposed method will work in the (sizable) applications mentioned in your introduction."} However, all mentions of such sizable applications have been completely removed from the introduction. The suggested references from referee 2 are also all regarding MCMC algorithms for high dimensional problems, when this method is more aimed at complex posteriors in lower dimensions. Referee 2 also suggest that ``The adaptation only aims at learning the scale of the proposal, whereas the shape can be as (or more) important''. We have no idea where this criticism comes from since the entire point of the paper is that the proposal distribution learns the shape of the posterior distribution from the ensemble of particles.

The changes to the paper since the initial submission have been comprehensive - almost all of the numerical experiments have been rerun, a new 2D example was added from the application of parameter determination for chemical networks. In particular we would highlight also that a new resampler has been developed in this second version, a greedy approximation of the full state-of-the-art resampler that was used in the original paper. The introduction was also completely rewritten.

We are however very keen to make adjustments to the paper in order to appease any legitimate concerns that the referees have highlighted. In particular, we have replaced the 2D problem at the end of the paper with a 4 dimensional example where the posterior exhibits nonlinear correlations between the parameters, again coming from an inverse problem arising from chemical kinetics. We have also completely rewritten the introduction for a second time, taking more care to reference more relevant literature from, among others, the adaptive importance sampling community. Hopefully the introduction now better reflects the content of the rest of the paper. Importantly, we have also added a short passage which directs interested/concerned readers to proofs of the consistency of ensemble adaptive importance sampling schemes.

We are keen to resubmit the manuscript to SIAM/ASA JUQ for a second time. In order to address the issues that have arisen, we have provided three copies of the paper. The first is the original submission to SISC. The second is the version of the paper that was submitted to SIAM UQ, with the changes made highlighted. The third version is our new submission, with the newest changes highlighted. We also provide detailed descriptions of how are changes address the concerns of both sets of referees.

However, if you still believe that this manuscript and its contents are not suitable for publication in your journal, we would very much appreciate a swift decision, in order that we are able to submit it elsewhere.

We thank you very much for your time and patience in dealing with this manuscript, and would like to apologise in our oversight in failing to provide these details at the time of the first submission to SIAM/ASA JUQ. We look forward to hearing from you in due course.

\closing{Yours sincerely,}
\end{letter}

\end{document}

### 2017-01-19 cover_letter.tex
Cover letter changes

@@
 The particle filter methodology has a large literature, all of which uses these motivations, when in fact particle filters, like all importance sampling-based algorithms, suffer badly from the curse of dimensionality.
-These papers often demonstrate their algorithm's efficacy by applying them to synthetic data sets arising from chaotic and low-dimensional dynamical systems, such as various iterations of the Lorenz equations.
+%These papers often demonstrate their algorithm's efficacy by applying them to synthetic data sets arising from chaotic and low-dimensional dynamical systems, such as various iterations of the Lorenz equations.
 The criticisms that were levelled at our paper could just as well be applied to this entire thriving area of mathematical enquiry.
 
-Despite this, the editor stated that our paper could only be accepted if we could demonstrate that the method works for a inverse problem with around $10^9$ unknowns.
-This request was insurmountable, unreasonable, and not one that anybody who is aware of the dimensionality restrictions of importance sampling/kernel estimation problems would ask.
-Besides this, even if we did have such an algorithm, the request would require us to have access to the sort of size of computing infrastructure that is available to the MET office.
+Unfortunately, one reviewer fixated on this, and as such we were asked not to resubmit unless we were able to demonstrate the performance of the algorithm in a very high dimensional setting, which of course is not feasible for this type of method.
 As such, we didn't feel that it would be worth our time in resubmitting to this particular journal.
@@
 In particular, they criticise us due to \emph{"the computed example provides no information about whether or not the proposed method will work in the (sizable) applications mentioned in your introduction."} However, all mentions of such sizable applications have been completely removed from the introduction.
-The suggested references from referee 2 are also all regarding MCMC algorithms for high dimensional problems, when this method is more aimed at complex posteriors in lower dimensions.
+% The suggested references from referee 2 are also all regarding MCMC algorithms for high dimensional problems, when this method is more aimed at complex posteriors in lower dimensions.
 Referee 2 also suggest that ``The adaptation only aims at learning the scale of the proposal, whereas the shape can be as (or more) important''.
@@
 We are however very keen to make adjustments to the paper in order to appease any legitimate concerns that the referees have highlighted.
-In particular, we have replaced the 2D problem at the end of the paper with a 4 dimensional example where the posterior exhibits nonlinear correlations between the parameters, again coming from an inverse problem arising from chemical kinetics.
+In particular, we have replaced the 2D problem at the end of the paper with a 6 dimensional multimodal posterior arising from inverse problems for mixture models.
 We have also completely rewritten the introduction for a second time, taking more care to reference more relevant literature from, among others, the adaptive importance sampling community.
@@
 The third version is our new submission, with the newest changes highlighted.
-We also provide detailed descriptions of how are changes address the concerns of both sets of referees.
+We also provide detailed descriptions of how the changes address the concerns of the referees.
 

### 2017-01-21 PAIS.tex
edits to introduction sections

--- 2017-01-16 PAIS.tex
@@
   computing increases. In this paper, we aim to develop a framework
-  for scalable parallel MCMC algorithms. At each iteration, an
+  for scalable parallel sampling algorithms. At each iteration, an
   importance sampling proposal distribution is formed using the
-  current states of all of the chains within an ensemble. Once
-  weighted samples have been produced from this, a state-of-the-art
+  an ensemble of particles. A stratified sample is taken from this
+  distribution and weighted under the posterior, a state-of-the-art
   resampling method is then used to create an evenly weighted sample
@@
   adaptive importance sampling (PAIS) method outperforms naive
-  parallelisation of serial MCMC methods using the same number of
-  ensemble members, for low dimensional problems, and in fact shows
+  parallelisation of serial MCMC methods for low dimensional problems,
+  and in fact shows
   better than linear improvements in convergence rates with respect to
@@
 
-Other methods are also designed for particular scenarios. For
-instance, the challenge of sampling from high/infinite-dimensional
-posterior distributions is a big one. The majority of
+A variety of other methods have been designed with particular scenarios in mind. For
+instance, sampling from high/infinite-dimensional
+posterior distributions is of interest in many applications. The majority of
 Metropolis-Hastings algorithms suffer from the curse of
 dimensionality, requiring more samples for a given degree of accuracy
-as the dimension of problem is increased. However some
+as the parameter dimension is increased. However some
 dimension-independent methods have been developed, based on
@@
 each part of the mixture\cite{marin2005bayesian}. The resulting posterior
-distribution is invariably a multimodal distribution, since it is
-usually invariant to permutations. Metropolis-Hastings algorithms, for
+distribution is invariably a multimodal distribution, since the likelihood is
+invariant to permutations. Metropolis-Hastings algorithms, for
 example, will often fail to converge in a reasonable time frame for
 problems such as this. Since the posterior may be multimodal,
-independent of this label switching, it is still important to be able
+independent of this label switching, it is important to be able
 to efficiently sample from the whole posterior.
 
-Importance sampling methods are another class of methods which allow
-us to sample from complex probability distribution. A related class of algorithms, adaptive importance sampling (AIS)
+Importance samplers are another class of methods which allow
+us to sample from complex probability distributions. A related class of algorithms, adaptive importance sampling (AIS)
 \cite{liu2008monte} reviewed in\cite{bugallo2015adaptive}, had
@@
 AIS methods produce a sequence of approximating distributions,
-constructed from standard distributions, from which samples can be
+constructed from mixtures of standard distributions, from which samples can be
 easily drawn. At each iteration the samples are weighted, often using
@@
 
-Ensemble importance sampling schemes also exist, such as population Monte Carlo (PMC)~\cite{cappe2012population}. This algorithm uses a population of
-samples to build a mixture or kernel density estimate (KDE) of the
+Ensemble importance sampling schemes also exist, e.g. population Monte Carlo (PMC)~\cite{cappe2012population}. PMC uses an ensemble to build a mixture or kernel density estimate (KDE) of the
 posterior distribution. The efficiency of
@@
 (AMIS)~\cite{cornuet2012adaptive}, and adaptive population importance sampling
-(APIS)~\cite{martino2015adaptive} have allowed these methods to
+(APIS)~\cite{martino2015adaptive} have enabled these methods to be
 applied to various applications, including population
@@
 Metropolis-based methodologies in order to create an efficient target
-proposal from the current state of all of the chains in the
+distribution from the current
 ensemble. The method makes use of a resampler based on optimal
@@
 
+%%%%%%
+
 \section{Preliminaries}\label{Sec:Prelim}
+
 In this Section we will introduce preliminary topics and algorithms
 that will be referred to throughout the paper.
+
+%%%%%%
+
 \subsection{Bayesian inverse problems}
+
 In this paper, we focus on the use of MCMC methods for characterising
 posterior probability distributions arising from Bayesian inverse problems. We
-wish to learn about a particular unknown quantity $u$, of which we are
+wish to learn about a particular unknown quantity $x$, of which we are
 able to make direct or indirect noisy observations. For now
@@
 
-The parameter $x$ is observed
-through the observation operator $\mathcal{G}:X \to\mathbb{R}^d$. 
-Since observations are never perfect, we
-assume that these measurements $\data$ are subject to Gaussian noise,
-so that
+The quantity $x$ is mapped on to observable space by the observation
+operator $\mathcal{G}:X \to\mathbb{R}^d$. We assume that the
+observations, $\data$, are subject to Gaussian noise,
 \begin{equation}\label{eqn:obs}
@@
 \end{equation}
-For example, if $x$ are the rates of reactions in a chemical system, $\G$ might
-return the quantities of each chemical species at a particular time, or some
-summary of this information.
+For example, if $x$ is an initial condition for the wave equation,
+$\mathcal{G}$ might calculate the height of a wave at a particular point and time.
 
 These modelling assumptions allow us to construct the 
-likelihood of observing the data $\data$ given the parameter $x =
+likelihood of observing the data $\data$ given the quantity $x =
 x^*$. Rearranging \eqref{eqn:obs} and using the distribution of
@@
 \end{equation}
-where $\| y_1 - y_2 \|_\Sigma$ is the Mahalanobis distance between $y_1$ and $y_2 \in \mathbb{R}^d$.
+where $\| y_1 - y_2 \|_\Sigma = (y_1-y_2)^\top\Sigma^{-1}(y_1-y_2)$ for $y_1,y_2 \in \mathbb{R}^d$.
 
@@
 
-% \subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}\label{Sec:pCNL}
-% In recent years, work has been carried out to frame MCMC proposal
-% distributions on function space\cite{cotter2013mcmc}. These new
-% discretisations perform comparably with the original versions in low
-% dimensions. If gradient information regarding the observation operator is
-% available, then a range of MCMC methods are available which exploit this information to improve mixing rates. One
-% example of such an algorithm is MALA. In \cite{cotter2013mcmc}, a function space version of this
-% method was presented, the pCNL algorithm, and is described in full in Table
-% \ref{tab:pCNL}. The proposal used in this method comes about through
-% a Crank-Nicolson approximation of the Langevin SDE, whose invariant
-% measure is the posterior measure $\mu_Y$.
-
-% \begin{table}
-% \begin{mdframed}
-% \begin{algorithmic}
-% \STATE $X = x_0$
-% \FOR{$i=1,2,3,\ldots$}
-% \STATE $Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
-% 2\delta\mathcal{C}\nabla \Phi(u)+
-% \sqrt{8\delta} W\right] $, $W \sim \mu_0$
-% \STATE $a(X_{i-1},Y) = \min \left \{ 1,  \exp(\Phi(X_{i-1}) - \Phi(Y) ) \right \}$.
-% \STATE $u \sim U([0,1])$
-% \IF{$u<a(X_{i-1},Y)$}
-% \STATE $X_i = Y$
-% \ELSE 
-% \STATE $X_i = X_{i-1}$
-% \ENDIF
-% \ENDFOR
-% \end{algorithmic}
-% \end{mdframed}\caption{A pseudo-code representation of the preconditioned Crank-Nicolson Langevin
-%    (pCNL) algorithm. $\delta \in (0,2]$ is a step size parameter.}
-% \label{tab:pCNL}
-% \end{table}
+%%%%%%
 
 \subsection{Particle filters and resamplers}\label{sec:filters}
-In several applications, data must be assimilated in an ``online''
-fashion, with up to date observations of the studied system being made
-available on a regular basis. In these contexts, such as in weather forecasting or
-oceanography, data is incorporated using a filtering
-methodology. One popular filtering method is the particle filter, the
-first of which was dubbed the Bootstrap
-filter\cite{gordon1993novel}. In this method, a set of weighted particles is
-used to represent the posterior distribution. The positions of the
-particles are updated using the model dynamics. Then, when more
-observations are made available, the relative weights of the particles
-are updated to take account of this data, using Bayes' formula. Other filtering methods, such
-as the Kalman filter\cite{kalman1960new} and ensemble Kalman filter\cite{evensen1994sequential}, have also been developed which are often used within the data
-assimilation community.
-
-One advantage of the particle filter is that there are convergence
-results for this method as the number of particles is increased.  One downside is that the required ensemble
-size increases quickly with dimension, making it difficult to use in
-high-dimensional problems. Another
-downside is that the effective sample size decreases at each
-iteration, resulting in degeneration of the approximation of the
-posterior. One way to tackle
-this is to employ a resampling scheme. The aim of a successful
-resample is to take the unevenly weighted ensemble and return a new
-ensemble of particles with even weights which is highly correlated to
-the original samples.
-
-The ensemble transform particle filter (ETPF) proposed by Reich
-\cite{reich2013nonparametric} makes use of optimal transportation as
-described in \cite{villani2003topics,villani2008optimal}. The transform
-takes a sample of weighted particles $\{y_i\}_{i=1}^M$ from $\mu_Y$
-and converts it into a sample of evenly weighted particles $\{x_i\}_{i=1}^M$
-from $\mu_X$, by means of defining a coupling $T^*$ between $Y$ and
-$X$. Given that a trivial coupling $T^t$ always exists in the space of
-transference plans, $\Pi(\mu_{X}, \mu_{Y})$, we can find a coupling
-$T^*$ which maximises the correlation between $X$ and $Y$
-\cite{cotter2012ensemble}. This coupling is the solution to a linear
-programming problem in $M^2$ variables with $2M-1$ constraints.
-Maximising the correlation ensures that the new sample is as much like
-the original sample as possible with the additional property that the sample
-is evenly weighted.
-
-A Monte Carlo algorithm can be implemented to resample from a weighted
-ensemble. We create a weighted sample, then solve the optimal
-transport problem which produces the coupling described above, we can
-draw a new sample from the evenly weighted distribution. Reich
-suggests using the mean of the evenly weighted distribution to produce
-a consistent estimator. Analysis of this method shows that as the
-ensemble size increases, the statistics of the evenly weighted sample
-approach those of the posterior distribution.
+
+Particle filters are a class of Monte Carlo algorithms designed to
+solve the filtering problem. That is, to find the best estimate of the
+true state of a system when given only noisy observations of the
+system. The solution of this problem has been of importance since the
+middle of the 20th century in fields such as molecular biology,
+computational physics and signal processing. In recent years the data
+assimilation community has contributed several efficient particle
+filters, including the ensemble Kalman filter
+(EnKF)~\cite{evensen1994sequential} and the ensemble transform
+particle filter (ETPF)~\cite{reich2013nonparametric}.
+
+The ETPF defines a coupling $T$ between two random variables $Y$ and
+$X$, allowing us to use the induced map as a resampler. An \emph{optimal
+coupling} $T^*$ is one which maximises the correlation between $X$ and
+$Y$ \cite{cotter2012ensemble}. This coupling is the solution to a
+linear programming problem in $M^2$ variables with $2M-1$ constraints,
+where $M$ is the size of the sample. Maximising the correlation
+preserves the statistics of $X$ in the new sample.
+
+In this work we use these particle filters as resampling methods. We
+approximate the posterior distribution, $\mu_Y$, with a small sample
+of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
+the weights would be found by incorporating new observed data whereas
+here we simply use importance weights. This distribution, $\hat{\mu}_Y
+\approx \mu_Y$, can then be resampled, using the ETPF or otherwise,
+into a new distribution, $\hat{\eta} \approx \mu_Y$,
+\begin{equation}\label{eqn:resampler}
+	\hat{\mu}_Y = \sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
+	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
+	 n_i\delta_{x_i}(\cdot) = \hat{\eta}.
+\end{equation}
+where $n_i \in \{0, 1, \dots,M\}$ and $\sum_{i=1}^M \! n_i = M$. This
+new sample $\{x_i\}_{i=1}^M$ is equally weighted.
+
+Particle filters such as the ETPF are well suited to this problem
+since it is easy to introduce conditions in the resampling to ensure
+you obtain the behaviour you require. One downside is that the
+required ensemble size increases quickly with dimension, making it
+difficult to use in high-dimensional problems.
+
+%%%%%%
 
@@
 All MCMC methods are naively parallelisable. One can take a method
-and simply implement it simultaneously over a set of processors in an ensemble. All
+and simply implement it simultaneously over an ensemble of processors. All
 of the states of all of the ensemble member can be recorded, and in the time
@@
 
+%%%%%%
+
 \section{The Parallel Adaptive Importance Sampler \allowbreak (PAIS)}\label{Sec:PAIS}
-
-% In a trivially parallelised MCMC scheme, if we have enough processes
-% all sampling from the target density at the same time, then we can
-% make a very rough approximation of that density by taking the current
-% states of all of the processes as a representative sample. In the
-% PAIS, the proposal density for the importance sampler stage is a
-% mixture of all of the proposal densities for a chosen MCMC scheme,
-% using the current states. If the proposal in the MCMC scheme is
-% appropriate and our number of samples is big enough, we might hope
-% that this mixture of densities is a good enough approximation of the
-% target density, and that new samples drawn from
-% the mixture proposal are approximately draws from the target
-% measure. If this is the case, then the variance of importance weights
-% for these samples will be small, leading to a highly efficient
-% sampler.
-
-% The problem at this stage, is that our current states all have varying
-% importance, but in order for us to iterate this whole process again,
-% we require them to be the same. Therefore, we employ a resampler to
-% take our new set of samples with differing weights, and return a set
-% of states with equal weights, which are chosen so that the correlation
-% between the two sets is maximised. At this stage we can start the whole
-% process again.
 
@@
 similar areas to the target density, and incredibly inefficient when
-this is not the case. The aim of the PAIS is to use an ensemble of states,
-each coming from a MCMC chain, to construct a proposal
+this is not the case. The aim of the PAIS is to use an ensemble of states to construct a proposal
 distribution which will be as close as possible to the target density. If this
@@
 but we choose to use a mixture distribution, made up of a sum of MCMC
-proposal distributions for each of the members of the ensemble. Once the proposal is
-constructed, we can sample a new set of states from the proposal
-distribution, and each is assigned a weight given by the ratio of the target
-density and the proposal mixture distribution density. Assuming that
-our proposal distribution is a good one, then the variance of the
-weights will be small, and we will have many useful samples. Finally, we
-need to create a set of evenly weighted samples which best represent
-this set of weighted samples. This is achieved by implementing a
-resampling algorithm. Initially we will use the ETPF
-algorithm\cite{reich2013nonparametric}, although we will suggest an
-alternative strategy in Section \ref{sec:AMR}. The resampling
-algorithm gives us a set of evenly weighted samples which represents
-the target distribution well, which we can use to iterate the process
-again. The algorithm is summarised in Algorithm \ref{alg:PAIS}.
+proposal kernels. This corresponds to replacing the Dirac distribution
+in the right-hand side of Equation~\ref{eqn:resampler} with a chosen
+distribution e.g. Gaussian. Once the proposal is constructed, we can
+sample a new ensemble from the proposal distribution, and each is
+assigned a weight given by the ratio of the target density and the
+proposal mixture density. Assuming that our proposal distribution is a
+good one, then the variance of the weights will be small, and we will
+have many useful samples. Finally, we need to create a set of evenly
+weighted samples which best represent this set of weighted samples.
+This is achieved by implementing a resampling algorithm. Initially we
+will use the ETPF algorithm\cite{reich2013nonparametric}, although we
+will suggest an alternative strategy in Section \ref{sec:AMR}. The
+resampling algorithm gives us a set of evenly weighted samples which
+represents the target distribution well, which we can use to iterate
+the process again. The algorithm is summarised in Algorithm
+\ref{alg:PAIS}.
 
@@
 represent the current state of all of the Markov chains as a vector
-$\X = [x_1,x_2,\ldots,x_M]^T$. We are also given a transition kernel
+$\X = [x_1,x_2,\ldots,x_M]^\top$. We are also given a transition kernel
 $\nu(\cdot,\cdot)$, which might come from an MCMC method, for example
@@
 variance of the proposal.
-
-%\begin{table}
-%\begin{mdframed}
-%\begin{algorithmic}
-%\STATE $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^T$
-%\FOR{$i=0,1,2, ..., N-1$}
-%\STATE $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^T, \quad y_j^{(i)} \sim
-%\nu(\cdot;x_j^{(i)})$
-%\STATE $\chi(y;\X^{(i)}) = \frac{1}{M}
-%\sum_{j=1}^M \nu(y;x_j^{(i)})$.
-%\STATE $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^T,$ \quad $w^{(i)}_j =
-%\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i)})}$.
-%\STATE Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i+1)})$
-%\ENDFOR 
-%\end{algorithmic}
-%\end{mdframed}
-%\caption{A pseudo-code representation of the Parallel Adaptive
-%  Importance Sampler (PAIS).}
-%\label{tab:PAIS}
-%\end{table}
 
@@
 		Sample $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^\top, \quad y_j^{(i)} \sim
-\nu(\cdot;x_j^{(i-1)})$.\;
+\nu(\cdot;x_j^{(i-1)})$.\label{algline:PAIS_propose}\;
 		Calculate $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^\top,$ \quad $w^{(i)}_j =
@@
 		\[
-			\chi(y;\X^{(i-1)}) = \frac{1}{M}\sum_{j=1}^M \nu(y;x_j^{(i-1)}).
-		\]
-
-		Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i)})$.
+			\chi(\cdot;\X^{(i-1)}) = \frac{1}{M}\sum_{j=1}^M \nu(\cdot;x_j^{(i-1)}).
+		\]
+
+		Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i)})$.\label{algline:PAIS_resample}\;
 	}
@@
 approximation of the posterior distribution. If this is the case,
-the newly proposed states $\Y^{(i)}$ will also be a good sample of the posterior with low variance in the
-weights $\W^{(i)}$.
+the newly proposed states $\Y^{(i)}$ will also be a good sample of the
+posterior with low variance in the weights $\W^{(i)}$.
 
 In Section \ref{Sec:Num}, we will demonstrate how the algorithm
-performs, primarily using RWMH proposals. We do not claim that this choice is optimal, but
+performs, primarily using random walk (RW) proposals. We do not claim that this choice is optimal, but
 is simply chosen as an example to show that sharing information across
-ensemble members can improve on the original MCMC algorithm and lead to
-convergence in fewer evaluations of $\G$. This is important since if
+ensemble members can improve on the original MH algorithm and lead to
+tolerances being achieved in fewer evaluations of $\G$. This is important since if
 the inverse problem being tackled involves computing
@@
 significantly improve the speed of convergence of the algorithm for
-simple examples\cite{Paul}.
+posteriors which are easy to approximate\cite{Paul}.
 
@@
 ensure that the proposals are absolutely continuous with respect to
-the posterior distribution. In Section \ref{sec:chem} we will consider
-an inverse problem with Gamma priors. Since the posterior distribution
-in this context is heavy-tailed, if we use a lighter tailed
-distribution for $\nu$, such as a Gaussian kernel, then importance
-weights in the tails will be unbounded, which will hamper convergence
-of the algorithm.
+the posterior distribution. If this is not the case, importance
+weights grow exponentially for samples in the tails. This will
+drastically decrease the efficiency of sampling.
+
+%%%%%%
 
@@
 
-%If the
-%proposal distribution is too over-dispersed, then the algorithm will often
-%propose states in the tails of the distribution, resulting in larger
-%variance of the weights, and therefore slower convergence to the
-%posterior distribution. Similarly, if the proposal distribution is
-%under-dispersed, the proposals will be highly correlated with the
-%previous states, and the algorithm will take a long time to fully explore
-%the parameter space, and worse, will lead regularly to states proposed
-%in the tails of the proposal distribution with very large weights. It is therefore necessary to find a proposal
-%distribution which is slightly over-dispersed to ensure the entire
-%posterior is explored\cite{gelman1992inference}, but is as close to the
-%posterior as possible.
-
-Most commonly used MCMC proposals have parametric dependence which
-allows the user to control their variance. For example, in the RWMH
-proposal $y = x + \beta \eta$, the parameter $\beta$ is the variance
+Most MCMC proposals have parametric dependence which
+allows the user to control their variance. For example, in the RW
+proposal $y = x + \beta \eta$, the parameter $\beta$ is the standard deviation
 of the proposal distribution. Therefore the proposal distributions can
@@
 
-%The alternative to using adaptive procedures to tune the scaling
-%parameters is to perform exploratory simulations to find the optimal
-%regimes by trial and error. This can be very costly and the optimal
-%parameters cannot realistically be found to more than a couple of
-%significant figures. It is therefore important that MCMC algorithms
-%provide a feasible adaptive strategy.
-
 In many MCMC algorithms such as the Random Walk Metropolis-Hastings
@@
 
+%%%%%%
+
 \subsection{Statistics for Determining the Optimal Scaling Parameter}\label{sec:statistics}
@@
 \subsubsection{Determining optimal scaling parameter using error analysis}
-
-%Sampling algorithms can be assessed by comparing their approximation of
-%the posterior to the analytic distribution, in cases where the
-%posterior distribution can be computed using alternative methods.
 
@@
 \end{equation}
-where $\{x_i\}_{i=1}^N$ is a sample of size $N$.
-
-The relative $L^2$ error, $E$, between a continuous function to a
-piecewise constant function, can be given by considering the
+and $\{x_i\}_{i=1}^N$ is a sample of size $N$.
+
+The relative $L^2$ error, $E$, between a continuous density function to a
+piecewise constant approximation of that density, can be defined by considering the
 difference in mass between the self-normalised histogram of the
@@
 These statistics cannot be used in general to find optimal values of
-$\beta$ since they require knowledge of the analytic solution, and
-the algorithm must be run for a long time to build up a
-sufficiently large sample. However they can be used to assess the
-ability of other indicators to find the optimal scaling parameters in
-a controlled setting.
-
-%\subsubsection{The variance of the weights}
-
-%Importance samplers assign a weight to each sample they produce based
-%on a ratio of the posterior to the proposal at that point. Importance
-%samplers are most efficient when the target is proportional to the
-%proposal distribution.
-In this case the weights are all equal, and so the variance of the weights, $var(w(y))$, is zero.
-Hence, we would like to %choose the value of $\beta$ which minimises the variance of the weights,
-%\[
-%	\beta^*_{\text{var}} = \argmin_{\beta} \text{var}(w(y)).
-%\]
-%In our experience, the mean of the estimator
-%$\text{var}(w(y))$ is a smooth enough function of $\beta$ that it can
-%be used to tune the proposal variance during the burn-in phase of MCMC
-%algorithms.
-However the variance of the estimator of the variance can be large, especially far away from the optimal value, so it can take a %large number of iterations to calculate descent directions.
+$\beta$ since they require the analytic solution, and are expensive to
+approximate. However they can be used to assess the ability of other
+indicators to find the optimal scaling parameters in a controlled
+setting.
+
+%%%%%%
 
@@
 sample size is more desirable.
-
-%In the numerics which follow, we use the effective sample size
-%calculated at each iteration and averaged across the entire run to
-%tune the scaling parameters. We do this because when we tune this
-%parameter on the fly, we use the single-iteration average to estimate
-%optimality. The optimal scaling parameter found using the global
-%optimum of the variance of the weights is also included for
-%comparison. Note that for the majority of iterations the value of
-%$\neff$ is an overestimate of this statistic over a larger number of
-%samples. Rare events which result in a large importance weight bring
-%this statistic down, and care must be made not to overfit the proposal
-%variance to $\neff$ on an iteration by iteration basis. This can be
-%accounted for by increasing the variance slightly once the adaptive
-%algorithm has arrived on a value using single iteration values of $\neff$.
 
@@
 $E$ during the first 30 iterations. These numerics are taken from the
-example in Section~\ref{sec:chem} using the PAIS-Gamma algorithm.}
+example in Section~\ref{sec:mixture_conv} using the PAIS-RW algorithm.}
 \label{fig:neff-burnin}
@@
 
-%If we are to use the effective sample size ratio as an indicator of
-%how to tune the proposal variance, we need to look at how it behaves
-%in different situations. Here we look at how the statistic behaves as
-%we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results
-%for the Gaussian posterior discussed in Section~\ref{sec:problem 1}
-%when using the PAIS algorithm with RW proposals, each with variance
-%$\beta^2 \in \mathbb{R}_{>0}$. Figure~\ref{fig:neff-M} (a) shows that
-%as the ensemble size increases, the scaling parameter which gives the
-%optimal effective sample size ratio decreases. This is to be expected
-%since if we are trying to approximate the posterior with a mixture
-%distribution of a small number of Gaussians, the optimal variance will
-%naturally be larger so that the whole of the significant regions are
-%covered by the proposal distribution. As the number increases, the
-%optimal variances decrease so that finer details in the posterior can
-%be better represented in the proposal. Figure~\ref{fig:neff-M} (b) shows that as the ensemble size increases, the efficiency of the sampler also increases.
-
-\subsection{Adaptive PAIS}\label{sec:adapt}
-
-%A popular approach for adaptive MCMC algorithms is to view the scaling
-%parameter as a random variable which we can sample during the course
-%of the MCMC iterations. However, it can be slow to converge to the
-%optimal value, and we may need an
-%uninformative prior for the scaling parameter. Alternatively, the
-%parameter may be randomly sampled at various points during the
-%evolution of the chain. This results in some iterations which make
-%larger global moves in the state space between modes in the target
-%distribution, and some which make local moves. Algorithms of this type
-%do not converge to an optimal value of the scaling parameter.
+\subsection{Adaptively Tuned PAIS}\label{sec:adapt}
 
@@
 \cite{roberts2009examples,Ji2013adaptive,andrieu2006ergodicity}.
-%Algorithm~\ref{tab:adapt} in Appendix~\ref{Sec:App}, is not
-%presented as an optimal strategy but as a demonstration that the PAIS
-%algorithm can be adaptively tuned effectively.
-
-%From here on in, we use the random walk proposal unless otherwise stated,
-%\begin{equation}y = x_{i-1}+\beta \omega_{i-1}, \qquad \omega_{i-1}
-%  \sim \mathcal{N}(0,\Sigma) \qquad {\rm i.i.d.}
-%\end{equation}
-%where $x_{i-1}$ is our current state, $y$ is our proposed state and
-%$\Sigma$ is a covariance operator which might come from the prior
-%distribution. In the numerics section we will refer to $\beta$ as the
-%scaling parameter.
-
-%Using an adaptive strategy, we calculate a sequence
-%$\{\beta^{(k)}\}_{k=1}$ which converges roughly to the optimal
-%scaling parameter $\beta^*$, resulting in the optimal transition
-%density $\chi$ for our MCMC algorithm. This optimal value will differ
-%depending on the criterion we are optimising. We must choose some sequence
-%of iterations, $\{n_k\}_{k=1}$, at which to update $\beta$, and due
-%to the constraints on adaptive MCMC
-%algorithms\cite{roberts2007coupling,roberts2009examples}, these $n_k$
-%must grow exponentially further apart. This same adaptive approach
-%can also be applied to the trivially parallelised MCMC algorithms to adaptively calculate their optimal scaling parameter $\beta^*$.
 
@@
 	\For{$n=1,\dots,N$}{
-		Compute normal PAIS steps.\;
+		Complete steps \ref{algline:PAIS_propose}-\ref{algline:PAIS_resample} of Algorithm~\ref{alg:PAIS}.\;
 		\If {$n \in \{n_k\}$}
@@
 	}
-\caption{The Adaptive PAIS Algorithm.\label{alg:adaptPAIS}}
+\caption{Adaptively tuned PAIS algorithm.\label{alg:adaptPAIS}}
 \end{algorithm}
@@
 
-% When implemented, there are a number of parameters which must be
-% chosen to ensure efficient tuning of $\beta$. Firstly, the initial
-% value of $\beta$ should be chosen so that the chains quickly explore the parameter space. However, if $\beta$ is chosen to be too
-% large, then the method may become unstable. We also choose two
-% iteration numbers at which we change our adaptive strategy. At
-% iteration $N_\text{join}$ we resample using the full ensemble instead
-% of using two subsets of the full ensemble. When using two subsets we
-% are training the scaling parameter using the wrong ensemble size, this
-% allows faster convergence but leads to a sub-optimal scaling
-% parameter. Resampling with the full ensemble towards the end of the
-% tuning allows fast convergence to the right order of magnitude,
-% followed by fine tuning to the optimal scaling parameter. Finally we
-% choose a time, $N_\text{stop}$, at which we stop updating our
-% parameter.
+%%%%%%
 
 \section{Approximate Multinomial Resampling}\label{sec:AMR}
+
 Although the ETPF is optimal in terms of preserving statistics of the
@@
 
+%%%%%%
+
 \section{Consistency of PAIS}\label{sec:consistency}
@@
 Since the normalisation constant is consistent, sums of the independent estimators are also consistent.
+
+%%%%%%
 
@@
 
-% One disadvantage of parallelised algorithms is that often different
-% processors will complete their tasks in different amount of times, for
-% a number of reasons, often due to communication between the
-% processors. One approach which could be applied to the PAIS to avoid
-% this is for each processor to immediately start a new iteration, only
-% using the last values of the other processors that were communicated
-% to it. This incomplete approach would still be valid, and could lead
-% to more efficient use of the computer architecture.
-
 One limitation of the PAIS approach as described above is that a
@@
 
-%  A second limitation is associated
-% with the problem of accurately approximating measures with empirical
-% measures in high dimensional phase space. This is a well-known issue
-% in filtering, where one possible solution is to ``localise'' the
-% impact of observations, as rigorously analysed in
-% \cite{rebeschini2013can}. An established technique in Ensemble Kalman
-% filters, localisation has also been incorporated into the Ensemble
-% Transform Particle Filter \cite{ChRe2015}. This issue could be
-% addressed in the PAIS by using localisation when computing the
-% transform, but keeping the unlocalised weights in the importance
-% sampling for consistency.
-
 \bibliographystyle{siam}
@@
 
-% \begin{appendix}
-% \section{The adaptive PAIS algorithm}\label{Sec:App}
-% \begin{table}[!h]
-% \begin{mdframed}
-% \begin{algorithmic}
-% \STATE $x_j^{(0)} \sim \mu_0$, for $j \in L\cup U$, where $L = \{j\}_{j=1}^{M/2},\ U = \{j\}_{j=M/2+1}^M$.
-% \STATE Choose $\beta^{(1)} \in (0, 2]$. Set $\beta_{L,U}^{(1)} = (1\pm 0.01)\beta^{(1)}\wedge 2$.
-% \FOR{$i=1,2, \ldots, N$}
-% \STATE $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\beta_L^{(i)})$ for $j \in L$, and $y_j^{(i)}~\sim~Q(x_j^{(i-1)},\beta_U^{(i)})$ for $j \in U$.
-% \STATE Calculate 
-% \[
-% 	w^{(i)}_j = \frac{\pi(y_j^{(i)})}{\nu(y_j^{(i)};X^{(i-1)})},
-% \]
-% where
-% \[
-% 	\nu(y;X) = \frac{1}{M}\sum_{j\in L} q(y;x_j,\beta_L)+\frac{1}{M}\sum_{j\in U} q(y;x_j,\beta_U).
-% \]
-
-% \IF{$i$ is in $\{n_k\}_{k=1}$}
-% 	\STATE For $w_{kj} = w_j^{(k)}$, $S = n_k - n_{k-1}$,
-% 	\STATE $T_L = SM(\sum_{k=i-S}^S\sum_{j\in L} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in L} w_{kj}^2)$.
-% 	\STATE $T_U = SM(\sum_{k=i-S}^S\sum_{j\in U} w_{kj})^2/(\sum_{k=i-S}^S\sum_{j\in U} w_{kj}^2)$.
-% 	\STATE $\beta^{(i+1)} = \beta^{(i)} - \Delta t \displaystyle\frac{T_U - T_L}{\beta_U^{(i)}-\beta_L^{(i)}}$.
-% \ELSE
-% 	\STATE $\beta^{(i+1)} = \beta^{(i)}$.
-% \ENDIF
-
-% \IF{$i < N_\text{stop}$}
-% 	\IF{$i < N_\text{join}$}
-% 		\STATE Resample 
-% 	\[
-% 		(w_j^{(i)},y_j^{(i)})_{j\in L} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in L}, \quad (w_j^{(i)},y_j^{(i)})_{j\in U} \rightarrow (\frac{1}{M}, x_j^{(i)})_{j\in U}.
-% \]
-% 	\ELSE
-% 		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
-% 	\ENDIF
-% 	\STATE $\beta_{L,U}^{(i)} = \beta^{(i)} \pm 2\sqrt{2}\beta^{(i)}/\sqrt{NM} \wedge 2$.
-% \ELSE
-% 		\STATE Resample $(w^{(i)},Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, X^{(i)})$.
-% 		\STATE $\beta_L^{(i+1)} = \beta_U^{(i+1)} = \beta^{(i+1)}$.
-% \ENDIF
-% \ENDFOR
-% \end{algorithmic}
-% \end{mdframed}\caption{A pseudo-code representation of the adaptive
-%   PAIS algorithm.}
-% \label{tab:adapt}
-% \end{table}
-% \end{appendix}
-
 \end{document}

### 2017-01-22 PAIS.tex
changes to numerics section - removal of B2 and Chem problem

@@
 \usepackage{cite}
-\usepackage{graphicx,bbm,pstricks}
+\usepackage{graphicx,bbm,pstricks,soul}
 \usepackage{pifont}
@@
  }
+\newcommand*{\centerfloat}{%
+  \parindent \z@
+  \leftskip \z@ \@plus 1fil \@minus \textwidth
+  \rightskip\leftskip
+  \parfillskip \z@skip}
+\makeatother
 \makeatother
@@
 
-\begin{table}[!h]
+\begin{table}[!ht]
 \centering
@@
 
-\section{Automated tuning of Algorithm Parameters}\label{Sec:adapt}
+\section{Automated tuning of algorithm parameters}\label{Sec:adapt}
 
@@
 		{
-			Divide ensemble into two halves. Use these halves to estimate the gradient in $\neff$ at $\delta_k$.\;
-			Update $\delta_k$ using gradient ascent,
+			Divide ensemble into two halves. Use these halves to estimate the gradient in $\neff$ at $\beta_k$.\;
+			Update $\beta_k$ using gradient ascent,
 				\[
-					\delta_{k+1} = \delta_k + \gamma\nabla\neff.
+					\beta_{k+1} = \beta_k + \gamma\nabla\neff.
 				\]
@@
 \subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX3"}}
-\subfigure[Cost of resamplers per significant figure]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
+\subfigure[\hl{Cost of resamplers per significant figure}]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
 \caption{Comparison of the performance between different resampling schemes. The example in Section~\ref{sec:problem 1} is implemented for this demonstration.}
@@
 sampler is presented for the first moment since both the ETPF and the
-AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
+AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} \hl{(d) shows how the computational cost,
 measured in seconds, scales with the ensemble size for the three
-different methods. These results demonstrate that the AMR behaves how we wish, and
+different methods}. These results demonstrate that the AMR behaves how we wish, and
 importantly ensures that exactly one sample of the output will lie in
@@
 performance of the PAIS algorithm using the relative $L^2$ error
-defined in \eqref{eqn:L2_error}, as well as the relative error in the
-first moment \eqref{eq:34567}.
-
-In the one-dimensional examples which follow, we have optimised the
+defined in Equation~\eqref{eqn:L2_error}, as well as the relative error in the
+first moment (Equation~\eqref{eq:34567}).
+
+%%%%%%
+
+\subsection{Numerical implementation}
+\label{sec:Implementation P1}
+
+For each example, we perform the following three tasks.
+
+\begin{enumerate}
+	\item {\bf Finding the optimal scaling parameters}: We chose 32 values of $\beta$ evenly spaced on a log scale in the interval $[10^{-5}, 2]$.
+We ran the PAIS and MH algorithms for one million iterations, each with an ensemble size of $M=50$.
+We took 32 repeats and used the geometric means of the sample statistics in Section~\ref{sec:statistics} to find the optimal scaling parameter.
+	\item {\bf Measuring convergence of nonadaptive algorithms}: We ran the PAIS and MH algorithms, using the scaling parameters found in Step 1, for one million iterations, again with $M=50$.
+The simulations are repeated 32 times.
+The performance is evaluated using the relative $L^2$ error defined in Equation~\eqref{eqn:L2_error}.
+	\item {\bf Measuring convergence of adaptive algorithms}: We run the
+adaptive algorithms under the same conditions as the nonadaptive
+algorithms, and again use the relative $L^2$ error to compare
+efficiency. The adaptive algorithms are initialised with $\beta_1$ = 1.
+\end{enumerate}
+
+%%%%%%
+
+\subsection{Sampling from a one dimensional Gaussian distribution}
+\label{sec:problem 1}
+
+This first example shows how PAIS handles searching for the mode of a
+Gaussian distribution when the initial state is a long way out in the
+tails. We optimise the
 naively parallelised RWMH algorithm using the optimal acceptance rate
 $\hat{\alpha} = 0.5$. This value differs from the theoretical
-asymptotic value of 0.234 which applies in higher dimensions, but this
+asymptotic value of 0.234 which applies in higher dimensions. This
 higher acceptance rate is commonly used for one dimensional Gaussian
-posteriors~\cite{rosenthal2011optimal}. To find the optimal scaling
-parameter we minimise the statistic
-\[
-	T_{\text{MH}}(\beta) = \left| \frac{N_{\text{acc}}(\beta)}{N_{\text{total}}} - \hat{\alpha} \right|,
-\]
-where $N_{\text{acc}}(\beta)$ is the number of accepted moves and
-$N_{\text{total}}$ is the total number of samples produced. For the
-PAIS algorithm, we maximise the effective sample size as
+posteriors~\cite{rosenthal2011optimal}. To tune the
+PAIS algorithm we maximise the effective sample size as
 discussed in Section~\ref{sec:ess}.
 
-\subsection{Sampling from a one dimensional Gaussian distribution}\label{sec:problem 1}
-
-This first example shows how PAIS handles searching for the mode of a
-Gaussian distribution when the initial state is a long way out in the
-tails.
+%%%%%%
 
@@
 
-Consider the simple case of a linear observation operator $\G(x) = x$,
-where the prior on $x$ and the observational noise follow centred
-Gaussian distributions. Then, following \eqref{eqn:like}, the Gaussian
-posterior has the form
+Consider the case of a linear observation operator which maps a one
+dimensional parameter onto the observale space, $\G\colon x \mapsto
+x$. We assign centred priors to the parameter $x$, as well as to the
+observational noise. We choose the true value of the parameter to be
+$x_\text{ref} = 4$, then following Equations~\eqref{eqn:obs} and
+\eqref{eqn:like},
 \begin{equation}\label{eqn:Gaussian posterior}
-	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right),
+	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x
+	 - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right).
 \end{equation}
-where $\sigma^2$ and $\tau^2$ are the variances of the
-observational noise and prior distributions respectively. In the
-numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$,
-and we observe $x_\text{ref}=4$ noisily such that
-\[
-D = \mathcal{G}(x) + \eta \sim \mathcal{N}(\mathcal{G}(x_\text{ref}),\sigma^2) = \mathcal{N}(4,0.01).
-\]
-These values result in a posterior density in which the vast majority
+
+This set-up results in a posterior density in which the vast majority
 of the density is out in the tails of the
@@
 
-\subsubsection{Numerical implementation}\label{sec:Implementation P1}
-
-In each of the following simulations, we perform three tasks. First we
-calculate the optimal value of $\beta$ by optimising the statistics
-described in Section~\ref{sec:statistics}. We then run the algorithms
-with optimal parameters to calculate and compare the convergence
-rates. Finally, we implement the adaptive algorithms described in
-Section~\ref{sec:adapt} and compare the convergence rates of these
-algorithms with the nonadaptive algorithms.
-
-{\bf (1) Finding the optimal parameters}: To find the optimal
-parameters we choose 32 values of $\beta$ evenly spaced on a log scale
-in the interval $[10^{-5}, 2]$. We run the PAIS-RW and RWMH algorithms
-for one million iterations, each with an ensemble size of $M=50$. We took
-32 repeats of both algorithms and then used the geometric means of the
-sample statistics to find the optimal parameters.
-
-{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the
-algorithms in Section~\ref{sec:problem 1} for one million iterations,
-again with $M=50$. The simulations are repeated 32 times using the
-optimal parameters found in (1). The performance of the algorithms is
-judged by the convergence of the relative $L^2$ error statistic in
-\eqref{eqn:L2_error}.
-
-{\bf (3) Measuring convergence of adaptive algorithms}: We run the
-adaptive algorithms under the same conditions as the nonadaptive
-algorithms, and again use the relative $L^2$ error to compare
-efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 1.
-
-\subsubsection{Optimal values of $\beta$}\label{sec:Optimal values P1}
+%%%%%%
+
+\subsubsection{Optimising the scaling parameter}\label{sec:Optimal values P1}
 
@@
 
+%%%%%%
+
 \subsubsection{Convergence of RWMH vs PAIS-RW}
@@
 
-Both adaptive algorithms are run with initial values of
-$\beta=1$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
+Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
 period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
 by the end of the simulation window is matching its performance. The
-ARWMH algorithm does not perform quite as well, this is possibly due
-to the fact that $T_{\text{MH}}(\beta)$ is not smooth at the optimal
-value making it difficult to minimise.
+ARWMH algorithm does not perform quite as well, this is possibly because
+it is optimising a cost functional which is not smooth.
+
+%%%%%%
 
@@
 
-Throughout this example, we use an ensemble size $M=50$, but it is
+In this and the next example we use an ensemble size $M=50$, however it is
 interesting to see how the PAIS algorithm scales when we increase the
-ensemble size, and if there is some limit below which the algorithm
-fails. We implement the problem in Section~\ref{sec:problem 1}, using
+ensemble size, and if there is a minimum required ensemble size. We
+implement the problem in Section~\ref{sec:problem 1}, using
 the RWMH and PAIS-RW algorithms with ensemble sizes in the interval $M
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
@@
 optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
-then running 32 repeats at each ensemble size. The convergence rates
+then running 32 repeats at each ensemble size
+
+We repeat the process described in Section~\ref{sec:Implementation P1}
+for each of the ensemble sizes we are interested in. The convergence rates
 are then found by regressing through the data. The graph is still very
@@
 
-\subsection{Sampling from Bimodal Distributions}\label{sec:bimodal}
-
-In this second example we investigate the behaviour of the PAIS algorithm
-when applied to bimodal problems. MH methods can
-struggle with multimodal problems, particularly where switches between
-the modes are rare, resulting in incorrectly proportioned modes in the
-histograms. This example demonstrates that the PAIS algorithm redistributes 
-chains to new modes as they are found. This means that we expect the number of chains in a mode to be
-approximately proportional to the probability density in that mode. As a result, reconstructed posteriors with disproportional modes, as is familiar with the MH algorithms, are not produced.
-
-\subsubsection{Target Distribution} \label{sec:tar}
-
-We look at an `easy' problem, $B_1$, which has a KL divergence of
-0.880, and a `harder' problem, $B_2$, which has a KL divergence of
-3.647. Problem $B_1$ has two modes which are not too far apart. In
-$B_2$ we increase the distance between the two modes which has the
-effect of increasing the expected number of iterations that it takes
-for a MCMC chain to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.
-
-The following setup is the same for both problems. We consider a
-non-linear observation operator $\G(x) = x^2$, and assign the prior
-$x \sim \mu_0 = \N(0, \tau^2=0.25)$. We assume that a noisy reading,
-$D$, is taken according to $D = \G(x_\text{ref}) + \varepsilon$, where
-$\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$. This results
-in the non-Gaussian posterior
-\[
-	\pi(x|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|x^2 - D\|^2 - \frac{1}{2\tau^2}\|x\|^2\right).
-\]
-To create the `easy' problem we say that the true value of
-$\G(x_\text{ref}) = 0.75$, and the `hard' problem is generated using
-$\G(x_\text{ref}) = 2$. In the numerics which follow we draw noise from
-$\mu_\varepsilon$ to generate our data point.
+%%%%%%
+
+\subsection{Sampling from Bimodal Distributions}
+\label{sec:bimodal}
+
+In this second example we investigate the behaviour of the PAIS
+algorithm when applied to bimodal problems. MH methods can struggle
+with multimodal problems, particularly where switches between the
+modes are rare, resulting in incorrectly proportioned modes in the
+histograms. This example demonstrates that the PAIS algorithm
+redistributes particles to new modes as they are found. This means
+that we expect the number of particles in a mode to be approximately
+proportional to the probability density in that mode, resulting in
+faster global convergence.
+
+%%%%%%
+
+\subsubsection{Target distribution} \label{sec:tar}
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{"figures/posteriors3"}
-\caption{Posterior distributions for problems $B_1$ and $B_2$ as
-  described in Section \ref{sec:tar}.}
-\label{fig:problem 3 posteriors}
+\includegraphics[width=\textwidth]{"figures/BM2_posterior"}
+\caption{Posterior distribution for the bimodal example in
+Section~\ref{sec:tar}.}
+\label{fig:BM2_posterior}
 \end{center}
@@
 
-\subsubsection{Calculating values of Optimal $\beta^*$}\label{sec:BM1_opt_beta}
+In this example we look at a bimodal posterior distribution which has
+a high energy barrier between the two modes. In the MH algorithm, the
+expected number of iterations for a switch between modes is a large
+proportion of the total number of iterations. This posterior is in
+Figure~\ref{fig:BM2_posterior}.
+
+We consider a non-linear observation operator $\G\colon x \mapsto
+x^2$, and assign the prior $\mu_0 = \N(0, \tau^2=0.25)$. Again, we
+assume the true parameter $x_\text{ref}=2$ is observed according to
+Equation~\eqref{eqn:obs} with observational noise $\sigma^2 = 0.1$,
+and the posterior is
+\[
+	\pi(x|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|x^2 - D\|^2
+		 - \frac{1}{2\tau^2}\|x\|^2\right).
+\]
+
+%%%%%%
+
+\subsubsection{Calculating values of the optimal scaling parameter}
+\label{sec:BM2_opt_beta}
 
 Calculating the optimal values of the scaling parameters for this
-problem is similar to the previous example; we check only the acceptance
-rate to find the optimal values for RWMH and we use the effective
-sample size to find the optimal values for PAIS-RW.
-Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both
-problems. The subscript on $\beta$ refers to the criterion which has been optimised.
+example is similar to the previous example. These values are given in
+Table~\ref{tab:BM2_opt_delta}. The subscript on $\beta$ refers to the
+criterion which has been optimised.
 
 \begin{table}[!htb]
-    \begin{minipage}{.5\linewidth}
-      \centering
-        \begin{tabular}{|l|r|r|}
-	\hline
-	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$ \\ \hline
-	RWMH								& 4.8e-1					& - \\
-	PAIS-RW								& -						& 1.0e-1\\
-	\hline
-	\end{tabular}
-    \end{minipage}%
-    \begin{minipage}{.5\linewidth}
       \centering
@@
 	\end{tabular}
-    \end{minipage}
-	\vspace{1mm}
-	\caption{Optimal values of $\beta$ for $B_1$ (left) and $B_2$ (right).}
-	\label{table:BM_opt_beta}
+	\vspace{2mm}
+	\caption{Optimal scaling parameters for the example in Section~\ref{sec:bimodal}.}
+	\label{tab:BM2_opt_delta}
 \end{table}
 
-It is relatively simple to find optimal scaling parameters for problem $B_1$.
-These values are given in Table~\ref{table:BM_opt_beta} (left). However
-problem $B_2$ is much harder as transitions between the modes are extremely
-unlikely for the standard RWMH algorithm. This means that we need to consider the
-convergence on two levels; we should consider the algorithm's ability
-to find both the modes, and also whether it can sample them in the correct proportions.
-
+Since transitions between the modes are extremely unlikely, we
+consider the convergence on two levels; convergence to equally
+proportioned modes, and convergence to smooth histograms.
 To get correctly proportioned modes with the RWMH algorithm it is
-important that the chains can transition between the modes frequently, which
-means that $\beta$ must be large. However, this leads to a lower
-acceptance rate, and so we sacrifice convergence locally. For this reason, the RWMH algorithm is very slow to
-converge for problems of this type.
-
-We can achieve these two regimes in RWMH by tuning $\beta$ using the acceptance rate for local convergence, and by $L^2$ error for global convergence.
-Similarly in PAIS-RW we can use the effective sample size for local convergence, and the $L^2$ error for global convergence.
-
-From Table~\ref{table:BM_opt_beta} (right) we see that there is a large difference between the optimal value of $\beta$ for each regime in RWMH, and so will result in inefficient sampling.
-The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\beta^*$; a clear advantage to using this algorithm for this problem.
+important that the chains can transition between the modes frequently,
+which requires a large proposal variance. However, this leads to a
+lower acceptance rate, and so we sacrifice convergence locally. For
+this reason, the RWMH algorithm is very slow to converge for problems
+of this type.
+
+We can achieve these two regimes in RWMH by tuning $\beta$ using
+different statistics; the acceptance rate for local convergence, and
+the $L^2$ error for global convergence. Similarly in PAIS-RW we can
+use the effective sample size for local convergence, and the $L^2$
+error for global convergence.
+
+%%%%%%
 
@@
 \centering
-\subfigure[Relative $L^2$ error.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_L2"}}
-\subfigure[Absolute error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_moments"}}
-\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_1$. Set up described in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
-\label{fig:BM1_L2}
-\end{figure}
-
-As in the Gaussian example we see a significant speed up with the
-PAIS-RW algorithm for problem $B_1$. Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates. We can see that the adaptive algorithms compare closely with the
-respective nonadaptive algorithms and the improvement PAIS offers remains significant.
-
-\begin{figure}[htb]
-\centering
-\subfigure[Relative $L^2$ error using the optimal scaling parameters.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
-\subfigure[Comparison of the relative $L^2$ error of the algorithms optimised by $L^2$ minimisation with locally optimised PAIS-RW with a scout chain.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
-\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_2$. Set up described in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
+\subfigure[Relative $L^2$ error using the optimal scaling parameters.]
+{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
+\subfigure[Convergence of locally optimised PAIS-RW with scout chain,
+and globally optimised PAIS-RW and RWMH.]
+{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
+\caption{Convergence of the PAIS-RW and RWMH algorithms for the
+bimodal example discussed in Section~\ref{sec:bimodal}. Set up
+described in Section~\ref{sec:Implementation P1}. Resampling is
+performed using the ETPF.}
 \label{fig:BM2_L2}
@@
 
-For $B_2$ the algorithms are run with the global optimal value of $\beta^*$, and with the local optimal value of $\beta^*$.
-Figure~\ref{fig:BM2_L2} (a) shows that the algorithms using the globally optimal $\beta^*$ convergence at the desired rate, whereas the algorithms optimised using the acceptance rate and effective sample size initially converge faster but at some point forget the location of one of the modes, causing the convergence to flatten out.
-
-Using the PAIS algorithm we can minimise the impact of forgotten modes by constantly allowing the algorithm to search them out.
-Since we have parallel chains, we can run PAIS-RW with the majority of chains using the locally optimal value of $\beta$, and one or two chains with a larger scaling parameter.
-These chains with larger scaling parameters act both as `scouts' for new modes, and act to aid in the over-dispersal of the proposal distribution.
-Figure~\ref{fig:BM2_L2} (b) shows the results of using 49 chains with the local optimal scaling parameter, and one chain with ten times the local optimal scaling parameter.
-We see that modes are not forgotten and the algorithm converges with the improvement we see from PAIS-RW in the other problems.
-Other methods of mode searches are described in \cite{lan2013wormhole}.
-
-The adaptive algorithm can just as easily be applied to the PAIS algorithm with the `scout' chains described in the previous paragraph.
-Since we need two equally sized sub-ensembles, we will use two groups of 24 ensembles with the same proposal distributions, and each group will also have a `scout' chain with a scaling parameter ten times that of the rest of the group.
+The MH and PAIS algorithms are run with both the globally and locally
+optimal scaling parameters. Figure~\ref{fig:BM2_L2} (a) shows that the
+globally optimal $\beta^*$ leads to normal convergence rates,
+$\mathcal{O}(n^{-1/2})$, whereas algorithms using locally optimal
+scaling parameters initially converge faster but do not sample
+proportionally for the whole simulation.
+
+Using the PAIS algorithm we can ensure that we do not lose modes as
+simulation progresses. Since we have an ensemble, we can sacrifice
+locally optimal sampling for one or two ensemble members in favour of
+a larger proposal variance. These chains with larger scaling
+parameters act both as `scouts' for new modes, and act to aid in the
+over-dispersal of the proposal distribution. Figure~\ref{fig:BM2_L2}
+(b) shows the results of using 49 chains with the local optimal
+scaling parameter, and one chain with ten times the local optimal
+scaling parameter. We see that modes are not forgotten and the
+algorithm converges with the improvement we see from PAIS-RW in the
+Gaussian example.
+
+The adaptive algorithm can just as easily be applied to the PAIS
+algorithm with the `scout' chains described in the previous paragraph.
+Since we need two equally sized sub-ensembles, we will use two groups
+of 25 ensemble members. Each group has 24 ensemble members tuned for
+locally optimal convergence and one `scout' chain with a scaling
+parameter ten times that of the rest of the group.
 
@@
 \includegraphics[width=0.6\textwidth]{"figures/BM2_AL2"}
-\caption{Convergence of the relative $L^2$ error for problem $B_2$, comparing the globally optimised nonadaptive algorithms with the locally optimised adaptive algorithms.
-The setup is as described in Section~\ref{sec:Implementation P1} (3).
-Resampling is performed using the ETPF.}
+\caption{Convergence of modified PAIS and RWMH algorithms, comparing
+the globally optimised nonadaptive algorithms with the locally
+optimised adaptive algorithms. The setup is as described in
+Section~\ref{sec:Implementation P1} (3). Resampling is performed using
+the ETPF.}
 \label{fig:BM2_AL2}
@@
 the acceptance rate converges at the same rate as the $L^2$ optimised
-algorithm until the scaling parameter gets small, and therefore
+algorithm until the scaling parameter gets too small, and therefore
 switches between the modes are rare, and the relative heights of the
 modes are decided by the arbitrary proportion of chains which are in
-each mode at this point. The adaptive PAIS-RW with scout chains tuned to the effective sample size converges at about the same rate as the locally optimised nonadaptive algorithm also with scouts.
-
-\subsubsection{Calculating the Speed Up in Convergence}\label{sec:calc_saving}
+each mode at this point. The adaptive PAIS-RW with scout chains tuned
+to the effective sample size converges at about the same rate as the
+locally optimised nonadaptive algorithm also with scouts.
+
+%%%%%%
+
+\subsubsection{Calculating the speed up in convergence}
+\label{sec:calc_saving}
 
@@
 \includegraphics[width=0.8\textwidth]{"figures/calc_saving"}
-\caption{Illustration of calculating the number of PAIS-pCNL iterations required to reach a tolerance of $~10^{-2.8}$ as a percentage of MH iterations.
-The relative $L^2$ error graphs are from the Gaussian problem in Section~\ref{sec:problem 1}.}
+\caption{Illustration of calculating the number of PAIS-RW iterations
+required to reach a tolerance of $10^{-2.8}$ as a percentage of MH
+iterations. The relative $L^2$ error graphs are from the Gaussian
+problem in Section~\ref{sec:problem 1}.}
 \label{fig:calc_saving}
@@
 
-The graphs in the previous section clearly show that the PAIS-RW algorithm converges faster than the RWMH algorithm when both are parallelised with the same ensemble size.
-We can calculate the number of iterations required to achieve a particular tolerance level in our solution for each algorithm and compare these to calculate a percentage saving.
-In Figure~\ref{fig:calc_saving} we demonstrate our calculation of the savings.
-The constants $c_1$ and $c_2$ are found by regressing through the data with a fixed exponent of $-1/2$ excluding the initial data points where the graph has not finished burning in.
-
-A summary of the percentage of iterations required using the PAIS algorithm compared with the respective Metropolis-Hastings algorithms is given in Table~\ref{table:calc_savings}.
-The blank entries correspond to occasions when either the MH algorithm or PAIS algorithm hasn't converged to the posterior distribution.
-
-\begin{table}[!h]
+The graphs in the previous section clearly show that the PAIS-RW
+algorithm converges faster than the RWMH algorithm when both are
+parallelised with the same ensemble size. We can calculate the number
+of iterations required to achieve a particular tolerance level in our
+solution for each algorithm and compare these to calculate a
+percentage saving. In Figure~\ref{fig:calc_saving} we demonstrate our
+calculation of the savings. The constants $c_1$ and $c_2$ are found by
+regressing through the data with a fixed exponent of $-1/2$ excluding
+the initial data points where the algorithm has not finished burning
+in.
+
+\begin{table}[!ht]
 \centering
@@
 \hline
-		& Gaussian & $B_1$ & $B_2$ \\ \hline
-	RWMH & 10\% & 32\% & 12.6\% (scout) \\
-	pCN & - & 32\% & - \\
-	MALA & 42\% & 36\% & 40\% \\
-	pCNL & 66\% & 56\% & - \\ \hline
+		& Gaussian & Bimodal \\ \hline
+	RWMH & 10\% & 12.6\% (scout) \\
+	MALA & 42\% & 40\% \\ \hline
 \end{tabular}
-\caption{Iterations for the PAIS algorithms required to achieve a desired tolerance as a percentage of the number of iterations required by the respective MH algorithms.
-The pCN and pCNL proposal distributions are taken from~\cite{cotter2013mcmc}.}
+\vspace{2mm}
+\caption{Iterations for the PAIS algorithms required to achieve a
+desired tolerance as a percentage of the number of iterations required
+by the respective MH algorithms.}
 \label{table:calc_savings}
@@
 
-\subsubsection{A Useful Property of the PAIS Algorithm for Multimodal Distributions}
-
-\begin{figure}[!h]
+A summary of the percentage of iterations required using the PAIS
+algorithm compared with the respective Metropolis-Hastings algorithms
+is given in Table~\ref{table:calc_savings}.
+
+\subsubsection{A useful property of the PAIS algorithm for multimodal distributions}
+
+\begin{figure}[!ht]
 \begin{center}
 \includegraphics[width=\textwidth]{"figures/BM2_suction"}
-\caption{This figure demonstrates the redistribution property of the PAIS algorithm. Initially there is one chain in the positive mode, and 49 chains in the negative mode.}
+\caption{This figure demonstrates the redistribution property of the
+PAIS algorithm. Initially there is one chain in the positive mode, and
+49 chains in the negative mode.}
 \label{fig:BM2_suction}
@@
 The biggest issue for the Metropolis-Hastings algorithms when sampling
-from a posterior such as the one in $B_2$ is that it is unlikely that
-the correct ratio of chains will be maintained in each of the modes, and since there is no interaction between the chains, there is no way to remedy this problem.
-The PAIS algorithm tackles this problem with its resampling step.
-The algorithm uses its dynamic kernel to build up an approximation of the posterior at each iteration, and then compares this to the posterior distribution via the weights function.
-Any large discrepancy in the approximation will result in a large or small weight being assigned to the relevant chain, meaning the chain will either pull other chains towards it or be sucked towards a chain with a larger weight.
-In this way, the algorithm allows chains to `teleport' to regions of the posterior which are in need of more exploration.
-Figure~\ref{fig:BM2_suction} shows Problem $B_2$ with initially 1 chain in the positive mode, and 49 chains in the negative mode.
-It takes only a handful of iterations for the algorithm to balance out the chains into 25 chains in each mode.
-The chains switch modes without having to climb the energy gradient in the middle.
-
-\subsection{Sampling from non-Gaussian bivariate distributions}\label{sec:chem}
-%***some introduction to chemical systems and QSSA
-In this section we apply the PAIS algorithm to a more complicated posterior distribution.
-The field of biochemical kinetics gives rise to multiscale stochastic problems which remain a challenge both theoretically and computationally.
-Biochemical reactions occur in single cells between a number of chemical populations and the rates of these reactions can vary on vastly different timescales.
-It is these reaction rates which we are interested in finding descriptions for.
-
-It is often possible to isolate which reactions are occurring more
-frequently (the fast reactions) and which are occurring less
-frequently (the slow reactions). The quasi-steady-state assumption
-(QSSA) is the assumption that the fast reactions converge in distribution
-on a timescale which is negligible with respect to the rate of
-occurrence of the slow reactions. This assumption allows us to
-approximate the dynamics of the slowly changing quantities in the
-system by assuming that the fast quantities are in equilibrium with
-respect to the fast reactions in isolation. This kind of model
-reduction can be used to approximate the likelihood in an inverse
-problem where we wish to recover the reaction parameters in the
-system.
-
-Let us consider a simple example by introducing the following simple
-chemical system involving two chemical species $S_1$ and $S_2$:
-\begin{equation}\label{eqn:full_system}
-	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
-\end{equation}
-Each arrow represents a reaction from a reactant to a product, with
-some rate constant $k_i$, and where the rates of the reactions are
-assumed to follow mass action kinetics. We denote the concentration of
-species $S_i$ by $X_i$. We assume that we are in a
-parameter regime such that the reactions $S_1\rightarrow S_2$ and $S_2\rightarrow S_1$ occur
-much much more frequently than the other reactions. Notice that both
-chemical species are involved in fast reactions. However, the quantity
-$\mathcal{S} = X_1 + X_2$ is conserved by both of the fast reactions,
-and as such, this is the slowly changing quantity in this system.
-The effective dynamics of $\mathcal{S}$ can be represented as follows.
-\begin{equation}\label{eqn:QSSA_system}
-	\emptyset \xrightarrow{k_1} \mathcal{S} \xrightarrow{\hat{k}_4} \emptyset.
-\end{equation}
-Here, the new reaction rate $\hat{k}_4$ is approximated through
-application of the QSSA to be
+from bimodal posterior distributions is that it is unlikely that the
+correct ratio of chains will be maintained in each of the modes, and
+since there is no interaction between the chains, there is no way to
+remedy this problem. The PAIS algorithm tackles this problem with its
+resampling step. The algorithm uses its dynamic kernel to build up an
+approximation of the posterior at each iteration, and then compares
+this to the posterior distribution via the weights function. Any large
+discrepancy in the approximation will result in a large or small
+weight being assigned to the relevant chain, meaning the chain will
+either pull other chains towards it or be sucked towards a chain with
+a larger weight. In this way, the algorithm allows chains to
+`teleport' to regions of the posterior which are in need of more
+exploration. Figure~\ref{fig:BM2_suction} shows the trace of a
+simulation of the PAIS-RW algorithm with initially 1 chain in the
+positive mode, and 49 chains in the negative mode. It takes only a
+handful of iterations for the algorithm to balance out the chains into
+25 chains in each mode. The chains switch modes without having to
+climb the energy gradient in the middle.
+
+%%%%%%
+
+\subsection{A Mixture Model}\label{sec:mixture}
+
+The technique of mixture modelling employs well known parametric
+families to construct an approximating nonparametric distribution
+which may have a complex structure. Most commonly, Gaussian kernels
+are used since underlying properties in the data can often be assumed
+to follow a Gaussian distribution. An example would be if a
+practitioner were to measure the heights of one hundred adults, but
+failed to record their gender. The data could be considered as one
+population with two sub populations, male and female. The problem then
+might be to find the average height of adult males from the data. In
+this case, since height is often considered to follow a Gaussian
+distribution, it makes sense to model the population as a mixture of
+two univariate Gaussian distributions.
+
+A well known problem in the Bayesian treatment of mixture modelling is
+that of identification, sometimes referred to as the label-switching
+phenomenon. The likelihood distribution for mixture models is
+invariant under permutations of the mixture labels. If a mixture has
+$n$ means and the point $(\mu_1, \dots, \mu_n)$ maximises the
+likelihood, then the likelihood will also be maximised by
+$(\mu_{\varphi(1)}, \dots, \mu_{\varphi(n)})$ for all permutations
+$\varphi(\cdot)$. This means that the number of modes in the posterior
+distribution is of order $\mathcal{O}(n!)$. As we have seen it can be
+hard for standard MH algorithms to obtain reliable inference for
+posterior distributions with a large number of modes, or even a small
+number of modes which are separated by a large distance.
+
+\subsubsection{Target Distribution}\label{sec:mixture_target}
+
+In particular we look at a data set where we assume that there are two
+subpopulations within the overall population. Since both
+subpopulations will be approximated by Gaussian distributions we have
+five parameters which we need to be estimated, two means $\mu_{1,2}$,
+two variances $\sigma^2_{1,2}$, and the probability, $p$, that an
+individual observation belongs to the first subpopulation. We have 100
+data points, $D_i$, which we assume to be distributed according to
 \[
-	\hat{k}_4(s) = \mathbb{E}\left[k_4X_2|\mathcal{S}=s\right] = \frac{k_2k_4\mathcal{S}}{k_2+k_3}.
+	D_i \sim p\mathcal{N}(\mu_1, \sigma^2_1) + (1-p)\mathcal{N}
+	(\mu_2, \sigma^2_2), \quad i = 1,\dots,100,
 \]
-The value of $\mathbb{E}\left[k_4X_2|\mathcal{S}=s\right]$ is
-approximated by finding the steady state of the ODE representing the fast subsystem of
-reactions:
-\[S_1 \xleftrightarrows[k_3]{k_2} S_2, \qquad X_1 + X_2 = s\]
-
-If we assume that we know the rate constants $k_1$ and $k_4$, and
-observe the system in \eqref{eqn:QSSA_system}, we indirectly
-observe the rates $k_2$ and $k_3$ through the effective rate
-$\hat{k}_4$ of the degradation of $\mathcal{S}$. Our
-observations are uninformative about these reaction rates, as there
-are surfaces in parameter space along which the effective rate
-$\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
-problem. Making the assumption that the errors in our observations of
-the value of $\mathcal{S}$ are Gamma distributed with some variance
-$\sigma^2$, this results in a long and thin posterior distribution on
-$k_2$ and $k_3$. This type of problem is notoriously difficult to
-sample from using standard MH algorithms, as the algorithms quickly
-find a point on this manifold on which $\hat{k}_4$ is invariant, but
-exploration along its length is slow.
-
-% solution to QSSA ode and Gamma noise assumption
-\subsubsection{Target Distribution}\label{sec:345}
-
-We now formalise the posterior of interest.
-We look for a distribution over the parameter $\mathbf{k} = (k_2, k_3)^T$, given that we know $k_1 = 100$ and $k_4=1$.
-We generate our data by making ten observations of the system at $t_i = 2, 4, \dots, 20$, here simulated by solving the full system, \eqref{eqn:full_system}, governed by the differential equations
-\begin{align*}
-	\frac{\text{d}S_1}{\text{d}t} &= k_1 - k_2X_1(t)+k_3X_2(t), \\
-	\frac{\text{d}S_2}{\text{d}t} &= k_2X_1(t) - (k_3+k_4)X_2(t),
-\end{align*}
-with initial conditions $X_1(0)=X_2(0) = 0$, and parameter values $\mathbf{k} = (50, 100)^T$.
-We then add noise taken from a Gamma distribution with variance $\sigma^2 = 225$ and centred at each $\mathcal{S}(t_i)=X_1(t_i)+X_2(t_i)$.
-
-In our modelling we use the QSSA to simplify this system of differential equations into the one dimensional system based on \eqref{eqn:QSSA_system},
+where $p \in [0, 1]$, $\mu_{1,2} \in \mathbb{R}$ and $\sigma^2_{1,2}
+\in \mathbb{R}^+$. Due to the domains of these parameters and also
+some prior knowledge, we assign the priors
 \[
-	\frac{\text{d}\mathcal{S}}{\text{d}t} = k_1 - \frac{k_2k_4}{k_2+k_3}\mathcal{S}(t).
+	p \sim \text{Beta}(1,1), \quad \mu_{1,2} \sim \mathcal{N}(0, 4)
+	 \quad \text{and} \quad \sigma^2_{1,2} \sim \text{Gamma}
+	(\alpha=2, \beta=1).
 \]
-The observation operator, $\mathcal{G}: (\mathbf{k},t) \mapsto S(t)$, maps from parameter space onto population space at a time $t$. This means that for the $i$th observation we assume,
-\[
-	 D_i \sim \text{Gamma}(\alpha_i, \beta_i),
-\]
-where
-\[
-	 \alpha_i = \frac{\mathcal{G}(\mathbf{k}, t_i)^2}{\sigma^2},\  \beta_i = \frac{\mathcal{G}(\mathbf{k}, t_i)}{\sigma^2}, \quad i = 1, \dots, 10.
-\]
-These are parameters required to give us a distribution with mean
-$\mathcal{G}(\mathbf{k}$ and variance $\sigma^2$.
-We assign Gamma priors to $\mathbf{k}$ with mean $\alpha_0/\beta_0 =
-75$ and variance $\alpha_0/\beta_0^2 = 100$ in both coordinates, resulting in the posterior
-\[
-	\pi(\mathbf{k}|\mathbf{D}) \propto \left[\prod\limits_{i=1}^{10} \text{Gamma}(D_i; \alpha_i, \beta_i)\right]\text{Gamma}(k_2; \alpha_0, \beta_0)\text{Gamma}(k_3; \alpha_0, \beta_0).
-\]
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.65\textwidth]{"figures/C1_posterior"}
-\caption{The posterior distribution on the parameters $k_2$ and $k_3$
-  given the data and priors described in Section \ref{sec:345}.}
-\label{fig:345}
-\end{figure}
-
-Figure \ref{fig:345} presents a visualisation of the posterior
-distribution for this problem, found by exhaustive MH simulation.
-
-\subsubsection{Implementation}\label{sec:chem_implementation}
-% explanation of l2 convergence measure
-For this problem there is no analytic form for the normalisation
-constant, and numerical methods implemented in MATLAB and Mathematica
-have proven to be unreliable. To demonstrate the convergence of the
-algorithm, we first run a MH algorithm for much longer than we
-normally would ($8\times 10^{10}$ samples), and consider the histogram
-produced to be a well-converged approximation of the true posterior
-distribution. We then perform our usual simulations with the PAIS and
-MH algorithms to compare the rate at which the histograms converge to
-the posterior over the same mesh.
-
-% modifications to algorithm + spikes
-In this problem we modify our PAIS algorithm to use a mixture of Gamma
-distributions in the proposal distribution instead of a Gaussian
-mixture. The PAIS-Gamma and MH-Gamma algorithms use a Gamma
-proposal distribution with mean centred at the previous state,
-\[
-	y \sim \text{Gamma}(\cdot;\alpha^*, \beta^*) \quad \text{where} \quad \frac{\alpha^*}{\beta^*} = x \ \text{and} \ \frac{\alpha^*}{(\beta^*)^2} = \beta^2.
-\]
-
-Similarly the PAIS-LA algorithm uses the Langevin proposal distribution
-from the standard MALA algorithm with a perturbationperturbation taken from a Gamma
-distribution,
-\[
-	y \sim \text{Gamma}(\cdot; \alpha^*, \beta^*), \quad \text{where} \quad \frac{\alpha^*}{\beta^*} = x+\frac{1}{2}\beta^2\nabla\log\pi(X) \ \text{and} \ \frac{\alpha^*}{(\beta^*)^2} = \beta^2.
-\]
-This ensures that we only propose positive
-parameters. Having a heavier right tail in the proposal distribution
-also means that the posterior is absolutely continuous with respect to
-the proposal. When this is not the case, the weight function can tend
-to zero or infinity when $k_i \rightarrow \infty$, which results in
-poor, very spiky approximations of the tails of the posterior distribution leading to slow convergence.
-
-% AMR and increase in sample size
-We now significantly increase the ensemble size we are using from $M=50$ to $M=2500$.
-This allows us to build a better approximation of the posterior distribution for our proposals.
-Following the discussion in Section~\ref{sec:AMR}, it is clear that to keep the runtime of the resampler negligible compared with the calculation of the posterior we must switch from the ETPF resampler to the AMR algorithm.
-
-% numerics.
-\subsubsection{Convergence of PAIS-Gamma and PAIS-LA vs MH with Gamma proposals}\label{sec:chem_conv}
-
-In the numerics which follow we have used the AMR algorithm to resample with an ensemble size of $M=2500$.
-The method otherwise remains the same as in previous sections.
-We perform test runs to find the optimal scaling parameters for both Gamma proposals and MALA-type proposals.
-We then calculate the convergence rates of the algorithms by producing 50 million samples from the posterior with each algorithm, and repeat the simulation 32 times.
-
-\begin{figure}[htb]
-\centering
-\subfigure[Optimal values of the scaling parameter.]{\includegraphics[width=0.45\textwidth]{"figures/C1_ESS"}}
-\subfigure[Relative $L^2$ error against number of iterations.]{\includegraphics[width=0.45\textwidth]{"figures/C1_L2"}}
-\caption{Convergence of the PAIS-Gamma and PAIS-LA algorithms for the chemical system problem described in Section~\ref{sec:chem}.
-Implementation described in Sections~\ref{sec:chem_implementation} and \ref{sec:chem_conv}.
-Resampling is performed using the AMR scheme.}
-\label{fig:C1_ESS}
-\end{figure}
-
-\begin{table}[!htb]
-      \centering
-        \begin{tabular}{|l|r|r|r|}
-	\hline
-	Algorithm	& MH-Gamma & PAIS-Gamma & PAIS-LA \\ \hline
-	$\beta^*$	& 2.7e-0     & 9.4e-1    & 1.0e-0 \\
-	\hline
-	\end{tabular}
-	\vspace{1mm}
-	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithms are optimised using the effective sample size.}
-	\label{table:C1_opt_beta}
-\end{table}
-
-In Figure~\ref{fig:C1_ESS} (a) we see that the two PAIS algorithms have similar optimal values of the scaling parameter, also shown in Table~\ref{table:C1_opt_beta}.
-However, the MALA-type proposal achieves a higher effective sample size with the same ensemble size.
-Convergence after the first 50 million samples is shown in Figure~\ref{fig:C1_ESS} (b) and demonstrates that the PAIS algorithm converges faster than the MH algorithm.
-The slightly unstable convergence in the PAIS-Gamma algorithm is the effect of spikes appearing in the tails of the posterior distribution.
-This is also what has caused the slightly lower effective sample size.
-
-\subsection{A Mixture Model}\label{sec:mixture}
-
-The technique of mixture modelling employs well known parametric families to construct an approximating nonparametric distribution which may have a complex structure.
-Most commonly, Gaussian kernels are used since underlying properties in the data can often be assumed to follow a Gaussian distribution.
-An example would be if a practitioner were to measure the heights of one hundred adults, but failed to record their gender.
-The data could be considered as one population with two sub populations, male and female.
-The problem then might be to find the average height of adult males from the data.
-In this case, since height is often considered to follow a Gaussian distribution, it makes sense to model the population as a mixture of two univariate Gaussian distributions.
-
-A well known problem in the Bayesian treatment of mixture modelling is that of identification, sometimes referred to as the label-switching phenomenon.
-The likelihood distribution for mixture models is invariant under permutations of the mixture labels.
-If a mixture has $n$ means and the point $(\mu_1, \dots, \mu_n)$ maximises the likelihood, then the likelihood will also be maximised by $(\mu_{\varphi(1)}, \dots, \mu_{\varphi(n)})$ for all permutations $\varphi(\cdot)$.
-This means that the number of modes in the posterior distribution is of order $\mathcal{O}(n!)$.
-As we have seen it can be hard for standard MCMC algorithms to obtain reliable inference for posterior distributions with a large number of modes, or even a small number of modes which are separated by a large distance.
-
-\subsubsection{Target Distribution}\label{sec:mixture_target}
-
-In particular we look at a Gaussian mixture where we assume that there are two subpopulations within the overall population.
-Since both subpopulations will be approximated by Gaussian distributions we have five parameters which we need to be estimated, two means $\mu_{1,2}$, two variances $\sigma^2_{1,2}$, and the probability, $p$, that an individual observation belongs to the first subpopulation.
-We have 100 data points, $D_i$, which we assume to be distributed according to
-\[
-	D_i \sim p\mathcal{N}(\mu_1, \sigma^2_1) + (1-p)\mathcal{N}(\mu_2, \sigma^2_2), \quad i = 1,\dots,100,
-\]
-where $p \in [0, 1]$, $\mu_{1,2} \in \mathbb{R}$ and $\sigma^2_{1,2} \in \mathbb{R}^+$. Due to the domains of these parameters and also some prior knowledge, we assign the priors
-\[
-	p \sim \text{Beta}(1,1), \quad \mu_{1,2} \sim \mathcal{N}(0, 4) \quad \text{and} \quad \sigma^2_{1,2} \sim \text{Gamma}(\alpha=2, \beta=1).
-\]
-If we collect these parameters in the vector $\theta = (p, \mu_1, \sigma^2_1, \mu_2, \sigma^2_2)^\top$, the resulting posterior distribution is
+If we collect these parameters in the vector $\theta = (p, \mu_1,
+\sigma^2_1, \mu_2, \sigma^2_2)^\top$, the resulting posterior
+distribution is
 \begin{equation}\label{eq:mixture_posterior}
-	\pi(\theta|D) \propto \prod\limits_{i=1}^{100} (p\mathcal{N}(D_i;\mu_1, \sigma^2_1) + (1-p)\mathcal{N}(D_i;\mu_2, \sigma^2_2))\prod\limits_{i=1}^5 \pi_0^i(\theta_i),
+	\pi(\theta|D) \propto \prod\limits_{i=1}^{100} (p\mathcal{N}
+	(D_i;\mu_1, \sigma^2_1) + (1-p)\mathcal{N}(D_i;\mu_2, \sigma^2_2))
+	\prod\limits_{i=1}^5 \pi_0^i(\theta_i),
 \end{equation}
@@
 \begin{figure}[htb]
-\centering
-\includegraphics[width=\textwidth]{"figures/PAIS_best_posterior"}
-\caption{The posterior distribution of $\theta$ as given in Equation~\eqref{eq:mixture_posterior} as found from 10 million samples from the PAIS algorithm.
-The main diagonal contains the marginal distributions of each $\theta_i$, and the lower triangular contours represents the correlation between pairs of parameters.}
+\centerfloat
+\includegraphics[width=1.2\textwidth]{"figures/PAIS_best_posterior"}
+\caption{The posterior distribution of $\theta$ as given in
+Equation~\eqref{eq:mixture_posterior} as found from 10 million samples
+from the PAIS algorithm. The main diagonal contains the marginal
+distributions of each $\theta_i$, and the lower triangular contours
+represents the correlation between pairs of parameters.}
 \label{fig:mixture_posterior}
@@
 
-Figure \ref{fig:mixture_posterior} presents a visualisation of the posterior distribution for this problem, created from a single PAIS simulation.
-
-\subsubsection{Implementation}\label{sec:mixture_implementation}
-
-We have no analytic form for the posterior distribution for this model, and MH cannot reliably sample from this type of bimodal distribution.
-We do however know that the probability density should be evenly divided between the two modes.
-This is due to the symmetric prior for $p$, and and the same priors being assigned to $\mu_1$ and $\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$.
-To decide which mode a sample belongs to we define a plane which bisects the posterior so that each point on this plane lies exactly halfway between the two true solutions to the inverse problem i.e.
-the value of $\theta$ used to generate the data, and also the $\theta$ obtained by a relabelling of the parameters.
-Now that we can assign a sample to a particular mode, we can calculate the density in each mode by summing the weights associated to all samples in that mode,
+Figure \ref{fig:mixture_posterior} presents a visualisation of the
+posterior distribution for this problem, created from 10million
+samples produced by the PAIS-RW algorithm.
+
+%%%%%%
+
+\subsubsection{Implementation}
+\label{sec:mixture_implementation}
+
+We have no analytic form for the posterior distribution for this
+model, and MH cannot reliably sample from this type of bimodal
+distribution. We do however know that the probability density should
+be evenly divided between the two modes. This is due to the symmetric
+prior for $p$, and and the same priors being assigned to $\mu_1$ and
+$\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$. To decide which
+mode a sample belongs to we define a plane which bisects the posterior
+so that each point on this plane lies exactly halfway between the two
+true solutions to the inverse problem i.e. the value of $\theta$ used
+to generate the data, and also the $\theta$ obtained by a relabelling
+of the parameters. Now that we can assign a sample to a particular
+mode, we can calculate the density in each mode by summing the weights
+associated to all samples in that mode,
 \[
@@
 and the relative error in the amount of density in each mode is then
-\[
+\begin{equation}\label{eqn:mode_prop}
 	w_\text{error} = 2\left|\frac{\bar{w}_1}{\bar{w}_1+\bar{w}_2} - \frac{1}{2}\right|.
-\]
-
-Since the probability $p$ is constrained to lie in the interval $[0,1]$, and the variances must be positive, it can be wasteful to use Gaussian proposal distributions.
-For this example we make proposals by matching the proposal distribution to the prior, scaling the variance and centring the proposal at the previous value.
-So
+\end{equation}
+
+Since the probability $p$ is constrained to lie in the interval
+$[0,1]$, and the variances must be positive, it can be wasteful to use
+Gaussian proposal distributions. For this example we make proposals by
+matching the proposal distribution to the prior, scaling the variance
+and centring the proposal at the previous value. So
 \[
@@
 \]
-where $\alpha^* = \sigma^2_{1,2}\beta^*$, $\beta^* = \sigma^2_{1,2}/2\delta^2$ and $\delta$ is a scaling parameter to be tuned.
-This means that our proposal distributions will not be a mixture of multivariate Gaussians, but independent mixtures of univariate Beta, Gamma and Gaussian distributions.
-
-In the numerics which follow we have increased the ensemble size from $M=50$ to $M=500$ to compensate for the increase in dimension.
-We also use the AMR algorithm for the resampling step because of the reduced computational cost.
-The method otherwise remains the same as in previous examples.
-We perform test runs to find the optimal scaling parameters considering convergence to modes with equal density.
-We then calculate the convergence rates of the algorithms by producing 10 million samples from the posterior with each algorithm, and repeat the simulation 32 times.
+where $\alpha^* = \sigma^2_{1,2}\beta^*$, $\beta^* =
+\sigma^2_{1,2}/2\delta^2$ and $\delta$ is a scaling parameter to be
+tuned. This means that our proposal distributions will not be a
+mixture of multivariate Gaussians, but independent mixtures of
+univariate Beta, Gamma and Gaussian distributions.
+
+In the numerics which follow we have increased the ensemble size from
+$M=50$ to $M=500$ to compensate for the increase in dimension. We also
+use the AMR algorithm for the resampling step because of the reduced
+computational cost. The method otherwise remains the same as in
+previous examples. We perform test runs to find the optimal scaling
+parameters considering convergence to modes with equal density. We
+then calculate the convergence rates of the algorithms by producing 10
+million samples from the posterior with each algorithm, and repeat the
+simulation 32 times.
 
@@
 \includegraphics[width=0.8\textwidth]{"figures/Mode_proportions"}
-\caption{Convergence of the PAIS algorithm for the mixture model described in Section~\ref{sec:mixture}.
+\caption{Convergence of the PAIS algorithm for the mixture model described in Section~\ref{sec:mixture}, convergence judged using the criterion in Equation~\eqref{eqn:mode_prop}.
 Implementation described in Sections~\ref{sec:mixture_implementation} and \ref{sec:mixture_conv}.

### 2017-01-24 PAIS.tex
fixed a missing reference, removed old image files
changed AMR back to time per iteration

@@
 \subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX3"}}
-\subfigure[\hl{Cost of resamplers per significant figure}]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
+\subfigure[Cost of resamplers per iteration]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
 \caption{Comparison of the performance between different resampling schemes. The example in Section~\ref{sec:problem 1} is implemented for this demonstration.}
@@
 sampler is presented for the first moment since both the ETPF and the
-AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} \hl{(d) shows how the computational cost,
+AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
 measured in seconds, scales with the ensemble size for the three
-different methods}. These results demonstrate that the AMR behaves how we wish, and
+different methods. These results demonstrate that the AMR behaves how we wish, and
 importantly ensures that exactly one sample of the output will lie in
@@
 
-We will use the AMR in the numerics in Section \ref{sec:chem_conv}
+We will use the AMR in the numerics in Section \ref{sec:mixture}
 where we have chosen to use a larger ensemble size. We do not claim

### 2017-01-24 PAIS.tex
Simon final changes
Small edits, typos, throughout. Couple of smallish additions.

@@
 This problem has been tackled in a variety of different ways. One
-approach is to constuct increasingly complex MCMC methods which are
+approach is to construct increasingly complex MCMC methods which are
 able to use the structure of the posterior to make more intelligent
@@
 fewer iterations. For example, the Hamiltonian or Hybrid Monte Carlo
-(HMC) algorithm uses gradient information and simplectic integrators
+(HMC) algorithm uses gradient information and symplectic integrators
 in order to make very large moves in state with relatively high
@@
 speed ups of up to two orders of magnitude when compared with serial
-methods.
+methods on a single core.
 
@@
 \subsection{Particle filters and resamplers}\label{sec:filters}
+In this subsection we briefly review particle filters, since the
+development of the resampler that we incorporate into the PAIS is
+motivated by this area.
 
@@
 distribution which will be as close as possible to the target density. If this
-ensemble is large enough, the distribution of states will be representative
+ensemble is large enough, the distribution of states will be
+approximately representative
 of the target density.
@@
 weighted samples which best represent this set of weighted samples.
-This is achieved by implementing a resampling algorithm. Initially we
+This is achieved by implementing a resampling algorithm. These samples
+are not stored in order to characterise the posterior density, since
+the resampling process inevitably adds some error/bias. They are
+simply needed in order to inform the mixture distribution for the next
+iteration of the algorithm.
+
+Initially we
 will use the ETPF algorithm\cite{reich2013nonparametric}, although we
@@
 significantly improve the speed of convergence of the algorithm for
-posteriors which are easy to approximate\cite{Paul}.
+posteriors that we have considered\cite{Paul}.
 
@@
 the last expression, if the variance of the weights is zero then
-$\neff = M$; this is our ideal scenario. Maximising the effective
+$\neff = M$; this is our ideal scenario. In the limit
+$M\rightarrow\infty$, maximising the effective
 sample size is equivalent to minimising the variance of the weights.
@@
 weights, as it takes values on $[1, M]$ while the variance of the
-weights takes values on $\mathbb{R}_\geq$. Since the variance can vary
-over many orders of magnitude bringing numerical issues, the effective
-sample size is more desirable.
+weights takes values on $\mathbb{R}_\geq$. Moreover the variance of
+the weights can vary
+over many orders of magnitude causing numerical issues, so that the effective
+sample size is more desirable as an indicator of optimality.
 
@@
 \subfigure[The highest effective sample size achieved for each ensemble size.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM"}}
-\caption{The behaviour of the effective sample size as the ensemble size increases. This analysis is of the example in Section~\ref{sec:problem 1} using the PAIS-RW algorithm.}
+\caption{The behaviour of the effective sample size as the ensemble
+  size increases, considering the example in Section~\ref{sec:problem 1} using the PAIS-RW algorithm.}
 \label{fig:neff-M}
@@
 as the ensemble size increases, the optimal scaling parameter
-decreases. This is expected since the scaling parameter must be
-reduced if we are to fit more kernels into the same target. We also
+decreases. This is expected since the larger ensemble allows for
+finer resolution in the proposal distribution approximation of the
+posterior distribution. We also
 see that the algorithm becomes less sensitive to changes in the
@@
 As outlined in~\cite{martino2015adaptive} consistency of population AIS algorithms can be considered in two different cases.
-In the first case we fix the number of iterations $N < \infty$, but allow the population size to grow to inifinity $M\rightarrow\infty$.
+In the first case we fix the number of iterations $N < \infty$, but allow the population size to grow to infinity $M\rightarrow\infty$.
 In the second case we hold the population size fixed $M<\infty$, and allow infinitely many iterations $N\rightarrow\infty$.
@@
 Consider the case of a linear observation operator which maps a one
-dimensional parameter onto the observale space, $\G\colon x \mapsto
+dimensional parameter onto the observable space, $\G\colon x \mapsto
 x$. We assign centred priors to the parameter $x$, as well as to the
@@
 ARWMH algorithm does not perform quite as well, this is possibly because
-it is optimising a cost functional which is not smooth.
+it is optimising a cost functional which is not smooth due to an
+increased sensitivity to Monte Carlo error.
 
@@
 expected number of iterations for a switch between modes is a large
-proportion of the total number of iterations. This posterior is in
+proportion of the total number of iterations. This posterior is presented in
 Figure~\ref{fig:BM2_posterior}.
@@
 subpopulations will be approximated by Gaussian distributions we have
-five parameters which we need to be estimated, two means $\mu_{1,2}$,
-two variances $\sigma^2_{1,2}$, and the probability, $p$, that an
+five parameters which we need to be estimated, two means $\{\mu_{1}, \mu_{2}\}$,
+two variances $\{\sigma^2_{1}, \sigma^2_{2}\}$, and the probability, $p$, that an
 individual observation belongs to the first subpopulation. We have 100
@@
 \]
-where $p \in [0, 1]$, $\mu_{1,2} \in \mathbb{R}$ and $\sigma^2_{1,2}
+where $p \in [0, 1]$, $\mu_{1},\mu_2 \in \mathbb{R}$ and $\sigma^2_{1},\sigma^2_2
 \in \mathbb{R}^+$. Due to the domains of these parameters and also
@@
 \[
-	\bar{w}_k = \sum\limits_{i=1}^N w_i\text{I}(X_i \in \text{Mode $k$}), \quad k = 1, 2,
+	\bar{w}_k = \sum\limits_{i=1}^N w_iI_{X_i \in \text{Mode $k$}}, \quad k = 1, 2,
 \]
@@
 proposal. There are a growing number of increasing sophisticated MCMC
-algorithms (HMC, Riemann manifold MCMC etc) which could be
+algorithms (non-reversible Langevin/HMC proposals, Riemann manifold MCMC etc) which could be
 incorporated into this framework, leading to even more efficient

### 2017-02-02 PAIS.tex
updated the figure for hte adaptive algorithms

@@
 observational noise. We choose the true value of the parameter to be
-$x_\text{ref} = 4$, then following Equations~\eqref{eqn:obs} and
-\eqref{eqn:like},
+$x_\text{ref} = 4$, and choose $\sigma^2=\tau^2=0.1$. Then
+following Equations~\eqref{eqn:obs} and \eqref{eqn:like},
 \begin{equation}\label{eqn:Gaussian posterior}
@@
 \centering
-\subfigure[Relative $L^2$ error.]{\includegraphics[width=0.45\textwidth]{"figures/G2_l2"}}
-\subfigure[Relative error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/G2_moments"}}
-\caption{Error analysis for the (A)RWMH and (A)PAIS-RW algorithms. The setup is as in Section~\ref{sec:Implementation P1} (2, 3). Resampling is performed using the ETPF.}
+\includegraphics[width=0.8\textwidth]{"figures/new_G2_l2"}
+\caption{Convergence of the (A)RWMH and (A)PAIS-RW algorithms. The
+setup is as in Section~\ref{sec:Implementation P1} (2, 3). Resampling
+is performed using the ETPF.}
 \label{fig:MH1 L2}
@@
 
-Figure~\ref{fig:MH1 L2} shows that the PAIS-RW algorithm converges to
-the posterior distribution significantly faster than the RWMH
+Figure~\ref{fig:MH1 L2} shows that the PAIS-RW algorithm converges
+tothe posterior distribution significantly faster than the RWMH
 algorithm, in both $L^2$ error and relative error in the moments. A
 description of the speed up attained by this algorithm is given in
-Section~\ref{sec:calc_saving}.
-
-Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
-period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
-by the end of the simulation window is matching its performance. The
-ARWMH algorithm does not perform quite as well, this is possibly because
-it is optimising a cost functional which is not smooth due to an
-increased sensitivity to Monte Carlo error.
+Section~\ref{sec:calc_saving}. The adaptive algorithms are also shown
+in Figure~\ref{fig:MH1 L2}. We see that both adaptive algorithms
+converge to the posterior at a similar speed to the respective
+optimised algorithm. This shows that, particularly for PAIS, we can
+optimise simulations efficiently on the fly.
 

### 2017-02-03 cover_letter.tex
cover letter edits

@@
 I am writing with regards to our previous submission to the SIAM/ASA Journal of Uncertainty Quantification, entitled ``Parallel Adaptive Importance Sampling''.
-Please forgive us for the length of this cover letter, but we felt strongly that we needed to address a few issues from our previous submission.
+Please forgive us for the length of this cover letter, but we felt that we needed to address a few issues from our previous submission.
 
@@
 We received 2 pretty positive reviews, although one review criticised our lack of a high dimensional example.
-The introduction at the time was possibly slightly misleading as application areas were mentioned (meteorology, oceanography, oil well diagnostic problems) where the unknowns in the inverse problem are high dimensional.
-We used these examples as these are the areas which are commonly used to motivate particle filters.
-Since our method uses a state-of-the-art resampler from a particle filter method, this felt like an appropriate setting for our method also.
-The particle filter methodology has a large literature, all of which uses these motivations, when in fact particle filters, like all importance sampling-based algorithms, suffer badly from the curse of dimensionality.
-%These papers often demonstrate their algorithm's efficacy by applying them to synthetic data sets arising from chaotic and low-dimensional dynamical systems, such as various iterations of the Lorenz equations.
-The criticisms that were levelled at our paper could just as well be applied to this entire thriving area of mathematical enquiry.
-
-Unfortunately, one reviewer fixated on this, and as such we were asked not to resubmit unless we were able to demonstrate the performance of the algorithm in a very high dimensional setting, which of course is not feasible for this type of method.
+The introduction at the time was misleading as application areas were mentioned (meteorology, oceanography, oil well diagnostic problems) where the unknowns in the inverse problem are high dimensional.
+One reviewer focussed on this, and as such we were asked not to resubmit unless we were able to demonstrate the performance of the algorithm in a very high dimensional setting, which of course is not feasible for this type of method.
 As such, we didn't feel that it would be worth our time in resubmitting to this particular journal.
 
-As a result of this, we undertook a thorough rewrite of the paper, addressing as many of the other concerns of the referees as we could.
+As a result of this, we undertook a thorough rewrite of the paper, addressing as many of the other concerns of the referees as we could, emphasising on low-dimensional but non-Gaussian distributions, one of which arises from inverse problems for biochemical networks.
 We then considered which other journals would be a good fit for this work, and felt that SIAM UQ would be perfect.
-At this stage, we did make an error, in that when submitting, we failed to disclose the previous submission to SISC.
-In particular, we should have submitted a detailed description of the changes we had made to the manuscipt in response to the referee's comments.
-This is entirely our own error.
-However, SIAM's own policy dictates that any resubmissions to a different SIAM journal will be treated as a new submission.
-Unfortunately, our experience with this second submission has not reflected this.
+The paper was substantially different, and as such we didn't view it as a simple resubmission to another journal.
+Our mistake was in not highlighting this in the coverletter to the submission to SIAM UQ.
 

### 2017-02-03 cover_letter.tex
Cover letter finished

@@
 In particular, they criticise us due to \emph{"the computed example provides no information about whether or not the proposed method will work in the (sizable) applications mentioned in your introduction."} However, all mentions of such sizable applications have been completely removed from the introduction.
-% The suggested references from referee 2 are also all regarding MCMC algorithms for high dimensional problems, when this method is more aimed at complex posteriors in lower dimensions.
-Referee 2 also suggest that ``The adaptation only aims at learning the scale of the proposal, whereas the shape can be as (or more) important''.
-We have no idea where this criticism comes from since the entire point of the paper is that the proposal distribution learns the shape of the posterior distribution from the ensemble of particles.
+The referees were responding to the reviews of the original paper which were sent to them, rather than the new paper, where these issues are not present.
 
@@
 In order to address the issues that have arisen, we have provided three copies of the paper.
-The first is the original submission to SISC.
-The second is the version of the paper that was submitted to SIAM UQ, with the changes made highlighted.
-The third version is our new submission, with the newest changes highlighted.
+The first is a latexdiff pdf of the changes that were made for our submission to SIAM/ASA JUQ from the original submission to SISC.
+The second is a latexdiff pdf of the changes we have made in our latest submission.
+The third version is our new submission.
 We also provide detailed descriptions of how the changes address the concerns of the referees.
@@
 
-We thank you very much for your time and patience in dealing with this manuscript, and would like to apologise in our oversight in failing to provide these details at the time of the first submission to SIAM/ASA JUQ.
-We look forward to hearing from you in due course.
+We thank you very much for your time and patience in dealing with this manuscript. We look forward to hearing from you in due course.
 

### 2017-02-03 response.tex
Paper changes + response
Added a reference to the paper as requested. First draft of response
letter.

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{amsmath,amssymb}
\usepackage{enumerate}
\usepackage{color}
\usepackage{amsthm,amsmath}
\usepackage{bbm}
\usepackage{mhchem}

%for stacking below a sum or max...
\usepackage{mathtools}

\usepackage[colorlinks]{hyperref}

\newcommand{\comment}[2]{\vspace{0.6cm}{\bf Comment:} {\it #1.}

\vspace{0.3cm}{\bf Answer:} #2}

\begin{document}
\title{Response to Referee Comments: Parallel Adaptive Importance Sampling}
\maketitle
%\section*{Introduction}
First of all, we would like to thank the referees for
their time in reading and commenting on our manuscript. In this document
we will address each of the points raised, and describe any relevant
changes that have been made to the paper. The changes made have been substantial, in particular with the addition of a 5-dimensional multimodal example arising from mixture models.

\section*{Referee 1 comments}
\comment{My strong suspicion is that the scheme is convergent when both the number of copies (N) and the number of MCMC iterations goes to infinity ... but this isn't shown here.}{We thank the referee for this comment. We have added a new section in the manuscript which points to proofs of both of the types of convergence that the referee alludes to.}

\comment{It seems likely that the paper was distilled from a Ph.D. thesis ... it's a bit longer than it has to be and includes some boilerplate background material. But other than that it's well written.}{The paper was certainly lengthy, and as such we have distilled it down somewhat. We have added a new example, but have removed the biochemical network example, which required lengthy introduction, and one of the earlier 1-dimensional examples. The introduction has been completely rewritten, and includes more material which is directly applicable to the paper. We hope that these extensive changes have made for a better paper with a slicker presentation. Even with other additions, such as the section on consistency, we have reduced the length of the paper to a leaner 24 pages.}

\comment{The examples are really basic and seem aimed to describe implementation rather than to convince us that the method will work. On more complicated problems I'd be concerned that the weights would be very degenerate and the resampling step would collapse all the walkers on top of each other ... in that case the different chains would be very highly correlated and you'd be doing a lot of work for no reason.}{We agree that if the proposal kernels are chosen poorly or with a poor choice of parameters, then problems can exist. All poor importance sampling schemes will result in the behaviour described. It is important that the proposal kernels are absolutely continuous with respect to the posterior distribution, and it is important that the variance of the kernels is appropriately chosen such that the mixture proposal distribution is slightly overdispersed. This has been further discussed and highlighted in the new submission.

We concede that the examples presented were relatively simple. We have removed a couple of these, and have included a new example arising from mixture models. The posterior in the new example is 5-dimensional, and has two modes which are well separated. Metropolis-Hasting algorithms really struggle in these scenarios, since the waiting time for the chains to switch modes is very large. Since many transitions are required for convergence, the convergence of these methods is glacial. We demonstrate the PAIS's aptitude for this type of problem with the new example, which we believe greatly improves the paper.}

\comment{Even if that work is done in parallel you have to compare to the same number of independent chains.}{As we hope is clear, all the comparisons made between Metropolis-Hastings and their PAIS equivalents are made with the same number of chains in both. We are not comparing our method with one serial chain of MH, as is common in most of the other parallel MCMC literature. What stands out with this method is the fact that we get superlinear improvement in the number of iterations that are required for convergence with respect to the number of chains used.}

\section*{Referee 2 comments}

\comment{The work in [1] on parallel samplers for climate model parameter estimation is definitely relevant here, but has not been cited. Same goes for the adaptive MCMC work of Haario, e.g. [2].}{We thank the referee for these suggestions which we have now included}

\comment{The "pre-fetching" way of parallelizing samplers (where the possible future acceptances/rejections are calculated in advance) would also be worth mentioning here.}{We have added a reference to this approach in the introduction.}

\comment{The numerical examples are very simple (only 1d and 2d), and somewhat fail to convince me that this is a good approach.}{As described above in the response to the first referee, a new 5-dimensional multimodal example has been added in which MH algorithms really struggle. We feel this demonstrates the significant advantages of this method.}

\comment{How would this scale to higher dimensional problems? Finding a good importance function is hard in high dimensions.}{This approach, like all importance sampling schemes, suffers from the curse of dimensionality. As the dimension increases, the number of particles which are required in order to achieve a good proposal distribution increases fast, and as such high-dimensional problems are beyond the scope of this work. We have rewritten the introduction to emphasise that this approach is more suited to low-dimensional challenging posterior distributions for which Metropolis-Hastings methods fail to converge quickly. This is demonstrated in our new 5-dimensional example as described above.}

\comment{The adaptation only aims at learning the scale of the proposal, whereas the shape can be as (or more) important.}{We partly disagree with this comment, since the positions of the chains themselves give us the shape of the posterior distribution. This said, maybe the referee was referring to the shape of the individual proposal kernels which together make up the mixture proposal distribution. In the situation where the posterior distribution is highly concentrated on a lower dimensional manifold, as is ubiquitous in many applications, such as inverse problems for biochemical networks and epidemiology, then if the shape of this correlation is not learned by the algorithm, then the result of the standard PAIS approach may not be optimal. We are currently working on a new paper where we use PAIS to sample from a transformation of the posterior distribution. This transformation is an approximation of the transformation which maps the posterior to a standardised N-dimensional Gaussian, which naturally is easy to sample from. This is an extensive piece of work, and beyond the scope of this paper.}

\end{document}

### 2017-02-03 response.tex
Final spellchecks

@@
 All poor importance sampling schemes will result in the behaviour described.
-It is important that the proposal kernels are absolutely continuous with respect to the posterior distribution, and it is important that the variance of the kernels is appropriately chosen such that the mixture proposal distribution is slightly overdispersed.
+It is important that the proposal kernels are absolutely continuous with respect to the posterior distribution, and it is important that the variance of the kernels is appropriately chosen such that the mixture proposal distribution is slightly over-dispersed.
 This has been further discussed and highlighted in the new submission.
@@
 
-\comment{The "pre-fetching" way of parallelizing samplers (where the possible future acceptances/rejections are calculated in advance) would also be worth mentioning here.}{We have added a reference to this approach in the introduction.}
+\comment{The "pre-fetching" way of parallelising samplers (where the possible future acceptances/rejections are calculated in advance) would also be worth mentioning here.}{We have added a reference to this approach in the introduction.}
 

### 2017-02-03 cover_letter.tex
Final spellchecks

@@
 Moreover, their reviews have in our view been unfairly coloured by the previous reviews from the SISC submission which were provided to them.
-In particular, they criticise us due to \emph{"the computed example provides no information about whether or not the proposed method will work in the (sizable) applications mentioned in your introduction."} However, all mentions of such sizable applications have been completely removed from the introduction.
+In particular, they criticise us due to \emph{"the computed example provides no information about whether or not the proposed method will work in the (sizeable) applications mentioned in your introduction."} However, all mentions of such sizeable applications have been completely removed from the introduction.
 The referees were responding to the reviews of the original paper which were sent to them, rather than the new paper, where these issues are not present.

### 2018-01-15 response_JUQ.tex
Starting edits

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{amsmath,amssymb}
\usepackage{enumerate}
\usepackage{color}
\usepackage{amsthm,amsmath}
\usepackage{bbm}
\usepackage{mhchem}

%for stacking below a sum or max...
\usepackage{mathtools}

\usepackage[colorlinks]{hyperref}

\newcommand{\comment}[2]{\vspace{0.6cm}{\bf Comment:} {\it #1.}

\vspace{0.3cm}{\bf Answer:} #2}

\begin{document}
\title{Response to Referee Comments: Parallel Adaptive Importance Sampling}
\maketitle
First of all, we would like to thank the referees for
their time in reading and commenting on our manuscript. In this document
we will address each of the points raised, and describe any relevant
changes that have been made to the paper.

\section*{Referee 1 comments}

\comment{The PAIS algorithm does exhibits interesting ideas to combine the information 
obtained from an ensemble of proposals. Furthermore, it seems quite flexible 
since several proposals and re-sampling methods can be used.

However, the examples somehow fail to convince that the algorithm can be used 
in a wide range of challenging inverse problems. The approach is limited by the 
dimensionality of the problem.}{We thank the referee for their comments, but regarding the final comments, we can't agree. We are very open about the fact that this approach cannot be used for high dimensional problems, but there are a plethora of lower dimensional inverse problems which exhibit structure in the posterior, such as multimodality or high correlations, for which existing MCMC methods struggle to converge in reasonable time. Dimensionality is not the only challenge faced in inverse problems, and these methods have been developed to tackle problems of a different nature. In particular, this method was designed with a view to sampling from inverse problems arising in Biology, where challenging posteriors exist on relatively low dimensional state spaces. We have added some words on this to clarify further the intended type of problem that this method is intended for in Section ****.}

\comment{For mixture problems, if the mixture has to many 
elements (or the number of elements is unknown), then this scheme will have 
difficulties.}{This is true, as would RWMH, MALA, HMC and a whole range of other methods. I would be happy for the referee to indicate a method which would comfortably deal with a mixture problem with a large number of possible known/unknown components. The example was chosen to demonstrate the remarkable performance of the algorithm (with respect to standard MCMC methods) for multimodal posteriors, and we believe it has achieved that. We have added a sentence to the end of the example to discuss the point that the referee makes.}

\comment{Also, it is not clear how it will behave in real life inverse 
problems where the likelihood is determined by complex physical systems that 
involve, for example, ODE's or PDE's. If the likelihood is expensive to 
evaluate, the tunning of the scaling parameters for the PAIS-RW might be a 
concern when the number of chains is big.}{The tuning of the scaling parameters is actually very fast, often converging within a few percentage points of the optimum within the order of $10^2$ iterations. We have included some plots to demonstrate this point, and we thank the referee for bringing this issue to our attention. This is quicker than one might expect for an adaptive serial algorithm. Moreover, in a previous iteration of the manuscript, we included an example where the dynamical system of concern was a continuous time Markov chain, but this was removed following previous referee comments; we would be more than happy to reintroduce this example should the referee wish us to.}

\comment{An real data example for the PAIS algorithm is much needed. It will be 
interesting to see how PAIS is compared not only against a naive 
parallelization, but also compared against a state of the art MCMC algorithm 
or/and a efficient ad hoc MCMC algorithm designed for such example.}{This is the problematic one I think. Very unhelpful. There are untold methodological papers without ``real life data'', such a problem requires building up a relationship with a non-mathematician, getting data, cleaning it up. It's a LOT of work.}

\comment{There aren't details about the architecture used to run the M chains in 
parallel. This is relevant since there are several options such as: multi-core 
and multi-processor computers, GPU cards, clusters, cloud computing, etc. 
Comments about the advantages and disadvantages of the architecture 
are also welcome. The same goes for details regarding the programming language 
used.}{We thank the referee for this comment, as we should have included these details in the manuscript. The method was implemented in C++, serially, and was run on a single core of a *****server spec here*****. We are not computer scientists, and as such, we are not in a position to comment on specific architectures. If this algorithm was to be spread over several cores, with each core computing the weight of one/more ensemble members, then issues of communication between cores would need to be addressed. Communication is only required at the point of resampling, where new positions of weights are communicated to a single core that computes the resampling, and the new ensemble positions are then sent back out. Some comments on this have been added to Section ****. However, following comments by the second referee, we have decided to reduce the emphasis on the parallelisation of the algorithm, not least because we have not implemented it in parallel ourselves. Please see our comments to the second referee for further details.}

\comment{CPU time must be reported to compare the gain of each algorithm in 
order to have a fair comparison. Authors claim that there is a significant 
gain in terms of the number of iterations needed to achieve a desire 
tolerance; examples in sections 7.2 and 7.3 show that PAIS saves up to 90\% of 
the iterations. However, if the CPU time needed to run 10 PAIS iterations is 
ten times the time needed to run 100 RWMH iterations, then the gain is not as 
good as presented. PAIS iterations are more time consuming than RWMH iterations 
due to the re-sampling step. The authors mentioned that the re-sampling step is 
computationally expensive and that is why they proposed the AMR re-sampling 
method. CPU time is of particular importance in example 7.4 since the number of 
parallel chains grows from 50 to 500.}{The referee may have missed Figure **** in which we show the computational cost, in time, of the resamplers that we use in the paper, both the optimal transport resampler, and the AMR. As mentioned before in response to another of the referees comments, we had neglected to share the specs of the machine that these times were recorded on, and this has now been added. There is an additional cost in the algorithm over and above the resampler, which is the computation of the denominator in the weights, as pointed out by the second referee. If we have $M$ members of our ensemble, then the cost of this denominator across the whole ensemble is $M^2$, but it can be easily parallelised. We include some timings of the cost of this aspect of the algorithm in Section ****, and a discussion of the overall additional cost of the approach. Coupled with the timings presented for the resampler, we believe that this now gives the reader a clear idea of the additional overhead costs which are incurred when using our method as opposed to a standard MCMC method. We thank the referee for their comments on this.}

\section*{Referee 2 comments}

\comment{This work in an interesting contribution in the field of Monte Carlo algorithms. The paper is technically sound and well-written. 
Moreover, it is possible to note the effort nd the care devoted by the authors.}{We thank the referee for their kind comments.}

\comment{However, I have some concerns.

1) The starting-point algorithm in Table 1, is very similar to the basic scheme suggested in, for instance,

V. Elvira, L. Martino, D. Luengo, M. F. Bugallo, Improving Population Monte Carlo: Alternative Weighting and Resampling Schemes, Signal Processing Vol. 131, pp. 77-91, 2017.

In practice, you use a deterministic approach for the IS weights. 
I believe that the authors have analyzed and studied this idea independently, but this similarity should be discussed in the introduction, at least. 
You can also point out that you also propose some different resampling or adaptation schemes that are not contained in the previous work.}{We agree that there are similarities to this other work, and we thank the referee for their comment. Additional references have been added, along with more detailed discussions about the novel aspects of this manuscript.}

\comment{2) I do not understand why you use the word "parallel" in the name of the algorithm. In Algorithm 1, it is difficult to recognize the parallelization. The denominator in the weight that you employ, in fact, is a bottleneck for a possible parallelization. Please clarify this point (even in the text).}{We agree with this comment from the referee, as it may not have been made clear enough. With an ensemble of states, it is possible to distribute the computation of the weights across a set of cores. We have added a comment on this to Section ****. However, as discussed in the response to the first referee, we ourselves implemented the algorithm serially. Therefore, we have decided to change the name of the algorithm to reflect this, to ``Optimal Transport Adaptive Importance Sampling'' or OTAIS. Changes throughout the manuscript have been made to reflect this.}

\comment{3) Please, add a table of acronyms that can help the reader. Now it is a bit confusing.}{We have added a glossary of acronyms to an appendix.}

\comment{4) The references should be completed with other important contributions (considering also the same kind of weights, in some cases):

- O.Cappe, R.Douc, A.Guillin, J.M.Marin, C.P.Robert, Adaptive importance sampling in general mixture classes, Statistics and Computing 18 (2008) 447-459. 
- R. Douc, A. Guillin, J. M. Marin, C. P. Robert, Convergence of adaptive mixtures of importance sampling schemes, Annals of Statistics 35 (2007) 420-448. 
- R. Douc, A. Guillin, J. M. Marin, C. P. Robert, Minimum variance importance sampling via population Monte Carlo, ESAIM: Probability and Statistics 11 (2007) 427-447. 
- L. Martino, V. Elvira, D. Luengo, J. Corander, Layered Adaptive Importance Sampling, Statistics and Computing, Vol. 27 (3), pp. 599-623, 2017.}{We thank the referee for thier suggestions for other highly relevant references.}

\end{document}

### 2018-01-25 response_JUQ.tex
Updated new response letter

@@
 \maketitle
-First of all, we would like to thank the referees for
+First of all, we would like to thank the referees and the associate editor for
 their time in reading and commenting on our manuscript. In this document
@@
 in a wide range of challenging inverse problems. The approach is limited by the 
-dimensionality of the problem.}{We thank the referee for their comments, but regarding the final comments, we can't agree.
+dimensionality of the problem.}{We thank the referee for their comments.
 We are very open about the fact that this approach cannot be used for high dimensional problems, but there are a plethora of lower dimensional inverse problems which exhibit structure in the posterior, such as multimodality or high correlations, for which existing MCMC methods struggle to converge in reasonable time.
 Dimensionality is not the only challenge faced in inverse problems, and these methods have been developed to tackle problems of a different nature.
-In particular, this method was designed with a view to sampling from inverse problems arising in Biology, where challenging posteriors exist on relatively low dimensional state spaces.
+In particular, this method was designed with a view to sampling from inverse problems arising in Biology, where challenging posteriors exist on relatively low dimensional state spaces, and for which the likelihood can be very expensive to compute.
 We have added some words on this to clarify further the intended type of problem that this method is intended for in Section ****.}
@@
 elements (or the number of elements is unknown), then this scheme will have 
-difficulties.}{This is true, as would RWMH, MALA, HMC and a whole range of other methods.
+difficulties.}{This is true, as would state-of-the-art Riemannian manifold MALA, HMC and a whole range of other methods.
 I would be happy for the referee to indicate a method which would comfortably deal with a mixture problem with a large number of possible known/unknown components.
-The example was chosen to demonstrate the remarkable performance of the algorithm (with respect to standard MCMC methods) for multimodal posteriors, and we believe it has achieved that.
+The example was chosen to demonstrate the remarkable performance of the algorithm (with respect to standard MCMC methods) for multimodal posteriors in low dimensions, and we believe it has achieved that.
 We have added a sentence to the end of the example to discuss the point that the referee makes.}
@@
 concern when the number of chains is big.}{The tuning of the scaling parameters is actually very fast, often converging within a few percentage points of the optimum within the order of $10^2$ iterations.
-We have included some plots to demonstrate this point, and we thank the referee for bringing this issue to our attention.
-This is quicker than one might expect for an adaptive serial algorithm.
-Moreover, in a previous iteration of the manuscript, we included an example where the dynamical system of concern was a continuous time Markov chain, but this was removed following previous referee comments; we would be more than happy to reintroduce this example should the referee wish us to.}
+We have included some plots and discussion to Section **** to demonstrate this point, and we thank the referee for bringing this issue to our attention.
+This is quicker than one might expect for an adaptive serial algorithm.}
 
@@
 parallelization, but also compared against a state of the art MCMC algorithm 
-or/and a efficient ad hoc MCMC algorithm designed for such example.}{This is the problematic one I think.
-Very unhelpful.
-There are untold methodological papers without ``real life data'', such a problem requires building up a relationship with a non-mathematician, getting data, cleaning it up.
-It's a LOT of work.}
+or/and a efficient ad hoc MCMC algorithm designed for such example.}{Unfortunately we do not have the resources to apply our method to a full-scale real data example.
+However, following additional comments from the editor, we have included a new example where the likelihood requires the solution of an ODE.
+In Section **** we include an example of data assimilation for the Lorenz*** equations.
+Expensive likelihoods actually make this method more applicable, since they make the extra overheads of the resampling step negligible.
+This is precisely what the method is designed for, and is the very reason for us to compare the algorithms by looking at error as a function of the number of samples/likelihood evaluations.
+We have also added further discussion regarding this to Section ****.}
 
@@
 used.}{We thank the referee for this comment, as we should have included these details in the manuscript.
-The method was implemented in C++, serially, and was run on a single core of a *****server spec here*****.
-We are not computer scientists, and as such, we are not in a position to comment on specific architectures.
-If this algorithm was to be spread over several cores, with each core computing the weight of one/more ensemble members, then issues of communication between cores would need to be addressed.
-Communication is only required at the point of resampling, where new positions of weights are communicated to a single core that computes the resampling, and the new ensemble positions are then sent back out.
-Some comments on this have been added to Section ****.
+The method was implemented in C++, serially, and was run on a single core of a *****server spec here*****, and these details have been added in Section ***.
+We are not computer scientists, and as such, we are not in a position to comment on specific architectures, especially as the optimal architecture would be problem specific.
+The optimal architecture would likely depend on the relative costs of communication of the states across the processors, the cost of the likelihood evaluation, and the cost of the resampler for the chosen ensemble size.
+We have added a discussion to Section ****.
 However, following comments by the second referee, we have decided to reduce the emphasis on the parallelisation of the algorithm, not least because we have not implemented it in parallel ourselves.
@@
 parallel chains grows from 50 to 500.}{The referee may have missed Figure **** in which we show the computational cost, in time, of the resamplers that we use in the paper, both the optimal transport resampler, and the AMR.
-As mentioned before in response to another of the referees comments, we had neglected to share the specs of the machine that these times were recorded on, and this has now been added.
-There is an additional cost in the algorithm over and above the resampler, which is the computation of the denominator in the weights, as pointed out by the second referee.
+As mentioned before in response to another of the referee's comments, we had neglected to share the specs of the machine that these times were recorded on, and this has now been added.
+% There is an additional cost in the algorithm over and above the resampler, which is the computation of the denominator in the weights, as pointed out by the second referee.
 If we have $M$ members of our ensemble, then the cost of this denominator across the whole ensemble is $M^2$, but it can be easily parallelised.
 We include some timings of the cost of this aspect of the algorithm in Section ****, and a discussion of the overall additional cost of the approach.
-Coupled with the timings presented for the resampler, we believe that this now gives the reader a clear idea of the additional overhead costs which are incurred when using our method as opposed to a standard MCMC method.
-We thank the referee for their comments on this.}
+% Coupled with the timings presented for the resampler, we believe that this now gives the reader a clear idea of the additional overhead costs which are incurred when using our method as opposed to a standard MCMC method.
+We thank the referee for their comments on this.
+
+The reason for analysing the error as a function of the number of the likelihood evaluations, is that this method is intended for problems with expensive likelihoods, for which these overhead costs would be dwarfed.
+We have added a discussion on this to Section ****.}
 
@@
 The denominator in the weight that you employ, in fact, is a bottleneck for a possible parallelization.
-Please clarify this point (even in the text).}{We agree with this comment from the referee, as it may not have been made clear enough.
+Please clarify this point (even in the text).}{We have added clarifications on this point to Section ****.
+The referee should note that in our convergence plots, the reason that we compare the algorithms with respect to the number of likelihood evaluations is that we envisaged that this method would be of most use when the likelihood is expensive, for example when it requires a high resolution approximation of the solution of a PDE, and/or there is a very large amount of data to be taken into account.
+In this context, it would be useful to spread the work of computing the likelihood for the ensemble members across a cluster, hence our use of the word ``parallel''.
+We acknowledge and thank the referee for indicating the issues with the denominator.
+In the case where the likelihood is expensive, these overhead costs, along with that of the resampler, should no longer be the bottleneck.
+This was not adequately explained in the text, and as such we have added a further discussion in Section ****.
+
+% We agree with this comment from the referee, as it may not have been made clear enough.
 With an ensemble of states, it is possible to distribute the computation of the weights across a set of cores.
@@
 Therefore, we have decided to change the name of the algorithm to reflect this, to ``Optimal Transport Adaptive Importance Sampling'' or OTAIS.
-Changes throughout the manuscript have been made to reflect this.}
+Changes throughout the manuscript have been made to reflect this.
+}
 
@@
 
+\section*{Associate Editor comments}
+\comment{A very similar version of the authors’ PAIS algorithm has been studied before, as explained by the second referee.
+Moreover, I could argue that the PAIS algorithm is a particular case since weights are considered deter- ministic.
+This is a serious issue in the paper, let alone that the authors are unaware of the series of papers mentioned by referee 2, that are closely related to their work.
+This should be explained in detail and the paper original perspective and contributions be further highlighted.}{Following suggestions by the second referee and the associate editor, we have added a significant new section to the introduction to highlight the other ensemble importance schemes already in the literature.
+We have also highlighted further the contribution made by this manuscript in Section ***.}
+
+\comment{An other serious concern is why not use PMC-MCMC for the updating scheme in Algorithm 1.
+That would be as costly (since it also involves evaluating the posterior) but the yis will have the correct distribution.}{This comment is not clear to us - what is the ``updating scheme'' in Algorithm 1??
+Also, the (weighted) $y_i$ are distributed according to the posterior, as discussed in Section ****.
+We'd be happy to address this point if the AE is able to clarify it for us.}
+
+\comment{The ”automatic” tuning will be as useful as the what traditional adaptive tuning is: in the limit, the proposal cov matrix converges to the cov matrix of the target.
+This is ok for Gaussian type targets but will fail for multimodal, multiscale distributions, mostly present in non-trivial forward maps, as those suggested in section 2.
+Something similar happens with gradient based proposals, since from the onset they are not conceived to scape local maxima.}{This is an interesting point, and the subject of our most recent paper in this area, in which transport maps are used to stabilise and accelerate the PAIS methodology, and also addressing exactly the point raised by the AE.
+We have added a discussion on this topic in Section ***.}
+
+\comment{In Algorithm 1, note that in step 5 the actual number of different particles could be drastically reduced and in some cases collapsed to 1.
+As mentioned by the authors, this is a basic concern in SIR type algorithms, that they may become totally inefficient.
+Then, why this seems not to bother the authors and no provision was included in their algorithm to mend this problem?}{Provision is made within the code for such a catastrophic collapse of the weights, and it was remiss of us not to include details on this.
+We have added details on this to Section ****.}
+
+\comment{The Radon-Nokodyn version of Bayes theorem in (2.3) is unnecessary in the actual finite dimensional context the authors work with.
+The authors should state that, although there is an infinite dimensional version for the posterior distribution, their discussion is restricted to the parametric, finite dimensional case.}{We thank the AE for this comment and have amended accordingly.}
+
+\comment{Note that in (2.4) the total measure of the space for ηˆ is M (!).}{Thanks to the AE for picking up on this typo, this has now been fixed.}
+
 \end{document}

### 2018-01-25 previous_submission_2017.tex
Created old version for difftex later on

--- 2017-02-02 PAIS.tex
@@
 speed ups of up to two orders of magnitude when compared with serial
-methods on a single core.
+methods on a single core. Another approach involves pre-fetching, where the possible future acceptances/rejections are calculated in advance\cite{angelino2014accelerating}.
 
@@
 Figure~\ref{fig:MH1 L2} shows that the PAIS-RW algorithm converges
-tothe posterior distribution significantly faster than the RWMH
+to the posterior distribution significantly faster than the RWMH
 algorithm, in both $L^2$ error and relative error in the moments. A

### 2018-01-25 response_JUQ.tex
Referee 1 comments addressed
Apart from the Lorenz example.

@@
 In particular, this method was designed with a view to sampling from inverse problems arising in Biology, where challenging posteriors exist on relatively low dimensional state spaces, and for which the likelihood can be very expensive to compute.
-We have added some words on this to clarify further the intended type of problem that this method is intended for in Section ****.}
+We have added some words on this to clarify further the intended type of problem that this method is in the introduction, and again in the discussion at the end.}  %done
 
@@
 The example was chosen to demonstrate the remarkable performance of the algorithm (with respect to standard MCMC methods) for multimodal posteriors in low dimensions, and we believe it has achieved that.
-We have added a sentence to the end of the example to discuss the point that the referee makes.}
+We have added a paragraph to the end of the example to discuss the point that the referee makes.}%done
 
@@
 concern when the number of chains is big.}{The tuning of the scaling parameters is actually very fast, often converging within a few percentage points of the optimum within the order of $10^2$ iterations.
-We have included some plots and discussion to Section **** to demonstrate this point, and we thank the referee for bringing this issue to our attention.
-This is quicker than one might expect for an adaptive serial algorithm.}
+We have included some plots and discussion to the end of Section 4.2 to demonstrate this point, and we thank the referee for highlighting this issue.
+This is quicker than one might expect for an adaptive serial algorithm.}%done
 
@@
 However, following additional comments from the editor, we have included a new example where the likelihood requires the solution of an ODE.
-In Section **** we include an example of data assimilation for the Lorenz*** equations.
-Expensive likelihoods actually make this method more applicable, since they make the extra overheads of the resampling step negligible.
+In Section **** we include an example of data assimilation for the Lorenz*** equations, which is a set of equations which is commonly used to test methods for inference in the applied mathematics community.
+
+Expensive likelihoods actually make this method more applicable, since they make the extra overheads of the resampling step relatively less important.
 This is precisely what the method is designed for, and is the very reason for us to compare the algorithms by looking at error as a function of the number of samples/likelihood evaluations.
@@
 used.}{We thank the referee for this comment, as we should have included these details in the manuscript.
-The method was implemented in C++, serially, and was run on a single core of a *****server spec here*****, and these details have been added in Section ***.
+The method was implemented in C++, serially, and was run on a single core of a Dell server with four 8-core 3.3GHz CPUs, 64Gb memory, and these details have been added in Section 7.
+
 We are not computer scientists, and as such, we are not in a position to comment on specific architectures, especially as the optimal architecture would be problem specific.
 The optimal architecture would likely depend on the relative costs of communication of the states across the processors, the cost of the likelihood evaluation, and the cost of the resampler for the chosen ensemble size.
-We have added a discussion to Section ****.
-However, following comments by the second referee, we have decided to reduce the emphasis on the parallelisation of the algorithm, not least because we have not implemented it in parallel ourselves.
-Please see our comments to the second referee for further details.}
+We have added a discussion to Section 3, following the presentation of the PAIS algorithm.}
 
@@
 method. CPU time is of particular importance in example 7.4 since the number of 
-parallel chains grows from 50 to 500.}{The referee may have missed Figure **** in which we show the computational cost, in time, of the resamplers that we use in the paper, both the optimal transport resampler, and the AMR.
+parallel chains grows from 50 to 500.}{The referee may have missed Figure 5.1 in which we show the computational cost, in time, of the resamplers that we use in the paper, both the optimal transport resampler, and the AMR.
 As mentioned before in response to another of the referee's comments, we had neglected to share the specs of the machine that these times were recorded on, and this has now been added.
@@
 The reason for analysing the error as a function of the number of the likelihood evaluations, is that this method is intended for problems with expensive likelihoods, for which these overhead costs would be dwarfed.
-We have added a discussion on this to Section ****.}
+We have added a discussion on this to the beginning of Section 7.}%done
 

### 2018-01-25 PAIS.tex
Paper changes + response
Added a reference to the paper as requested. First draft of response
letter.
Referee 1 comments addressed
Apart from the Lorenz example.

@@
 transport which has been used in the context of particle
-filters\cite{reich2013nonparametric}. In particular we demonstrate the
+filters\cite{reich2013nonparametric}. We also detail how algorithmic
+parameters can be quickly tuned into optimal regimes, with respect to
+the effective sample size statistic. In particular we demonstrate the
 advantages of this method when attempting to sample from multimodal
 posterior distributions, such as those arising from inference for
-mixture models.
+mixture models. The method that we present is specifically designed to
+tackle the challenges of certain types of low-dimensional inverse
+problem which have complex posterior structure, for example
+multimodality or strong correlations, and expensive
+likelihood evaluations. We will demonstrate that this methodology
+allows us to sample from such a posterior distribution with greater
+precision per likelihood evaluation than an uncorrelated ensemble of
+standard MCMC chains of the same type.
 
@@
 we will analyse.
+
+In each iteration of the algorithm, we are required to compute the
+likelihood for each member of the ensemble. In the case where the
+likelihood is expensive to compute, for example because it requires
+the numerical approximation of an ODE or PDE, and/or there is a very
+large amount of observations in the data, it will be expedient to
+split this computational effort across multiple cores. Whether this is
+efficient will depend on the architecture of the machine, and the
+relative cost of communication across the cores, to the cost of the
+likelihood evaluation. Evaluation of the denominator in the weights
+also contributes to the overhead costs of this approach over and above
+a standard MCMC method, but this too can be parallelised efficiently,
+depending on communication costs of the architecture of the machine.
 
@@
 
+\begin{figure}[!htb]
+\centering
+\subfigure[APAIS-RW algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISRW_adapt_var"}}
+\subfigure[APAIS-RW algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISRW_adapt_ess"}}
+\subfigure[APAIS-MALA algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISMALA_adapt_var"}}
+\subfigure[APAIS-MALA algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISMALA_adapt_ess"}}
+\caption{Demonstration of the convergence of the adaptive scaling parameter in the adaptively tuned PAIS algorithms.}
+\label{fig:G1_adapt_trace_PAIS}
+\end{figure}
+
+In the context of an expensive likelihood evaluation, slow convergence
+of the algorithmic parameters in this adaptive regime would be a
+concern. However, this is rarely an issue, due in part to the speed at
+which the PAIS algorithm completes the burn-in phase. Figure
+\ref{fig:G1_adapt_trace_PAIS} demonstrates this for two different
+versions of the PAIS methodology with a RW proposal and a MALA\cite{roberts1998optimal}
+proposal. This picture is typical in our experience, with the scaling
+parameters converging in $\mathcal{O}(10^2)$ iterations to an optimal regime.
 %%%%%%
@@
 measured in seconds, scales with the ensemble size for the three
-different methods. These results demonstrate that the AMR behaves how we wish, and
+different methods, where timings have been taken from simulations on a Dell
+server with four 8-core 3.3GHz CPUs and 64Gb memory. These results demonstrate that the AMR behaves how we wish, and
 importantly ensures that exactly one sample of the output will lie in
@@
 \section{Numerical Examples}\label{Sec:Num}
+The numerical examples in this section were all computed on a single
+core of Dell
+server with four 8-core 3.3GHz CPUs and 64Gb memory. Throughout we
+measure the efficiency of the algorithms in terms of the number of
+iterations (i.e. likelihood evaluations) which are required in order
+to reach a given order of accuracy. The reason for this is that this
+method is designed with a particular type of challenging inverse
+problem in mind, namely one which is low dimensional, has an expensive
+likelihood (such as a PDE solve, and/or very large data) which dwarfs
+the overhead cost of PAIS, and which could have complex posterior
+structure, such as correlations, multimodality or other highly
+non-Gaussian features.
 
@@
 
-Convergence of the relative error for the two algorithms is displayed in Figure~\ref{fig:mixture_modes}.
-PAIS converges at the expected $\mathcal{O}(1/\sqrt{N})$ rate, whereas the MH algorithm converges to locally smooth histograms but with the wrong proportion of samples in each mode.
-The relatively low value of the error for the MH example is due to the priors covering the sample space evenly, however since transitions are near impossible with a small value of the scaling parameter, this error will take a very long time to reduce.
-This problem was discussed in Section~\ref{sec:bimodal}.
+Convergence of the relative error for the two algorithms is displayed
+in Figure~\ref{fig:mixture_modes}. PAIS converges at the expected
+$\mathcal{O}(1/\sqrt{N})$ rate, whereas the MH algorithm converges to
+locally smooth histograms but with the wrong proportion of samples in
+each mode. The relatively low value of the error for the MH example is
+due to the priors covering the sample space evenly, however since
+transitions are near impossible with a small value of the scaling
+parameter, this error will take a very long time to reduce. This
+problem was discussed in Section~\ref{sec:bimodal}.
+
+In this example, we have only considered a relatively low dimensional
+mixture model problem, with only two distributions in the
+mixture. With more elements in the mixture, and/or an undefined number
+in the mixture, the dimension of this problem will quickly
+increase. Importance sampling schemes such as this suffer from the
+curse of dimensionality, limiting the size of the dimension of target
+density that it can efficiently sample from. However, this example
+does demonstrate the remarkable convergence of the PAIS for multimodal
+target distributions.
 
@@
 the ensemble members to regions which require more exploration. This allows the
-method to be used to sample from complex multimodal distribution.
+method to be used to sample from complex multimodal distributions.
 
@@
 
+We have demonstrated that the framework that we have considered, with
+the use of state-of-the-art optimal transport-based resampling, can
+reduce the number of likelihood evaluations required to characterise
+complex posterior distributions in low dimensions to a given
+degree. We have also introduced a greedy approximation to this
+resampler, which drastically reduces the cost, at the loss of some
+accuracy, which can be mitigated with the use of larger ensembles.
+
 \bibliographystyle{siam}

### 2018-01-25 response_JUQ.tex
Referee 2 and AE comments addressed

@@
 You can also point out that you also propose some different resampling or adaptation schemes that are not contained in the previous work.}{We agree that there are similarities to this other work, and we thank the referee for their comment.
-Additional references have been added, along with more detailed discussions about the novel aspects of this manuscript.}
+Additional references have been added, along with more detailed discussions about the novel aspects of this manuscript in the introduction and discussion.}
 
@@
 The denominator in the weight that you employ, in fact, is a bottleneck for a possible parallelization.
-Please clarify this point (even in the text).}{We have added clarifications on this point to Section ****.
+Please clarify this point (even in the text).}{We have added clarifications on this point to Section 3.
 The referee should note that in our convergence plots, the reason that we compare the algorithms with respect to the number of likelihood evaluations is that we envisaged that this method would be of most use when the likelihood is expensive, for example when it requires a high resolution approximation of the solution of a PDE, and/or there is a very large amount of data to be taken into account.
@@
 In the case where the likelihood is expensive, these overhead costs, along with that of the resampler, should no longer be the bottleneck.
-This was not adequately explained in the text, and as such we have added a further discussion in Section ****.
+This was not adequately explained in the text, and as such we have added a further discussion in Section 3.
 
@@
 
-\comment{3) Please, add a table of acronyms that can help the reader. Now it is a bit confusing.}{We have added a glossary of acronyms to an appendix.}
+\comment{3) Please, add a table of acronyms that can help the reader. Now it is a bit confusing.}{We have added a glossary of acronyms to an appendix.} %half done -  complete
 
@@
 27 (3), pp.
-599-623, 2017.}{We thank the referee for thier suggestions for other highly relevant references.}
+599-623, 2017.}{We thank the referee for thier suggestions for other highly relevant references.} %done
 
@@
 This should be explained in detail and the paper original perspective and contributions be further highlighted.}{Following suggestions by the second referee and the associate editor, we have added a significant new section to the introduction to highlight the other ensemble importance schemes already in the literature.
-We have also highlighted further the contribution made by this manuscript in Section ***.}
+We have also highlighted further the contribution made by this manuscript in the introduction and discussion sections.}%done
 
@@
 That would be as costly (since it also involves evaluating the posterior) but the yis will have the correct distribution.}{This comment is not clear to us - what is the ``updating scheme'' in Algorithm 1??
-Also, the (weighted) $y_i$ are distributed according to the posterior, as discussed in Section ****.
-We'd be happy to address this point if the AE is able to clarify it for us.}
+Also, the (weighted) $y_i$ are distributed according to the posterior, as discussed in Section 6.
+We'd be happy to address this point if the AE is able to clarify it for us.}%nothing to do
 
@@
 Something similar happens with gradient based proposals, since from the onset they are not conceived to scape local maxima.}{This is an interesting point, and the subject of our most recent paper in this area, in which transport maps are used to stabilise and accelerate the PAIS methodology, and also addressing exactly the point raised by the AE.
-We have added a discussion on this topic in Section ***.}
+We have added a discussion on this topic in the discussion section, proposing it as an area for future study (our new manuscript will be completed in the next few weeks).}%done
 
@@
 Then, why this seems not to bother the authors and no provision was included in their algorithm to mend this problem?}{Provision is made within the code for such a catastrophic collapse of the weights, and it was remiss of us not to include details on this.
-We have added details on this to Section ****.}
+We have added details on this to the end of Section 3.}%done
 
 \comment{The Radon-Nokodyn version of Bayes theorem in (2.3) is unnecessary in the actual finite dimensional context the authors work with.
-The authors should state that, although there is an infinite dimensional version for the posterior distribution, their discussion is restricted to the parametric, finite dimensional case.}{We thank the AE for this comment and have amended accordingly.}
+The authors should state that, although there is an infinite dimensional version for the posterior distribution, their discussion is restricted to the parametric, finite dimensional case.}{We thank the AE for this comment.
+There is no need for this version of Bayes' theorem here, and so we have switched to the finite dimensional version.}%done
 
-\comment{Note that in (2.4) the total measure of the space for ηˆ is M (!).}{Thanks to the AE for picking up on this typo, this has now been fixed.}
+\comment{Note that in (2.4) the total measure of the space for ηˆ is M (!).}{Thanks to the AE for picking up on this typo, this has now been fixed.}%done
 

### 2018-01-25 PAIS.tex
Referee 2 and AE comments addressed

@@
 this optimisation is restricted by the component kernel(s) chosen, and
-the quality of the current sample from the posterior. Extensions, such
-as adaptive multiple importance sampling algorithm
+the quality of the current sample from the posterior\cite{cappe2008adaptive,douc2007convergence,douc2007minimum}. Extensions, such
+as layered adaptive importance sampling\cite{martino2017layered}, adaptive multiple importance sampling algorithm
 (AMIS)~\cite{cornuet2012adaptive}, and adaptive population importance sampling
@@
 
+The method considered is similar to population Monte Carlo methods
+which have been previously studied, for example in
+\cite{elvira2017improving}. However our approach exploits the current
+state-of-the-art resamplers. Higher accuracy resamplers, that provide
+a deterministic resample which optimally reflects the statistics of
+the original sample, lead to better mixture approximations of the
+posterior. These approximations can then be used to construct stable
+and efficient importance sampling proposals. The cost of the
+state-of-art resamplers can be prohibitive for large ensemble sizes,
+which are necessary for sampling in moderately higher dimensions, and
+as such we also present a greedy approximation of the optimal
+transport resampler, which provides us with good results for a
+fraction of the cost. We also detail how scaling parameters in the
+MCMC proposals can be quickly tuned, leading us to an efficient and
+fully-automated algorithm.
+
 In Section \ref{Sec:Prelim} we introduce some mathematical
@@
 \end{equation}
-where $\| y_1 - y_2 \|_\Sigma = (y_1-y_2)^\top\Sigma^{-1}(y_1-y_2)$ for $y_1,y_2 \in \mathbb{R}^d$.
-
-As discussed in \cite{stuart2010inverse,cotter2009bayesian},
-in order for this inverse problem to be well-posed in the Bayesian
-sense, we require the posterior distribution, $\mu_Y$, to be absolutely
-continuous with respect to the prior, $\mu_0$. A
-minimal regularity prior can be chosen informed by regularity results
-of the observational operator $\mathcal{G}$. Given such a prior, then
-the Radon-Nikodym derivative of the posterior measure, $\mu_Y$, with
-respect to the prior measure, $\mu_0$, is proportional to the
-likelihood:
-\begin{equation}\label{eqn:RND}
-	\frac{d\mu_Y}{d\mu_0} \propto \exp \left ( -\Phi(x^*) \right ).
-\end{equation}
-
-%%%%%%
+where $\| y_1 - y_2 \|_\Sigma = (y_1-y_2)^\top\Sigma^{-1}(y_1-y_2)$
+for $y_1,y_2 \in \mathbb{R}^d$.
+
+Then, according to Bayes' theorem, the posterior distribution that we
+are interested in characterising is given, up to a constant of proportionality, by
+\begin{equation*}
+	\pi_Y \propto \exp \left ( -\Phi(x^*) \right )\pi_0(x^*),
+\end{equation*}
+where $\pi_0$ is the prior density on the quantity of interest $x^*$.
+
+% As discussed in \cite{stuart2010inverse,cotter2009bayesian},
+% in order for this inverse problem to be well-posed in the Bayesian
+% sense, we require the posterior distribution, $\mu_Y$, to be absolutely
+% continuous with respect to the prior, $\mu_0$. A
+% minimal regularity prior can be chosen informed by regularity results
+% of the observational operator $\mathcal{G}$. Given such a prior, then
+% the Radon-Nikodym derivative of the posterior measure, $\mu_Y$, with
+% respect to the prior measure, $\mu_0$, is proportional to the
+% likelihood:
+% \begin{equation}\label{eqn:RND}
+% 	\frac{d\mu_Y}{d\mu_0} \propto \exp \left ( -\Phi(x^*) \right ).
+% \end{equation}
+
+% %%%%%%
 
@@
 	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
-	 n_i\delta_{x_i}(\cdot) = \hat{\eta}.
+	 \frac{1}{M} \delta_{x_i}(\cdot) = \hat{\eta}.
 \end{equation}
-where $n_i \in \{0, 1, \dots,M\}$ and $\sum_{i=1}^M \! n_i = M$. This
+where $n_i \in \{0, 1, \dots,M\}$. This
 new sample $\{x_i\}_{i=1}^M$ is equally weighted.
@@
 a standard MCMC method, but this too can be parallelised efficiently,
-depending on communication costs of the architecture of the machine.
+depending on communication costs of the architecture of the machine,
+reducing the $\mathcal{O}(M^2)$ cost to $\mathcal{O}(M)$ over $M$
+cores, for example.
 
@@
 drastically decrease the efficiency of sampling.
+
+It is also possible, especially during the early stages of the
+algorithm, where the mixture proposal is a poor approximation of the
+target distribution, for a sample to be produced with a weight which
+is orders of magnitude bigger than the rest. This is a problem, since
+the resampling step will then lead to a vector of particles all
+centred at the outlier sample. Fortunately these spikes in weights can
+easily be detected, and in such a case, the sample for this iteration
+can be removed. The ETPF can then be used to formulate a new mixture
+sample with one particle placed at the outlier point, and the others
+distributed according to the particle positions before the problematic
+iteration. This ensures that the region where the problem occurred is
+better represented by the mixture, and prevents further spikes in this region.
 
@@
 resampler, which drastically reduces the cost, at the loss of some
-accuracy, which can be mitigated with the use of larger ensembles.
+accuracy, which can be mitigated with the use of larger
+ensembles. Lastly, we have detailed how scaling parameters in the MCMC
+proposals that are used within the mixture distribution can be quickly
+and efficiently
+tuned in an automated way. However, tuning the variances and
+covariances of the mixture proposal components
+globally is likely to be of limited use for multimodal problems where
+the modes have very different covariances, or where there is curved
+ridges in the density. There are a variety of possible solutions to
+this problem that would be worthy of future consideration, not least
+the potential to use transport maps\cite{el2012bayesian,parno2014transport} to map Gaussian mixtures to
+highly complex non-Gaussian approximations of the posterior. Such a
+map could encode local covariance information, and lead to accelerated
+and stable PAIS sampling. This is an interesting area for future study.
+
+\begin{appendix}
+
+\section{Glossary of acronyms}
+
+MCMC
+RWMH
+MALA
+AIS
+PMC
+PAIS
+ETPF
+AMR
+PAIS-RW
+PAIS-MALA
+APAIS-RW
+APAIS-MALA
+
+\end{appendix}
 

### 2018-02-10 PAIS.tex
Added Lorenz example

@@
 
+\subsection{Data assimilation with Lorenz `63 trajectories}\label{sec:lorenz}
+
+The sampling algorithm we have introduced in this paper is designed for problems where the likelihood density function is expensive to calculate, and the computational overhead involved in the calculation of the importance weights is dwarfed.
+This situation commonly occurs in inverse problems where the model being investigated involves the solution of a differential equation.
+In this example we attempt to recover the initial position of a particle with motion governed by Lorenz's 1963 atmospheric convection equations~\cite{lorenz1963deterministic}:
+\begin{align*}
+	\frac{\text{d}x}{\text{d}t} &= \sigma(y-x), \\
+	\frac{\text{d}y}{\text{d}t} &= x(\rho-z)-y, \\
+	\frac{\text{d}z}{\text{d}t} &= xy - \beta z.
+\end{align*}
+When the parameters are chosen to be $\rho = 28$, $\sigma=10$ and $\beta=8/3$ this system has chaotic solutions.
+If we are interesting in calculating the initial position by observing the location of the particle at certain points in time, it can quickly become intractable when there is any noise in the observations.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.8\textwidth]{"figures/Lorenz_xtraj"}
+\caption{Trajectory of the $x$ component of a particles position varying in time when motion is governed by the Lorenz `63 equations.
+The two trajectories are calculated using an initial position of $[-0.587,-0.563,16.870]$ and $[-0.590,-0.563,16.870]$.}
+\label{fig:Lorenz_xtraj}
+\end{figure}
+
+To demonstrate how small errors in the initial condition of a Lorenz solution can affect the trajectory of a particle, Figure~\ref{fig:Lorenz_xtraj} shows the path taken by two particles with very similar initial conditions,
+\[
+	\mathbf{x}_0^1 = \begin{pmatrix} -0.587\\-0.563\\16.870 \end{pmatrix} \quad \text{and} \quad \mathbf{x}_0^2 = \begin{pmatrix} -0.590\\-0.563\\16.870 \end{pmatrix}.
+\]
+This small difference in the $x$-dimension of the initial condition leads to the trajectories of the two particles diverging near $t=5$.
+This chaotic behaviour means that if we take only a few observations over a long time period we will find that many trajectories which may be vastly different achieve similar values of the likelihood function and so the posterior distribution for the initial condition can become very complex.%
+
+\subsubsection{Target Distribution}\label{sec:lorenz_target}
+
+For this example, we observe noisily the position of a particle at ten equally spaced points in the time interval $t \in (0, 1]$.
+The Lorenz equations are solved with the chaotic parameters given above and the initial condition $\mathbf{x}_0^1$.
+A time step of $h=1\times 10^{-3}$ was used to evolve the equations numerically with the Euler method.
+The noise added to each observation was taken from the distribution $\mathcal{N}(0, 0.1^2)$.
+An informative prior distribution was chosen to limit the complications arising from multiple initial conditions resulting in point-wise matching trajectories,
+\[
+	x_0 \sim \mathcal{N}(-0.5, 0.03^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.03^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.03^2).
+\]
+The posterior density function takes the form
+\[
+	\pi(\mathbf{x}_0|D) \propto \exp\left\{-\frac{1}{2}\|\mathcal{G}(\mathbf{x}_0) - D\|^2_\Sigma\right\}\mathcal{N}(\mathbf{x}_0; m, 0.03^2\text{I}),
+\]
+where $\mathcal{G}$ calculates the solution to the Lorenz equations, using the Euler method with a time step of $h=1\times 10^{-3}$ starting at the initial condition $\mathbf{x}_0$, and returns the position at $t = 0.1, ..., 1$.
+The mean of the prior distribution $m = [-0.5, -0.5, 15]$.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.8\textwidth]{"figures/Lorenz_posterior"}
+\caption{Posterior distribution for the initial condition of the particle with motion governed by the Lorenz equations as described in Section~\ref{sec:lorenz_target}.}
+\label{fig:Lorenz_posterior}
+\end{figure}
+
+Figure~\ref{fig:Lorenz_posterior} shows the marginal distributions of the posterior distribution for the initial condition $\mathbf{x}_0$ as found using the MH algorithm.
+While the $z$-dimension is largely uncorrelated with the other two dimensions, the correlation between $x$ and $y$ is -0.98, which makes it a challenging posterior distribution to explore without a transformation of the parameter space.
+
+%%%%%%
+\subsubsection{Implementation}\label{sec:lorenz_implementation}
+
+As in the previous example, we have no analytic form for the normalisation constant of the posterior and so we will measure convergence of the algorithms by calculating convergence to the posterior mean, where the truth is calculated using a very long chain produced by the MH algorithm.
+
+Both the MH and PAIS algorithms use an ensemble size of $M=500$.
+The PAIS algorithm uses the AMR resampling algorithm.
+The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations.
+Convergence graphs are produced using 32 repeats of each algorithm with each simulation producing one million samples.
+
+% numerics.
+\subsubsection{Convergence of MH vs PAIS}\label{sec:lorenz_conv}
+
+The optimal scaling parameters for the MH and PAIS algorithms are given in Table~\ref{table:Lorenz_opt_parameters}.
+
+\begin{table}[!htb]
+      \centering
+        \begin{tabular}{|l|r|r|}
+	\hline
+	Algorithm	& MH & PAIS \\ \hline
+	$\delta^*$	& 1.0e-1     & 1.3e-2 \\
+	\hline
+	\end{tabular}
+	\vspace{1mm}
+	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithm is optimised using the effective sample size.}
+	\label{table:Lorenz_opt_parameters}
+\end{table}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.8\textwidth]{"figures/Lorenz_convergence"}
+\caption{Convergence of the sample mean to the posterior mean for the MH and PAIS algorithms. Posterior distribution as described in Section~\ref{sec:lorenz_target}.}
+\label{fig:Lorenz_convergence}
+\end{figure}
+
+As is clear from Figure~\ref{fig:Lorenz_convergence} the PAIS-RW has a vastly superior burn-in time and achieves improved accuracy on the RWMH algorithm of three orders of magnitude after one million samples.
+
+To produce one sample from the posterior distribution for this example using the PAIS algorithm takes on average $5.5\times 10^{-4}$ seconds, while a sample can be produced in $1.7\times 10^{-5}$ seconds using the MH algorithm.
+However; the quality of the sample produced by PAIS is much higher.
+To reach a tolerance of $5\times 10^{-5}$, it takes the PAIS algorithm roughly 16.5 seconds and the MH algorithm roughly 1020 seconds.
+
 \section{Discussion and Conclusions}\label{Sec:Conc}

### 2018-02-16 response_JUQ.tex
Simon final additions

@@
 changes that have been made to the paper.
+We have made some substantial changes, not least with the inclusion of a new ``large scale'' example where the likelihood requires a solve of a differential equation.
+We demonstrate in this example that PAIS is 2 orders of magnitude faster, in terms of CPU time, than the equivalent standard MCMC.
 
@@
 However, following additional comments from the editor, we have included a new example where the likelihood requires the solution of an ODE.
-In Section **** we include an example of data assimilation for the Lorenz*** equations, which is a set of equations which is commonly used to test methods for inference in the applied mathematics community.
+In Section 7.5 we include an example of data assimilation for the Lorenz '63 equations, which is a set of equations which is commonly used to test methods for inference in the applied mathematics community.
 
@@
 This is precisely what the method is designed for, and is the very reason for us to compare the algorithms by looking at error as a function of the number of samples/likelihood evaluations.
-We have also added further discussion regarding this to Section ****.}
+This is seen even for a simple 3D ODE as we see in our example in Section 7.5, where PAIS reaches a relative error in the mean of $10^{-5}$ in a time which is 2 orders of magnitude quicker in CPU time than for trivially parallelised RWMH chains.}%done
 
@@
 The optimal architecture would likely depend on the relative costs of communication of the states across the processors, the cost of the likelihood evaluation, and the cost of the resampler for the chosen ensemble size.
-We have added a discussion to Section 3, following the presentation of the PAIS algorithm.}
+We have added a discussion to Section 3, following the presentation of the PAIS algorithm.}%done
 
@@
 The reason for analysing the error as a function of the number of the likelihood evaluations, is that this method is intended for problems with expensive likelihoods, for which these overhead costs would be dwarfed.
-We have added a discussion on this to the beginning of Section 7.}%done
+We have added a discussion on this to the beginning of Section 7.
+
+Additionally to this, in Section 7.5, as detailed above, we have compared the CPU time that it takes the PAIS to reach a relative error in the mean of the posterior of $10^{-5}$ against that for trivially parallelised (non-communicative) RWMH.
+Even for a simple example such as this, PAIS outperforms standard parallelised MCMC by two orders of magnitude.}%done
 
@@
 \comment{An other serious concern is why not use PMC-MCMC for the updating scheme in Algorithm 1.
-That would be as costly (since it also involves evaluating the posterior) but the yis will have the correct distribution.}{This comment is not clear to us - what is the ``updating scheme'' in Algorithm 1??
+That would be as costly (since it also involves evaluating the posterior) but the yis will have the correct distribution.}{This comment is not clear to us - what is the ``updating scheme'' in Algorithm 1?
 Also, the (weighted) $y_i$ are distributed according to the posterior, as discussed in Section 6.

### 2018-02-16 PAIS.tex
Simon final additions

@@
   superior sampling for multimodal problems, such as those arising
-  from inference for mixture models.
+  from inference for mixture models, and for problems with expensive
+  likelihoods requiring the solution of a differential equation.
 \end{abstract}
@@
 
-To produce one sample from the posterior distribution for this example using the PAIS algorithm takes on average $5.5\times 10^{-4}$ seconds, while a sample can be produced in $1.7\times 10^{-5}$ seconds using the MH algorithm.
-However; the quality of the sample produced by PAIS is much higher.
-To reach a tolerance of $5\times 10^{-5}$, it takes the PAIS algorithm roughly 16.5 seconds and the MH algorithm roughly 1020 seconds.
+To produce one sample from the posterior distribution for this example
+using the PAIS algorithm takes on average $5.5\times 10^{-4}$ seconds,
+while a sample can be produced in $1.7\times 10^{-5}$ seconds using
+the MH algorithm. However; the quality of the sample produced by PAIS
+is much higher. To reach a tolerance of $5\times 10^{-5}$, it takes
+the PAIS algorithm roughly 16.5 seconds and the MH algorithm roughly
+1020 seconds. Although we do not claim that either codes are optimally
+efficient, this give an indication of the saving that can be achieved
+through this type of methodology.
 
@@
 accuracy, which can be mitigated with the use of larger
-ensembles. Lastly, we have detailed how scaling parameters in the MCMC
+ensembles. We have detailed how scaling parameters in the MCMC
 proposals that are used within the mixture distribution can be quickly
 and efficiently
-tuned in an automated way. However, tuning the variances and
+tuned in an automated way. Lastly we have demonstrated that when the
+likelihood is expensive, for instance because it involves the
+numerical approximation of the solution of a differential equation, the savings made in terms of the number
+iterations required in order reach a particular degree of accuracy
+outweight the additional overhead costs that are incurred
+over-and-above standard MCMC.
+
+However, tuning the variances and
 covariances of the mixture proposal components
@@
 
-MCMC
-RWMH
-MALA
-AIS
-PMC
-PAIS
-ETPF
-AMR
-PAIS-RW
-PAIS-MALA
-APAIS-RW
-APAIS-MALA
+\begin{table}[h!]
+\centering
+\begin{tabular}{|l|l|}
+\hline
+Acronym & Full name                              \\ \hline
+MCMC    & Markov chain Monte Carlo               \\
+RWMH    & Random walk Metropolis-Hastings        \\
+MALA    & Metropolis-adjusted Langevin algorithm \\
+AIS     & Adaptive importance sampling           \\
+PMC     & Population Monte Carlo                 \\
+PAIS    & Parallel adaptive importance sampling  \\
+ETPF    & Ensemble transport particle filter     \\
+AMR     & Approximate multinomial resampler      \\
+PAIS-X  & PAIS with X kernels                    \\
+APAIS-X & Adaptive PAIS with X kernels          \\ \hline
+\end{tabular}
+\end{table}
 

### 2018-02-28 PAIS.tex
Final edits and spellcheck complete

@@
   importance sampling proposal distribution is formed using the
-  an ensemble of particles. A stratified sample is taken from this
+  ensemble of particles. A stratified sample is taken from this
   distribution and weighted under the posterior, a state-of-the-art
@@
   ensemble sizes, which can then be used in conjunction with PAIS for
-  complex problems. In particular, we demonstrate this methodology's
+  complex problems. We also focus on how algorithmic parameters
+  regarding the mixture proposal can be quickly tuned to optimise
+  performance. In particular, we demonstrate this methodology's
   superior sampling for multimodal problems, such as those arising
@@
 in order to make very large moves in state with relatively high
-acceptance probability\cite{sexton1992hamiltonian}. Non-reversible
+acceptance probabilities\cite{sexton1992hamiltonian}. Non-reversible
 methods are also becoming quite popular as they can improve
@@
 related to epidemiology and
-other biological applications\cite{house2016bayesian}. Multimodality of the posterior can also lead to
+other biological applications\cite{house2016bayesian}.
+
+Multimodality of the posterior can also lead to
 incredibly slow convergence in many methods. Many methods allow for
@@
 mode, and for all of the modes to be explored fully, this presents a
-significant challenge.
-
-One application where this is an ever-present problem is that of
+significant challenge. One application where this is an ever-present problem is that of
 mixture models. Given a dataset, where we know that the data is from
@@
 MCMC proposals can be quickly tuned, leading us to an efficient and
-fully-automated algorithm.
+fully-automated algorithm. We will demonstrate that despite additional
+overheads, the PAIS methodology can outperform its plain
+Metropolis-Hastings cousin by orders of magnitude to reach a histogram
+of a particular degree of convergence.
 
@@
 \end{equation}
-For example, if $x$ is an initial condition for the wave equation,
-$\mathcal{G}$ might calculate the height of a wave at a particular point and time.
 
@@
 \end{equation}
-where $\| y_1 - y_2 \|_\Sigma = (y_1-y_2)^\top\Sigma^{-1}(y_1-y_2)$
+where $\| y_1 - y_2 \|^2_\Sigma = (y_1-y_2)^\top\Sigma^{-1}(y_1-y_2)$
 for $y_1,y_2 \in \mathbb{R}^d$.
@@
 
-In this work we use these particle filters as resampling methods. We
-approximate the posterior distribution, $\mu_Y$, with a small sample
+In this work we use the resampler used within these particle filters. We
+approximate the posterior distribution, $\mu_Y$, with a sample
 of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
@@
 
-Particle filters such as the ETPF are well suited to this problem
+The resampler used within particle filters such as the ETPF are well suited to this problem
 since it is easy to introduce conditions in the resampling to ensure
@@
 All MCMC methods are naively parallelisable. One can take a method
-and simply implement it simultaneously over an ensemble of processors. All
+and simply implement it multiple times over an ensemble of processors. All
 of the states of all of the ensemble member can be recorded, and in the time
@@
 Moreover, many MCMC algorithms suffer from poor mixing, especially in
-multimodal systems. The amount of samples that it takes for an MCMC
+multimodal systems. The number of samples that it takes for an MCMC
 trajectory to switch between modes can be large, and given that a large
@@
 since all samples are saved along with their relative weighting.
-
-Moreover, a naively parallelised MCMC scheme is exactly that -
-naive. Intuition suggests that we can gain some speed up by sharing
-information across the ensemble members, and this is
-what we wish to demonstrate in this paper.
 
@@
 but we choose to use a mixture distribution, made up of a sum of MCMC
-proposal kernels. This corresponds to replacing the Dirac distribution
-in the right-hand side of Equation~\ref{eqn:resampler} with a chosen
-distribution e.g. Gaussian. Once the proposal is constructed, we can
+proposal kernels  e.g. Gaussians centred at the current state, as in RWMH. Once the proposal is constructed, we can
 sample a new ensemble from the proposal distribution, and each is
@@
 are not stored in order to characterise the posterior density, since
-the resampling process inevitably adds some error/bias. They are
-simply needed in order to inform the mixture distribution for the next
+the resampling process inevitably adds some error. They are
+simply needed in order to inform the mixture proposal distribution for the next
 iteration of the algorithm.
@@
 Initially we
-will use the ETPF algorithm\cite{reich2013nonparametric}, although we
-will suggest an alternative strategy in Section \ref{sec:AMR}. The
+will use the ETPF resampler algorithm\cite{reich2013nonparametric}, although we
+will suggest an alternative strategy in Section \ref{sec:AMR} where a
+large ensemble is required, and for which the ETPF may become more expensive. The
 resampling algorithm gives us a set of evenly weighted samples which
@@
 Since the resampling does not give us a statistically identical sample
-to that which is inputted, we cannot assume that the samples $\X^{(i)}$
+to that which is input, we cannot assume that the samples $\X^{(i)}$
 are samples from the posterior. Therefore, as with serial
 importance samplers, the weighted samples
-$(\W^{(i)},\Y^{(i)})_{i=1}^N$ are the samples from the posterior that
+$(\W,\Y)_{i=1}^N$ are the samples from the posterior that
 we will analyse.
@@
 The key is to choose a suitable transition kernel $\nu$ such that
-if $X^{(i)}$ is a representative sample of the posterior,
+if $\mathbf{X}^{(i)}$ is a representative sample of the posterior,
 then the mixture density $\chi(\cdot;\X^{(i)})$ is a good
@@
 the inverse problem being tackled involves computing
-the likelihood from a very large data set this could lead to a
+the likelihood from a very large data set, or where the likelihood
+requires the numerical solution of a differential equation, this could lead to a
 large saving of computational cost. We have observed that
@@
 significantly improve the speed of convergence of the algorithm for
-posteriors that we have considered\cite{Paul}.
-
-Care needs to be taken when choosing the proposal distribution to
-ensure that the proposals are absolutely continuous with respect to
+posteriors that we have considered, although kernels such as those
+used in MALA can be more stable for certain problems\cite{Paul}.
+
+Care needs to be taken when choosing kernel(s) for the proposal distribution to
+ensure that the overall mixture is absolutely continuous with respect to
 the posterior distribution. If this is not the case, importance
-weights grow exponentially for samples in the tails. This will
+weights can become high variance where samples are made away from
+modes in the distribution. This will
 drastically decrease the efficiency of sampling.
@@
 of the proposal distribution. Therefore the proposal distributions can
-be tuned such that they are are slightly over-dispersed. This tuning
+be tuned such that they are slightly over-dispersed. This tuning
 can take place during the burn-in phase of the algorithm. Algorithms
@@
 would like all $M$ of our samples to provide us with new information
-about the posterior distribution. In practise, we cannot achieve a
+about the posterior distribution. In practice, we cannot achieve a
 perfect effective sample size of $M$.
@@
 the weights can vary
-over many orders of magnitude causing numerical issues, so that the effective
+over many orders of magnitude causing numerical instabilities, so that the effective
 sample size is more desirable as an indicator of optimality.
@@
 
-Although the ETPF is optimal in terms of preserving statistics of the
-sample, it can also become quite costly as the number of ensemble members
+Although the ETPF produces the optimal linear coupling, it can also become quite costly as the number of ensemble members
 is increased. It is arguable that in the context of PAIS, we do not
@@
 harder for the new samples to forget a significant region in the
-density. This motivates the following algorithm, which we refer to as
-approximate multinomial resampling (AMR).
+density. This motivates the following algorithm, which we refer to as the
+approximate multinomial resampler (AMR), which is a greedy
+approximation of the ETPF resampler.
 
@@
 three moments of the samples changes with ensemble size $M$ for the three different
-samplers. As expected, the AMR lies somewhere between the high
+resamplers. As expected, the AMR lies somewhere between the high
 accuracy of the ETPF and the less accurate bootstrap
@@
 favourable features, and demonstrates how different choices of
-resampler can affect the speed and accuracy of the PAIS algorithm.
+resampler can affect the speed and efficiency of the PAIS algorithm.
 
@@
 The numerical examples in this section were all computed on a single
-core of Dell
+core of a Dell
 server with four 8-core 3.3GHz CPUs and 64Gb memory. Throughout we
@@
 \includegraphics[width=0.6\textwidth]{"figures/PAIS_saving"}
-\caption{Ratio of PAIS-RW samples required to reach the same tolerance as the RWMH algorithm.}
+\caption{Ratio of PAIS-RW samples required to reach the same tolerance
+  as the RWMH algorithm, using the ETPF resampler.}
 \label{fig:PAIS_saving}
@@
 In this example we look at a bimodal posterior distribution which has
-a high energy barrier between the two modes. In the MH algorithm, the
+a region of very low probability density between the two modes. In the MH algorithm, the
 expected number of iterations for a switch between modes is a large
@@
 	\hline
-	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$	& $\beta^*_{\text{L2}}$ \\ \hline
+	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$	& $\beta^*_{\text{L}^2}$ \\ \hline
 	RWMH								& 2.3e-1					& - 						& 9.3e-1\\
@@
 globally optimal $\beta^*$ leads to normal convergence rates,
-$\mathcal{O}(n^{-1/2})$, whereas algorithms using locally optimal
-scaling parameters initially converge faster but do not sample
+$\mathcal{O}(1/\sqrt{N})$, whereas algorithms using locally optimal
+scaling parameters do not sample
 proportionally for the whole simulation.
 
-Using the PAIS algorithm we can ensure that we do not lose modes as
+Using the PAIS algorithm we can ensure that we do not lose modes as the
 simulation progresses. Since we have an ensemble, we can sacrifice
@@
 remedy this problem. The PAIS algorithm tackles this problem with its
-resampling step. The algorithm uses its dynamic kernel to build up an
+resampling step. The algorithm uses its dynamic mixture proposal distribution to build up an
 approximation of the posterior at each iteration, and then compares
@@
 The technique of mixture modelling employs well known parametric
-families to construct an approximating nonparametric distribution
+families to construct an approximating distribution
 which may have a complex structure. Most commonly, Gaussian kernels
@@
 Figure \ref{fig:mixture_posterior} presents a visualisation of the
-posterior distribution for this problem, created from 10million
+posterior distribution for this problem, created from 10 million
 samples produced by the PAIS-RW algorithm.
@@
 \]
-This small difference in the $x$-dimension of the initial condition leads to the trajectories of the two particles diverging near $t=5$.
+This small difference in the $x$-dimension of the initial condition leads to the trajectories of the two particles decoupling near $t=5$.
 This chaotic behaviour means that if we take only a few observations over a long time period we will find that many trajectories which may be vastly different achieve similar values of the likelihood function and so the posterior distribution for the initial condition can become very complex.%
@@
 
-For this example, we observe noisily the position of a particle at ten equally spaced points in the time interval $t \in (0, 1]$.
-The Lorenz equations are solved with the chaotic parameters given above and the initial condition $\mathbf{x}_0^1$.
-A time step of $h=1\times 10^{-3}$ was used to evolve the equations numerically with the Euler method.
-The noise added to each observation was taken from the distribution $\mathcal{N}(0, 0.1^2)$.
-An informative prior distribution was chosen to limit the complications arising from multiple initial conditions resulting in point-wise matching trajectories,
+For this example, we observe noisily the position of a particle at ten
+equally spaced points in the time interval $t \in (0, 1]$. The Lorenz
+equations are solved with the chaotic parameters given above and the
+initial condition $\mathbf{x}_0^1$. A time step of $h=1\times 10^{-3}$
+was used to evolve the equations numerically with the explicit Euler
+method. The noise added to each observation was taken from the
+distribution $\mathcal{N}(0, 0.1^2I$. Priors for the initial condition
+coordinates were given by
 \[
@@
 1020 seconds. Although we do not claim that either codes are optimally
-efficient, this give an indication of the saving that can be achieved
+efficient, this gives an indication of the saving that can be achieved
 through this type of methodology.
@@
 iterations required in order reach a particular degree of accuracy
-outweight the additional overhead costs that are incurred
+outweighs the additional overhead costs that are incurred
 over-and-above standard MCMC.

### 2018-03-05 cover_letter.tex
Final changes
Final small change, and added the diff file to the response to referees
document.

--- 2017-02-03 cover_letter.tex
@@
 
-I am writing with regards to our previous submission to the SIAM/ASA Journal of Uncertainty Quantification, entitled ``Parallel Adaptive Importance Sampling''.
-Please forgive us for the length of this cover letter, but we felt that we needed to address a few issues from our previous submission.
-
-This manuscript had previously been submitted to the SIAM Journal for Scientific Computing, back in August 2015.
-We received 2 pretty positive reviews, although one review criticised our lack of a high dimensional example.
-The introduction at the time was misleading as application areas were mentioned (meteorology, oceanography, oil well diagnostic problems) where the unknowns in the inverse problem are high dimensional.
-One reviewer focussed on this, and as such we were asked not to resubmit unless we were able to demonstrate the performance of the algorithm in a very high dimensional setting, which of course is not feasible for this type of method.
-As such, we didn't feel that it would be worth our time in resubmitting to this particular journal.
-
-As a result of this, we undertook a thorough rewrite of the paper, addressing as many of the other concerns of the referees as we could, emphasising on low-dimensional but non-Gaussian distributions, one of which arises from inverse problems for biochemical networks.
-We then considered which other journals would be a good fit for this work, and felt that SIAM UQ would be perfect.
-The paper was substantially different, and as such we didn't view it as a simple resubmission to another journal.
-Our mistake was in not highlighting this in the coverletter to the submission to SIAM UQ.
-
-In the response letter, the associate editor stated that: \emph{"You will see that both referees reports are quite short -- that is my doing.
-I was aware (through the SIAM grapevine) that you had previously submitted this paper to SISC and that lengthy reviews had been obtained.
-So, in the interest of not using too much of people's time, I asked for indicative reviews."} This clearly indicates that this submission was not treated as a new submission, and the referees were instructed not to give a full review.
-Moreover, their reviews have in our view been unfairly coloured by the previous reviews from the SISC submission which were provided to them.
-In particular, they criticise us due to \emph{"the computed example provides no information about whether or not the proposed method will work in the (sizable) applications mentioned in your introduction."} However, all mentions of such sizable applications have been completely removed from the introduction.
-The referees were responding to the reviews of the original paper which were sent to them, rather than the new paper, where these issues are not present.
-
-The changes to the paper since the initial submission have been comprehensive - almost all of the numerical experiments have been rerun, a new 2D example was added from the application of parameter determination for chemical networks.
-In particular we would highlight also that a new resampler has been developed in this second version, a greedy approximation of the full state-of-the-art resampler that was used in the original paper.
-The introduction was also completely rewritten.
-
-We are however very keen to make adjustments to the paper in order to appease any legitimate concerns that the referees have highlighted.
-In particular, we have replaced the 2D problem at the end of the paper with a 6 dimensional multimodal posterior arising from inverse problems for mixture models.
-We have also completely rewritten the introduction for a second time, taking more care to reference more relevant literature from, among others, the adaptive importance sampling community.
-Hopefully the introduction now better reflects the content of the rest of the paper.
-Importantly, we have also added a short passage which directs interested/concerned readers to proofs of the consistency of ensemble adaptive importance sampling schemes.
-
-We are keen to resubmit the manuscript to SIAM/ASA JUQ for a second time.
-In order to address the issues that have arisen, we have provided three copies of the paper.
-The first is a latexdiff pdf of the changes that were made for our submission to SIAM/ASA JUQ from the original submission to SISC.
-The second is a latexdiff pdf of the changes we have made in our latest submission.
-The third version is our new submission.
-We also provide detailed descriptions of how the changes address the concerns of the referees.
-
-However, if you still believe that this manuscript and its contents are not suitable for publication in your journal, we would very much appreciate a swift decision, in order that we are able to submit it elsewhere.
-
-We thank you very much for your time and patience in dealing with this manuscript. We look forward to hearing from you in due course.
+Please find attached the latest version of our manuscript, ``Parallel Adaptive Importance Sampling''.
+We have taken onboard the comments of the referees and the associate editor, and we hope that they will be satisfied with the substantial changes and additions that we have made.
+We very much look forward to hearing from you in due course regarding this paper.
 

### 2018-12-05 MATLAB/density.m
Old unsubmitted changes :(

@@ -1,0 +1,7 @@
+function pi = density(X)
+mu1 = [1;1];
+sig1 = [0.1 0; 0 0.1];
+mu2 = [-5;-5];
+sig2 = [2.75 -2.25; -2.25 2.75];
+pi = 0.2*mvnpdf(X,mu1,sig1) + 0.8*mvnpdf(X,mu2,sig2);
+end

### 2018-12-05 MATLAB/plot_density.m

@@ -1,0 +1,10 @@
+x = -10:1e-1:10;
+y = x;
+[X,Y] = meshgrid(x,y);
+Z = 0*X;
+for i=1:numel(x)
+for j=1:numel(y)
+Z(i,j) = density([X(i,j);Y(i,j)]);
+end
+end
+pcolor(X,Y,Z);

### 2018-12-05 figures/data/read_data.m

@@ -1,0 +1,6 @@
+f = fopen('trace_X_l_1', 'r');
+X = fread(f, [3 inf], 'double')';
+fclose(f);
+f = fopen('trace_w_l_1', 'r');
+w = fread(f, [1 inf], 'double')';
+fclose(f);

### 2018-12-07 response_JUQ_2.tex
Added response document to 2nd JUQ review

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{pdfpages}
\usepackage{amsmath,amssymb}
\usepackage{enumerate}
\usepackage{color}
\usepackage{amsthm,amsmath}
\usepackage{bbm}
\usepackage{mhchem}

%for stacking below a sum or max...
\usepackage{mathtools}

\usepackage[colorlinks]{hyperref}

\newcommand{\comment}[2]{\vspace{0.6cm}{\bf Comment:} {\it #1.}

\vspace{0.3cm}{\bf Answer:} #2}

\begin{document}
\title{Response to Referee Comments: Parallel Adaptive Importance Sampling}
\maketitle
First of all, we would like to thank the referees and the associate editor for
their time in reading and commenting on our manuscript. In this document
we will address each of the points raised, and describe any relevant
changes that have been made to the paper. ****INSERT SUMMARY OF MAJOR CHANGES****

\section*{Referee 1 comments}

\comment{This is supposed to be a parallel algorithm and yet it's implemention is serial. Furthermore, the authors don't seem to be entirely sure that parallelization will work since they comment that it's effectiveness will depend on various factors. The title seems somehow misleading to me. Maybe if the "parallel" part of the title is removed and the discussion about parallelization is treated as a possibility instead of a critical part of the algorithm, then the document will be less confusing.}{Despite the fact that we felt that we had demonstrated the feasibility of the parallelisation of this method, we agree with the referee that we have not demonstrated this in the numerics. Furthermore, this algorithm delivers a speed-up even when run in serial. As such, we are happy to change the name of the algorithm to ``Ensemble Transport Adaptive Importance Sampling'' (ETAIS). This required some rewriting throughout the paper. In particular, we have toned down the emphasis on parallelisation, and put more emphasis on the fact that this algorithm also delivers speed-up for serial implementaions.}

\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{The naive parallel chains are run for long enough that they are equivalent to one long chain. We have added a discussion on this point to *****}

\comment{It will be interesting that the authors show the PAIS using AMS with the examples. They show the difference between the AMS and the ETPF but not within the PAIS.}{We assume that the referee meant the AMR? As the AE asked for a new example to replace the first two examples, we took this as an opportunity to compare ETAIS with the ETPF, AMR and bootstrap resamplers, which is now presented in the new first example. The numerics demonstrate that superior samplers lead to more stable and accurate sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear.}

\comment{There is an error at the end of page 24.}{Many thanks for spotting this, this has now been corrected.*******}

\section*{Referee 2 comments}
\comment{The paper is ready for publication.}{Many thanks. We hope you agree that the additional changes have further improved the paper.}

\section*{Associate Editor comments}

\comment{The authors have made a substantial improvement over previous version of the paper. However, there are a number of points that I would like to see clarified before the paper is accepted for publication.\\
1) The posterior sometimes is called $\mu$ sometimes $\pi$ some times $\pi(x|D)$ etc. In fact, there is a confusion in Alg. 1 in p.6 with this.}{We have carefully gone through the paper and clarified where appropriate whether we are referring the posterior measure $\mu$ or its density $\pi$, and unified the notation as much as possible.}

\comment{2) Choosing the proposal (kernel) in a MH is all the main and commonly the only problem in MCMC. PAIS also needs a proposal of exactly the same kind. Why can we expect that within PAIS this will be easier? An indeed, PAIS will perform better that a naively parallelized MH, both using the same proposal. How PAIS saves some of the problems in finding good proposals?}{As discussed in the paper, there are many possible choices for the proposal kernel around each of the ensemble members. The only thing that we require to ensure convergence of an AIS algorithm, is the absolute continuity of the kernel with respect to the posterior. However, to ensure stability of the algorithm, we also require that the tails of the proposal distributions are at least as fat as the target. In practice, this means that picking kernels which are of the same type as the prior is usually a good idea. The ETAIS methodology does not, however, greatly benefit from the use of more informed and expensive proposals, such as are seen in the standard M-H framework, where gradient information can be used to improve proposals (MALA, HMC, etc), since the information about the whole target is represented by the ensemble. As mentioned in the paper, we have conducted similar studies with MALA proposals, and although we sometimes observe a slightly quicker burn-in period, overall the performance is largely the similar to that of a RW proposal, at twice the cost. We have added a detailed discussion around this to ******}

\comment{3) It might be trivially to do so, but the authors only briefly mention a prove for the convergence of their algorithm. In fact, is it ok to use a dependent proposal in importance sampling? ... no regularity conditions are needed? Why Alg. 3 works? resampling "introduces error" ... is this relevant? does the error cascades down in each iteration? In the examples the PAIS seem to work ok, but, do we need any regularity condition? Please make your convergence claims more explicit and take more time to explain them. In particular, do we have the correct invariant distribution?}{This algorithm belongs to a family of methods for which convergence has been proven in the literature, and as such, we have referenced this proof. We have clarified this a little further in *******}

\comment{4) There are too many examples. The univariate gaussian is not very informative and the univariate bimodal, is not either since both modes are of the same scale. I would rather see these two examples turned into one single one with a 2D mixture of two separated gaussians with contrasting scales (variance covariance matrices).}{At the request of the AE, we have removed the first two examples, and replaced them with a single example, which is a two dimensional bimodal density with differing scales in the covariances of the two modes. We have also taken this opportunity to demonstrate the differences in results when using the different resamplers discussed in the paper. We have also included a comparison with a standard bootstrap resampler. The new numerics demonstrate the improvement in stability and convergence of the method with the use of more advanced resampling methods. The remaining examples say **** and BLAH and so we're keeping them.*****}

\comment{5) In the examples the authors include very ad hoc strategies like the 'scout' chain and specially the very ad hoc strategy described in section 7.4.2 (!). The authors, in my opinion, do not really discuss the implementation of the PAIS algorithm but a case-by-case ad hoc version of it. This is clearly unacceptable.}{We strongly disagree that the approaches used are in any way ``ad hoc''. As per the discussion following the AE's 2nd comment, it is important that the overall proposal has tails that are as fat as those of the target. The ``scout'' chain, as also used in [ref], ensures two things. 1, it allows for larger moves into unexplored regions of the state-space, and it also fattens the tails of the proposal distribution. As for the proposal kernels used in 7.4.2, these are chosen simply because the proposal distribution must only be supported by [0,1]. We have gone into more details regarding this and explaining the choices, and how users might make their own choices for different scenarios, in *********.}

\comment{6) Please clarify the first phrase of section 7.4.2, which, as far as I see, is simply twice wrong. There is a (very simple!) analytic version of the posterior AND with the right proposal MH can sample from it (!).}{We have amended the statement about the analytic form of the posterior. We are not aware of any M-H methods which can reliably sample from multimodal densities with well-separated modes. HMC trajectories can be hand-tuned to make transitions between modes, but only with information regarding the separation. Ensemble-based methods appear to us to be the only candidate for reliable sampling from such distributions without such a hand-tuning.}

\comment{7) Although the chaotic Lorenz 93 ODE is used, it is used in such a short time scale that the posterior is quite well behaved (unimodal and fairly gaussian and only some correlation is seen between to parameters). Since the priors are normal, this regime suggest a linear part of the FM. The authors should attempt something more challenging, by running the system into a highly nonlinear regime. Otherwise, why consider this example?}{The reason for considering the example was that the AE had previously asked us to produce an example in the regime where we suggest that we get the biggest win using our proposed method, i.e. those where the likelihood evaluations have significant cost due to requirement of the solution of a differential equation. In this example, we have demonstrated a significant speed-up over the equivalent M-H algorithm. We do understand however the AE's desire for a more challenging inverse problem in itself, and so we have created a new example that *********. The new example still demonstrates a speed-up of ********.}

\comment{8) Do not use an explicit Euler solver, use any higher order, off the shelve solver.}{Many of the off-the-shelf solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations, since they are chaotic, and as such \emph{any} numerical error is quickly amplified through the dynamics. Moreover, the focus of this paper is not on these applications, but in the statistical algorithm. In this example, we are simply looking for a non-linear mapping such that we can demonstrate the speed-up that is gained from using ETAIS as opposed to standard M-H sampling, and we believe that a simple Euler approximation provides such a mapping.}
\end{document}

### 2018-12-13 ETAIS.tex
Started new document "ETAIS"

--- 2018-02-28 PAIS.tex
@@
     Mathematics, University of Manchester, Manchester, UK}}
-\title{Parallel Adaptive Importance Sampling}
+\title{Ensemble Transport Adaptive Importance Sampling}
 \begin{document}
@@
   for scalable parallel sampling algorithms. At each iteration, an
-  importance sampling proposal distribution is formed using the
+  importance sampling proposal distribution is formed using an
   ensemble of particles. A stratified sample is taken from this
   distribution and weighted under the posterior, a state-of-the-art
-  resampling method is then used to create an evenly weighted sample
-  ready for the next iteration. We demonstrate that this parallel
-  adaptive importance sampling (PAIS) method outperforms naive
+  ensemble transport resampling method is then used to create an evenly weighted sample
+  ready for the next iteration. We demonstrate that this ensemble transport
+  adaptive importance sampling (ETAIS) method outperforms naive
   parallelisation of serial MCMC methods for low dimensional problems,
@@
   strategy, approximate multinomial resampling (AMR), which while not
-  as accurate as other schemes is substantially less costly for large
-  ensemble sizes, which can then be used in conjunction with PAIS for
+  as accurate as the ensemble transport resampler, is substantially less costly for large
+  ensemble sizes, and can then be used in conjunction with ETAIS for
   complex problems. We also focus on how algorithmic parameters
@@
   from inference for mixture models, and for problems with expensive
-  likelihoods requiring the solution of a differential equation.
+  likelihoods requiring the solution of a differential equation, for
+  which significant speed-ups are demonstrated.
 \end{abstract}
-\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
+\begin{keywords}MCMC, importance sampling, Bayesian, inverse problems,
+  parallel.
 \end{keywords}
@@
 \cite{elvira2017improving}. However our approach exploits the current
-state-of-the-art resamplers. Higher accuracy resamplers, that provide
+state-of-the-art resamplers, and adaptively optimises algorithmic parameters. Higher accuracy resamplers, that provide
 a deterministic resample which optimally reflects the statistics of
@@
 fully-automated algorithm. We will demonstrate that despite additional
-overheads, the PAIS methodology can outperform its plain
+overheads, the ETAIS methodology can outperform its plain
 Metropolis-Hastings cousin by orders of magnitude to reach a histogram
@@
 In Section \ref{Sec:Prelim} we introduce some mathematical
-preliminaries upon which we will later rely. In Section \ref{Sec:PAIS}
-we present the general framework of the PAIS algorithm. In Section
-\ref{Sec:adapt} we consider adaptive versions of PAIS which
+preliminaries upon which we will later rely. In Section \ref{Sec:ETAIS}
+we present the general framework of the ETAIS algorithm. In Section
+\ref{Sec:adapt} we consider adaptive versions of ETAIS which
 automatically tune algorithmic parameters concerned with the proposal
@@
 problem exactly. In Section~\ref{sec:consistency} we consider
-consistency of the PAIS algorithm. In Section \ref{Sec:Num} we present
+consistency of the ETAIS algorithm. In Section \ref{Sec:Num} we present
 some numerical examples, before a brief conclusion and discussion in
@@
 In this subsection we briefly review particle filters, since the
-development of the resampler that we incorporate into the PAIS is
+development of the resampler that we incorporate into the ETAIS is
 motivated by this area.
@@
 	 \frac{1}{M} \delta_{x_i}(\cdot) = \hat{\eta}.
-\end{equation}
-where $n_i \in \{0, 1, \dots,M\}$. This
+\end{equation}This
 new sample $\{x_i\}_{i=1}^M$ is equally weighted.
@@
 motivated the development of the Parallel Adaptive Importance Sampler
-(PAIS). In the next section we will introduce the method in its most
+(ETAIS). In the next section we will introduce the method in its most
 general form.
@@
 
-\section{The Parallel Adaptive Importance Sampler \allowbreak (PAIS)}\label{Sec:PAIS}
+\section{The Ensemble Transform Adaptive Importance Sampler \allowbreak (ETAIS)}\label{Sec:ETAIS}
 
@@
 similar areas to the target density, and incredibly inefficient when
-this is not the case. The aim of the PAIS is to use an ensemble of states to construct a proposal
+this is not the case. The aim of the ETAIS is to use an ensemble of states to construct a proposal
 distribution which will be as close as possible to the target density. If this
@@
 the process again. The algorithm is summarised in Algorithm
-\ref{alg:PAIS}.
+\ref{alg:ETAIS}.
 
@@
 		Sample $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^\top, \quad y_j^{(i)} \sim
-\nu(\cdot;x_j^{(i-1)})$.\label{algline:PAIS_propose}\;
+\nu(\cdot;x_j^{(i-1)})$.\label{algline:ETAIS_propose}\;
 		Calculate $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^\top,$ \quad $w^{(i)}_j =
@@
 
-		Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i)})$.\label{algline:PAIS_resample}\;
+		Resample: $(\W^{(i)},\Y^{(i)}) \rightarrow (\frac{1}{M}\mathbf{1}, \X^{(i)})$.\label{algline:ETAIS_resample}\;
 	}
 	Output $(\W, \Y)$.\;
-\caption{The parallel adaptive importance sampler (PAIS).\label{alg:PAIS}}
+\caption{The ensemble transform adaptive importance sampler (ETAIS).\label{alg:ETAIS}}
 \end{algorithm}
@@
 chain with high autocorrelation. One aspect worthy of consideration
-with the PAIS, is finding an appropriate proposal kernel $\nu$ such
+with the ETAIS, is finding an appropriate proposal kernel $\nu$ such
 that the mixture distribution $\chi$ is a close approximation to the
@@
 
-Algorithms which use mixture proposals, e.g. PAIS, must tune the
+Algorithms which use mixture proposals, e.g. ETAIS, must tune the
 variance of the individual kernels within the proposal mixture. This
@@
 RWMH and 57.4\% for MALA\cite{roberts2001optimal}. This method
-is not applicable to PAIS so we must use other statistics to optimise
+is not applicable to ETAIS so we must use other statistics to optimise
 the scaling parameter. Section~\ref{sec:statistics} gives some
@@
 $E$ during the first 30 iterations. These numerics are taken from the
-example in Section~\ref{sec:mixture_conv} using the PAIS-RW algorithm.}
+example in Section~\ref{sec:mixture_conv} using the ETAIS-RW algorithm.}
 \label{fig:neff-burnin}
@@
 \caption{The behaviour of the effective sample size as the ensemble
-  size increases, considering the example in Section~\ref{sec:problem 1} using the PAIS-RW algorithm.}
+  size increases, considering the example in Section~\ref{sec:problem 1} using the ETAIS-RW algorithm.}
 \label{fig:neff-M}
@@
 Figure~\ref{fig:neff-M} shows how the effective sample size used in
-PAIS is affected by the ensemble size. We see from subfigure (a) that
+ETAIS is affected by the ensemble size. We see from subfigure (a) that
 as the ensemble size increases, the optimal scaling parameter
@@
 
-\subsection{Adaptively Tuned PAIS}\label{sec:adapt}
+\subsection{Adaptively Tuned ETAIS}\label{sec:adapt}
 
@@
 	\For{$n=1,\dots,N$}{
-		Complete steps \ref{algline:PAIS_propose}-\ref{algline:PAIS_resample} of Algorithm~\ref{alg:PAIS}.\;
+		Complete steps \ref{algline:ETAIS_propose}-\ref{algline:ETAIS_resample} of Algorithm~\ref{alg:ETAIS}.\;
 		\If {$n \in \{n_k\}$}
@@
 	}
-\caption{Adaptively tuned PAIS algorithm.\label{alg:adaptPAIS}}
+\caption{Adaptively tuned ETAIS algorithm.\label{alg:adaptETAIS}}
 \end{algorithm}
 \end{table}
-Our adaptive algorithm is given in Algorithm~\ref{alg:adaptPAIS}.
+Our adaptive algorithm is given in Algorithm~\ref{alg:adaptETAIS}.
 We choose update times which, as suggested in Section~\ref{sec:ess} allow for a reasonable estimate of the effective sample size, but do not waste too many iterations.
@@
 \centering
-\subfigure[APAIS-RW algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISRW_adapt_var"}}
-\subfigure[APAIS-RW algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISRW_adapt_ess"}}
-\subfigure[APAIS-MALA algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISMALA_adapt_var"}}
-\subfigure[APAIS-MALA algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISMALA_adapt_ess"}}
-\caption{Demonstration of the convergence of the adaptive scaling parameter in the adaptively tuned PAIS algorithms.}
-\label{fig:G1_adapt_trace_PAIS}
+\subfigure[AETAIS-RW algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_ETAISRW_adapt_var"}}
+\subfigure[AETAIS-RW algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_ETAISRW_adapt_ess"}}
+\subfigure[AETAIS-MALA algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_ETAISMALA_adapt_var"}}
+\subfigure[AETAIS-MALA algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_ETAISMALA_adapt_ess"}}
+\caption{Demonstration of the convergence of the adaptive scaling parameter in the adaptively tuned ETAIS algorithms.}
+\label{fig:G1_adapt_trace_ETAIS}
 \end{figure}
@@
 concern. However, this is rarely an issue, due in part to the speed at
-which the PAIS algorithm completes the burn-in phase. Figure
-\ref{fig:G1_adapt_trace_PAIS} demonstrates this for two different
-versions of the PAIS methodology with a RW proposal and a MALA\cite{roberts1998optimal}
+which the ETAIS algorithm completes the burn-in phase. Figure
+\ref{fig:G1_adapt_trace_ETAIS} demonstrates this for two different
+versions of the ETAIS methodology with a RW proposal and a MALA\cite{roberts1998optimal}
 proposal. This picture is typical in our experience, with the scaling
@@
 Although the ETPF produces the optimal linear coupling, it can also become quite costly as the number of ensemble members
-is increased. It is arguable that in the context of PAIS, we do not
+is increased. It is arguable that in the context of ETAIS, we do not
 require this degree of accuracy, and that a faster more approximate
@@
 The resulting sample has several properties which are advantageous in
-the context of being used with the PAIS algorithm. Firstly, we have
+the context of being used with the ETAIS algorithm. Firstly, we have
 effectively chopped up the multinomial distribution used in the
@@
 where we have chosen to use a larger ensemble size. We do not claim
-that the AMR is the optimal choice within PAIS, but it does have
+that the AMR is the optimal choice within ETAIS, but it does have
 favourable features, and demonstrates how different choices of
-resampler can affect the speed and efficiency of the PAIS algorithm.
-
-%%%%%%
-
-\section{Consistency of PAIS}\label{sec:consistency}
+resampler can affect the speed and efficiency of the ETAIS algorithm.
+
+%%%%%%
+
+\section{Consistency of ETAIS}\label{sec:consistency}
 
@@
 Estimation of the normalisation constant $Z$ is biased, and so estimates of statistics are sums of independent but biased estimators.
-Since the estimators are independent, the proof of consistency of PAIS in this second limit can be approached in the same way as the PMC algorithm, where it has been shown that $\hat{Z}\rightarrow Z$~\cite{robert2013monte}.
+Since the estimators are independent, the proof of consistency of ETAIS in this second limit can be approached in the same way as the PMC algorithm, where it has been shown that $\hat{Z}\rightarrow Z$~\cite{robert2013monte}.
 Since the normalisation constant is consistent, sums of the independent estimators are also consistent.
@@
 likelihood (such as a PDE solve, and/or very large data) which dwarfs
-the overhead cost of PAIS, and which could have complex posterior
+the overhead cost of ETAIS, and which could have complex posterior
 structure, such as correlations, multimodality or other highly
@@
 
-In this section we demonstrate convergence of the PAIS algorithm. We
-also investigate the convergence properties of PAIS compared with
+In this section we demonstrate convergence of the ETAIS algorithm. We
+also investigate the convergence properties of ETAIS compared with
 naively parallelised Metropolis-Hastings samplers. We assess the
-performance of the PAIS algorithm using the relative $L^2$ error
+performance of the ETAIS algorithm using the relative $L^2$ error
 defined in Equation~\eqref{eqn:L2_error}, as well as the relative error in the
@@
 	\item {\bf Finding the optimal scaling parameters}: We chose 32 values of $\beta$ evenly spaced on a log scale in the interval $[10^{-5}, 2]$.
-We ran the PAIS and MH algorithms for one million iterations, each with an ensemble size of $M=50$.
+We ran the ETAIS and MH algorithms for one million iterations, each with an ensemble size of $M=50$.
 We took 32 repeats and used the geometric means of the sample statistics in Section~\ref{sec:statistics} to find the optimal scaling parameter.
-	\item {\bf Measuring convergence of nonadaptive algorithms}: We ran the PAIS and MH algorithms, using the scaling parameters found in Step 1, for one million iterations, again with $M=50$.
+	\item {\bf Measuring convergence of nonadaptive algorithms}: We ran the ETAIS and MH algorithms, using the scaling parameters found in Step 1, for one million iterations, again with $M=50$.
 The simulations are repeated 32 times.
@@
 
-This first example shows how PAIS handles searching for the mode of a
+This first example shows how ETAIS handles searching for the mode of a
 Gaussian distribution when the initial state is a long way out in the
@@
 posteriors~\cite{rosenthal2011optimal}. To tune the
-PAIS algorithm we maximise the effective sample size as
+ETAIS algorithm we maximise the effective sample size as
 discussed in Section~\ref{sec:ess}.
@@
 \subfigure[RWMH algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_rwmho"}}
-\subfigure[PAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
+\subfigure[ETAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
 \caption{Finding optimal values of $\beta$ for the example in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
@@
 	\hline
-	Statistic							& PAIS-RW \\ \hline
+	Statistic							& ETAIS-RW \\ \hline
 	$\beta_{\text{eff}}^*$				& 4.7e-2 \\
@@
 
-\subsubsection{Convergence of RWMH vs PAIS-RW}
+\subsubsection{Convergence of RWMH vs ETAIS-RW}
 
@@
 \includegraphics[width=0.8\textwidth]{"figures/new_G2_l2"}
-\caption{Convergence of the (A)RWMH and (A)PAIS-RW algorithms. The
+\caption{Convergence of the (A)RWMH and (A)ETAIS-RW algorithms. The
 setup is as in Section~\ref{sec:Implementation P1} (2, 3). Resampling
@@
 
-Figure~\ref{fig:MH1 L2} shows that the PAIS-RW algorithm converges
+Figure~\ref{fig:MH1 L2} shows that the ETAIS-RW algorithm converges
 to the posterior distribution significantly faster than the RWMH
@@
 converge to the posterior at a similar speed to the respective
-optimised algorithm. This shows that, particularly for PAIS, we can
+optimised algorithm. This shows that, particularly for ETAIS, we can
 optimise simulations efficiently on the fly.
@@
 
-\subsubsection{Scaling of the PAIS algorithm with ensemble size}
+\subsubsection{Scaling of the ETAIS algorithm with ensemble size}
 
 In this and the next example we use an ensemble size $M=50$, however it is
-interesting to see how the PAIS algorithm scales when we increase the
+interesting to see how the ETAIS algorithm scales when we increase the
 ensemble size, and if there is a minimum required ensemble size. We
 implement the problem in Section~\ref{sec:problem 1}, using
-the RWMH and PAIS-RW algorithms with ensemble sizes in the interval $M
+the RWMH and ETAIS-RW algorithms with ensemble sizes in the interval $M
 \in [1, 160]$.
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{"figures/PAIS_saving"}
-\caption{Ratio of PAIS-RW samples required to reach the same tolerance
+\includegraphics[width=0.6\textwidth]{"figures/ETAIS_saving"}
+\caption{Ratio of ETAIS-RW samples required to reach the same tolerance
   as the RWMH algorithm, using the ETPF resampler.}
-\label{fig:PAIS_saving}
+\label{fig:ETAIS_saving}
 \end{center}
@@
 
-Figure~\ref{fig:PAIS_saving} was produced using the method of finding
+Figure~\ref{fig:ETAIS_saving} was produced using the method of finding
 optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
@@
 parallelised MH. The decreasing trend shows superlinear improvement of
-PAIS with respect to ensemble size, in terms of the number of
+ETAIS with respect to ensemble size, in terms of the number of
 iterations required, which is a demonstration of our belief that
@@
 
-In this second example we investigate the behaviour of the PAIS
+In this second example we investigate the behaviour of the ETAIS
 algorithm when applied to bimodal problems. MH methods can struggle
@@
 modes are rare, resulting in incorrectly proportioned modes in the
-histograms. This example demonstrates that the PAIS algorithm
+histograms. This example demonstrates that the ETAIS algorithm
 redistributes particles to new modes as they are found. This means
@@
 	RWMH								& 2.3e-1					& - 						& 9.3e-1\\
-	PAIS-RW								& -						& 5.1e-2 					& 1.3e-1\\
+	ETAIS-RW								& -						& 5.1e-2 					& 1.3e-1\\
 	\hline
@@
 different statistics; the acceptance rate for local convergence, and
-the $L^2$ error for global convergence. Similarly in PAIS-RW we can
+the $L^2$ error for global convergence. Similarly in ETAIS-RW we can
 use the effective sample size for local convergence, and the $L^2$
@@
 {\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
-\subfigure[Convergence of locally optimised PAIS-RW with scout chain,
-and globally optimised PAIS-RW and RWMH.]
+\subfigure[Convergence of locally optimised ETAIS-RW with scout chain,
+and globally optimised ETAIS-RW and RWMH.]
 {\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
-\caption{Convergence of the PAIS-RW and RWMH algorithms for the
+\caption{Convergence of the ETAIS-RW and RWMH algorithms for the
 bimodal example discussed in Section~\ref{sec:bimodal}. Set up
@@
 
-The MH and PAIS algorithms are run with both the globally and locally
+The MH and ETAIS algorithms are run with both the globally and locally
 optimal scaling parameters. Figure~\ref{fig:BM2_L2} (a) shows that the
@@
 
-Using the PAIS algorithm we can ensure that we do not lose modes as the
+Using the ETAIS algorithm we can ensure that we do not lose modes as the
 simulation progresses. Since we have an ensemble, we can sacrifice
@@
 scaling parameter. We see that modes are not forgotten and the
-algorithm converges with the improvement we see from PAIS-RW in the
+algorithm converges with the improvement we see from ETAIS-RW in the
 Gaussian example.
 
-The adaptive algorithm can just as easily be applied to the PAIS
+The adaptive algorithm can just as easily be applied to the ETAIS
 algorithm with the `scout' chains described in the previous paragraph.
@@
 \includegraphics[width=0.6\textwidth]{"figures/BM2_AL2"}
-\caption{Convergence of modified PAIS and RWMH algorithms, comparing
+\caption{Convergence of modified ETAIS and RWMH algorithms, comparing
 the globally optimised nonadaptive algorithms with the locally
@@
 modes are decided by the arbitrary proportion of chains which are in
-each mode at this point. The adaptive PAIS-RW with scout chains tuned
+each mode at this point. The adaptive ETAIS-RW with scout chains tuned
 to the effective sample size converges at about the same rate as the
@@
 \includegraphics[width=0.8\textwidth]{"figures/calc_saving"}
-\caption{Illustration of calculating the number of PAIS-RW iterations
+\caption{Illustration of calculating the number of ETAIS-RW iterations
 required to reach a tolerance of $10^{-2.8}$ as a percentage of MH
@@
 
-The graphs in the previous section clearly show that the PAIS-RW
+The graphs in the previous section clearly show that the ETAIS-RW
 algorithm converges faster than the RWMH algorithm when both are
@@
 \vspace{2mm}
-\caption{Iterations for the PAIS algorithms required to achieve a
+\caption{Iterations for the ETAIS algorithms required to achieve a
 desired tolerance as a percentage of the number of iterations required
@@
 
-A summary of the percentage of iterations required using the PAIS
+A summary of the percentage of iterations required using the ETAIS
 algorithm compared with the respective Metropolis-Hastings algorithms
@@
 
-\subsubsection{A useful property of the PAIS algorithm for multimodal distributions}
+\subsubsection{A useful property of the ETAIS algorithm for multimodal distributions}
 
@@
 \caption{This figure demonstrates the redistribution property of the
-PAIS algorithm. Initially there is one chain in the positive mode, and
+ETAIS algorithm. Initially there is one chain in the positive mode, and
 49 chains in the negative mode.}
@@
 since there is no interaction between the chains, there is no way to
-remedy this problem. The PAIS algorithm tackles this problem with its
+remedy this problem. The ETAIS algorithm tackles this problem with its
 resampling step. The algorithm uses its dynamic mixture proposal distribution to build up an
@@
 exploration. Figure~\ref{fig:BM2_suction} shows the trace of a
-simulation of the PAIS-RW algorithm with initially 1 chain in the
+simulation of the ETAIS-RW algorithm with initially 1 chain in the
 positive mode, and 49 chains in the negative mode. It takes only a
@@
 \centerfloat
-\includegraphics[width=1.2\textwidth]{"figures/PAIS_best_posterior"}
+\includegraphics[width=1.2\textwidth]{"figures/ETAIS_best_posterior"}
 \caption{The posterior distribution of $\theta$ as given in
 Equation~\eqref{eq:mixture_posterior} as found from 10 million samples
-from the PAIS algorithm. The main diagonal contains the marginal
+from the ETAIS algorithm. The main diagonal contains the marginal
 distributions of each $\theta_i$, and the lower triangular contours
@@
 posterior distribution for this problem, created from 10 million
-samples produced by the PAIS-RW algorithm.
+samples produced by the ETAIS-RW algorithm.
 
@@
 % numerics.
-\subsubsection{Convergence of MH vs PAIS}\label{sec:mixture_conv}
+\subsubsection{Convergence of MH vs ETAIS}\label{sec:mixture_conv}
 
@@
 	\hline
-	Algorithm	& MH & PAIS \\ \hline
+	Algorithm	& MH & ETAIS \\ \hline
 	$\delta^*$	& 1.3e-1     & 2.3e-1 \\
@@
 	\vspace{1mm}
-	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithm is optimised using the effective sample size.}
+	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the ETAIS algorithm is optimised using the effective sample size.}
 	\label{table:mixture_opt_beta}
@@
 
-The optimal scaling parameters for the MH and PAIS algorithms with the proposal distributions described in Section~\ref{sec:mixture_implementation} are given in Table~\ref{table:mixture_opt_beta}.
+The optimal scaling parameters for the MH and ETAIS algorithms with the proposal distributions described in Section~\ref{sec:mixture_implementation} are given in Table~\ref{table:mixture_opt_beta}.
 
@@
 \includegraphics[width=0.8\textwidth]{"figures/Mode_proportions"}
-\caption{Convergence of the PAIS algorithm for the mixture model described in Section~\ref{sec:mixture}, convergence judged using the criterion in Equation~\eqref{eqn:mode_prop}.
+\caption{Convergence of the ETAIS algorithm for the mixture model described in Section~\ref{sec:mixture}, convergence judged using the criterion in Equation~\eqref{eqn:mode_prop}.
 Implementation described in Sections~\ref{sec:mixture_implementation} and \ref{sec:mixture_conv}.
@@
 Convergence of the relative error for the two algorithms is displayed
-in Figure~\ref{fig:mixture_modes}. PAIS converges at the expected
+in Figure~\ref{fig:mixture_modes}. ETAIS converges at the expected
 $\mathcal{O}(1/\sqrt{N})$ rate, whereas the MH algorithm converges to
@@
 density that it can efficiently sample from. However, this example
-does demonstrate the remarkable convergence of the PAIS for multimodal
+does demonstrate the remarkable convergence of the ETAIS for multimodal
 target distributions.
@@
 
-Both the MH and PAIS algorithms use an ensemble size of $M=500$.
-The PAIS algorithm uses the AMR resampling algorithm.
+Both the MH and ETAIS algorithms use an ensemble size of $M=500$.
+The ETAIS algorithm uses the AMR resampling algorithm.
 The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations.
@@
 % numerics.
-\subsubsection{Convergence of MH vs PAIS}\label{sec:lorenz_conv}
-
-The optimal scaling parameters for the MH and PAIS algorithms are given in Table~\ref{table:Lorenz_opt_parameters}.
+\subsubsection{Convergence of MH vs ETAIS}\label{sec:lorenz_conv}
+
+The optimal scaling parameters for the MH and ETAIS algorithms are given in Table~\ref{table:Lorenz_opt_parameters}.
 
@@
 	\hline
-	Algorithm	& MH & PAIS \\ \hline
+	Algorithm	& MH & ETAIS \\ \hline
 	$\delta^*$	& 1.0e-1     & 1.3e-2 \\
@@
 	\vspace{1mm}
-	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithm is optimised using the effective sample size.}
+	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the ETAIS algorithm is optimised using the effective sample size.}
 	\label{table:Lorenz_opt_parameters}
@@
 \includegraphics[width=0.8\textwidth]{"figures/Lorenz_convergence"}
-\caption{Convergence of the sample mean to the posterior mean for the MH and PAIS algorithms. Posterior distribution as described in Section~\ref{sec:lorenz_target}.}
+\caption{Convergence of the sample mean to the posterior mean for the MH and ETAIS algorithms. Posterior distribution as described in Section~\ref{sec:lorenz_target}.}
 \label{fig:Lorenz_convergence}
@@
 
-As is clear from Figure~\ref{fig:Lorenz_convergence} the PAIS-RW has a vastly superior burn-in time and achieves improved accuracy on the RWMH algorithm of three orders of magnitude after one million samples.
+As is clear from Figure~\ref{fig:Lorenz_convergence} the ETAIS-RW has a vastly superior burn-in time and achieves improved accuracy on the RWMH algorithm of three orders of magnitude after one million samples.
 
 To produce one sample from the posterior distribution for this example
-using the PAIS algorithm takes on average $5.5\times 10^{-4}$ seconds,
+using the ETAIS algorithm takes on average $5.5\times 10^{-4}$ seconds,
 while a sample can be produced in $1.7\times 10^{-5}$ seconds using
-the MH algorithm. However; the quality of the sample produced by PAIS
+the MH algorithm. However; the quality of the sample produced by ETAIS
 is much higher. To reach a tolerance of $5\times 10^{-5}$, it takes
-the PAIS algorithm roughly 16.5 seconds and the MH algorithm roughly
+the ETAIS algorithm roughly 16.5 seconds and the MH algorithm roughly
 1020 seconds. Although we do not claim that either codes are optimally
@@
 
-Importantly, we have compared the efficiency of our parallel scheme
-with a naive parallelisation of serial methods. Thus our increase in
-efficiency is over and above an $M$-fold increase, where $M$ is the
-number of ensemble members used. Our approach
-demonstrates a better-than-linear speed-up with the number of ensemble
-members used.
-
-The PAIS has a number of favourable features, for example the
+\red{The implementations of the ETAIS that have been presented in this
+  paper, have been run in serial, but we argue that, particularly for
+  inverse problems with expensive likelihoods, for example in the case
+  where this requires the approximation of the solution of a
+  differential equation, that this method is an excellent candidate
+  for parallelisation, with the likelihood evaluations for each
+  particle being computed on a different processor. This said, even in
+  parallel, the numerical results that we have presented demonstrate
+  that even a serial implementation can outperform standard
+  Metropolis-Hastings methods.}
+
+% Importantly, we have compared the efficiency of our scheme
+% with a naive parallelisation of serial methods. Thus our increase in
+% efficiency is over and above an $M$-fold increase, where $M$ is the
+% number of ensemble members used. Our approach
+% demonstrates a better-than-linear speed-up with the number of ensemble
+% members used.
+
+The ETAIS has a number of favourable features, for example the
 algorithm's ability to redistribute, through the resampling regime,
@@
 
-Another strength of the PAIS is that it can also be used with any MCMC
+Another strength of the ETAIS is that it can also be used with any MCMC
 proposal. There are a growing number of increasing sophisticated MCMC
@@
 
-One limitation of the PAIS approach as described above is that a
+One limitation of the ETAIS approach as described above is that a
 direct solver of the ETPF problem (such as FastEMD \cite{FastEMD}) has computational cost
@@
 approach to the limit
-with much larger ensemble sizes. The PAIS framework is very flexible
+with much larger ensemble sizes. The ETAIS framework is very flexible
 in terms of being able to use any combination of proposal
@@
 map could encode local covariance information, and lead to accelerated
-and stable PAIS sampling. This is an interesting area for future study.
+and stable ETAIS sampling. This is an interesting area for future study.
 
@@
 PMC     & Population Monte Carlo                 \\
-PAIS    & Parallel adaptive importance sampling  \\
+ETAIS    & Ensemble transport adaptive importance sampling  \\
 ETPF    & Ensemble transport particle filter     \\
 AMR     & Approximate multinomial resampler      \\
-PAIS-X  & PAIS with X kernels                    \\
-APAIS-X & Adaptive PAIS with X kernels          \\ \hline
+ETAIS-X  & ETAIS with X kernels                    \\
+AETAIS-X & Adaptive ETAIS with X kernels          \\ \hline
 \end{tabular}

### 2018-12-13 response_JUQ_2.tex
Added new bimodal example
Various edits, including replacing PAIS with ETAIS everywhere (still
needs to be done in the figure). Replaced the old bimodal example, and
substantially cut down on details for example 1.

@@
 we will address each of the points raised, and describe any relevant
-changes that have been made to the paper. ****INSERT SUMMARY OF MAJOR CHANGES****
+changes that have been made to the paper.
+We have made several substantial changes, including the renaming of the algorithm and paper from ``parallel adaptive importance sampling'' (PAIS) to ``ensemble transport adaptive importance sampling'' (ETAIS).
+We have also replaced three of the old numerical examples with two ones, one a bimodal example, and the second a more challenging problem for Bayesian inversion of the Lorenz '63 equations.
+We believe that the paper is stronger, and thank the reviewers and associate editor for their input.
 
@@
 
-\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{The naive parallel chains are run for long enough that they are equivalent to one long chain.
+\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{The naive parallel chains are run for long enough that they are statistically equivalent to one long chain.
 We have added a discussion on this point to *****}
@@
 They show the difference between the AMS and the ETPF but not within the PAIS.}{We assume that the referee meant the AMR?
-As the AE asked for a new example to replace the first two examples, we took this as an opportunity to compare ETAIS with the ETPF, AMR and bootstrap resamplers, which is now presented in the new first example.
+As the AE asked for a new example to replace the second example, we took this as an opportunity to compare ETAIS with the ETPF, AMR and bootstrap resamplers, which is now presented in the new first example.
 The numerics demonstrate that superior samplers lead to more stable and accurate sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear.}
@@
 The univariate gaussian is not very informative and the univariate bimodal, is not either since both modes are of the same scale.
-I would rather see these two examples turned into one single one with a 2D mixture of two separated gaussians with contrasting scales (variance covariance matrices).}{At the request of the AE, we have removed the first two examples, and replaced them with a single example, which is a two dimensional bimodal density with differing scales in the covariances of the two modes.
+I would rather see these two examples turned into one single one with a 2D mixture of two separated gaussians with contrasting scales (variance covariance matrices).}{We agree that the numerical experiments section is very long.
+However the examples presented do each demonstrate a different aspect of the advantages of our approach.
+We have taken this comment on board, however, and have drastically reduced the length of example 1, in which we demonstrate that tuning the proposal variances using the effective sample size is approximately equivalent to using the $L^2$ errors in the target, which is only available to us as we have picked a trivial target.
+We have replaced the second bimodal example, with a bimodal example as requested by the AE with differing scales in the covariances of the two modes.
 We have also taken this opportunity to demonstrate the differences in results when using the different resamplers discussed in the paper.
@@
 The new numerics demonstrate the improvement in stability and convergence of the method with the use of more advanced resampling methods.
-The remaining examples say **** and BLAH and so we're keeping them.*****}
+The mixture model example demonstrates the algorithm working in a higher dimensional multimodal setting.
+The final new example (see later comments) demonstrates the speed-ups achieved using our approach for a challenging problem with an expensive likelihood, requiring the approximation of a differential equation.
+We believe that each example serves a purpose, and we have also reduced the size of this section by ***** pages.}
 
@@
 The authors, in my opinion, do not really discuss the implementation of the PAIS algorithm but a case-by-case ad hoc version of it.
-This is clearly unacceptable.}{We strongly disagree that the approaches used are in any way ``ad hoc''.
-As per the discussion following the AE's 2nd comment, it is important that the overall proposal has tails that are as fat as those of the target.
-The ``scout'' chain, as also used in [ref], ensures two things.
-1, it allows for larger moves into unexplored regions of the state-space, and it also fattens the tails of the proposal distribution.
-As for the proposal kernels used in 7.4.2, these are chosen simply because the proposal distribution must only be supported by [0,1].
+This is clearly unacceptable.}{The scout chains were only used in the old bimodal example, and have since been cut.
+This approach has already been used in the pMC literature to stabilise the importance sampling by fattening the tails of the proposal distribution, but we have no problem with removing it from this paper.
+
+  As for the proposal kernels used in 7.4.2, we disagree that this approach is ``ad-hoc''.
+These are chosen simply because the proposal distribution must only be supported by [0,1].
 We have gone into more details regarding this and explaining the choices, and how users might make their own choices for different scenarios, in *********.}
@@
 HMC trajectories can be hand-tuned to make transitions between modes, but only with information regarding the separation.
-Ensemble-based methods appear to us to be the only candidate for reliable sampling from such distributions without such a hand-tuning.}
+Ensemble-based methods appear to us to be the only candidate for reliable sampling from such distributions without such a hand-tuning.
+However, we have agreed to remove the claim.}
 
@@
 In this example, we have demonstrated a significant speed-up over the equivalent M-H algorithm.
-We do understand however the AE's desire for a more challenging inverse problem in itself, and so we have created a new example that *********.
-The new example still demonstrates a speed-up of ********.}
+We do understand however the AE's desire for a more challenging inverse problem in itself, and so we have created a new example which has a bigger prior distribution.
+The push-forward of this prior distribution is multimodal on the timescale of the observations (see the figure below), showing that trajectories do pass close to a hyperbolic point in the assimilation window.
+This also results in a longer thinner posterior, which as can be seen from the numerics, RWMH really struggles to sample from efficiently.
+The ETAIS implementation handles this problem with ease, for a big speed-up.}
 
-\comment{8) Do not use an explicit Euler solver, use any higher order, off the shelve solver.}{Many of the off-the-shelf solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations, since they are chaotic, and as such \emph{any} numerical error is quickly amplified through the dynamics.
+\comment{8) Do not use an explicit Euler solver, use any higher order, off the shelve solver.}{Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations, since they are chaotic, and as such \emph{any} numerical error is quickly amplified through the dynamics.
 Moreover, the focus of this paper is not on these applications, but in the statistical algorithm.

### 2018-12-13 ETAIS.tex
Added new bimodal example
Various edits, including replacing PAIS with ETAIS everywhere (still
needs to be done in the figure). Replaced the old bimodal example, and
substantially cut down on details for example 1.

@@
 \centering
-\subfigure[AETAIS-RW algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_ETAISRW_adapt_var"}}
-\subfigure[AETAIS-RW algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_ETAISRW_adapt_ess"}}
-\subfigure[AETAIS-MALA algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_ETAISMALA_adapt_var"}}
-\subfigure[AETAIS-MALA algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_ETAISMALA_adapt_ess"}}
+\subfigure[AETAIS-RW algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISRW_adapt_var"}}
+\subfigure[AETAIS-RW algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISRW_adapt_ess"}}
+\subfigure[AETAIS-MALA algorithm adapting variance.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISMALA_adapt_var"}}
+\subfigure[AETAIS-MALA algorithm maximising ESS.]{\includegraphics[width=0.495\textwidth]{"figures/G1_PAISMALA_adapt_ess"}}
 \caption{Demonstration of the convergence of the adaptive scaling parameter in the adaptively tuned ETAIS algorithms.}
@@
 
+\section{A useful property of the ETAIS algorithm for multimodal distributions}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=\textwidth]{"figures/BM2_suction"}
+\caption{This figure demonstrates the redistribution property of the
+ETAIS algorithm. Initially there is one chain in the positive mode, and
+49 chains in the negative mode.}
+\label{fig:BM2_suction}
+\end{center}
+\end{figure}
+
+The biggest issue for the Metropolis-Hastings algorithms when sampling
+from multimodal posterior distributions is that frequent switches
+between the modes are required if the method is to converge
+efficiently. The ETAIS algorithm tackles this problem with its
+resampling step. The algorithm uses its dynamic mixture proposal distribution to build up an
+approximation of the posterior at each iteration, and then compares
+this to the posterior distribution via the weights function. Any large
+discrepancy in the approximation will result in a large or small
+weight being assigned to the relevant chain, meaning the chain will
+either pull other chains towards it or be sucked towards a chain with
+a larger weight. In this way, the algorithm allows chains to
+`teleport' to regions of the posterior which are in need of more
+exploration. Figure~\ref{fig:BM2_suction} shows the trace of a
+simulation of the ETAIS-RW algorithm for a bimodal example with
+equally weighted modes, with initially 1 chain in the
+positive mode, and 49 chains in the negative mode. It takes only a
+handful of iterations for the algorithm to balance out the chains into
+25 chains in each mode. The chains switch modes without having to
+climb the energy gradient in the middle.
+
 \section{Numerical Examples}\label{Sec:Num}
@@
 
+*******
 In this section we demonstrate convergence of the ETAIS algorithm. We
@@
 first moment (Equation~\eqref{eq:34567}).
-
+********
 %%%%%%
 
-\subsection{Numerical implementation}
-\label{sec:Implementation P1}
-
-For each example, we perform the following three tasks.
-
-\begin{enumerate}
-	\item {\bf Finding the optimal scaling parameters}: We chose 32 values of $\beta$ evenly spaced on a log scale in the interval $[10^{-5}, 2]$.
+% \subsection{Numerical implementation}
+% \label{sec:Implementation P1}
+
+% For each example, we perform the following three tasks.
+
+% \begin{enumerate}
+% 	\item {\bf Finding the optimal scaling parameters}: We chose 32 values of $\beta$ evenly spaced on a log scale in the interval $[10^{-5}, 2]$.
 We ran the ETAIS and MH algorithms for one million iterations, each with an ensemble size of $M=50$.
 We took 32 repeats and used the geometric means of the sample statistics in Section~\ref{sec:statistics} to find the optimal scaling parameter.
-	\item {\bf Measuring convergence of nonadaptive algorithms}: We ran the ETAIS and MH algorithms, using the scaling parameters found in Step 1, for one million iterations, again with $M=50$.
+% 	\item {\bf Measuring convergence of nonadaptive algorithms}: We ran the ETAIS and MH algorithms, using the scaling parameters found in Step 1, for one million iterations, again with $M=50$.
 The simulations are repeated 32 times.
 The performance is evaluated using the relative $L^2$ error defined in Equation~\eqref{eqn:L2_error}.
-	\item {\bf Measuring convergence of adaptive algorithms}: We run the
-adaptive algorithms under the same conditions as the nonadaptive
-algorithms, and again use the relative $L^2$ error to compare
-efficiency. The adaptive algorithms are initialised with $\beta_1$ = 1.
-\end{enumerate}
-
-%%%%%%
-
-\subsection{Sampling from a one dimensional Gaussian distribution}
+% 	\item {\bf Measuring convergence of adaptive algorithms}: We run the
+% adaptive algorithms under the same conditions as the nonadaptive
+% algorithms, and again use the relative $L^2$ error to compare
+% efficiency. The adaptive algorithms are initialised with $\beta_1$ = 1.
+% \end{enumerate}
+
+% %%%%%%
+
+\subsection{Automated variance tuning}
 \label{sec:problem 1}
-
-This first example shows how ETAIS handles searching for the mode of a
-Gaussian distribution when the initial state is a long way out in the
-tails. We optimise the
-naively parallelised RWMH algorithm using the optimal acceptance rate
-$\hat{\alpha} = 0.5$. This value differs from the theoretical
-asymptotic value of 0.234 which applies in higher dimensions. This
-higher acceptance rate is commonly used for one dimensional Gaussian
-posteriors~\cite{rosenthal2011optimal}. To tune the
-ETAIS algorithm we maximise the effective sample size as
-discussed in Section~\ref{sec:ess}.
-
-%%%%%%
-
-\subsubsection{Target distribution}
-
-Consider the case of a linear observation operator which maps a one
-dimensional parameter onto the observable space, $\G\colon x \mapsto
-x$. We assign centred priors to the parameter $x$, as well as to the
-observational noise. We choose the true value of the parameter to be
-$x_\text{ref} = 4$, and choose $\sigma^2=\tau^2=0.1$. Then
-following Equations~\eqref{eqn:obs} and \eqref{eqn:like},
-\begin{equation}\label{eqn:Gaussian posterior}
-	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x
-	 - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right).
-\end{equation}
-
-This set-up results in a posterior density in which the vast majority
-of the density is out in the tails of the
-prior distribution. The Kullback-Leibler (KL)
-divergence, which gives us a measure of how different the prior and
-posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
-problem. A KL divergence of zero indicates that two distributions are
-identical almost everywhere.
-
-%%%%%%
-
-\subsubsection{Optimising the scaling parameter}\label{sec:Optimal values P1}
+{\red In this first example, we target a simple 1D Gaussian distribution, in
+order to show the approximate equivalence of optimising the variance
+of the proposal kernels using the effective sample size, and the $L^2$
+error, which of course is not available to us in practical
+applications.}
 
@@
 \subfigure[RWMH algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_rwmho"}}
-\subfigure[ETAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
-\caption{Finding optimal values of $\beta$ for the example in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
+\subfigure[ETAIS-RW
+algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
+
+\caption{Finding optimal values of $\beta$ for a 1D Gaussian density. Bottom left: Convergence of the (A)RWMH and (A)ETAIS-RW algorithms.}
 \label{fig:P1 opt beta}
@@
 
-%%%%%%
-
-\subsubsection{Convergence of RWMH vs ETAIS-RW}
-
 \begin{figure}[htb]
 \centering 
-\includegraphics[width=0.8\textwidth]{"figures/new_G2_l2"}
-\caption{Convergence of the (A)RWMH and (A)ETAIS-RW algorithms. The
-setup is as in Section~\ref{sec:Implementation P1} (2, 3). Resampling
-is performed using the ETPF.}
+\subfigure[Convergence]{\includegraphics[width=0.45\textwidth]{"figures/new_G2_l2"}}
+\subfigure[Scaling of ETAIS-RW]{\includegraphics[width=0.45\textwidth]{"figures/PAIS_saving"}}
+
+\caption{Left: Convergence of the (A)RWMH and (A)ETAIS-RW algorithms
+  for a 1D Gaussian. Computed using 50 ensemble members, and the
+  ETPF resampler. Right: Ratio of ETAIS-RW samples required to reach the same tolerance
+  as the RWMH algorithm for a range of ensemble sizes. }
 \label{fig:MH1 L2}
@@
 
-Figure~\ref{fig:MH1 L2} shows that the ETAIS-RW algorithm converges
+Figure~\ref{fig:MH1 L2} (left) shows that the ETAIS-RW algorithm converges
 to the posterior distribution significantly faster than the RWMH
-algorithm, in both $L^2$ error and relative error in the moments. A
-description of the speed up attained by this algorithm is given in
-Section~\ref{sec:calc_saving}. The adaptive algorithms are also shown
+algorithm, in both $L^2$ error and relative error in the moments. The adaptive algorithms are also shown
 in Figure~\ref{fig:MH1 L2}. We see that both adaptive algorithms
@@
 
-%%%%%%
-
-\subsubsection{Scaling of the ETAIS algorithm with ensemble size}
-
-In this and the next example we use an ensemble size $M=50$, however it is
-interesting to see how the ETAIS algorithm scales when we increase the
-ensemble size, and if there is a minimum required ensemble size. We
-implement the problem in Section~\ref{sec:problem 1}, using
-the RWMH and ETAIS-RW algorithms with ensemble sizes in the interval $M
-\in [1, 160]$.
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=0.6\textwidth]{"figures/ETAIS_saving"}
-\caption{Ratio of ETAIS-RW samples required to reach the same tolerance
-  as the RWMH algorithm, using the ETPF resampler.}
-\label{fig:ETAIS_saving}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:ETAIS_saving} was produced using the method of finding
-optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
-then running 32 repeats at each ensemble size
-
-We repeat the process described in Section~\ref{sec:Implementation P1}
-for each of the ensemble sizes we are interested in. The convergence rates
-are then found by regressing through the data. The graph is still very
-noisy but demonstrates that increasing the ensemble size continues to
-reduce the number of iterations required in comparison with naively
-parallelised MH. The decreasing trend shows superlinear improvement of
+Figure~\ref{fig:MH1 L2} (right) was produced by identifying the
+optimal value of the scaling parameter for a range of ensemble sizes,
+and then computing the ratio of ETAIS samples needed in comparison
+with RWMH samples for the same level of error. The decreasing trend shows superlinear improvement of
 ETAIS with respect to ensemble size, in terms of the number of
 iterations required, which is a demonstration of our belief that
-parallelism of MCMC should give us added value over and above that
+communication between the ensemble members should give us added value over and above that
 provided by naive parallelism. This decrease is due to the increasing
@@
 
-%%%%%%
-
-\subsection{Sampling from Bimodal Distributions}
+\subsection{Effect of resampler quality}
 \label{sec:bimodal}
-
+{\red
 In this second example we investigate the behaviour of the ETAIS
-algorithm when applied to bimodal problems. MH methods can struggle
+algorithm when applied to a bimodal problem, given by a mixture of two
+Gaussians,
+\[\pi(x) = 0.2\pi_1(x) + 0.8\pi_2(x),\]
+where $\pi_1$ is the density of a $\mathcal{N}\left (\begin{pmatrix} 1
+    \\ 1 \end{pmatrix}, \begin{pmatrix} 0.1 & 0 
+    \\ 0 & 0.1 \end{pmatrix} \right )$ random variable and $\pi_2$ is the
+  density of a $\mathcal{N}\left (\begin{pmatrix} -5
+    \\ -5 \end{pmatrix}, \begin{pmatrix} 2.75 & -2.25 
+    \\ -2.25 & 2.75 \end{pmatrix} \right )$ random variable.}
+
+MH methods can struggle
 with multimodal problems, particularly where switches between the
@@
 proportional to the probability density in that mode, resulting in
-faster global convergence.
-
-%%%%%%
-
-\subsubsection{Target distribution} \label{sec:tar}
-
-\begin{figure}[htpb]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/BM2_posterior"}
-\caption{Posterior distribution for the bimodal example in
-Section~\ref{sec:tar}.}
-\label{fig:BM2_posterior}
-\end{center}
+faster global convergence. {\red In particular, we will look at the effect
+of using either the ETPF, AMR or a standard bootstrap resampler.
+
+\begin{figure}[htb]
+\centering 
+\subfigure[Target density]{\includegraphics[width=0.45\textwidth]{"figures/bimodal"}}
+\subfigure[Comparison of resamplers]{\includegraphics[width=0.45\textwidth]{"figures/Bivariate_L2"}}
+
+\caption{Left: Bimodal target density. Right: Convergence of RWMH and
+  ETAIS-RW with ETPF, AMR and bootstrap resamplers. }
+\label{kshdbf}
 \end{figure}
 
-In this example we look at a bimodal posterior distribution which has
-a region of very low probability density between the two modes. In the MH algorithm, the
-expected number of iterations for a switch between modes is a large
-proportion of the total number of iterations. This posterior is presented in
-Figure~\ref{fig:BM2_posterior}.
-
-We consider a non-linear observation operator $\G\colon x \mapsto
-x^2$, and assign the prior $\mu_0 = \N(0, \tau^2=0.25)$. Again, we
-assume the true parameter $x_\text{ref}=2$ is observed according to
-Equation~\eqref{eqn:obs} with observational noise $\sigma^2 = 0.1$,
-and the posterior is
-\[
-	\pi(x|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|x^2 - D\|^2
-		 - \frac{1}{2\tau^2}\|x\|^2\right).
-\]
-
-%%%%%%
-
-\subsubsection{Calculating values of the optimal scaling parameter}
-\label{sec:BM2_opt_beta}
-
-Calculating the optimal values of the scaling parameters for this
-example is similar to the previous example. These values are given in
-Table~\ref{tab:BM2_opt_delta}. The subscript on $\beta$ refers to the
-criterion which has been optimised.
-
-\begin{table}[!htb]
-      \centering
-        \begin{tabular}{|l|r|r|r|}
-	\hline
-	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$	& $\beta^*_{\text{L}^2}$ \\ \hline
-	RWMH								& 2.3e-1					& - 						& 9.3e-1\\
-	ETAIS-RW								& -						& 5.1e-2 					& 1.3e-1\\
-	\hline
-	\end{tabular}
-	\vspace{2mm}
-	\caption{Optimal scaling parameters for the example in Section~\ref{sec:bimodal}.}
-	\label{tab:BM2_opt_delta}
-\end{table}
-
-Since transitions between the modes are extremely unlikely, we
-consider the convergence on two levels; convergence to equally
-proportioned modes, and convergence to smooth histograms.
-To get correctly proportioned modes with the RWMH algorithm it is
-important that the chains can transition between the modes frequently,
-which requires a large proposal variance. However, this leads to a
-lower acceptance rate, and so we sacrifice convergence locally. For
-this reason, the RWMH algorithm is very slow to converge for problems
-of this type.
-
-We can achieve these two regimes in RWMH by tuning $\beta$ using
-different statistics; the acceptance rate for local convergence, and
-the $L^2$ error for global convergence. Similarly in ETAIS-RW we can
-use the effective sample size for local convergence, and the $L^2$
-error for global convergence.
-
-%%%%%%
-
-\subsubsection{Convergence of RWMH vs ETAIS-RW}
-
-\begin{figure}[htb]
-\centering
-\subfigure[Relative $L^2$ error using the optimal scaling parameters.]
-{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
-\subfigure[Convergence of locally optimised ETAIS-RW with scout chain,
-and globally optimised ETAIS-RW and RWMH.]
-{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
-\caption{Convergence of the ETAIS-RW and RWMH algorithms for the
-bimodal example discussed in Section~\ref{sec:bimodal}. Set up
-described in Section~\ref{sec:Implementation P1}. Resampling is
-performed using the ETPF.}
-\label{fig:BM2_L2}
-\end{figure}
-
-The MH and ETAIS algorithms are run with both the globally and locally
-optimal scaling parameters. Figure~\ref{fig:BM2_L2} (a) shows that the
-globally optimal $\beta^*$ leads to normal convergence rates,
-$\mathcal{O}(1/\sqrt{N})$, whereas algorithms using locally optimal
-scaling parameters do not sample
-proportionally for the whole simulation.
-
-Using the ETAIS algorithm we can ensure that we do not lose modes as the
-simulation progresses. Since we have an ensemble, we can sacrifice
-locally optimal sampling for one or two ensemble members in favour of
-a larger proposal variance. These chains with larger scaling
-parameters act both as `scouts' for new modes, and act to aid in the
-over-dispersal of the proposal distribution. Figure~\ref{fig:BM2_L2}
-(b) shows the results of using 49 chains with the local optimal
-scaling parameter, and one chain with ten times the local optimal
-scaling parameter. We see that modes are not forgotten and the
-algorithm converges with the improvement we see from ETAIS-RW in the
-Gaussian example.
-
-The adaptive algorithm can just as easily be applied to the ETAIS
-algorithm with the `scout' chains described in the previous paragraph.
-Since we need two equally sized sub-ensembles, we will use two groups
-of 25 ensemble members. Each group has 24 ensemble members tuned for
-locally optimal convergence and one `scout' chain with a scaling
-parameter ten times that of the rest of the group.
-
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=0.6\textwidth]{"figures/BM2_AL2"}
-\caption{Convergence of modified ETAIS and RWMH algorithms, comparing
-the globally optimised nonadaptive algorithms with the locally
-optimised adaptive algorithms. The setup is as described in
-Section~\ref{sec:Implementation P1} (3). Resampling is performed using
-the ETPF.}
-\label{fig:BM2_AL2}
-\end{center}
-\end{figure}
-
-Comparing the convergence of the adaptive algorithms against the
-nonadaptive algorithms in Figure~\ref{fig:BM2_AL2} shows that the
-algorithms behave as expected. The adaptive RWMH algorithm tuned using
-the acceptance rate converges at the same rate as the $L^2$ optimised
-algorithm until the scaling parameter gets too small, and therefore
-switches between the modes are rare, and the relative heights of the
-modes are decided by the arbitrary proportion of chains which are in
-each mode at this point. The adaptive ETAIS-RW with scout chains tuned
-to the effective sample size converges at about the same rate as the
-locally optimised nonadaptive algorithm also with scouts.
-
-%%%%%%
-
-\subsubsection{Calculating the speed up in convergence}
-\label{sec:calc_saving}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.8\textwidth]{"figures/calc_saving"}
-\caption{Illustration of calculating the number of ETAIS-RW iterations
-required to reach a tolerance of $10^{-2.8}$ as a percentage of MH
-iterations. The relative $L^2$ error graphs are from the Gaussian
-problem in Section~\ref{sec:problem 1}.}
-\label{fig:calc_saving}
-\end{center}
-\end{figure}
-
-The graphs in the previous section clearly show that the ETAIS-RW
-algorithm converges faster than the RWMH algorithm when both are
-parallelised with the same ensemble size. We can calculate the number
-of iterations required to achieve a particular tolerance level in our
-solution for each algorithm and compare these to calculate a
-percentage saving. In Figure~\ref{fig:calc_saving} we demonstrate our
-calculation of the savings. The constants $c_1$ and $c_2$ are found by
-regressing through the data with a fixed exponent of $-1/2$ excluding
-the initial data points where the algorithm has not finished burning
-in.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{|l|r|r|r|}
-\hline
-		& Gaussian & Bimodal \\ \hline
-	RWMH & 10\% & 12.6\% (scout) \\
-	MALA & 42\% & 40\% \\ \hline
-\end{tabular}
-\vspace{2mm}
-\caption{Iterations for the ETAIS algorithms required to achieve a
-desired tolerance as a percentage of the number of iterations required
-by the respective MH algorithms.}
-\label{table:calc_savings}
-\end{table}
-
-A summary of the percentage of iterations required using the ETAIS
-algorithm compared with the respective Metropolis-Hastings algorithms
-is given in Table~\ref{table:calc_savings}.
-
-\subsubsection{A useful property of the ETAIS algorithm for multimodal distributions}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=\textwidth]{"figures/BM2_suction"}
-\caption{This figure demonstrates the redistribution property of the
-ETAIS algorithm. Initially there is one chain in the positive mode, and
-49 chains in the negative mode.}
-\label{fig:BM2_suction}
-\end{center}
-\end{figure}
-
-The biggest issue for the Metropolis-Hastings algorithms when sampling
-from bimodal posterior distributions is that it is unlikely that the
-correct ratio of chains will be maintained in each of the modes, and
-since there is no interaction between the chains, there is no way to
-remedy this problem. The ETAIS algorithm tackles this problem with its
-resampling step. The algorithm uses its dynamic mixture proposal distribution to build up an
-approximation of the posterior at each iteration, and then compares
-this to the posterior distribution via the weights function. Any large
-discrepancy in the approximation will result in a large or small
-weight being assigned to the relevant chain, meaning the chain will
-either pull other chains towards it or be sucked towards a chain with
-a larger weight. In this way, the algorithm allows chains to
-`teleport' to regions of the posterior which are in need of more
-exploration. Figure~\ref{fig:BM2_suction} shows the trace of a
-simulation of the ETAIS-RW algorithm with initially 1 chain in the
-positive mode, and 49 chains in the negative mode. It takes only a
-handful of iterations for the algorithm to balance out the chains into
-25 chains in each mode. The chains switch modes without having to
-climb the energy gradient in the middle.
-
-%%%%%%
-
+Figure \ref{kshdbf} shows the bimodal target density, and convergence
+plots for the RWMH and ETAIS-RW with ETPF, AMR and bootstrap
+resamplers. As expected, a higher quality resampler leads to better
+proposal distributions, which in turn leads to greater stability and
+more reliable convergence. However, this also shows that the AMR is a
+good greedy approximation of the ETPF, and for a fraction of the cost
+when the number of ensemble members is larger, as shown in Section
+\ref{sec:AMR}. The RWMH algorithm fails to converge efficiently since chains only very rarely
+make switches between the two modes, leading to very slow mixing. This
+demonstrates a big advantage of ensemble-based methods when the target
+is multimodal.
+
+}
 \subsection{A Mixture Model}\label{sec:mixture}
@@
 \centerfloat
-\includegraphics[width=1.2\textwidth]{"figures/ETAIS_best_posterior"}
+\includegraphics[width=1.2\textwidth]{"figures/PAIS_best_posterior"}
 \caption{The posterior distribution of $\theta$ as given in

### 2019-01-04 response_JUQ_2.tex
A lot of changes
Did most of the changes. Things still to do: The plots and changes to
the final example.

@@
 This required some rewriting throughout the paper.
-In particular, we have toned down the emphasis on parallelisation, and put more emphasis on the fact that this algorithm also delivers speed-up for serial implementaions.}
+In particular, we have toned down the emphasis on parallelisation, and put more emphasis on the fact that this algorithm also delivers speed-up for serial implementaions.
+See the second paragraph of the discussion.}
 
-\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{The naive parallel chains are run for long enough that they are statistically equivalent to one long chain.
-We have added a discussion on this point to *****}
+\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{The naive parallel chains are run for long enough that they are reasonably similar to one long chain, and as such we didn't feel that it would be worthwhile to rerun all of these experiments.}
 
@@
 As the AE asked for a new example to replace the second example, we took this as an opportunity to compare ETAIS with the ETPF, AMR and bootstrap resamplers, which is now presented in the new first example.
-The numerics demonstrate that superior samplers lead to more stable and accurate sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear.}
+The numerics demonstrate that superior samplers lead to more stable and accurate sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear.
+This helps support our claim that state-of-the-art resampling methods can lead to more efficient AIS algorithms.}
 
-\comment{There is an error at the end of page 24.}{Many thanks for spotting this, this has now been corrected.*******}
+\comment{There is an error at the end of page 24.}{Many thanks for spotting the missing parenthesis, this has now been corrected.}
 
@@
 In fact, there is a confusion in Alg.
-1 in p.6 with this.}{We have carefully gone through the paper and clarified where appropriate whether we are referring the posterior measure $\mu$ or its density $\pi$, and unified the notation as much as possible.}
+1 in p.6 with this.}{Thanks for noticing this.
+We have carefully gone through the paper and removed all uses of $\mu$/measures in reference to distributions, and used only $\pi$ for posterior density throughout.}
 
@@
 As mentioned in the paper, we have conducted similar studies with MALA proposals, and although we sometimes observe a slightly quicker burn-in period, overall the performance is largely the similar to that of a RW proposal, at twice the cost.
-We have added a detailed discussion around this to ******}
+There was already a discussion on this at the end of Section 3, but we have added some more detail.}
 
@@
 Please make your convergence claims more explicit and take more time to explain them.
-In particular, do we have the correct invariant distribution?}{This algorithm belongs to a family of methods for which convergence has been proven in the literature, and as such, we have referenced this proof.
-We have clarified this a little further in *******}
+In particular, do we have the correct invariant distribution?}{This algorithm belongs to a family of methods for which convergence has been proven in the literature, and as such, we have referenced this proof, and as such we did not feel it appropriate to provide further details.
+One detail was missing from our statement, which is that the posterior distribution must be absolutely continuous with respect to the proposal distribution, which we have now added.}
 
@@
 The final new example (see later comments) demonstrates the speed-ups achieved using our approach for a challenging problem with an expensive likelihood, requiring the approximation of a differential equation.
-We believe that each example serves a purpose, and we have also reduced the size of this section by ***** pages.}
+We believe that each example serves a purpose, but we have also reduced the size of this section by 3-4 pages.}
 
@@
 These are chosen simply because the proposal distribution must only be supported by [0,1].
-We have gone into more details regarding this and explaining the choices, and how users might make their own choices for different scenarios, in *********.}
+This relates to the previous answers we have given to comments made by the referee above.
+We have gone into more details regarding this and explaining the choices.}
 
 \comment{6) Please clarify the first phrase of section 7.4.2, which, as far as I see, is simply twice wrong.
-There is a (very simple!) analytic version of the posterior AND with the right proposal MH can sample from it (!).}{We have amended the statement about the analytic form of the posterior.
-We are not aware of any M-H methods which can reliably sample from multimodal densities with well-separated modes.
+There is a (very simple!) analytic version of the posterior AND with the right proposal MH can sample from it (!).}{We have removed this phrase from the paper.}
+% {We have amended the statement about the analytic form of the posterior.
+We are not aware of any M-H methods which can reliably sample from multimodal densities with well-separated modes which are not known a priori.
 HMC trajectories can be hand-tuned to make transitions between modes, but only with information regarding the separation.
-Ensemble-based methods appear to us to be the only candidate for reliable sampling from such distributions without such a hand-tuning.
-However, we have agreed to remove the claim.}
+Ensemble-based methods appear to us to be a good candidate for reliable sampling from such distributions without such a hand-tuning.
+However, we agree to remove the claim.}
 
@@
 In this example, we have demonstrated a significant speed-up over the equivalent M-H algorithm.
-We do understand however the AE's desire for a more challenging inverse problem in itself, and so we have created a new example which has a bigger prior distribution.
+We do understand however the AE's desire for a more challenging inverse problem in itself, and so we have created a new example which has a prior distribution with larger variance.
 The push-forward of this prior distribution is multimodal on the timescale of the observations (see the figure below), showing that trajectories do pass close to a hyperbolic point in the assimilation window.
 This also results in a longer thinner posterior, which as can be seen from the numerics, RWMH really struggles to sample from efficiently.
-The ETAIS implementation handles this problem with ease, for a big speed-up.}
+The ETAIS implementation handles this problem with ease, for a big speed-up.
+The posterior is not multimodal, but it is even thinner/more elongated than previously, and thus more challenging Chaotic behaviour will in general lead to only a very small area of the state space being consistent with the observations, and so often (as in this case), the posterior is challenging because it is very thin in certain directions.
+Overall, the example demonstrates the speed-up of our approach for a problem where the likelihood is expensive.}
 
-\comment{8) Do not use an explicit Euler solver, use any higher order, off the shelve solver.}{Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations, since they are chaotic, and as such \emph{any} numerical error is quickly amplified through the dynamics.
+\begin{figure}[htp]
+\includegraphics[width=0.3\textwidth]{figures/xhist}
+\includegraphics[width=0.3\textwidth]{figures/yhist}
+\includegraphics[width=0.3\textwidth]{figures/zhist}
+\caption{Histograms from 1e7 Monte Carlo runs of the push forward of the new prior for the Lorenz '63 problem}
+  \end{figure}
+
+  \comment{8) Do not use an explicit Euler solver, use any higher order, off the shelve solver.}{Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations.
+We agree that the explicit Euler method with a small timestep is not an optimal choice, but it is a simple choice, which still leads to a chaotic nonlinear system.
+We argue that it is sufficient for the purpose of comparing the efficiency of two different Monte Carlo methods, both of which are using the same approximation of the likelihood.}
+
+  %{Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations, since they are chaotic, and as such \emph{any} numerical error is quickly amplified through the dynamics.
 Moreover, the focus of this paper is not on these applications, but in the statistical algorithm.

### 2019-01-04 ETAIS.tex
A lot of changes
Did most of the changes. Things still to do: The plots and changes to
the final example.

@@
 \subsection{Bayesian inverse problems}
-
 In this paper, we focus on the use of MCMC methods for characterising
@@
 \begin{equation}\label{eqn:obs}
-	\data = \mathcal{G}(x) + \varepsilon, \qquad \varepsilon \sim \mu_{\varepsilon} = \mathcal{N}(0,\Sigma).
+	\data = \mathcal{G}(x) + \varepsilon, \qquad \varepsilon \sim \mathcal{N}(0,\Sigma).
 \end{equation}
@@
 
-Then, according to Bayes' theorem, the posterior distribution that we
+Then, according to Bayes' theorem, the posterior density that we
 are interested in characterising is given, up to a constant of proportionality, by
@@
 In this work we use the resampler used within these particle filters. We
-approximate the posterior distribution, $\mu_Y$, with a sample
+approximate the posterior {\red density $\pi$,} with a sample
 of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
 the weights would be found by incorporating new observed data whereas
-here we simply use importance weights. This distribution, $\hat{\mu}_Y
-\approx \mu_Y$, can then be resampled, using the ETPF or otherwise,
-into a new distribution, $\hat{\eta} \approx \mu_Y$,
+here we simply use importance weights. These weighted states can then be resampled, using the ETPF or otherwise,
+into a {\red set of new equally-weighted states, such that
 \begin{equation}\label{eqn:resampler}
-	\hat{\mu}_Y = \sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
+	\sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
 	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
-	 \frac{1}{M} \delta_{x_i}(\cdot) = \hat{\eta}.
-\end{equation}This
-new sample $\{x_i\}_{i=1}^M$ is equally weighted.
+	 \frac{1}{M} \delta_{x_i}(\cdot).
+\end{equation}}
 
@@
 We wish to sample states $x \in X$ from a posterior
-probability distribution $\mu_D$, where $D$ represents our data which
-we wish to use to infer $x$. Since we have $M$ ensemble members, we
+probability distribution with density {\red $\pi$}. Since we have $M$ ensemble members, we
 represent the current state of all of the Markov chains as a vector
@@
 ensure that the overall mixture is absolutely continuous with respect to
-the posterior distribution. If this is not the case, importance
-weights can become high variance where samples are made away from
-modes in the distribution. This will
-drastically decrease the efficiency of sampling.
+the posterior distribution. {\red For example, in many inverse problems we
+may be looking to find the value of certain physical parameters which
+may be strictly positive, or even bounded on an interval. In
+this case, proposal kernels $\nu$ should be picked which have the same
+support as the target. We will see an example of an unknown parameter
+being bounded in Section \ref{sec:mixture}, where
+$\beta$-distributed proposals are made to ensure that the distribution
+is supported on $[0,1]$.}
 
@@
 \section{Consistency of ETAIS}\label{sec:consistency}
+{\red As with all importance sampling schemes, we must have absolute
+  continuity of target distribution with respect to the proposal
+  distribution if we wish to achieve convergence. This can usually be
+  ensured by picking the proposal kernels $\nu$ to be from the same
+  distribution type as the prior distribution.}
 
@@
 
-\section{A useful property of the ETAIS algorithm for multimodal distributions}
+{\red
+\section{A useful property of the ETAIS algorithm for multimodal
+  distributions} }
 
@@
 
-\subsection{Effect of resampler quality}
+\subsection{Multimodal targets and the effect of resampler quality}
 \label{sec:bimodal}
@@
 make switches between the two modes, leading to very slow mixing. This
-demonstrates a big advantage of ensemble-based methods when the target
+demonstrates the advantage that ensemble-based methods can have over
+other methods when the target
 is multimodal.
@@
 \label{sec:mixture_implementation}
-
-We have no analytic form for the posterior distribution for this
-model, and MH cannot reliably sample from this type of bimodal
-distribution. We do however know that the probability density should
+{\red In order to quantify error, we note that} the probability density should
 be evenly divided between the two modes. This is due to the symmetric
@@
 $[0,1]$, and the variances must be positive, it can be wasteful to use
-Gaussian proposal distributions. For this example we make proposals by
-matching the proposal distribution to the prior, scaling the variance
-and centring the proposal at the previous value. So
+Gaussian proposal distributions, {\red which will produce samples outside of
+the support of the posterior. Moreover, the value of the variances of
+the unknown distributions are strictly non-negative. The algorithm
+will be most efficient if the proposal and posterior distributions are
+mutually absolutely continuous. It is also useful to be able to pick
+proposal densities for which the variance is easily scaled so that we
+can tune them to optimise efficiency. Thus we pick the following
+proposal distributions for the $p$, $\mu$ and $\sigma^2$ parameters, respectively;}
 \[
@@
 method. The noise added to each observation was taken from the
-distribution $\mathcal{N}(0, 0.1^2I$. Priors for the initial condition
+distribution {\red $\mathcal{N}(0, 0.1^2I)$.} Priors for the initial condition
 coordinates were given by
 \[
-	x_0 \sim \mathcal{N}(-0.5, 0.03^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.03^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.03^2).
+	x_0 \sim \mathcal{N}(-0.5, 0.03^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.03^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.4^2).
 \]
@@
 
+*********
 \begin{figure}[htb]
@@
 \end{figure}
+********* NEW FIGURE HERE
 
@@
 
-Both the MH and ETAIS algorithms use an ensemble size of $M=500$.
-The ETAIS algorithm uses the AMR resampling algorithm.
+For this example, we use an ensemble size of $M=1500$, and the AMR resampling algorithm.
 The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations.
@@
 
-The optimal scaling parameters for the MH and ETAIS algorithms are given in Table~\ref{table:Lorenz_opt_parameters}.
-
-\begin{table}[!htb]
-      \centering
-        \begin{tabular}{|l|r|r|}
-	\hline
-	Algorithm	& MH & ETAIS \\ \hline
-	$\delta^*$	& 1.0e-1     & 1.3e-2 \\
-	\hline
-	\end{tabular}
-	\vspace{1mm}
-	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the ETAIS algorithm is optimised using the effective sample size.}
-	\label{table:Lorenz_opt_parameters}
-\end{table}
-
+% The optimal scaling parameters for the MH and ETAIS algorithms are given in Table~\ref{table:Lorenz_opt_parameters}.
+
+% \begin{table}[!htb]
+%       \centering
+%         \begin{tabular}{|l|r|r|}
+% 	\hline
+% 	Algorithm	& MH & ETAIS \\ \hline
+% 	$\delta^*$	& 1.0e-1     & 1.3e-2 \\
+% 	\hline
+% 	\end{tabular}
+% 	\vspace{1mm}
+% 	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the ETAIS algorithm is optimised using the effective sample size.}
+% 	\label{table:Lorenz_opt_parameters}
+% \end{table}
+
+**********
 \begin{figure}[htb]
@@
 \end{figure}
+****************** NEW FIGURE HERE
 
@@
 
-\red{The implementations of the ETAIS that have been presented in this
+{\red The implementations of the ETAIS that have been presented in this
   paper, have been run in serial, but we argue that, particularly for

### 2019-01-07 response_JUQ_2.tex
Lorenz example added + other edits
Draft nearly ready.

@@
 In particular, do we have the correct invariant distribution?}{This algorithm belongs to a family of methods for which convergence has been proven in the literature, and as such, we have referenced this proof, and as such we did not feel it appropriate to provide further details.
-One detail was missing from our statement, which is that the posterior distribution must be absolutely continuous with respect to the proposal distribution, which we have now added.}
+One detail was missing from our statement, which is that the posterior distribution must be absolutely continuous with respect to the proposal distribution, which we have now added.
+
+Regarding the error mentioned in the comment, I believe the AE is referring to a comment regarding the fact that the importance samples are stored for Monte Carlo estimates rather than the resampled samples, since the resampling process is not exact.
+Any introduction of errors in the resampling process simply leads to a proposal distribution which is less representative of the target distribution, and as such the resulting importance sample will have larger variance in the weights.
+The method will still converge to the correct distribution, but larger variance leads to a less efficient method.
+We have added a paragraph to clarify this.}
 
@@
 The posterior is not multimodal, but it is even thinner/more elongated than previously, and thus more challenging Chaotic behaviour will in general lead to only a very small area of the state space being consistent with the observations, and so often (as in this case), the posterior is challenging because it is very thin in certain directions.
-Overall, the example demonstrates the speed-up of our approach for a problem where the likelihood is expensive.}
+Overall, the example demonstrates the speed-up of our approach for a problem where the likelihood is expensive.
+
+Arbitrarily complex posteriors can be efficiently sampled using ETAIS if used in conjunction with a transport map to simplify the target.
+This is the subject of our next paper which we will submit in the coming weeks.}
 

### 2019-01-07 ETAIS.tex
Lorenz example added + other edits
Draft nearly ready.

@@
   from inference for mixture models, and for problems with expensive
-  likelihoods requiring the solution of a differential equation, for
-  which significant speed-ups are demonstrated.
+  likelihoods requiring the solution of a differential equation, {\red
+    for
+  which speed-ups of orders of magnitude are demonstrated.}
 \end{abstract}
@@
 problem exactly. In Section~\ref{sec:consistency} we consider
-consistency of the ETAIS algorithm. In Section \ref{Sec:Num} we present
+consistency of the ETAIS algorithm. {\red In Section~\ref{sec:useful} we briefly look at one
+particular advantageous property of this approach.} In Section \ref{Sec:Num} we present
 some numerical examples, before a brief conclusion and discussion in
@@
 are not stored in order to characterise the posterior density, since
-the resampling process inevitably adds some error. They are
+the resampling process {\red is not exact}. They are
 simply needed in order to inform the mixture proposal distribution for the next
@@
 \ref{alg:ETAIS}.
+
+{\red The choice of resampling algorithm is important since its
+  output is used to formulate the proposal distribution for the next
+  iteration. A basic resampler may be cheap to implement, but the
+  resulting sample, and in turn the next iteration's proposal distribution, may not be as representative of the target
+  distribution. This leads to higher variances of the weights in the
+  importance sampler, and therefore to slower convergence.}
 
@@
 \section{A useful property of the ETAIS algorithm for multimodal
-  distributions} }
+  distributions} } \label{sec:useful}
 
@@
 \caption{This figure demonstrates the redistribution property of the
-ETAIS algorithm. Initially there is one chain in the positive mode, and
+ETAIS algorithm for a bimodal target with equally weighted modes. Initially there is one chain in the positive mode, and
 49 chains in the negative mode.}
@@
 
-*******
-In this section we demonstrate convergence of the ETAIS algorithm. We
-also investigate the convergence properties of ETAIS compared with
-naively parallelised Metropolis-Hastings samplers. We assess the
-performance of the ETAIS algorithm using the relative $L^2$ error
-defined in Equation~\eqref{eqn:L2_error}, as well as the relative error in the
-first moment (Equation~\eqref{eq:34567}).
-********
+% In this section we demonstrate convergence of the ETAIS algorithm. We
+% also investigate the convergence properties of ETAIS compared with
+% naively parallelised Metropolis-Hastings samplers. We assess the
+% performance of the ETAIS algorithm using the relative $L^2$ error
+% defined in Equation~\eqref{eqn:L2_error}, as well as the relative error in the
+% first moment (Equation~\eqref{eq:34567}).
 %%%%%%
@@
 
-*********
 \begin{figure}[htb]
 \centering
-\includegraphics[width=0.8\textwidth]{"figures/Lorenz_posterior"}
+\includegraphics[width=0.8\textwidth]{"figures/Lorenz_posterior_b"}
 \caption{Posterior distribution for the initial condition of the particle with motion governed by the Lorenz equations as described in Section~\ref{sec:lorenz_target}.}
@@
 \end{figure}
-********* NEW FIGURE HERE
-
-Figure~\ref{fig:Lorenz_posterior} shows the marginal distributions of the posterior distribution for the initial condition $\mathbf{x}_0$ as found using the MH algorithm.
-While the $z$-dimension is largely uncorrelated with the other two dimensions, the correlation between $x$ and $y$ is -0.98, which makes it a challenging posterior distribution to explore without a transformation of the parameter space.
+
+Figure~\ref{fig:Lorenz_posterior} shows the marginal distributions of
+the posterior distribution for the initial condition $\mathbf{x}_0$ as
+found using the MH algorithm. While the $z$-dimension is largely
+uncorrelated with the other two dimensions, the correlation between
+$x$ and $y$ is {\red -0.97}, which makes it a challenging posterior distribution to explore without a transformation of the parameter space.
 
@@
 
-For this example, we use an ensemble size of $M=1500$, and the AMR resampling algorithm.
+For this example, we use an ensemble size of {\red $M=1500$}, and the AMR resampling algorithm.
 The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations.
@@
 
-**********
 \begin{figure}[htb]
 \centering
-\includegraphics[width=0.8\textwidth]{"figures/Lorenz_convergence"}
+\includegraphics[width=0.7\textwidth]{"figures/Lorenz_convergence"}
 \caption{Convergence of the sample mean to the posterior mean for the MH and ETAIS algorithms. Posterior distribution as described in Section~\ref{sec:lorenz_target}.}
@@
 \end{figure}
-****************** NEW FIGURE HERE
-
-As is clear from Figure~\ref{fig:Lorenz_convergence} the ETAIS-RW has a vastly superior burn-in time and achieves improved accuracy on the RWMH algorithm of three orders of magnitude after one million samples.
-
-To produce one sample from the posterior distribution for this example
-using the ETAIS algorithm takes on average $5.5\times 10^{-4}$ seconds,
-while a sample can be produced in $1.7\times 10^{-5}$ seconds using
-the MH algorithm. However; the quality of the sample produced by ETAIS
-is much higher. To reach a tolerance of $5\times 10^{-5}$, it takes
-the ETAIS algorithm roughly 16.5 seconds and the MH algorithm roughly
-1020 seconds. Although we do not claim that either codes are optimally
-efficient, this gives an indication of the saving that can be achieved
-through this type of methodology.
+
+{\red Figure \ref{fig:Lorenz_convergence} shows convergence plots for
+  this problem using RWMH and ETAIS. The burn-in of the ETAIS is
+  remarkably better than MH, as the ensemble quickly covers the
+  manifold on which the majority of the density lies close to. The
+  plot shows convergence with respect to number of samples, but the
+  two approaches have differing costs, due to the calculation of the
+  denominator in the weights in the ETAIS, and the resampling step
+  (AMR). Assuming that both error curves decay like $N^{-1/2}$ from
+  $N=10^6$, then for a given error tolerance, the number $N_{\rm MH}$
+  of MH samples required is given by
+\[N_{\rm MH} = N_{\rm ETAIS} \times 3.838 \times 10^4,\]
+where $N_{\rm ETAIS}$ is the number of ETAIS samples required for the
+same error tolerance. The cost per sample for the MH and ETAIS methods
+for this problem are $1.64 \times 10^{-5}$ and $1.46 \times 10^{-3}$
+respectively. Putting this together, we arrive at a speed-up factor of
+over 430. This demonstrates the benefits of this approach for
+problems with expensive likelihoods and challenging posteriors such as
+this one.
+}
 
@@
 likelihood is expensive, for instance because it involves the
-numerical approximation of the solution of a differential equation, the savings made in terms of the number
-iterations required in order reach a particular degree of accuracy
-outweighs the additional overhead costs that are incurred
-over-and-above standard MCMC.
+numerical approximation of the solution of a differential equation, {\red
+that we can achieve orders of magnitude reductions in cost to reach a
+given error tolerance in comparison with standard Metropolis-Hastings approaches.}
 
@@
 globally is likely to be of limited use for multimodal problems where
-the modes have very different covariances, or where there is curved
+the modes have very different covariances, or where there are curved
 ridges in the density. There are a variety of possible solutions to

### 2019-01-09 response_JUQ_2.tex
Draft complete
Done I hope!!!!!!!!!!!

@@
 First of all, we would like to thank the referees and the associate editor for
-their time in reading and commenting on our manuscript. In this document
+their time in reading and commenting on our manuscript. We thank all of you and Professor Higdon for your patience in waiting for these delayed corrections. In this document
 we will address each of the points raised, and describe any relevant
@@
 We have made several substantial changes, including the renaming of the algorithm and paper from ``parallel adaptive importance sampling'' (PAIS) to ``ensemble transport adaptive importance sampling'' (ETAIS).
-We have also replaced three of the old numerical examples with two ones, one a bimodal example, and the second a more challenging problem for Bayesian inversion of the Lorenz '63 equations.
-We believe that the paper is stronger, and thank the reviewers and associate editor for their input.
+We have also replaced two of the numerical examples.
+In particular we have incorporated a new bimodal example with differing scales in the two modes, and for which we have tested the ETAIS with three different resampling algorithms.
+This in particular we feel makes the paper stronger, and we hope that the referees and editors will now agree that the paper is ready for publication.
 
@@
 This required some rewriting throughout the paper.
-In particular, we have toned down the emphasis on parallelisation, and put more emphasis on the fact that this algorithm also delivers speed-up for serial implementaions.
+In particular, we have toned down the emphasis on parallelisation, and put more emphasis on the fact that this algorithm also delivers speed-up for serial implementations.
 See the second paragraph of the discussion.}
 
-\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{The naive parallel chains are run for long enough that they are reasonably similar to one long chain, and as such we didn't feel that it would be worthwhile to rerun all of these experiments.}
+\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{Most of the numerics compares our methodology to Metropolis-Hastings, based on the number of likelihood evaluations required, rather than runtime, with the argument being based on the idea that this approach is most suitable for targets with expensive likelihood evaluations.
+Since the only difference between implementations in parallel or serial of our method will be in CPU runtime (due to extra overheads due to communications between processors), it is still valid for us to compare to a naively parallelised Metropolis-Hastings approach, since our choice of serial/parallel implementation makes no difference to these results.
+For the final example where we consider runtime, a sentence has been added to state that our results will be slightly overoptimistic based on the lack of communication overheads.}
 
@@
 As the AE asked for a new example to replace the second example, we took this as an opportunity to compare ETAIS with the ETPF, AMR and bootstrap resamplers, which is now presented in the new first example.
-The numerics demonstrate that superior samplers lead to more stable and accurate sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear.
+The numerics demonstrate that superior resamplers lead to more stable and accurate importance sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear.
 This helps support our claim that state-of-the-art resampling methods can lead to more efficient AIS algorithms.}
@@
 Any introduction of errors in the resampling process simply leads to a proposal distribution which is less representative of the target distribution, and as such the resulting importance sample will have larger variance in the weights.
-The method will still converge to the correct distribution, but larger variance leads to a less efficient method.
+The method will still converge to the correct distribution, but larger variance in the weights leads to a less efficient method.
 We have added a paragraph to clarify this.}
@@
 The new numerics demonstrate the improvement in stability and convergence of the method with the use of more advanced resampling methods.
-The mixture model example demonstrates the algorithm working in a higher dimensional multimodal setting.
+The mixture model example demonstrates the algorithm working in a higher dimensional multimodal setting on a popular practical problem.
 The final new example (see later comments) demonstrates the speed-ups achieved using our approach for a challenging problem with an expensive likelihood, requiring the approximation of a differential equation.
-We believe that each example serves a purpose, but we have also reduced the size of this section by 3-4 pages.}
+We believe that each example serves a purpose, but we have also reduced the size of this section by several pages.}
 
@@
   As for the proposal kernels used in 7.4.2, we disagree that this approach is ``ad-hoc''.
-These are chosen simply because the proposal distribution must only be supported by [0,1].
+These are chosen simply because the proposal distribution must only be supported by [0,1], and the distribution chosen give us a way to make proposals which are sure to be in the this interval but with tuneable mean and variance.
 This relates to the previous answers we have given to comments made by the referee above.
@@
 Otherwise, why consider this example?}{The reason for considering the example was that the AE had previously asked us to produce an example in the regime where we suggest that we get the biggest win using our proposed method, i.e.
-those where the likelihood evaluations have significant cost due to requirement of the solution of a differential equation.
+those where the likelihood evaluations have significant cost due to requirement of, for example, the solution of a differential equation.
 In this example, we have demonstrated a significant speed-up over the equivalent M-H algorithm.
@@
 The ETAIS implementation handles this problem with ease, for a big speed-up.
-The posterior is not multimodal, but it is even thinner/more elongated than previously, and thus more challenging Chaotic behaviour will in general lead to only a very small area of the state space being consistent with the observations, and so often (as in this case), the posterior is challenging because it is very thin in certain directions.
+The posterior is challenging because it is concentrated very close to a lower dimensional manifold.
 Overall, the example demonstrates the speed-up of our approach for a problem where the likelihood is expensive.
-
-Arbitrarily complex posteriors can be efficiently sampled using ETAIS if used in conjunction with a transport map to simplify the target.
-This is the subject of our next paper which we will submit in the coming weeks.}
 
@@
 \end{figure}
+
+The more dimensions and posterior complexity that exists in a problem, the bigger the ensemble will need to be to have a stable ETAIS implementation.
+After a certain point, due to the extra overheads with a large ensemble due to for example the resampling, ETAIS will no longer be competitive.
+This said, in our most recent work, arbitrarily complex posteriors can be efficiently sampled using ETAIS if used in conjunction with a transport map to simplify the target.
+This is the subject of our next paper which we will be submitting imminently.
+We have added a paragraph to this end to the discussion.}
 

### 2019-01-09 ETAIS.tex
Draft complete
Done I hope!!!!!!!!!!!

@@
 for this problem are $1.64 \times 10^{-5}$ and $1.46 \times 10^{-3}$
-respectively. Putting this together, we arrive at a speed-up factor of
-over 430. This demonstrates the benefits of this approach for
+respectively. Here the ETAIS results were produced with a serial
+implementation, so this cost may be slightly overoptimistic, since the
+runtimes do not include extra overheads of communicating the ensemble
+states between processors. However, putting this together, we arrive at a speed-up factor of
+over 430, which will undoubtedly overshadow any such underestimate of
+the cost-per-sample of the ETAIS approach. This demonstrates the benefits of this approach for
 problems with expensive likelihoods and challenging posteriors such as
@@
 the modes have very different covariances, or where there are curved
-ridges in the density. There are a variety of possible solutions to
+ridges in the density. {\red With increased dimensionality and/or
+  complexity of the target the distribution, we also require increases
+  in the size of the ensemble if we wish to have a stable ETAIS
+  implementation. After a certain point, the extra overheads
+  associated with a very large ensemble will outweigh the advantages
+  of this approach. However, there} are a variety of possible solutions to
 this problem that would be worthy of future consideration, not least
@@
 map could encode local covariance information, and lead to accelerated
-and stable ETAIS sampling. This is an interesting area for future study.
+and stable ETAIS sampling {\red with much smaller ensemble sizes}.
 

### 2019-01-09 figures/createfigure.m
Fixed figures
Edited figures to ETAIS instead of PAIS

@@ -1,0 +1,43 @@
+function createfigure(X1, YMatrix1, X2, Y1)
+%CREATEFIGURE1(X1, YMATRIX1, X2, Y1)
+%  X1:  vector of x data
+%  YMATRIX1:  matrix of y data
+%  X2:  vector of x data
+%  Y1:  vector of y data
+
+%  Auto-generated by MATLAB on 09-Jan-2019 15:35:10
+
+% Create figure
+figure1 = figure();
+
+% Create axes
+axes1 = axes('Parent',figure1,'YMinorTick','on','YScale','log',...
+    'XMinorTick','on',...
+    'XScale','log',...
+    'FontSize',17);
+%% Uncomment the following line to preserve the X-limits of the axes
+% xlim(axes1,[2000 10000000]);
+box(axes1,'on');
+hold(axes1,'on');
+
+% Create multiple lines using matrix input to loglog
+loglog1 = loglog(X1,YMatrix1,'LineWidth',2);
+set(loglog1(1),'DisplayName','ETAIS-RW','Color',[0 0.447 0.741]);
+set(loglog1(2),'DisplayName','RWMH','LineStyle','--',...
+    'Color',[0.85 0.325 0.098]);
+
+% Create loglog
+loglog(X2,Y1,'ZDataSource','','DisplayName','$\mathcal{O}(1/\sqrt{N})$',...
+    'Color',[0.392156862745098 0.392156862745098 0.392156862745098]);
+
+% Create xlabel
+xlabel('Number of Samples','FontSize',18.7,'Interpreter','latex');
+
+% Create ylabel
+ylabel('$w_{\mbox{error}}$','FontSize',17,'Interpreter','latex');
+
+% Create legend
+legend1 = legend(axes1,'show');
+set(legend1,'Interpreter','latex','FontSize',15.3,...
+    'EdgeColor',[0.15 0.15 0.15]);
+

### 2019-01-10 ETAIS.tex
Made changes suggested by Colin
Changes to abstract, intro and conclusions. AMR -> MT

@@
   family of numerical methods for sampling from complex probability
-  distributions. As applications of these methods increase in size and
-  complexity, the need for efficient methods which can exploit the
-  parallel architectures which are prevalent in high performance
-  computing increases. In this paper, we aim to develop a framework
-  for scalable parallel sampling algorithms. At each iteration, an
+  distributions. {\red As applications of these methods increase in size and
+  complexity, the need for efficient methods increases. In this
+paper, we present a particle ensemble algorithm.} At each iteration, an
   importance sampling proposal distribution is formed using an
@@
   ready for the next iteration. We demonstrate that this ensemble transport
-  adaptive importance sampling (ETAIS) method outperforms naive
-  parallelisation of serial MCMC methods for low dimensional problems,
+  adaptive importance sampling (ETAIS) method outperforms MCMC methods
+  {\red with equivalent proposal distributions} for low dimensional problems,
   and in fact shows
@@
   the number of ensemble members. We also introduce a new resampling
-  strategy, approximate multinomial resampling (AMR), which while not
+  strategy, {\red multinomial transformation (MT),} which while not
   as accurate as the ensemble transport resampler, is substantially less costly for large
@@
     for
-  which speed-ups of orders of magnitude are demonstrated.}
+  which speed-ups of orders of magnitude are demonstrated. Likelihood
+  evaluations of the ensemble could be computed in a distributed
+  manner, suggesting that this methodology could be a good candidate
+  for parallel Bayesian computations.}
 \end{abstract}
 \begin{keywords}MCMC, importance sampling, Bayesian, inverse problems,
-  parallel.
+  {\red ensemble, resampling}.
 \end{keywords}
@@
 
-In this paper, we present a framework for parallelisation of
+{\red In this paper, we present a framework for ensemble
 importance sampling, which can be built around many of the current
@@
 distribution from the current
-ensemble. The method makes use of a resampler based on optimal
+ensemble.} The method makes use of a resampler based on optimal
 transport which has been used in the context of particle
@@
 allows us to sample from such a posterior distribution with greater
-precision per likelihood evaluation than an uncorrelated ensemble of
+precision per likelihood evaluation than 
 standard MCMC chains of the same type.
@@
 automatically tune algorithmic parameters concerned with the proposal
-distributions. In Section \ref{sec:AMR} we introduce the approximate
-multinomial resampling (AMR) algorithm which is a less accurate but
+distributions. In Section \ref{sec:MT} we introduce the {\red multinomial transformation (MT)} algorithm which is a less accurate but
 faster alternative to resamplers which solve the optimal transport
@@
 
-These deficiencies of the naive method of parallelising MCMC methods
-motivated the development of the Parallel Adaptive Importance Sampler
-(ETAIS). In the next section we will introduce the method in its most
+{\red These deficiencies of MCMC methods
+motivated the development of the Ensemble Transport Adaptive Importance Sampler
+(ETAIS).}
+In the next section we will introduce the method in its most
 general form.
@@
 will use the ETPF resampler algorithm\cite{reich2013nonparametric}, although we
-will suggest an alternative strategy in Section \ref{sec:AMR} where a
+will suggest an alternative strategy in Section \ref{sec:MT} where a
 large ensemble is required, and for which the ETPF may become more expensive. The
@@
 
-\section{Approximate Multinomial Resampling}\label{sec:AMR}
+\section{Approximate Multinomial Resampling}\label{sec:MT}
 
@@
 density. This motivates the following algorithm, which we refer to as the
-approximate multinomial resampler (AMR), which is a greedy
+multinomial transformation (MT), which is a greedy
 approximation of the ETPF resampler.
@@
 \subset \mathbb{R}^M_{\ge 0}$
-such that  $\frac{1}{M} \sum {\bf p}_i = \bar{\bf w}$. The AMR is then
+such that  $\frac{1}{M} \sum {\bf p}_i = \bar{\bf w}$. The MT is then
 given by a sample from each of the multinomials defined by the vectors
@@
 	}
-\caption{The approximate multinomial resampler (AMR).\label{alg:AMR}}
+\caption{The approximate multinomial resampler (MT).\label{alg:MT}}
 \end{algorithm}
@@
 
-Algorithm~\ref{alg:AMR} describes the basis of the algorithm with
+Algorithm~\ref{alg:MT} describes the basis of the algorithm with
 deterministic resampling, using the means of each of the
@@
 \caption{Comparison of the performance between different resampling schemes. The example in Section~\ref{sec:problem 1} is implemented for this demonstration.}
-\label{fig:AMR}
+\label{fig:MT}
 \end{figure}
@@
 To test the accuracy and speed of the three resamplers (ETPF,
-bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
+bootstrap and MT), we drew a sample of size $M$ from the proposal distribution
 $\mathcal{N}(1,2)$. Importance weights were assigned, based on a
 target distribution of $\mathcal{N}(2,3)$. The statistics of the
-resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
+resampled outputs were compared with the original weighted samples. Figure \ref{fig:MT} (a)-(c) show how the relative errors in the first
 three moments of the samples changes with ensemble size $M$ for the three different
-resamplers. As expected, the AMR lies somewhere between the high
+resamplers. As expected, the MT lies somewhere between the high
 accuracy of the ETPF and the less accurate bootstrap
@@
 sampler is presented for the first moment since both the ETPF and the
-AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
+MT preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:MT} (d) shows how the computational cost,
 measured in seconds, scales with the ensemble size for the three
 different methods, where timings have been taken from simulations on a Dell
-server with four 8-core 3.3GHz CPUs and 64Gb memory. These results demonstrate that the AMR behaves how we wish, and
+server with four 8-core 3.3GHz CPUs and 64Gb memory. These results demonstrate that the MT behaves how we wish, and
 importantly ensures that exactly one sample of the output will lie in
@@
 
-We will use the AMR in the numerics in Section \ref{sec:mixture}
+We will use the MT in the numerics in Section \ref{sec:mixture}
 where we have chosen to use a larger ensemble size. We do not claim
-that the AMR is the optimal choice within ETAIS, but it does have
+that the MT is the optimal choice within ETAIS, but it does have
 favourable features, and demonstrates how different choices of
@@
 
+{\red The numerics contained herein compare the ETAIS approach that we
+  have presented against Metropolis-Hastings methods with
+  the same type of proposal as is used by the kernels within
+  ETAIS. Since we propose that ETAIS is a potential framework for
+  parallelised Bayesian computations, we compare ETAIS with $N$
+  ensemble members against an independent ensemble of $N$
+  Metropolis-Hastings chains. All implementations were
+  computed in serial, but the results demonstrate a clear speed-up
+  which could be further exploited through parallelisation in the future.
+  }
+
 % In this section we demonstrate convergence of the ETAIS algorithm. We
@@
 faster global convergence. {\red In particular, we will look at the effect
-of using either the ETPF, AMR or a standard bootstrap resampler.
+of using either the ETPF, MT or a standard bootstrap resampler.
 
@@
 \caption{Left: Bimodal target density. Right: Convergence of RWMH and
-  ETAIS-RW with ETPF, AMR and bootstrap resamplers. }
+  ETAIS-RW with ETPF, MT and bootstrap resamplers. }
 \label{kshdbf}
@@
 Figure \ref{kshdbf} shows the bimodal target density, and convergence
-plots for the RWMH and ETAIS-RW with ETPF, AMR and bootstrap
+plots for the RWMH and ETAIS-RW with ETPF, MT and bootstrap
 resamplers. As expected, a higher quality resampler leads to better
 proposal distributions, which in turn leads to greater stability and
-more reliable convergence. However, this also shows that the AMR is a
+more reliable convergence. However, this also shows that the MT is a
 good greedy approximation of the ETPF, and for a fraction of the cost
 when the number of ensemble members is larger, as shown in Section
-\ref{sec:AMR}. The RWMH algorithm fails to converge efficiently since chains only very rarely
+\ref{sec:MT}. The RWMH algorithm fails to converge efficiently since chains only very rarely
 make switches between the two modes, leading to very slow mixing. This
@@
 $M=50$ to $M=500$ to compensate for the increase in dimension. We also
-use the AMR algorithm for the resampling step because of the reduced
+use the MT algorithm for the resampling step because of the reduced
 computational cost. The method otherwise remains the same as in
@@
 Implementation described in Sections~\ref{sec:mixture_implementation} and \ref{sec:mixture_conv}.
-Resampling is performed using the AMR scheme.}
+Resampling is performed using the MT scheme.}
 \label{fig:mixture_modes}
@@
 
-For this example, we use an ensemble size of {\red $M=1500$}, and the AMR resampling algorithm.
+For this example, we use an ensemble size of {\red $M=1500$}, and the MT resampling algorithm.
 The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations.
@@
   denominator in the weights in the ETAIS, and the resampling step
-  (AMR). Assuming that both error curves decay like $N^{-1/2}$ from
+  (MT). Assuming that both error curves decay like $N^{-1/2}$ from
   $N=10^6$, then for a given error tolerance, the number $N_{\rm MH}$
@@
 
-We have explored the application of parallelised MCMC algorithms in
-low dimensional inverse problems. We have demonstrated numerically
-that these algorithms converge faster than the analogous naively parallelised
-Metropolis-Hastings algorithms. Further experimentation with the Metropolis
+{\red We have explored the application of 
+low dimensional Bayesian inverse problems. We have demonstrated numerically
+that this method converges faster than the analogous naively parallelised
+Metropolis-Hastings algorithms.} Further experimentation with the Metropolis
 Adjusted Langevin Algorithm (MALA), preconditioned Crank-Nicolson (pCN),
@@
 ETPF    & Ensemble transport particle filter     \\
-AMR     & Approximate multinomial resampler      \\
+MT     & Multinomial transformation      \\
 ETAIS-X  & ETAIS with X kernels                    \\

### 2019-01-10 figures/createfigure.m
Removed AMR from figures

@@ -1,43 +1,39 @@
-function createfigure(X1, YMatrix1, X2, Y1)
-%CREATEFIGURE1(X1, YMATRIX1, X2, Y1)
+function createfigure(X1, YMatrix1)
+%CREATEFIGURE1(X1, YMATRIX1)
 %  X1:  vector of x data
 %  YMATRIX1:  matrix of y data
-%  X2:  vector of x data
-%  Y1:  vector of y data
 
-%  Auto-generated by MATLAB on 09-Jan-2019 15:35:10
+%  Auto-generated by MATLAB on 10-Jan-2019 12:15:01
 
 % Create figure
-figure1 = figure();
+figure1 = figure;
 
 % Create axes
 axes1 = axes('Parent',figure1,'YMinorTick','on','YScale','log',...
+    'YTick',[1e-07 1e-05 0.001 0.1 10],...
     'XMinorTick','on',...
     'XScale','log',...
     'FontSize',17);
 %% Uncomment the following line to preserve the X-limits of the axes
-% xlim(axes1,[2000 10000000]);
+% xlim(axes1,[0 500]);
+%% Uncomment the following line to preserve the Y-limits of the axes
+% ylim(axes1,[1e-07 10]);
 box(axes1,'on');
 hold(axes1,'on');
 
 % Create multiple lines using matrix input to loglog
-loglog1 = loglog(X1,YMatrix1,'LineWidth',2);
-set(loglog1(1),'DisplayName','ETAIS-RW','Color',[0 0.447 0.741]);
-set(loglog1(2),'DisplayName','RWMH','LineStyle','--',...
-    'Color',[0.85 0.325 0.098]);
-
-% Create loglog
-loglog(X2,Y1,'ZDataSource','','DisplayName','$\mathcal{O}(1/\sqrt{N})$',...
-    'Color',[0.392156862745098 0.392156862745098 0.392156862745098]);
+loglog1 = loglog(X1,YMatrix1,'LineWidth',2,'Parent',axes1);
+set(loglog1(1),'DisplayName','ETPF');
+set(loglog1(2),'DisplayName','MT','LineStyle','--');
+set(loglog1(3),'DisplayName','Bootstrap','LineStyle','-.','Color',[0 0.4 0]);
 
 % Create xlabel
-xlabel('Number of Samples','FontSize',18.7,'Interpreter','latex');
+xlabel('Ensemble size','FontSize',18.7,'Interpreter','latex');
 
 % Create ylabel
[…]

### 2019-01-14 response_JUQ_2.tex
Made changes suggested by Colin
Changes to abstract, intro and conclusions. AMR -> MT
Final version complete

@@
 In particular we have incorporated a new bimodal example with differing scales in the two modes, and for which we have tested the ETAIS with three different resampling algorithms.
-This in particular we feel makes the paper stronger, and we hope that the referees and editors will now agree that the paper is ready for publication.
+We feel that this, in particular, makes the paper stronger, and we hope that the referees and editors will now agree that the paper is ready for publication.
 
@@
 The numerics demonstrate that superior resamplers lead to more stable and accurate importance sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear.
-This helps support our claim that state-of-the-art resampling methods can lead to more efficient AIS algorithms.}
+This helps support our claim that state-of-the-art resampling methods can lead to more efficient AIS algorithms.
+Note that we have changed the name of the AMR to the multinomial transform (MT).}
 
@@
 
-  \comment{8) Do not use an explicit Euler solver, use any higher order, off the shelve solver.}{Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations.
-We agree that the explicit Euler method with a small timestep is not an optimal choice, but it is a simple choice, which still leads to a chaotic nonlinear system.
-We argue that it is sufficient for the purpose of comparing the efficiency of two different Monte Carlo methods, both of which are using the same approximation of the likelihood.}
+  \comment{8) Do not use an explicit Euler solver, use any higher order, off the shelve solver.}{In this example, we use explicit Euler applied to Lorenz '63 as a simple way to generate a discrete chaotic map so that the inverse problem is challenging.
+This is an identical twin experiment assuming that the discrete system is the perfect one from which the data is obtained (following the presentation of the textbook of Reich and Cotter), so changing to another integration scheme doesn't really change anything.}
 
-  %{Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations, since they are chaotic, and as such \emph{any} numerical error is quickly amplified through the dynamics.
+  % {Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations, since they are chaotic, and as such \emph{any} numerical error is quickly amplified through the dynamics.
 Moreover, the focus of this paper is not on these applications, but in the statistical algorithm.
 In this example, we are simply looking for a non-linear mapping such that we can demonstrate the speed-up that is gained from using ETAIS as opposed to standard M-H sampling, and we believe that a simple Euler approximation provides such a mapping.}
+
+  \section{Other changes}
+  We have also changed the name of the approximate multinomial resampler (AMR) to multinomial transformation (MT).
+
 \end{document}

### 2019-01-14 ETAIS.tex
Removed AMR from figures
Final version complete

@@
   evaluations of the ensemble could be computed in a distributed
-  manner, suggesting that this methodology could be a good candidate
+  manner, suggesting that this methodology is a good candidate
   for parallel Bayesian computations.}
@@
 \begin{equation*}
-	\pi_Y \propto \exp \left ( -\Phi(x^*) \right )\pi_0(x^*),
+	\pi \propto \exp \left ( -\Phi(x^*) \right )\pi_0(x^*),
 \end{equation*}
@@
 development of the resampler that we incorporate into the ETAIS is
-motivated by this area.
-
-Particle filters are a class of Monte Carlo algorithms designed to
+motivated by this area. Particle filters are a class of Monte Carlo algorithm designed to
 solve the filtering problem. That is, to find the best estimate of the
@@
 In this work we use the resampler used within these particle filters. We
-approximate the posterior {\red density $\pi$,} with a sample
-of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
+approximate the posterior {\red density $\pi$, with an ensemble
+of weighted samples,} $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
 the weights would be found by incorporating new observed data whereas
-here we simply use importance weights. These weighted states can then be resampled, using the ETPF or otherwise,
-into a {\red set of new equally-weighted states, such that
+here we simply use importance weights. These weighted samples can then be resampled, using the ETPF or otherwise,
+into a {\red set of new equally-weighted samples $\{x_i\}_{i=1}^M$, such that
 \begin{equation}\label{eqn:resampler}
@@
 	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
-	 \frac{1}{M} \delta_{x_i}(\cdot).
+	 \frac{1}{M} \delta_{x_i}(\cdot),
 \end{equation}}
-
-The resampler used within particle filters such as the ETPF are well suited to this problem
+{\red where $\delta_{x}(\cdot)$ is the Dirac delta measure.}
+
+The resamplers used within particle filters such as the ETPF are well suited to this problem
 since it is easy to introduce conditions in the resampling to ensure
@@
 will use the ETPF resampler algorithm\cite{reich2013nonparametric}, although we
-will suggest an alternative strategy in Section \ref{sec:MT} where a
+will suggest an alternative strategy in Section \ref{sec:MT}, for
+examples where a
 large ensemble is required, and for which the ETPF may become more expensive. The
@@
   distribution. This leads to higher variances of the weights in the
-  importance sampler, and therefore to slower convergence.}
+  importance sampler, and therefore to slower convergence to the
+  target density.}
 
@@
 is orders of magnitude bigger than the rest. This is a problem, since
-the resampling step will then lead to a vector of particles all
+the resampling step will then lead to a vector of samples all
 centred at the outlier sample. Fortunately these spikes in weights can
@@
 can be removed. The ETPF can then be used to formulate a new mixture
-sample with one particle placed at the outlier point, and the others
-distributed according to the particle positions before the problematic
+sample with one sample placed at the outlier point, and the others
+distributed according to the sample positions before the problematic
 iteration. This ensures that the region where the problem occurred is
@@
 weights, as it takes values on $[1, M]$ while the variance of the
-weights takes values on $\mathbb{R}_\geq$. Moreover the variance of
+weights takes values on $\mathbb{R}^+$. Moreover the variance of
 the weights can vary
@@
 consider the algorithm in the early stages, for example, we have all
-$M$ particles in the tails of the target searching for a mode. The
-particle closest to the mode will have an exponentially higher weight
+$M$ samples in the tails of the target searching for a mode. The
+sample closest to the mode will have an exponentially higher weight
 and the effective sample size will be close to 1. In later stages the
@@
 
-\section{Approximate Multinomial Resampling}\label{sec:MT}
+\section{Multinomial Transformation}\label{sec:MT}
 
@@
 	}
-\caption{The approximate multinomial resampler (MT).\label{alg:MT}}
+\caption{The  multinomial transformation (MT).\label{alg:MT}}
 \end{algorithm}
@@
 Estimation of the normalisation constant $Z$ is biased, and so estimates of statistics are sums of independent but biased estimators.
-Since the estimators are independent, the proof of consistency of ETAIS in this second limit can be approached in the same way as the PMC algorithm, where it has been shown that $\hat{Z}\rightarrow Z$~\cite{robert2013monte}.
+Since the estimators are independent, the proof of consistency of ETAIS in this second limit can be approached in the same way as the pMC algorithm, where it has been shown that $\hat{Z}\rightarrow Z$~\cite{robert2013monte}.
 Since the normalisation constant is consistent, sums of the independent estimators are also consistent.
@@
 
-{\red
-\section{A useful property of the ETAIS algorithm for multimodal
-  distributions} } \label{sec:useful}
+\section{{\red A useful property of the ETAIS algorithm for multimodal
+  distributions } }\label{sec:useful}
 
@@
   computed in serial, but the results demonstrate a clear speed-up
-  which could be further exploited through parallelisation in the future.
+  which could be further exploited through parallelisation.
   }
@@
 histograms. This example demonstrates that the ETAIS algorithm
-redistributes particles to new modes as they are found. This means
-that we expect the number of particles in a mode to be approximately
+redistributes samples to new modes as they are found. This means
+that we expect the number of samples in a mode to be approximately
 proportional to the probability density in that mode, resulting in
@@
 plots for the RWMH and ETAIS-RW with ETPF, MT and bootstrap
-resamplers. As expected, a higher quality resampler leads to better
+resamplers, averaged over 32 repeats. As expected, a higher quality resampler leads to better
 proposal distributions, which in turn leads to greater stability and
@@
 \[
-	x_0 \sim \mathcal{N}(-0.5, 0.03^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.03^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.4^2).
+	x_0 \sim \mathcal{N}(-0.5, 0.4^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.4^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.4^2).
 \]
@@
 \[
-	\pi(\mathbf{x}_0|D) \propto \exp\left\{-\frac{1}{2}\|\mathcal{G}(\mathbf{x}_0) - D\|^2_\Sigma\right\}\mathcal{N}(\mathbf{x}_0; m, 0.03^2\text{I}),
+	\pi(\mathbf{x}_0|D) \propto \exp\left\{-\frac{1}{2}\|\mathcal{G}(\mathbf{x}_0) - D\|^2_\Sigma\right\}\mathcal{N}(\mathbf{x}_0; m, 0.4^2\text{I}),
 \]
@@
   for parallelisation, with the likelihood evaluations for each
-  particle being computed on a different processor. This said, even in
-  parallel, the numerical results that we have presented demonstrate
+  particle being computed on a different processor. This said, the numerical results that we have presented demonstrate
   that even a serial implementation can outperform standard

### 2019-01-15 cover_letter.tex
Added data
Cover letter added

@@
 
-\signature{Dr Simon Cotter (corresponding author), Dr Colin Cotter and Paul Russell}
+\signature{Dr Simon Cotter (corresponding author), Dr Colin Cotter and
+  Dr Paul Russell}
 \address{School of Mathematics \\ Alan Turing Building \\ Oxford
@@
 
-Please find attached the latest version of our manuscript, ``Parallel Adaptive Importance Sampling''.
-We have taken onboard the comments of the referees and the associate editor, and we hope that they will be satisfied with the substantial changes and additions that we have made.
-We very much look forward to hearing from you in due course regarding this paper.
+Please find attached the latest version of our manuscript, now
+entitled``Ensemble Transport Adaptive Importance Sampling''. We have
+taken onboard the comments of the referees and the associate editor,
+and we hope that they will be satisfied with the substantial changes
+and additions that we have made. We apologise once more for the delay
+in completing these revisions. We very much look forward to hearing from you in due course regarding this paper.
 

### 2019-01-30 ETAIS.tex
remove red and fix bib for archiving

@@
 \title{Ensemble Transport Adaptive Importance Sampling}
+%\newcommand{\revised}{\red}
+\newcommand{\revised}{}
 \begin{document}
@@
   family of numerical methods for sampling from complex probability
-  distributions. {\red As applications of these methods increase in size and
+  distributions. {\revised As applications of these methods increase in size and
   complexity, the need for efficient methods increases. In this
@@
   adaptive importance sampling (ETAIS) method outperforms MCMC methods
-  {\red with equivalent proposal distributions} for low dimensional problems,
+  {\revised with equivalent proposal distributions} for low dimensional problems,
   and in fact shows
@@
   the number of ensemble members. We also introduce a new resampling
-  strategy, {\red multinomial transformation (MT),} which while not
+  strategy, {\revised multinomial transformation (MT),} which while not
   as accurate as the ensemble transport resampler, is substantially less costly for large
@@
   from inference for mixture models, and for problems with expensive
-  likelihoods requiring the solution of a differential equation, {\red
+  likelihoods requiring the solution of a differential equation, {\revised
     for
@@
 \begin{keywords}MCMC, importance sampling, Bayesian, inverse problems,
-  {\red ensemble, resampling}.
+  {\revised ensemble, resampling}.
 \end{keywords}
@@
 
-{\red In this paper, we present a framework for ensemble
+{\revised In this paper, we present a framework for ensemble
 importance sampling, which can be built around many of the current
@@
 automatically tune algorithmic parameters concerned with the proposal
-distributions. In Section \ref{sec:MT} we introduce the {\red multinomial transformation (MT)} algorithm which is a less accurate but
+distributions. In Section \ref{sec:MT} we introduce the {\revised multinomial transformation (MT)} algorithm which is a less accurate but
 faster alternative to resamplers which solve the optimal transport
 problem exactly. In Section~\ref{sec:consistency} we consider
-consistency of the ETAIS algorithm. {\red In Section~\ref{sec:useful} we briefly look at one
+consistency of the ETAIS algorithm. {\revised In Section~\ref{sec:useful} we briefly look at one
 particular advantageous property of this approach.} In Section \ref{Sec:Num} we present
@@
 In this work we use the resampler used within these particle filters. We
-approximate the posterior {\red density $\pi$, with an ensemble
+approximate the posterior {\revised density $\pi$, with an ensemble
 of weighted samples,} $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
@@
 here we simply use importance weights. These weighted samples can then be resampled, using the ETPF or otherwise,
-into a {\red set of new equally-weighted samples $\{x_i\}_{i=1}^M$, such that
+into a {\revised set of new equally-weighted samples $\{x_i\}_{i=1}^M$, such that
 \begin{equation}\label{eqn:resampler}
@@
 \end{equation}}
-{\red where $\delta_{x}(\cdot)$ is the Dirac delta measure.}
+{\revised where $\delta_{x}(\cdot)$ is the Dirac delta measure.}
 
@@
 
-{\red These deficiencies of MCMC methods
+{\revised These deficiencies of MCMC methods
 motivated the development of the Ensemble Transport Adaptive Importance Sampler
@@
 are not stored in order to characterise the posterior density, since
-the resampling process {\red is not exact}. They are
+the resampling process {\revised is not exact}. They are
 simply needed in order to inform the mixture proposal distribution for the next
@@
 
-{\red The choice of resampling algorithm is important since its
+{\revised The choice of resampling algorithm is important since its
   output is used to formulate the proposal distribution for the next
@@
 We wish to sample states $x \in X$ from a posterior
-probability distribution with density {\red $\pi$}. Since we have $M$ ensemble members, we
+probability distribution with density {\revised $\pi$}. Since we have $M$ ensemble members, we
 represent the current state of all of the Markov chains as a vector
@@
 ensure that the overall mixture is absolutely continuous with respect to
-the posterior distribution. {\red For example, in many inverse problems we
+the posterior distribution. {\revised For example, in many inverse problems we
 may be looking to find the value of certain physical parameters which
@@
 \section{Consistency of ETAIS}\label{sec:consistency}
-{\red As with all importance sampling schemes, we must have absolute
+{\revised As with all importance sampling schemes, we must have absolute
   continuity of target distribution with respect to the proposal
@@
 
-\section{{\red A useful property of the ETAIS algorithm for multimodal
+\section{{\revised A useful property of the ETAIS algorithm for multimodal
   distributions } }\label{sec:useful}
@@
 
-{\red The numerics contained herein compare the ETAIS approach that we
+{\revised The numerics contained herein compare the ETAIS approach that we
   have presented against Metropolis-Hastings methods with
@@
 \label{sec:problem 1}
-{\red In this first example, we target a simple 1D Gaussian distribution, in
+{\revised In this first example, we target a simple 1D Gaussian distribution, in
 order to show the approximate equivalence of optimising the variance
@@
 \label{sec:bimodal}
-{\red
+{\revised
 In this second example we investigate the behaviour of the ETAIS
@@
 proportional to the probability density in that mode, resulting in
-faster global convergence. {\red In particular, we will look at the effect
+faster global convergence. {\revised In particular, we will look at the effect
 of using either the ETPF, MT or a standard bootstrap resampler.
@@
 \label{sec:mixture_implementation}
-{\red In order to quantify error, we note that} the probability density should
+{\revised In order to quantify error, we note that} the probability density should
 be evenly divided between the two modes. This is due to the symmetric
@@
 $[0,1]$, and the variances must be positive, it can be wasteful to use
-Gaussian proposal distributions, {\red which will produce samples outside of
+Gaussian proposal distributions, {\revised which will produce samples outside of
 the support of the posterior. Moreover, the value of the variances of
@@
 method. The noise added to each observation was taken from the
-distribution {\red $\mathcal{N}(0, 0.1^2I)$.} Priors for the initial condition
+distribution {\revised $\mathcal{N}(0, 0.1^2I)$.} Priors for the initial condition
 coordinates were given by
@@
 uncorrelated with the other two dimensions, the correlation between
-$x$ and $y$ is {\red -0.97}, which makes it a challenging posterior distribution to explore without a transformation of the parameter space.
+$x$ and $y$ is {\revised -0.97}, which makes it a challenging posterior distribution to explore without a transformation of the parameter space.
 
@@
 
-For this example, we use an ensemble size of {\red $M=1500$}, and the MT resampling algorithm.
+For this example, we use an ensemble size of {\revised $M=1500$}, and the MT resampling algorithm.
 The optimal scaling parameters are calculated for both algorithms using test runs which are not included in the convergence cost calculations.
@@
 
-{\red Figure \ref{fig:Lorenz_convergence} shows convergence plots for
+{\revised Figure \ref{fig:Lorenz_convergence} shows convergence plots for
   this problem using RWMH and ETAIS. The burn-in of the ETAIS is
@@
 
-{\red We have explored the application of 
+{\revised We have explored the application of 
 low dimensional Bayesian inverse problems. We have demonstrated numerically
@@
 
-{\red The implementations of the ETAIS that have been presented in this
+{\revised The implementations of the ETAIS that have been presented in this
   paper, have been run in serial, but we argue that, particularly for
@@
 likelihood is expensive, for instance because it involves the
-numerical approximation of the solution of a differential equation, {\red
+numerical approximation of the solution of a differential equation, {\revised
 that we can achieve orders of magnitude reductions in cost to reach a
@@
 the modes have very different covariances, or where there are curved
-ridges in the density. {\red With increased dimensionality and/or
+ridges in the density. {\revised With increased dimensionality and/or
   complexity of the target the distribution, we also require increases
@@
 map could encode local covariance information, and lead to accelerated
-and stable ETAIS sampling {\red with much smaller ensemble sizes}.
+and stable ETAIS sampling {\revised with much smaller ensemble sizes}.
 
